### 2021-02-09 exploratory/exploratory.Rmd
add exploratory analysis

---
title: "OptiFit GLNE Exploratory"
author: "Courtney R Armour"
date: "February 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,message=FALSE)

library(tidyverse)
library(caret)
library(kableExtra)
library(pROC)
```

## Percent of reads mapped with OptiFit

```{r reads_mapped}

counts <- read_csv("./counts/all_counts.csv",col_names = c("sample","count"))

counts %>% 
  mutate(pct_mapped = count/10000) %>% 
  ggplot(aes(x=pct_mapped)) +
    geom_histogram(fill="grey",color="black") +
    theme_bw() + 
    scale_x_continuous(labels=scales::percent) +
    xlab("Percent of reads mapped")

# counts %>% 
#   mutate(pct_mapped = count/10000) %>% 
#   ggplot(aes(x="",y=pct_mapped)) +
#     geom_boxplot(fill="grey",color="black") +
#     theme_bw() +
#     xlab("")
```

## MCC OptiFit vs OptiClust

```{r mcc}

optifit_in_mcc <- read_tsv("./optifit_in_mcc.tsv")
optifit_out_mcc <- read_tsv("./optifit_out_mcc.tsv")
opticlust_mcc <- read_tsv("./opticlust_mcc.tsv")

mcc_data <- tibble(type=c(rep("OptiClust_489",nrow(optifit_out_mcc)),
                          rep("OptiFit_1",nrow(optifit_in_mcc)),
                          rep("OptiClust_all",nrow(opticlust_mcc))),
                   mcc=c(optifit_out_mcc$mcc,optifit_in_mcc$mcc,opticlust_mcc$mcc))

mcc_data$type <- factor(mcc_data$type,levels=c("OptiClust_all","OptiClust_489","OptiFit_1"))
mcc_data %>% 
  ggplot(aes(x=type,y=mcc,color=type)) + 
    geom_jitter(height=0) +
    xlab("") +
    theme_bw()

```

# Model performance during training
```{r data}

optifit_summary <- read_tsv("../data/learning/summary/optifit/model_results.tsv",
                            col_types = cols(sample = col_character(),
                                             cv_auc = col_double(),
                                             cancer = col_double(),
                                             normal = col_double(),
                                             dx_diff = col_double(),
                                             dx_flag = col_factor(),
                                             pred_class = col_number(),
                                             meta_class = col_number()
                                             ))  %>% 
  mutate(pred_class = case_when(pred_class == 0 ~ "normal",
                                TRUE ~ "cancer")) %>% 
  mutate(meta_class = case_when(meta_class == 0 ~ "normal",
                                TRUE ~ "cancer"))

opticlust_summary <- read_tsv("../data/learning/summary/opticlust/model_results.tsv",
                              col_types = cols(sample = col_character(),
                                               cv_auc = col_double(),
                                               cancer = col_double(),
                                               normal = col_double(),
                                               dx_diff = col_double(),
                                               dx_flag = col_factor(),
                                               pred_class = col_factor(),
                                               meta_class = col_factor()
                                               )) %>% 
  mutate(pred_class = case_when(pred_class == 0 ~ "normal",
                                TRUE ~ "cancer")) %>% 
  mutate(meta_class = case_when(meta_class == 0 ~ "normal",
                                TRUE ~ "cancer"))

opticlust_summary %>% mutate(method="OptiClust") %>% 
  bind_rows(optifit_summary %>% mutate(method="OptiFit")) %>% 
  ggplot(aes(x=method,y=cv_auc,fill=method)) +
    geom_boxplot() +
    #geom_jitter(height=0,alpha = 0.5) +
    labs(title = "Cross Validation Performance") +
    theme_bw() +
    theme(legend.position = "none") 
ggsave("./figures/cv_auc_performance.png")

```

# Prediction Probability 
```{r }

opticlust_summary %>% 
  ggplot(aes(x=cancer,fill=meta_class)) +
    geom_histogram(color="black",alpha=0.8,binwidth = 0.05) + theme_bw() +
    labs(xlab = "Probability of Cancer") + 
    facet_wrap(~meta_class) +
    xlab("probability of cancer") +
    geom_vline(xintercept = 0.5,lty="dashed") +
    scale_y_continuous(expand=c(0,0)) +
    scale_fill_manual(values=c("red","blue"),guide=NULL) +
    ggtitle("OptiClust")

optifit_summary %>% 
  ggplot(aes(x=cancer,fill=meta_class)) +
    geom_histogram(color="black",alpha=0.8,binwidth = 0.05) + theme_bw() +
    labs(xlab = "Probability of Cancer") + 
    facet_wrap(~meta_class) +
    xlab("probability of cancer") +
    geom_vline(xintercept = 0.5,lty="dashed") +
    scale_y_continuous(expand=c(0,0)) +
    scale_fill_manual(values=c("red","blue"),guide=NULL) +
    ggtitle("OptiFit")

```

### Difference in prediction probablity  
```{r hist}

# ggplot(opticlust_summary,aes(x=dx_diff)) +
#   geom_histogram(binwidth = 0.05,color="black",fill="grey") +
#   labs(title = "OptiClust: Difference in DX Prediction",
#        x="Difference in Probability") +
#   theme_bw() +
#   scale_x_continuous(expand=c(0,0)) +
#   scale_y_continuous(expand=c(0,0))
# 
# ggplot(optifit_summary,aes(x=dx_diff)) +
#   geom_histogram(binwidth = 0.05,color="black",fill="grey") +
#   labs(title = "OptiFit: Difference in DX Prediction",
#        x="Difference in Probability") +
#   theme_bw() +
#   scale_x_continuous(expand=c(0,0)) +
#   scale_y_continuous(expand=c(0,0))

opticlust_summary %>% mutate(method="OptiClust") %>% 
  bind_rows(optifit_summary %>% mutate(method="OptiFit")) %>% 
  ggplot(aes(x=dx_diff)) +
    geom_histogram(binwidth = 0.05,color="black",fill="grey") +
    labs(title = "Difference in DX Prediction",
         x="Difference in Probability") +
    theme_bw() +
    scale_x_continuous(expand=c(0,0)) +
    scale_y_continuous(expand=c(0,0)) +
    facet_wrap(~method)

```

# Summary of performance with a threshold of 0.5
```{r confusionMatrix}

opticlust_cm <- confusionMatrix(data=factor(opticlust_summary$pred_class),reference = factor(opticlust_summary$meta_class))
optifit_cm <- confusionMatrix(data=factor(optifit_summary$pred_class),reference = factor(optifit_summary$meta_class))

kable(opticlust_cm$table,caption="OptiClust Confusion Matrix") %>% 
  kable_styling(full_width = F,position = "float_left")
kable(optifit_cm$table,caption="OptiFit Confusion Matrix") %>% 
  kable_styling(full_width = F,position = "left")

cm_summary <- round(opticlust_cm$byClass,digits=3) %>% 
  bind_rows(round(optifit_cm$byClass,digits=3)) %>% 
  mutate(Method=c("OptiClust","OptiFit")) %>% 
  select(Method,everything())

kable(cm_summary,align="c",booktabs=TRUE) %>% 
  kable_classic()
```

### ROC plot

```{r roc}

#opticlust
response <- ifelse(opticlust_summary$meta_class == "cancer",1,0)
predictor <- opticlust_summary$cancer

clust_roc <- roc(response,predictor)
opticlust_thr <- coords(clust_roc, "best", ret = "threshold",transpose = TRUE)

#optifit
response <- ifelse(optifit_summary$meta_class == "normal",0,1)
predictor <- optifit_summary$normal

fit_roc <- roc(response,predictor)
optifit_thr <- coords(fit_roc, "best", ret = "threshold",transpose = TRUE)

#plot
ggroc(list(OptiClust=clust_roc,OptiFit=fit_roc),legacy.axes=TRUE) +
  theme_bw() + 
  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1),
                 color="darkgrey", linetype="dashed") +
  scale_y_continuous(expand=c(0,0)) +
  scale_x_continuous(expand=c(0,0))

```

OptiClust AUC: `r round(clust_roc$auc,digits=3)`  
OptiFit AUC: `r round(fit_roc$auc,digits=3)`

Optimal threshold for OptiClust: `r round(opticlust_thr,digits=3)`  
Optimal threshold for OptiFit: `r round(optifit_thr,digits=3)`

### 2021-02-09 exploratory/counts/get_count.R

@@ -1,0 +1,17 @@
+#!/usr/bin/env Rscript
+library(tidyverse)
+
+args = commandArgs(trailingOnly=TRUE)
+
+#file <- read_tsv("/Users/armourc/Desktop/opticlust/data/process/optifit/2003650/in/glne.precluster.pick.subsample.optifit_mcc.shared")
+file <- read_tsv(args[1])
+
+group <- file$Group
+
+count <- file %>% 
+  select(-label,-Group,-numOtus) %>% 
+  rowSums()
+
+results <- as.data.frame(t(c(group,count)))
+
+write_csv(results,paste0("./",group,"_count.csv"),col_names = FALSE)

### 2021-02-09 code/learning/model_pipeline_deployed.R
new results with fraction mapped

@@ -145,7 +145,7 @@ pipeline <- function(dataset, test, model, outcome=NA, hyperparameters=NULL){
   # ---------------------------------------------------------------------------------->
 
   # Calculate the best decision threshold for this model
-  test_roc <- roc(ifelse(testTransformed[,outcome] == first_outcome, 1, 0), rpartProbs[[1]])
+  #test_roc <- roc(ifelse(testTransformed[,outcome] == first_outcome, 1, 0), rpartProbs[[1]])
   #thr <- coords(test_roc, "best", ret = "threshold")
 
   # ----------------------------Save metrics as vector ------------------------------->

### 2021-04-01 code/bash/mothurPrecluster.sh
update filenames in mothurPrecluster, add file check

@@ -36,34 +36,43 @@ if [ -n "$(ls -A "${OUTDIR}")" ]; then
 	rm "${OUTDIR}"/*
 fi
 
-mothur "#make.contigs(file="${FILESFILE}", outputdir="${TMP}"/, processors="${NPROC}");
-	screen.seqs(fasta=current, group=current, maxambig=0, maxlength=275, maxhomop=8);
-	unique.seqs(fasta=current);
-	count.seqs(name=current, group=current);
-	align.seqs(fasta=current, reference="${SILVAV4}");
-	screen.seqs(fasta=current, count=current, start=1968, end=11550);
-	filter.seqs(fasta=current, vertical=T, trump=.);
-	unique.seqs(fasta=current, count=current);
-	pre.cluster(fasta=current, count=current, diffs=2);
-	chimera.vsearch(fasta=current, count=current, dereplicate=T);
-	remove.seqs(fasta=current, accnos=current);
-	classify.seqs(fasta=current, count=current, reference="${RDPFASTA}", taxonomy="${RDPTAX}", cutoff=80);
-	remove.lineage(fasta=current, count=current, taxonomy=current, taxon=Chloroplast-Mitochondria-unknown-Archaea-Eukaryota);
-	dist.seqs(fasta=current,cutoff=0.03)"
+#mothur "#make.contigs(file="${FILESFILE}", outputdir="${TMP}"/, processors="${NPROC}");
+#	screen.seqs(fasta=current, group=current, maxambig=0, maxlength=275, maxhomop=8);
+#	unique.seqs(fasta=current);
+#	count.seqs(name=current, group=current);
+#	align.seqs(fasta=current, reference="${SILVAV4}");
+#	screen.seqs(fasta=current, count=current, start=1968, end=11550);
+#	filter.seqs(fasta=current, vertical=T, trump=.);
+#	unique.seqs(fasta=current, count=current);
+#	pre.cluster(fasta=current, count=current, diffs=2);
+#	chimera.vsearch(fasta=current, count=current, dereplicate=T);
+#	remove.seqs(fasta=current, accnos=current);
+#	classify.seqs(fasta=current, count=current, reference="${RDPFASTA}", taxonomy="${RDPTAX}", cutoff=80);
[…]

### 2021-04-01 code/bash/mothurPrecluster.sh
update mothurPrecluster file check

@@ -36,20 +36,20 @@ if [ -n "$(ls -A "${OUTDIR}")" ]; then
 	rm "${OUTDIR}"/*
 fi
 
-#mothur "#make.contigs(file="${FILESFILE}", outputdir="${TMP}"/, processors="${NPROC}");
-#	screen.seqs(fasta=current, group=current, maxambig=0, maxlength=275, maxhomop=8);
-#	unique.seqs(fasta=current);
-#	count.seqs(name=current, group=current);
-#	align.seqs(fasta=current, reference="${SILVAV4}");
-#	screen.seqs(fasta=current, count=current, start=1968, end=11550);
-#	filter.seqs(fasta=current, vertical=T, trump=.);
-#	unique.seqs(fasta=current, count=current);
-#	pre.cluster(fasta=current, count=current, diffs=2);
-#	chimera.vsearch(fasta=current, count=current, dereplicate=T);
-#	remove.seqs(fasta=current, accnos=current);
-#	classify.seqs(fasta=current, count=current, reference="${RDPFASTA}", taxonomy="${RDPTAX}", cutoff=80);
-#	remove.lineage(fasta=current, count=current, taxonomy=current, taxon=Chloroplast-Mitochondria-unknown-Archaea-Eukaryota);
-#	dist.seqs(fasta=current,cutoff=0.03)"
+mothur "#make.contigs(file="${FILESFILE}", outputdir="${TMP}"/, processors="${NPROC}");
+	screen.seqs(fasta=current, group=current, maxambig=0, maxlength=275, maxhomop=8);
+	unique.seqs(fasta=current);
+	count.seqs(name=current, group=current);
+	align.seqs(fasta=current, reference="${SILVAV4}");
+	screen.seqs(fasta=current, count=current, start=1968, end=11550);
+	filter.seqs(fasta=current, vertical=T, trump=.);
+	unique.seqs(fasta=current, count=current);
+	pre.cluster(fasta=current, count=current, diffs=2);
+	chimera.vsearch(fasta=current, count=current, dereplicate=T);
+	remove.seqs(fasta=current, accnos=current);
+	classify.seqs(fasta=current, count=current, reference="${RDPFASTA}", taxonomy="${RDPTAX}", cutoff=80);
[…]

### 2021-04-03 exploratory/exploratory.Rmd

@@
 
-counts <- read_csv("./counts/all_counts.csv",col_names = c("sample","count"))
+fraction_mapped <- read_tsv("../results/tables/fraction_reads_mapped.tsv")
 
-counts %>% 
-  mutate(pct_mapped = count/10000) %>% 
-  ggplot(aes(x=pct_mapped)) +
+fraction_mapped %>% 
+  ggplot(aes(x=fraction_mapped)) + 
     geom_histogram(fill="grey",color="black") +
@@
   scale_x_continuous(labels=scales::percent) +
-    xlab("Percent of reads mapped")
+  xlab("Percent of reads mapped with cluster.fit") +
+  ylab("number of samples")
 
-# counts %>% 
-#   mutate(pct_mapped = count/10000) %>% 
-#   ggplot(aes(x="",y=pct_mapped)) +
-#     geom_boxplot(fill="grey",color="black") +
-#     theme_bw() +
-#     xlab("")
 ```
@@
 
-optifit_in_mcc <- read_tsv("./optifit_in_mcc.tsv")
-optifit_out_mcc <- read_tsv("./optifit_out_mcc.tsv")
-opticlust_mcc <- read_tsv("./opticlust_mcc.tsv")
+sensspec <- read_csv("../results/tables/merged_sensspec.csv")
 
-mcc_data <- tibble(type=c(rep("OptiClust_489",nrow(optifit_out_mcc)),
-                          rep("OptiFit_1",nrow(optifit_in_mcc)),
-                          rep("OptiClust_all",nrow(opticlust_mcc))),
-                   mcc=c(optifit_out_mcc$mcc,optifit_in_mcc$mcc,opticlust_mcc$mcc))
+mcc_data <- sensspec %>% 
+  select(type,sample,mcc) %>% 
+  mutate(type=factor(type,levels=c("opticlust","optifit_out","optifit_in"),labels=c("OptiClust_all","OptiClust_489","OptiFit_1")))
 
-mcc_data$type <- factor(mcc_data$type,levels=c("OptiClust_all","OptiClust_489","OptiFit_1"))
+#mcc_data$type <- factor(mcc_data$type,levels=c("OptiClust_all","OptiClust_489","OptiFit_1"))
 mcc_data %>% 

### 2021-07-02 code/slurm/snakemake.sh
config files

@@ -19,19 +19,19 @@
 #SBATCH --time=5-00:00:00
 
 # Account
-#SBATCH --account=ACCOUNT
+#SBATCH --account=pschloss1
 #SBATCH --partition=standard
 
 # Logs
-#SBATCH --mail-user=EMAIL
-#SBATCH --mail-type=BEGIN,END,FAIL
+#SBATCH --mail-user=armourc@umich.edu
+#SBATCH --mail-type=FAIL
 #SBATCH --output=%x-%j.out
 
 # Environment
 #SBATCH --export=ALL
 
 # Connecting node to internet
-source /etc/profile.d/http_proxy.sh 
+source /etc/profile.d/http_proxy.sh
 
 # List compute nodes allocated to the job
 if [[ $SLURM_JOB_NODELIST ]] ; then

### 2021-07-02 exploratory/exploratory.Rmd

@@
 
+med <- median(fraction_mapped$fraction_mapped)
+med_frac <- paste0(100*median(fraction_mapped$fraction_mapped),"%")
+min_frac <- paste0(100*min(fraction_mapped$fraction_mapped),"%")
+max_frac <- paste0(100*max(fraction_mapped$fraction_mapped),"%")
+
 fraction_mapped %>% 
@@
   xlab("Percent of reads mapped with cluster.fit") +
-  ylab("number of samples")
+  ylab("number of samples") +
+  geom_vline(xintercept = med,lty="dashed",lwd=1) +
+  annotate("text", x=.90, y=55, label= paste0("Median: ",med_frac)) +   
+  annotate("text", x=.90, y=50, label= paste0("Max: ",max_frac)) +
+  annotate("text", x=.90, y=45, label= paste0("Min: ",min_frac))
 
@@
   ggplot(aes(x=type,y=mcc,color=type)) + 
-    geom_jitter(height=0) +
+    geom_jitter(height=0,size=3,alpha=0.6) +
     xlab("") +
-    theme_bw()
+    theme_bw() +
+    stat_summary(fun=median,geom="crossbar",color="black",lwd=0.3)
 

### 2021-11-16 code/bash/mothurOptiFitLOO.sh
remove LOO scripts

@@ -1,41 +1,0 @@
-#! /bin/bash
-# mothurLOO.sh
-# William L. Close
-# Begum Topcuoglu
-# Schloss Lab
-# University of Michigan
-
-##################
-# Set Script Env #
-##################
-
-# Set the variables to be used in this script
-FASTA=${1:?ERROR: Need to define FASTA.} # Preclustered fasta file
-COUNT=${2:?ERROR: Need to define COUNT.} # Preclustered count file
-TAXONOMY=${3:?ERROR: Need to define TAXONOMY.} # Preclustered tax file
-DIST=${4:?ERROR: Need to define DIST.} # Preclustered dist file
-SAMPLE=${5:?ERROR: Need to define SAMPLE.} # Sample to be removed
-
-# Other variables
-OUTDIR=data/process/optifit/"${SAMPLE}"/out/ # Output dir based on sample name to keep things separate during parallelization/organized
-NPROC=$(nproc) # Setting number of processors to use based on available resources
-SUBSIZE=10000 # Number of reads to subsample to, based on Baxter, et al., Genome Med, 2016
-
-############################################
-# Generate Shared After Leaving Sample Out #
-############################################
-
-# Make output dir if it doesn't exist
-mkdir -p "${OUTDIR}"/
-
-# Removing old files if they exist
-if [ -n "$(ls -A "${OUTDIR}")" ]; then
-	rm "${OUTDIR}"/*
-fi
-
-# Cluster all sequences after leaving out the specified sample
-mothur "#set.current(outputdir="${OUTDIR}"/, processors="${NPROC}");
-	remove.groups(fasta="${FASTA}", count="${COUNT}", taxonomy="${TAXONOMY}",  groups="${SAMPLE}", column="${DIST}");
-	cluster(column=current, count=current);
-	make.shared(list=current, count=current, label=0.03);
-	sub.sample(shared=current, label=0.03, size="${SUBSIZE}")"

### 2021-11-16 code/bash/mothurClusterOptiFit.sh
add scripts to cluster and fit with optifit

@@ -1,0 +1,62 @@
+#! /bin/bash
+# mothurOpitFit.sh
+# Courtney R. Armour
+# William L. Close
+# Begum Topcuoglu
+# Schloss Lab
+# University of Michigan
+
+##################
+# Set Script Env #
+##################
+
+# Set the variables to be used in this script
+FASTA=${1:?ERROR: Need to define FASTA.} # Preclustered fasta file
+COUNT=${2:?ERROR: Need to define COUNT.} # Preclustered count file
+TAXONOMY=${3:?ERROR: Need to define TAXONOMY.} # Preclustered tax file
+DIST=${4:?ERROR: Need to define DIST.} # Preclustered distance file
+SPLIT=${5:?ERROR: Need to define SPLIT.} #file defining test/train split
+NUM=`basename $SPLIT .csv`
+
+# Other variables
+OUTDIR=data/process/optifit/$NUM/train/ # Output dir based on split number
+NPROC=$(nproc) # Setting number of processors to use based on available resources
+SUBSIZE=10000 # Number of reads to subsample to, based on Baxter, et al., Genome Med, 2016
+
+###################
+### GROUP SETUP ###
+###################
+
+#initialize train arrays
+trainIDS=()
+
+#assign IDS to train array
+{
+    read
+    while IFS="," read -a line; do    
+        if [[ "${line[1]}" == "train" ]]; then
+            trainIDS+="${line[0]}-"
+        fi
+    done 
+}< "$SPLIT"
+
+echo "training: ${trainIDS%?}" # %? removes last character (extra "-")
+
+############################################
+# Generate Shared After Leaving Sample Out #
+############################################
+
+# Make output dir if it doesn't exist
+mkdir -p "${OUTDIR}"/
+
+# Removing old files if they exist
+if [ -n "$(ls -A "${OUTDIR}")" ]; then
+	rm "${OUTDIR}"/*
+fi
+
+# Cluster all sequences for the training set
+mothur "#set.current(outputdir="${OUTDIR}"/, processors="${NPROC}");
+	get.groups(fasta="${FASTA}", count="${COUNT}", taxonomy="${TAXONOMY}",  groups=${trainIDS%?}, column="${DIST}");
+	cluster(column=current, count=current);
+	make.shared(list=current, count=current, label=0.03);
+	sub.sample(shared=current, label=0.03, size="${SUBSIZE}")"

### 2021-11-16 code/bash/mothurFitOptiFit.sh

@@ -1,0 +1,70 @@
+#! /bin/bash
+# mothurFitOptiFit.sh
+# Courtney R. Armour
+# William L. Close
+# Begum Topcuoglu
+# Schloss Lab
+# University of Michigan
+
+##################
+# Set Script Env #
+##################
+
+# Set the variables to be used in this script
+FASTA=${1:?ERROR: Need to define FASTA.} # Preclustered fasta file
+COUNT=${2:?ERROR: Need to define COUNT.} # Preclustered count file
+TAXONOMY=${3:?ERROR: Need to define TAXONOMY.} # Preclustered tax file
+DIST=${4:?ERROR: Need to define DIST.} # Preclustered distance file
+SPLIT=${5:?ERROR: Need to define SPLIT.} # Split of test and training sets
+REFFASTA=${6:?ERROR: Need to define TRAINFASTA.} # Reference fasta file with test data removed
+REFDIST=${7:?ERROR: Need to define TRAINDIST.} # Reference dist file with test data removed
+REFLIST=${8:?ERROR: Need to define TRAINLIST.} # Reference list file with test data removed
+
+# Other variables
+NUM=`basename $SPLIT .csv`
+OUTDIR=data/process/optifit/$NUM/test/ # Output dir based on split number 
+NPROC=$(nproc) # Setting number of processors to use based on available resources
+SUBSIZE=10000 # Number of reads to subsample to, based on Baxter, et al., Genome Med, 2016
+
+echo $OUTDIR
+
+###################
+### GROUP SETUP ###
+###################
+
+#initialize test arrays
+testIDS=()
+
+#assign IDS to test array
+{
+    read
+    while IFS="," read -a line; do    
+        if [[ "${line[1]}" == "test" ]]; then
+            testIDS+="${line[0]}-"
+        fi
+    done 
+}< "$SPLIT"
+
+echo "testing: ${testIDS%?}" # %? removes last character (extra "-")
+
+##########################################
+# Clustering the test data Using OptiFit #
+##########################################
+
+# Make output dir if it doesn't exist
+mkdir -p "${OUTDIR}"/
+
+# Removing old files if they exist
+if [ -n "$(ls -A "${OUTDIR}")" ]; then
+	rm "${OUTDIR}"/*
+fi
+
+# Selecting test data to be fit, subsampling input files to account for 'closed' clustering method, and fitting sample to previous shared file
[…]

### 2021-11-16 code/bash/splitOptiClust.sh
script to split opticlust data

@@ -1,0 +1,73 @@
+#! /bin/bash
+# splitOptiClust.sh
+# Courtney Armour
+# Schloss Lab
+# University of Michigan
+
+# Purpose: use split file to create train/test shared files
+
+#argvals=("$@")
+
+SHARED=${1:?ERROR: Need to define SHARED.}
+SPLIT=${2:?ERROR: Need to define SPLITS.} 
+NUM=`basename $SPLIT .csv`
+
+#SHARED=${argvals[0]}
+#SPLITS=(${argvals[@]:1}) 
+
+OUTDIR=data/process/opticlust/$NUM/ # Output dir based on sample name to keep things separate during parallelization/organized
+NPROC=$(nproc) # Setting number of processors to use based on available resources
+
+###################
+### GROUP SETUP ###
+###################
+
+#initialize train and test arrays
+trainIDS=()
+testIDS=()
+
+#assign IDS to train or test array
+
+#for f in ${SPLITS[@]}
+#do
+    #echo "processing $f"
+{
+    read
+    while IFS="," read -a line; do    
+        if [[ "${line[1]}" == "train" ]]; then
+            trainIDS+="${line[0]}-"
+        else
+            testIDS+="${line[0]}-"
+        fi
+    done 
+}< "$SPLIT"
+#done
+echo "training: ${trainIDS%?}" # %? removes last character (extra "-")
+echo "testing: ${testIDS%?}"
+
+##############
+### MOTHUR ###
+##############
+
+# Make output dirs if they don't exist
+mkdir -p "${OUTDIR}"/ "${OUTDIR}"/train/ "${OUTDIR}"/test/
+
+### TRAINING DATA
+# Removing old files if they exist
+if [ -n "$(ls -A "${OUTDIR}"/train)" ]; then
+	rm "${OUTDIR}"/train/*
+fi
+
+# # Create OptiClust shared file for training data
+ mothur "#set.current(outputdir="${OUTDIR}"/train/, processors="${NPROC}");
+ 	get.groups(shared="${SHARED}", groups=${trainIDS%?})"
+     
+### TESTING DATA
+# Removing old files if they exist
+if [ -n "$(ls -A "${OUTDIR}"/test)" ]; then
+	rm "${OUTDIR}"/test/*
+fi
+
+# Create OptiClust shared file for training data
+ mothur "#set.current(outputdir="${OUTDIR}"/test/, processors="${NPROC}");
+ 	get.groups(shared="${SHARED}", groups=${testIDS%?})"

### 2021-11-16 code/R/run_model.R
code to run mikropml for 80 20 splits

@@ -1,0 +1,193 @@
+##################################################################
+# Author: Courtney Armour (adapted from Begum main.R script)
+# Date: November 2021
+# Description: Script to run mikropml pipeline
+#
+##################################################################
+
+library(mikropml)
+library(tidyverse)
+library(caret)
+
+# User defined variables
+input <- commandArgs(trailingOnly=TRUE)
+
+metadata <- input[1] # Metadata containing classification to predict
+training <- input[2] # Subsampled shared from training samples 
+testing  <- input[3] # Subsampled shared from testing samples 
+model <- input[4] # Type of model to use
+outcome <- input[5] # Classifaction to predict
+
+#for testing the pipeline
+# metadata <- "data/metadata/metadata.tsv"
+# # training <- "data/process/opticlust/split_1/train/glne.precluster.opti_mcc.0.03.subsample.0.03.pick.shared"
+# training <- "data/process/optifit/split_5/train/glne.precluster.pick.opti_mcc.0.03.subsample.shared"
+# # testing  <- "data/process/opticlust/split_1/test/glne.precluster.opti_mcc.0.03.subsample.0.03.pick.shared"
+# testing <- "data/process/optifit/split_5/test/glne.precluster.pick.subsample.renamed.fit.optifit_mcc.shared"
+# model <- "glmnet"
+# outcome <- "dx"
+
+split <- unlist(str_split(training,"/"))[4]
+
+print(paste0("metadata file: ",metadata))
+print(paste0("training file: ",training))
+print(paste0("testing file: ",testing))
+print(paste0("model: ",model))
+print(paste0("outcome: ",outcome))
+print(paste0("split: ",split))
+
+# Other variables
+if (str_detect(training, "optifit")) { # Setting output dir based on source of input
+  outDir <- "data/learning/results/optifit/"
+} else {
+  outDir <- "data/learning/results/opticlust/"
+}
+
+######################## DATA PREPARATION ########################
+# Features: 16S rRNA gene sequences(OTUs) in the stool
+
+message("PROGRESS: Preparing data.")
+
+# Reading in shared file from training data
+train_shared <- read_tsv(training, col_types = cols(Group=col_character(),
+                                                    .default = col_double())) %>%
+  select(-label, -numOtus)
+
+# Reading in shared file of the testing data
+if (str_detect(testing, "optifit")) { 
+  test_shared <- read_tsv(testing, col_types = cols(Group=col_character(),
[…]

### 2021-11-16 code/R/view_splits.R
script to visualize data splits

@@ -1,0 +1,28 @@
+library(tidyverse)
+
+outdir <- "analysis/"
+if(!dir.exists(outdir)){dir.create(outdir)}
+
+files <- commandArgs(trailingOnly = TRUE)
+
+split_files <- files  %>% 
+    lapply(read_csv)  %>% 
+    bind_rows()  %>% 
+    mutate(group=as.character(group))
+
+counts <- split_files  %>% 
+    group_by(group)  %>% 
+    count(train_test)  
+
+counts  %>% 
+    ggplot(aes(x=train_test,y=n,color=train_test)) +
+        geom_jitter(size=2,alpha=0.5,width=0.4,height = 0) +
+        theme_classic() +
+        scale_y_continuous(limits=c(0,100),breaks=seq(0,100,10)) +
+        theme(panel.grid.major.y = element_line(color="grey"),
+              axis.text = element_text(size=14),
+              axis.title = element_text(size=14),
+              legend.text = element_text(size=14),
+              legend.title = element_text(size=14)) + 
+        xlab("Group") + ylab("Number of times sample is in group") 
+ggsave("analysis/view_splits.pdf",width = 6, height= 6)

### 2021-11-16 code/R/generate_splits.R
script to create train test splits

@@ -1,0 +1,42 @@
+library(tidyverse)
+
+### FUNCTIONS #############
+create_split <- function(groups,train_frac,split_num){
+    
+    groups_shuffled <- sample(groups,length(groups))
+    
+    train_n <- as.integer(length(groups)*train_frac)
+    test_n <- length(groups) - train_n
+    
+    data <- tibble(group=groups_shuffled,
+                   train_test=c(rep("train",train_n),
+                                rep("test",test_n)))  %>% 
+        arrange(group)
+    
+    return(data)
+}
+
+### VARIABLES ################
+outdir <- "data/process/splits/"
+if(!dir.exists(outdir)){dir.create(outdir)}
+
+# Parsing command line arguments
+args <- commandArgs(trailingOnly=TRUE)
+
+# Variables defined by user
+shared <- read_tsv(args[1],col_types = cols(Group=col_character(),
+                                            .default=col_double()))
+
+n_splits <- args[2]
+
+### MAIN ########################
+# get the list of sampels from the shared file
+groups <- shared  %>% 
+    pull(Group)
+
+#generate n random 80/20 splits, write files to outdir
+set.seed(2021)
+for(i in 1:n_splits){
+    create_split(groups,0.8,1)  %>% 
+        write_csv(.,paste0(outdir,"split_",i,".csv"))
+}

### 2021-11-16 code/bash/mothurPrecluster.sh
update paths

@@ -53,16 +53,10 @@ mothur "#make.contigs(file="${FILESFILE}", outputdir="${TMP}"/, processors="${NP
 
 
 # Moving and renaming important files for downstream use
-#mv "${TMP}"/*.precluster.pick.pick.fasta "${OUTDIR}"/glne.precluster.fasta
-#mv "${TMP}"/*.precluster.denovo.vsearch.pick.pick.count_table "${OUTDIR}"/glne.precluster.count_table
-#mv "${TMP}"/*.precluster.pick.pds.wang.pick.taxonomy "${OUTDIR}"/glne.precluster.taxonomy
-#mv "${TMP}"/*.precluster.pick.pick.dist "${OUTDIR}"/glne.precluster.dist
-
-#new filenames with 1.45
-mv "${TMP}"/*.precluster.pick.fasta "${OUTDIR}"/glne.precluster.fasta
-mv "${TMP}"/*.precluster.pick.count_table "${OUTDIR}"/glne.precluster.count_table
-mv "${TMP}"/*.precluster.pds.wang.taxonomy "${OUTDIR}"/glne.precluster.taxonomy
-mv "${TMP}"/*.precluster.pick.dist "${OUTDIR}"/glne.precluster.dist
+mv "${TMP}"/*.precluster.pick.pick.fasta "${OUTDIR}"/glne.precluster.fasta
+mv "${TMP}"/*.precluster.denovo.vsearch.pick.pick.count_table "${OUTDIR}"/glne.precluster.count_table
+mv "${TMP}"/*.precluster.pick.pds.wang.pick.taxonomy "${OUTDIR}"/glne.precluster.taxonomy
+mv "${TMP}"/*.precluster.pick.pick.dist "${OUTDIR}"/glne.precluster.dist
 
 
 ###############

### 2021-11-16 exploratory/exploratory.Rmd
add split plot to exploratory

--- 2021-04-03 exploratory/exploratory.Rmd
@@
 author: "Courtney R Armour"
-date: "February 2021"
+date: "Updated: `r Sys.Date()`"
 output: html_document
@@
 library(tidyverse)
-library(caret)
 library(kableExtra)
-library(pROC)
+#library(caret)
+#library(pROC)
+```
+### Table of Contents
+
+-   [Project Summary](#summary)
+-   [Overview of data splits](#splits)
+
+### Project Summary {#summary}
+...explanation of project....
+
+### Overview of data splits {#splits}
+
+Since the random 80% training 20% testing splits are generated 100 times, we anticipate that if the samples are randomly assigned each one should appear in the training set about 80 of the 100 times (80%) and testing about 20 of the 100 times (20%).
+The plot below visualizes how many times each sample appears in the training and testing datasets.
+
+```{r plot_splits}
+knitr::include_graphics("../analysis/view_splits.png")
+```
+### Percent of reads mapped with OptiFit
+
+...in progress...
+
+### Model performance during training
+
+```{r cv_auc}
+
+optifit_cv_files <- list.files(path="../data/learning/results/optifit/",
+                               pattern="cv_results_split_.*.csv",
+                               full.path=TRUE)
+
 ```
 
-## Percent of reads mapped with OptiFit
+### ...THE REST IS MISC LEFTOVER FROM PREVIOUS ANALYSIS...
+
+### Percent of reads mapped with OptiFit
 
@@
 
-fraction_mapped <- read_tsv("../results/tables/fraction_reads_mapped.tsv")
+# fraction_mapped <- read_tsv("../results/tables/fraction_reads_mapped.tsv")
 
-fraction_mapped %>% 
-  ggplot(aes(x=fraction_mapped)) + 
-    geom_histogram(fill="grey",color="black") +
-  theme_bw() + 
-  scale_x_continuous(labels=scales::percent) +
-  xlab("Percent of reads mapped with cluster.fit") +
-  ylab("number of samples")
+# med <- median(fraction_mapped$fraction_mapped)
+# med_frac <- paste0(100*median(fraction_mapped$fraction_mapped),"%")
+# min_frac <- paste0(100*min(fraction_mapped$fraction_mapped),"%")
+# max_frac <- paste0(100*max(fraction_mapped$fraction_mapped),"%")
+
+# fraction_mapped %>% 
+#   ggplot(aes(x=fraction_mapped)) + 
+#     geom_histogram(fill="grey",color="black") +
+#   theme_bw() + 
+#   scale_x_continuous(labels=scales::percent) +
+#   xlab("Percent of reads mapped with cluster.fit") +
+#   ylab("number of samples") +
+#   geom_vline(xintercept = med,lty="dashed",lwd=1) +
+#   annotate("text", x=.90, y=55, label= paste0("Median: ",med_frac)) +   
+#   annotate("text", x=.90, y=50, label= paste0("Max: ",max_frac)) +
+#   annotate("text", x=.90, y=45, label= paste0("Min: ",min_frac))
 
@@
 
-## MCC OptiFit vs OptiClust
+### MCC OptiFit vs OptiClust
 
@@
 
-sensspec <- read_csv("../results/tables/merged_sensspec.csv")
+# sensspec <- read_csv("../results/tables/merged_sensspec.csv")
 
-mcc_data <- sensspec %>% 
-  select(type,sample,mcc) %>% 
-  mutate(type=factor(type,levels=c("opticlust","optifit_out","optifit_in"),labels=c("OptiClust_all","OptiClust_489","OptiFit_1")))
+# mcc_data <- sensspec %>% 
+#   select(type,sample,mcc) %>% 
+#   mutate(type=factor(type,levels=c("opticlust","optifit_out","optifit_in"),labels=c("OptiClust_all","OptiClust_489","OptiFit_1")))
 
-#mcc_data$type <- factor(mcc_data$type,levels=c("OptiClust_all","OptiClust_489","OptiFit_1"))
-mcc_data %>% 
-  ggplot(aes(x=type,y=mcc,color=type)) + 
-    geom_jitter(height=0) +
-    xlab("") +
-    theme_bw()
+# #mcc_data$type <- factor(mcc_data$type,levels=c("OptiClust_all","OptiClust_489","OptiFit_1"))
+# mcc_data %>% 
+#   ggplot(aes(x=type,y=mcc,color=type)) + 
+#     geom_jitter(height=0,size=3,alpha=0.6) +
+#     xlab("") +
+#     theme_bw() +
+#     stat_summary(fun=median,geom="crossbar",color="black",lwd=0.3)
 
@@
 
-# Model performance during training
+### Model performance during training
 ```{r data}
 
-optifit_summary <- read_tsv("../data/learning/summary/optifit/model_results.tsv",
-                            col_types = cols(sample = col_character(),
-                                             cv_auc = col_double(),
-                                             cancer = col_double(),
-                                             normal = col_double(),
-                                             dx_diff = col_double(),
-                                             dx_flag = col_factor(),
-                                             pred_class = col_number(),
-                                             meta_class = col_number()
-                                             ))  %>% 
-  mutate(pred_class = case_when(pred_class == 0 ~ "normal",
-                                TRUE ~ "cancer")) %>% 
-  mutate(meta_class = case_when(meta_class == 0 ~ "normal",
-                                TRUE ~ "cancer"))
+# optifit_summary <- read_tsv("../data/learning/summary/optifit/model_results.tsv",
+#                             col_types = cols(sample = col_character(),
+#                                              cv_auc = col_double(),
+#                                              cancer = col_double(),
+#                                              normal = col_double(),
+#                                              dx_diff = col_double(),
+#                                              dx_flag = col_factor(),
+#                                              pred_class = col_number(),
+#                                              meta_class = col_number()
+#                                              ))  %>% 
+#   mutate(pred_class = case_when(pred_class == 0 ~ "normal",
+#                                 TRUE ~ "cancer")) %>% 
+#   mutate(meta_class = case_when(meta_class == 0 ~ "normal",
+#                                 TRUE ~ "cancer"))
 
-opticlust_summary <- read_tsv("../data/learning/summary/opticlust/model_results.tsv",
-                              col_types = cols(sample = col_character(),
-                                               cv_auc = col_double(),
-                                               cancer = col_double(),
-                                               normal = col_double(),
-                                               dx_diff = col_double(),
-                                               dx_flag = col_factor(),
-                                               pred_class = col_factor(),
-                                               meta_class = col_factor()
-                                               )) %>% 
-  mutate(pred_class = case_when(pred_class == 0 ~ "normal",
-                                TRUE ~ "cancer")) %>% 
-  mutate(meta_class = case_when(meta_class == 0 ~ "normal",
-                                TRUE ~ "cancer"))
+# opticlust_summary <- read_tsv("../data/learning/summary/opticlust/model_results.tsv",
+#                               col_types = cols(sample = col_character(),
+#                                                cv_auc = col_double(),
+#                                                cancer = col_double(),
+#                                                normal = col_double(),
+#                                                dx_diff = col_double(),
+#                                                dx_flag = col_factor(),
+#                                                pred_class = col_factor(),
+#                                                meta_class = col_factor()
+#                                                )) %>% 
+#   mutate(pred_class = case_when(pred_class == 0 ~ "normal",
+#                                 TRUE ~ "cancer")) %>% 
+#   mutate(meta_class = case_when(meta_class == 0 ~ "normal",
+#                                 TRUE ~ "cancer"))
 
-opticlust_summary %>% mutate(method="OptiClust") %>% 
-  bind_rows(optifit_summary %>% mutate(method="OptiFit")) %>% 
-  ggplot(aes(x=method,y=cv_auc,fill=method)) +
-    geom_boxplot() +
-    #geom_jitter(height=0,alpha = 0.5) +
-    labs(title = "Cross Validation Performance") +
-    theme_bw() +
-    theme(legend.position = "none") 
-ggsave("./figures/cv_auc_performance.png")
+# opticlust_summary %>% mutate(method="OptiClust") %>% 
+#   bind_rows(optifit_summary %>% mutate(method="OptiFit")) %>% 
+#   ggplot(aes(x=method,y=cv_auc,fill=method)) +
+#     geom_boxplot() +
+#     #geom_jitter(height=0,alpha = 0.5) +
+#     labs(title = "Cross Validation Performance") +
+#     theme_bw() +
+#     theme(legend.position = "none") 
+# ggsave("./figures/cv_auc_performance.png")
 
@@
 
-# Prediction Probability 
+### Prediction Probability 
 ```{r }
 
-opticlust_summary %>% 
-  ggplot(aes(x=cancer,fill=meta_class)) +
-    geom_histogram(color="black",alpha=0.8,binwidth = 0.05) + theme_bw() +
-    labs(xlab = "Probability of Cancer") + 
-    facet_wrap(~meta_class) +
-    xlab("probability of cancer") +
-    geom_vline(xintercept = 0.5,lty="dashed") +
-    scale_y_continuous(expand=c(0,0)) +
-    scale_fill_manual(values=c("red","blue"),guide=NULL) +
-    ggtitle("OptiClust")
+# opticlust_summary %>% 
+#   ggplot(aes(x=cancer,fill=meta_class)) +
+#     geom_histogram(color="black",alpha=0.8,binwidth = 0.05) + theme_bw() +
+#     labs(xlab = "Probability of Cancer") + 
+#     facet_wrap(~meta_class) +
+#     xlab("probability of cancer") +
+#     geom_vline(xintercept = 0.5,lty="dashed") +
+#     scale_y_continuous(expand=c(0,0)) +
+#     scale_fill_manual(values=c("red","blue"),guide=NULL) +
+#     ggtitle("OptiClust")
 
-optifit_summary %>% 
-  ggplot(aes(x=cancer,fill=meta_class)) +
-    geom_histogram(color="black",alpha=0.8,binwidth = 0.05) + theme_bw() +
-    labs(xlab = "Probability of Cancer") + 
-    facet_wrap(~meta_class) +
-    xlab("probability of cancer") +
-    geom_vline(xintercept = 0.5,lty="dashed") +
-    scale_y_continuous(expand=c(0,0)) +
-    scale_fill_manual(values=c("red","blue"),guide=NULL) +
-    ggtitle("OptiFit")
+# optifit_summary %>% 
+#   ggplot(aes(x=cancer,fill=meta_class)) +
+#     geom_histogram(color="black",alpha=0.8,binwidth = 0.05) + theme_bw() +
+#     labs(xlab = "Probability of Cancer") + 
+#     facet_wrap(~meta_class) +
+#     xlab("probability of cancer") +
+#     geom_vline(xintercept = 0.5,lty="dashed") +
+#     scale_y_continuous(expand=c(0,0)) +
+#     scale_fill_manual(values=c("red","blue"),guide=NULL) +
+#     ggtitle("OptiFit")
 
@@
 
-opticlust_summary %>% mutate(method="OptiClust") %>% 
-  bind_rows(optifit_summary %>% mutate(method="OptiFit")) %>% 
-  ggplot(aes(x=dx_diff)) +
-    geom_histogram(binwidth = 0.05,color="black",fill="grey") +
-    labs(title = "Difference in DX Prediction",
-         x="Difference in Probability") +
-    theme_bw() +
-    scale_x_continuous(expand=c(0,0)) +
-    scale_y_continuous(expand=c(0,0)) +
-    facet_wrap(~method)
+# opticlust_summary %>% mutate(method="OptiClust") %>% 
+#   bind_rows(optifit_summary %>% mutate(method="OptiFit")) %>% 
+#   ggplot(aes(x=dx_diff)) +
+#     geom_histogram(binwidth = 0.05,color="black",fill="grey") +
+#     labs(title = "Difference in DX Prediction",
+#          x="Difference in Probability") +
+#     theme_bw() +
+#     scale_x_continuous(expand=c(0,0)) +
+#     scale_y_continuous(expand=c(0,0)) +
+#     facet_wrap(~method)
 
@@
 
-# Summary of performance with a threshold of 0.5
+### Summary of performance with a threshold of 0.5
 ```{r confusionMatrix}
 
-opticlust_cm <- confusionMatrix(data=factor(opticlust_summary$pred_class),reference = factor(opticlust_summary$meta_class))
-optifit_cm <- confusionMatrix(data=factor(optifit_summary$pred_class),reference = factor(optifit_summary$meta_class))
+# opticlust_cm <- confusionMatrix(data=factor(opticlust_summary$pred_class),reference = factor(opticlust_summary$meta_class))
+# optifit_cm <- confusionMatrix(data=factor(optifit_summary$pred_class),reference = factor(optifit_summary$meta_class))
 
-kable(opticlust_cm$table,caption="OptiClust Confusion Matrix") %>% 
-  kable_styling(full_width = F,position = "float_left")
-kable(optifit_cm$table,caption="OptiFit Confusion Matrix") %>% 
-  kable_styling(full_width = F,position = "left")
+# kable(opticlust_cm$table,caption="OptiClust Confusion Matrix") %>% 
+#   kable_styling(full_width = F,position = "float_left")
+# kable(optifit_cm$table,caption="OptiFit Confusion Matrix") %>% 
+#   kable_styling(full_width = F,position = "left")
 
-cm_summary <- round(opticlust_cm$byClass,digits=3) %>% 
-  bind_rows(round(optifit_cm$byClass,digits=3)) %>% 
-  mutate(Method=c("OptiClust","OptiFit")) %>% 
-  select(Method,everything())
+# cm_summary <- round(opticlust_cm$byClass,digits=3) %>% 
+#   bind_rows(round(optifit_cm$byClass,digits=3)) %>% 
+#   mutate(Method=c("OptiClust","OptiFit")) %>% 
+#   select(Method,everything())
 
-kable(cm_summary,align="c",booktabs=TRUE) %>% 
-  kable_classic()
+# kable(cm_summary,align="c",booktabs=TRUE) %>% 
+#   kable_classic()
 ```
@@
 #opticlust
-response <- ifelse(opticlust_summary$meta_class == "cancer",1,0)
-predictor <- opticlust_summary$cancer
+# response <- ifelse(opticlust_summary$meta_class == "cancer",1,0)
+# predictor <- opticlust_summary$cancer
 
-clust_roc <- roc(response,predictor)
-opticlust_thr <- coords(clust_roc, "best", ret = "threshold",transpose = TRUE)
+# clust_roc <- roc(response,predictor)
+# opticlust_thr <- coords(clust_roc, "best", ret = "threshold",transpose = TRUE)
 
-#optifit
-response <- ifelse(optifit_summary$meta_class == "normal",0,1)
-predictor <- optifit_summary$normal
+# #optifit
+# response <- ifelse(optifit_summary$meta_class == "normal",0,1)
+# predictor <- optifit_summary$normal
 
-fit_roc <- roc(response,predictor)
-optifit_thr <- coords(fit_roc, "best", ret = "threshold",transpose = TRUE)
+# fit_roc <- roc(response,predictor)
+# optifit_thr <- coords(fit_roc, "best", ret = "threshold",transpose = TRUE)
 
-#plot
-ggroc(list(OptiClust=clust_roc,OptiFit=fit_roc),legacy.axes=TRUE) +
-  theme_bw() + 
-  geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1),
-                 color="darkgrey", linetype="dashed") +
-  scale_y_continuous(expand=c(0,0)) +
-  scale_x_continuous(expand=c(0,0))
+# #plot
+# ggroc(list(OptiClust=clust_roc,OptiFit=fit_roc),legacy.axes=TRUE) +
+#   theme_bw() + 
+#   geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1),
+#                  color="darkgrey", linetype="dashed") +
+#   scale_y_continuous(expand=c(0,0)) +
+#   scale_x_continuous(expand=c(0,0))
 
@@
 
-OptiClust AUC: `r round(clust_roc$auc,digits=3)`  
-OptiFit AUC: `r round(fit_roc$auc,digits=3)`
+```{r}
+# OptiClust AUC: `r round(clust_roc$auc,digits=3)`  
+# OptiFit AUC: `r round(fit_roc$auc,digits=3)`
 
-Optimal threshold for OptiClust: `r round(opticlust_thr,digits=3)`  
-Optimal threshold for OptiFit: `r round(optifit_thr,digits=3)`
+# Optimal threshold for OptiClust: `r round(opticlust_thr,digits=3)`  
+# Optimal threshold for OptiFit: `r round(optifit_thr,digits=3)`  
+```

### 2021-11-16 code/R/view_splits.R
change pdf to png

@@ -25,4 +25,4 @@ counts  %>%
               legend.text = element_text(size=14),
               legend.title = element_text(size=14)) + 
         xlab("Group") + ylab("Number of times sample is in group") 
-ggsave("analysis/view_splits.pdf",width = 6, height= 6)
+ggsave("analysis/view_splits.png",width = 6, height= 6)

### 2021-11-17 code/R/run_model.R
cleanup

@@ -75,7 +75,6 @@ train_labels <- names(train_shared)
 test_labels <- names(test_shared)
 
 # Finding list of otus common to both shared files
-# common_labels <- intersect(opti_labels, loo_labels)
 common_labels <- intersect(train_labels,test_labels)
 
 # Finding otus that are missing in sample shared
@@ -110,8 +109,8 @@ data <- inner_join(train_shared, meta, by=c("Group"="sample")) %>%
                         Dx_Bin == "Cancer" ~ "cancer",
                         TRUE ~ NA_character_),
          dx = as.factor(dx)) %>% # Encoding dx as factor
-  select(-Group, -Dx_Bin, -fit_result) %>%
-  #select(-Dx_Bin, -fit_result) %>%
+  #select(-Group, -Dx_Bin, -fit_result) %>%
+  select(-Dx_Bin, -fit_result) %>%
   drop_na() %>%
   select(dx, everything()) %>%
   as.data.frame()
@@ -124,11 +123,27 @@ test <- inner_join(test, meta, by=c("Group"="sample"))  %>%
                         Dx_Bin == "Cancer" ~ "cancer",
                         TRUE ~ NA_character_),
          dx = as.factor(dx)) %>% # Encoding dx as factor
-  select(-Group, -Dx_Bin, -fit_result) %>%
-  #select(-Dx_Bin, -fit_result) %>%
+  #select(-Group, -Dx_Bin, -fit_result) %>%
+  select(-Dx_Bin, -fit_result) %>%
   drop_na() %>%
   select(dx, everything()) %>%
   as.data.frame()
+
+# check that there is no overlap in samples between test and train 
+if( length(intersect(data$Group,test$Group)) > 0 ){
+  stop("something went wrong with the data split, there are samples in testing and training.")
+}
+
+# save ID and dx
+trainIDS <- data %>% select(dx,Group) 
+testIDS <- test  %>% select(dx,Group)
+
+#remove group from data/test
+data <- data %>% 
+  select(-Group)
+  
+test <- test %>% 
+  select(-Group)
 
 ######################## RUN PIPELINE ###########################
 
@@ -179,7 +194,9 @@ results <- mikropml::run_ml(dataset = allData,
                             seed = 2021)
 
 cv_auc <- results$performance$cv_metric_AUC
-prediction <- predict(results$trained_model,testTransformed,type = "prob")
[…]

### 2021-11-17 code/R/merge_predictions.R
script to merge predictions

@@ -1,0 +1,53 @@
+##################################################################
+# Author: Courtney Armour 
+# Date: November 2021
+# Description: merge prediction results from run_model.R
+##################################################################
+
+library(tidyverse)
+
+outDir <- "data/learning/summary/"
+
+# get files from command line
+input <- commandArgs(trailingOnly=TRUE)
+
+#separate optifit and opticlust files
+opticlust_files <- input[grepl(pattern="opticlust",input)]
+optifit_files <- input[grepl(pattern="optifit",input)]
+
+print("opticlust files: ")
+print(opticlust_files)
+print("optifit files: ")
+print(optifit_files)
+
+#################################
+### FUNCTIONS -------------------
+#################################
+read_files <- function(file){
+    #get split and algorithm info from filename
+    file_parts <- unlist(str_split(file,"/|\\."))
+    split <- file_parts[grepl(pattern="split",file_parts)] 
+    split <- gsub(pattern="prediction_results_",replacement="",split)
+    algorithm <- file_parts[grepl(pattern="opti",file_parts)]
+    
+    pred <- read_csv(file) %>% 
+        mutate(split=split,algorithm=algorithm)
+        
+    return(pred)
+}
+
+
+#################################
+### MAIN -------------------
+#################################
+
+# merge opticlust files
+opticlust_pred <- map_dfr(opticlust_files, read_files) 
+
+# merge optifit files
+optifit_pred <- map_dfr(optifit_files, read_files) 
+
+# merge all files together
+all_pred <- bind_rows(opticlust_pred,optifit_pred)
+
+write_csv(all_pred,paste0(outDir,"merged_predictions.csv"))

### 2021-11-17 code/fraction_reads_mapped.py
change wildcard from sample to num

@@ -46,7 +46,7 @@ def main():
 
     wildcards = snakemake.wildcards
     #header_line = "sample\tfraction_mapped\n"
-    data_str = f"{wildcards.sample}\t{fraction_mapped}\n"
+    data_str = f"{wildcards.num}\t{fraction_mapped}\n"
     with open(snakemake.output.tsv, "w") as output_file:
         #output_file.write(header_line)
         output_file.write(data_str)

### 2021-11-18 code/R/get_sensspec.R
update code for mcc scores

@@ -1,7 +1,6 @@
-#!/usr/bin/env Rscript
 # get_sensspec.R
 # Courtney Armour
-# March 2021
+# November 2021
 # Pat Schloss Lab
 # University of Michigan
 
@@ -11,36 +10,38 @@
 #############################################################################
 library(tidyverse)
 
-#only cluster one time with opticlust so only one sensspec file
-opticlust_file <- "data/process/opticlust/shared/glne.precluster.opti_mcc.sensspec"
+outDir <- "results/tables/"
 
-# get directory paths for all samples from optifit
-optifit_sample_dirs <- list.dirs(path="data/process/optifit",recursive=F)
+# grab input files
+input <- commandArgs(trailingOnly=TRUE)
+# input <- c("data/process/opticlust/shared/glne.precluster.opti_mcc.sensspec",
+#            "data/process/optifit/split_1/train/glne.precluster.pick.opti_mcc.sensspec",
+#            "data/process/optifit/split_2/train/glne.precluster.pick.opti_mcc.sensspec",
+#            "data/process/optifit/split_1/test/glne.precluster.pick.subsample.renamed.fit.optifit_mcc.sensspec",
+#            "data/process/optifit/split_2/test/glne.precluster.pick.subsample.renamed.fit.optifit_mcc.sensspec")
 
-# get file lists for all samples and optifit/opticlust
-# IN refers to cluster.fit on the left out sample
-# OUT is cluster on the other 489 samples
-optifit_in_files <- list.files(path=paste0(optifit_sample_dirs,"/in"),pattern="*.sensspec",full.names=T)
-optifit_out_files <- list.files(path=paste0(optifit_sample_dirs,"/out"),pattern="*.sensspec",full.names=T)
-
-#function to read files and add sample and type column
-read_optifit_sensspec_files <- function(filenames){
-  for(file in filenames){
-    # Read the sensspec files and add type(e.g. optifit_in) and sampleID
-    data <- read_tsv(file,col_types = cols(.default=col_double())) %>%
-      mutate(sample=unlist(strsplit(file,"/"))[4],
-             type=paste0(unlist(strsplit(file,"/"))[3],"_",unlist(strsplit(file,"/"))[5]))
-  }
-  return(data)
+#function to read in mcc files
+read_mcc <- function(file){
+    file_parts <- unlist(str_split(file,"/|\\."))
+    
[…]

### 2021-11-18 code/R/mergeCV.R
code to merge CV results

@@ -1,0 +1,34 @@
+################################
+# Courtney R Armour
+# November 2021
+# Schloss Lab
+# University of Michigan
+################################
+
+### SETUP -------------------
+library(tidyverse)
+
+outDir <- "data/learning/summary/"
+
+input <- commandArgs(trailingOnly = TRUE)
+
+### FUNCTIONS -------------------
+read_cv <- function(file){
+    file_parts <- unlist(str_split(file,"/|\\."))
+    
+    split <- file_parts[grepl(pattern="split",file_parts)]
+    split <- gsub("cv_results_","",split)
+    
+    algorithm <- file_parts[grepl(pattern="^opticlust$|^optifit$",file_parts)]
+    
+    cv <- read_csv(file,na = c("","NA","NaN")) %>% #some of the Pos/Neg Pred Values were NaN 
+        mutate(split=split,algorithm=algorithm)
+  
+  return(cv)
+}
+
+### MAIN -------------------
+
+mergedCV <- map_dfr(input,read_cv)
+
+write.csv(mergedCV,paste0(outDir,"merged_CV.csv"))

### 2021-11-19 code/R/get_sensspec.R
fix typo

@@ -37,7 +37,7 @@ read_mcc <- function(file){
     mcc <- read_tsv(file,col_types = cols(.default = col_double()))  %>% 
         select(mcc)  %>% 
         mutate(split=split,
-               algorith=algorithm,
+               algorithm=algorithm,
                subset=subset)
     return(mcc)
 }

### 2022-01-25 code/R/get_mcc.R
update outputs

@@ -9,17 +9,18 @@
 # to output a concatenated file of values for all samples and methods
 #############################################################################
 library(tidyverse)
+library(assertthat)
 
-
-outDir <- "results/tables/"
+outDir <- "data/learning/summary/"
+if(!dir.exists(outDir)){dir.create(outDir)}
 
 # grab input files
 input <- commandArgs(trailingOnly=TRUE)
 # input <- c("data/process/opticlust/shared/glne.precluster.opti_mcc.sensspec",
 #            "data/process/optifit/split_1/train/glne.precluster.pick.opti_mcc.sensspec",
 #            "data/process/optifit/split_2/train/glne.precluster.pick.opti_mcc.sensspec",
-#            "data/process/optifit/split_1/test/glne.precluster.pick.subsample.renamed.fit.optifit_mcc.sensspec",
-#            "data/process/optifit/split_2/test/glne.precluster.pick.subsample.renamed.fit.optifit_mcc.sensspec")
+#            "data/process/optifit/split_1/test/glne.precluster.pick.subsample.renamed.fit.optifit_mcc.steps",
+#            "data/process/optifit/split_2/test/glne.precluster.pick.subsample.renamed.fit.optifit_mcc.steps")
 
 #function to read in mcc files
 read_mcc <- function(file){
@@ -27,22 +28,47 @@ read_mcc <- function(file){
     
     algorithm <- file_parts[grepl(pattern="^opticlust$|^optifit$",file_parts)]
     if(algorithm == "opticlust"){
-        split <- NA
-        subset <- NA
+
+        #get just the mcc score and add the other info
+        mcc <- read_tsv(file,col_types = cols(.default = col_double()))  %>% 
+            select(mcc)  %>% 
+            mutate(split=NA,
+                   algorithm=algorithm,
+                   subset=NA,
+                   state=NA)
     }else{
         split <- file_parts[grepl(pattern="split",file_parts)]
         subset <- file_parts[grepl(pattern="train|test",file_parts)]
+        
+        if(subset == "test"){
+            #select the last iteration of the steps file
+            steps <- read_tsv(file)  %>% 
+                select(state,iter,mcc) %>% 
+                filter( iter == max(iter)) 
+            #make sure we have both states
+            assert_that(nrow(steps) == 2)
+            #get mcc score
+            mcc <- steps  %>% 
+                select(state,mcc)  %>% 
+                mutate(split=split,
+                       algorithm=algorithm,
[…]

### 2022-01-25 code/R/get_pct_correct.R

@@ -1,5 +1,7 @@
 library(tidyverse)
 
+outDir <- "results/tables/"
+
 ### FUNCTIONS ##########################################
 get_pred_file_list <- function(algorithm){
   list.files(path=paste0("data/learning/results/",algorithm),
@@ -46,7 +48,7 @@ algorithms=c("opticlust","optifit")
 pct_correct <- map_dfr(algorithms, get_pct_correct, threshold=0.5) %>%
   mutate(Group=as.character(Group))
   
-write_csv(pct_correct,"analysis/pct_class_correct.csv")
+write_csv(pct_correct,paste0(outDir,"pct_class_correct.csv"))
 
 
 ### TEST ########

### 2022-01-25 code/R/mergeHP.R

@@ -12,9 +12,9 @@ outDir <- "data/learning/summary/"
 if(!dir.exists(outDir)){dir.create(outDir)}
 
 input <- commandArgs(trailingOnly = TRUE)
-input1 <- list.files(path="data/learning/results/opticlust/",pattern="hp*",full.names=T)
-input2 <- list.files(path="data/learning/results/optifit/",pattern="hp*",full.names=T)
-input <- c(input1,input2)
+#  input1 <- list.files(path="data/learning/results/opticlust/",pattern="hp*",full.names=T)
+#  input2 <- list.files(path="data/learning/results/optifit/",pattern="hp*",full.names=T)
+# input <- c(input1,input2)
 
 ### FUNCTIONS -------------------
 read_hp <- function(file){

### 2022-01-25 code/R/run_model.R

@@ -74,6 +74,9 @@ grps <- c(paste0("TR",seq(1,nrow(train))),
 #set seed to split number
 sd <- as.numeric(str_replace(split,"split_",""))
 
+#set hyperparameters
+new_hp = list(mtry = c(68,137,274,350))
+
 # Run the model
 results <- mikropml::run_ml(dataset = allData,
                             method = model,
@@ -82,7 +85,8 @@ results <- mikropml::run_ml(dataset = allData,
                             #group_partitions = list(train = c("TR"),test = c("TE")),
                             group_partitions = list(train = c(paste0("TR",seq(1,nrow(train)))),
                                                     test = c(paste0("TE",seq(1,nrow(test))))),
-                            seed = sd)
+                            seed = sd,
+                            hyperparameters = new_hp)
 
 performance <- results$performance
 

### 2022-01-25 code/R/view_splits.R

@@ -1,9 +1,9 @@
 library(tidyverse)
 
-outdir <- "analysis/"
+outdir <- "results/figures/"
 if(!dir.exists(outdir)){dir.create(outdir)}
 
-files <- commandArgs(trailingOnly = TRUE)
+files <- snakemake@input[["splits"]]
 
 split_files <- files  %>% 
     lapply(read_csv)  %>% 
@@ -25,4 +25,4 @@ counts  %>%
               legend.text = element_text(size=14),
               legend.title = element_text(size=14)) + 
         xlab("Group") + ylab("Number of times sample is in group") 
-ggsave("analysis/view_splits.png",width = 6, height= 6)
+ggsave(paste0(outdir,"view_splits.png"),width = 6, height= 6)

### 2022-01-25 code/bash/mothurFitOptiFit.sh
reorganize

@@ -65,7 +65,7 @@ mothur "#set.current(outputdir="${OUTDIR}"/, processors="${NPROC}");
 	get.groups(fasta="${FASTA}", count="${COUNT}", taxonomy="${TAXONOMY}",  groups=${testIDS%?}, column="${DIST}");
 	sub.sample(fasta=current, count=current, taxonomy=current, size="${SUBSIZE}");
 	rename.seqs(fasta=current,count=current);
-	cluster.fit(fasta=current, count=current, reffasta="${REFFASTA}", refcolumn="${REFDIST}", reflist="${REFLIST}", method=closed, printref = t);
+	cluster.fit(fasta=current, count=current, reffasta="${REFFASTA}", refcolumn="${REFDIST}", reflist="${REFLIST}", method=closed, printref=f);
 	remove.seqs(count=current, accnos=current);
 	make.shared(list=current, count=current, label=0.03);
 	list.seqs(list=current)"  # creates accnos file of seqs in OTUs

### 2022-01-25 code/slurm/snakemake.sh

@@ -31,4 +31,4 @@ conda activate glne
 mkdir -p logs/slurm/
 
 # Initiating snakemake and running workflow in cluster mode
-snakemake --profile config/slurm/ --latency-wait 90 
+snakemake --profile config/slurm/ --latency-wait 90 --rerun-incomplete

### 2022-01-25 code/R/plot_HP_performance.R
script to plot hp

@@ -1,0 +1,25 @@
+library(tidyverse)
+
+outDir <- "results/figures/"
+if(!dir.exists(outDir)){dir.create(outDir)}
+
+mergedHP <- read_csv(snakemake@input[["hp"]])
+#mergedHP <- read_csv("data/learning/summary/merged_HP.csv")
+
+mergedHP  %>% 
+  group_by(mtry,algorithm)  %>% 
+  summarise("mean_AUC" = mean(AUC),
+            "sd_AUC" = sd(AUC),
+            ymin_metric = mean_AUC - sd_AUC,
+            ymax_metric = mean_AUC + sd_AUC)  %>% 
+  ggplot(aes(x=mtry,y=mean_AUC,color=algorithm)) +  
+    geom_point(size=2,alpha=0.5) + geom_line(alpha=0.5) +
+    theme_bw() +
+    #facet_wrap(.~algorithm,nrow=2) +
+    geom_errorbar(aes(
+      ymin = .data$ymin_metric,
+      ymax = .data$ymax_metric),
+      width = .001) +
+    ylab("Mean AUC") +
+    ggtitle("Hyperparameter Performance")
+ggsave(paste0(outDir,"hp_performance.png"))

### 2022-01-25 exploratory/exploratory.Rmd
updated results

@@
 ```
+
 ### Table of Contents
@@
 -   [Overview of data splits](#splits)
+-   [Percent of reads mapped with optifit](#pct_mapped)
+-   [MCC Optifit vs OptiClust](#mcc)
+-   [Percent of reads mapped vs MCC](#pct_v_mcc_ref)
+-   [Hyperparameter Performance](#hp_performance)
+-   [Performance during training](#cv_auc)
+-   [Performance on test data](#test_auc)
+-   [Average ROC curves](#avg_roc)
+-   [Correct Classification Frequency](#class_correct_freq)
 
@@
 
-```{r plot_splits}
-knitr::include_graphics("../analysis/view_splits.png")
+```{r plot_splits,out.width="50%"}
+knitr::include_graphics("../results/figures/view_splits.png")
 ```
-### Percent of reads mapped with OptiFit
 
-...in progress...
+### Percent of reads mapped with OptiFit {#pct_mapped}
 
-### Model performance during training
+```{r pct_mapped}
+pct_mapped <- read_tsv("../results/tables/fraction_reads_mapped.tsv") %>% 
+  rename(split=sample) %>% 
+  mutate(algorithm="optifit") %>% 
+  mutate(split=paste0("split_",split))
 
-```{r cv_auc}
+# ggplot(pct_mapped,aes(x=algorithm,y=fraction_mapped)) + 
+#   geom_jitter(width=0.2,color="blue",size=3,alpha = 0.5) +
+#   stat_summary(fun.data=median_hilow, fun.args=0.5, geom="crossbar", width=0.5, color="black") +
+#   theme_bw() +
+#   xlab("") + ylab("Fraction of Reads Mapped with OptiFit") +
+#   scale_y_continuous(limits=c(.98,1),labels = scales::percent_format(accuracy=0.1))
 
-optifit_cv_files <- list.files(path="../data/learning/results/optifit/",
-                               pattern="cv_results_split_.*.csv",
-                               full.path=TRUE)
+highest_pct_mapped <- pct_mapped %>% 
+  filter(fraction_mapped == max(fraction_mapped)) %>% 
+  pull(split)
+lowest_pct_mapped <- pct_mapped %>% 
+  filter(fraction_mapped == min(fraction_mapped)) %>% 
+  pull(split)
+
+ggplot(pct_mapped,aes(x=fraction_mapped)) + 
+  geom_histogram(color="black",fill="grey") +
+  theme_bw() +
+  xlab("Percent of Reads Mapped with OptiFit") +
+  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
+  scale_y_continuous(expand=c(0,0)) +
+  geom_text(x=max(pct_mapped$fraction_mapped),y=1.5,label=highest_pct_mapped,hjust=0.25,size=3) +
+  geom_text(x=min(pct_mapped$fraction_mapped),y=1.5,label=lowest_pct_mapped,hjust=0.5,size=3)
 
@@
 
-### ...THE REST IS MISC LEFTOVER FROM PREVIOUS ANALYSIS...
+### MCC OptiFit vs OptiClust {#mcc}
 
-### Percent of reads mapped with OptiFit
+```{r mcc}
+merged_mcc <- read_csv("../data/learning/summary/merged_MCC.csv") %>% 
+  mutate(label = case_when(algorithm == "opticlust" ~ algorithm,
+                           TRUE ~ paste0(algorithm,"_",subset))) %>% 
+  mutate(color=case_when(algorithm == "opticlust" ~ "OptiClust",
+                         subset == "train" ~ "OptiFit Reference",
+                         state == "fit" ~ "Optifit Fit without Reference",
+                         TRUE ~ "Optifit Fit with Reference")) %>% 
+  mutate(label = factor(label,levels=c("opticlust","optifit_train","optifit_test"),
+                        labels = c("OptiClust","OptiFit Train","OptiFit Test")))
 
-```{r reads_mapped}
-
-# fraction_mapped <- read_tsv("../results/tables/fraction_reads_mapped.tsv")
-
-# med <- median(fraction_mapped$fraction_mapped)
-# med_frac <- paste0(100*median(fraction_mapped$fraction_mapped),"%")
-# min_frac <- paste0(100*min(fraction_mapped$fraction_mapped),"%")
-# max_frac <- paste0(100*max(fraction_mapped$fraction_mapped),"%")
-
-# fraction_mapped %>% 
-#   ggplot(aes(x=fraction_mapped)) + 
-#     geom_histogram(fill="grey",color="black") +
-#   theme_bw() + 
-#   scale_x_continuous(labels=scales::percent) +
-#   xlab("Percent of reads mapped with cluster.fit") +
-#   ylab("number of samples") +
-#   geom_vline(xintercept = med,lty="dashed",lwd=1) +
-#   annotate("text", x=.90, y=55, label= paste0("Median: ",med_frac)) +   
-#   annotate("text", x=.90, y=50, label= paste0("Max: ",max_frac)) +
-#   annotate("text", x=.90, y=45, label= paste0("Min: ",min_frac))
+merged_mcc %>% 
+  ggplot(aes(x=label,y=mcc,color=color)) +
+    geom_jitter(height=0,width=0.2,alpha=0.5,size=3) +
+    theme_bw() +
+    xlab("") + ylab("MCC") +
+    theme(panel.grid.major.y = element_line(color="grey85"),
+          panel.grid.minor.y = element_line(color="grey85"),
+          panel.grid.major.x = element_blank()) +
+    geom_text(aes(label,mcc,label=split),data=subset(merged_mcc, mcc < 0.7),color="black",hjust=0) #+
+    # geom_text(aes(label,mcc,label=split),data=subset(merged_mcc, split == lowest_pct_mapped),color="black",hjust=0) +
+    # geom_text(aes(label,mcc,label=split),data=subset(merged_mcc, split == highest_pct_mapped),color="black",hjust=0)
 
@@
 
-### MCC OptiFit vs OptiClust
+### Percent Mapped VS MCC (without reference) {#pct_v_mcc_ref}
+```{r pct_map_v_mcc_ref}
 
-```{r mcc}
+sub_pct <- pct_mapped %>% 
+  select(split,fraction_mapped)
 
-# sensspec <- read_csv("../results/tables/merged_sensspec.csv")
+mcc_pct <- merged_mcc %>% 
+  filter(algorithm == "optifit" & subset == "test",state == "fit") %>% 
+  select(split,mcc) %>% 
+  inner_join(., sub_pct ,by="split")
 
-# mcc_data <- sensspec %>% 
-#   select(type,sample,mcc) %>% 
-#   mutate(type=factor(type,levels=c("opticlust","optifit_out","optifit_in"),labels=c("OptiClust_all","OptiClust_489","OptiFit_1")))
+mcc_pct %>% 
+  ggplot(aes(x=fraction_mapped,y=mcc)) +
+    geom_point(size=2,alpha = 0.5) +
+    theme_bw() +
+    xlab("Percent of Reads Mapped with OptiFit") +
+    ylab("OptiFit MCC (printref=f)") +
+    geom_text(aes(fraction_mapped,mcc,label=split),
+              data=subset(mcc_pct, (fraction_mapped > 0.997 | fraction_mapped < 0.98)),
+              color="black",vjust=-1,size=3) +
+    geom_text(aes(fraction_mapped,mcc,label=split),
+              data=subset(mcc_pct, (mcc > 0.775 | mcc < 0.7)),
+              color="black",vjust=-1,size=3)
+```
 
-# #mcc_data$type <- factor(mcc_data$type,levels=c("OptiClust_all","OptiClust_489","OptiFit_1"))
-# mcc_data %>% 
-#   ggplot(aes(x=type,y=mcc,color=type)) + 
-#     geom_jitter(height=0,size=3,alpha=0.6) +
-#     xlab("") +
+### Percent Mapped VS MCC (WITH reference) {#pct_v_mcc_noref}
+```{r pct_map_v_mcc}
+
+sub_pct <- pct_mapped %>% 
+  select(split,fraction_mapped)
+
+mcc_pct <- merged_mcc %>% 
+  filter(algorithm == "optifit" & subset == "test",state == "combo") %>% 
+  select(split,mcc) %>% 
+  inner_join(., sub_pct ,by="split")
+
+mcc_pct %>% 
+  ggplot(aes(x=fraction_mapped,y=mcc)) +
+    geom_point(size=2,alpha = 0.5) +
+    theme_bw() +
+    xlab("Percent of Reads Mapped with OptiFit") +
+    ylab("OptiFit MCC (printref=t)") +
+    geom_text(aes(fraction_mapped,mcc,label=split),
+              data=subset(mcc_pct, (fraction_mapped > 0.997 | fraction_mapped < 0.98)),
+              color="black",vjust=-1,size=3) +
+    geom_text(aes(fraction_mapped,mcc,label=split),
+              data=subset(mcc_pct, (mcc > 0.8961 | mcc < 0.8905)),
+              color="black",vjust=-1,size=3)
+```
+
+### HP performance {#hp_performance}
+
+```{r hp}
+# hp_performance <- read_csv("../data/learning/summary/merged_HP.csv",
+#                            col_types = cols(split=col_character(),
+#                                             algorithm=col_character(),
+#                                             .default=col_double()))
+# hp_performance %>% 
+#   group_by(mtry,algorithm) %>% 
+#   summarise("mean_AUC" = mean(AUC),
+#             "sd_AUC" = sd(AUC),
+#             ymin_metric = mean_AUC - sd_AUC,
+#             ymax_metric = mean_AUC + sd_AUC)
+# ggplot(hp_performance,aes(x=mtry,y=AUC,color=algorithm)) +
+#   geom_point() +
+#   facet_wrap(.~algorithm)
+knitr::include_graphics("../results/figures/hp_performance.png")
+```
+
+### Model performance during training {#cv_auc}
+
+```{r cv_auc}
+#allCV <- read_csv("../data/learning/summary/merged_CV.csv")
+allCV <- read.csv("../data/learning/summary/merged_performance.csv")
+
+meds <- allCV %>% 
+  group_by(algorithm) %>% 
+  summarise(median_CV_AUC = round(median(cv_metric_AUC),digits=3))
+
+ggplot(allCV,aes(x=algorithm,y=cv_metric_AUC,color=algorithm)) +
+  geom_jitter(width = 0.2,size=3,alpha = 0.5) +
+  stat_summary(fun.data=median_hilow, fun.args=0.5, geom="crossbar", width=0.5, color="black") +
+  theme_bw() +
+  ylab("Cross-Validation AUC") +
+  geom_text(data=meds,aes(x=algorithm,y=median_CV_AUC,label=median_CV_AUC),
+            color="black",hjust=-2)
+
+allCV %>% 
+  select(cv_metric_AUC,split,algorithm) %>% 
+  pivot_wider(id_cols = split,names_from = algorithm,values_from = cv_metric_AUC) %>% 
+  mutate(diff = opticlust-optifit) %>% 
+  mutate(change_from_opticlust = case_when(diff >= 0.01 ~ "decrease",
+                                           diff <= -0.01 ~ "increase",
+                                           TRUE ~ "less than |0.01|")) %>% 
+  pivot_longer(cols = c("opticlust","optifit"),names_to="algorithm",values_to = "cv_metric_AUC") %>% 
+  ggplot(aes(x=algorithm,y=cv_metric_AUC,color=change_from_opticlust,group=split)) +
+    geom_point(size=3,alpha = 0.5) +
+    geom_line(alpha = 0.5) +
+    theme_bw() +
+    scale_color_manual(values = c("red","blue","grey")) +
+    ylab("Cross-Validation AUC")
+
+# ggplot(allCV,aes(x=algorithm,y=cv_metric_AUC,color=split,group=split)) +
+#     geom_point(size=3,alpha = 0.5) +
+#     geom_line() +
 #     theme_bw() +
-#     stat_summary(fun=median,geom="crossbar",color="black",lwd=0.3)
+#     ylab("Cross-Validation AUC") +
+#     theme(legend.position = "none")
+```
+
+### Model performance on test data {#test_auc}
+
+```{r sens_spec}
+all_sens_spec <- read_csv("../data/learning/summary/all_sens_spec.csv",
+                          col_types = cols(algorithm=col_character(),
+                                           .default=col_double()))
+```
+
+```{r test_auc}
+allAUCs <- read_csv("../data/learning/summary/merged_performance.csv",
+                    col_types = cols(algorithm=col_character(),
+                                     method=col_character(),
+                                     split=col_character(),
+                                     .default=col_double())) 
+meds <- allAUCs %>% 
+  group_by(algorithm) %>% 
+  summarise(median_AUC = round(median(AUC),digits=3))
+
+ggplot(allAUCs,aes(x=algorithm,y=AUC,color=algorithm)) +
+  geom_jitter(height=0,width=0.2,alpha = 0.7,size=2) + 
+  stat_summary(fun.data=median_hilow, fun.args=0.5, geom="crossbar", width=0.5, color="black") +
+  theme_bw() +
+  ylab("AUC") +
+  xlab("") +
+  theme(legend.position = "top") +
+  geom_text(data=meds,aes(x=algorithm,y=median_AUC,label=median_AUC),
+            color="black",hjust=-2.8)
+```
+
+### Averaged ROC Curve {#avg_roc}
+
+```{r avg_roc}
+# calculate average/sd sensitivity
+avg_sens <- all_sens_spec %>%
+  group_by(specificity,algorithm) %>%
+  summarise(mean_sensitivity = mean(sensitivity),
+            sd_sensitivity = sd(sensitivity),
+            .groups = "drop") %>%
+  mutate(upper_sens = mean_sensitivity + sd_sensitivity,
+         lower_sens = mean_sensitivity - sd_sensitivity) %>%
+  mutate(upper_sens = case_when(upper_sens > 1 ~ 1,
+                                TRUE ~ upper_sens),
+         lower_sens = case_when(upper_sens < 0 ~ 0,
+                                TRUE ~ lower_sens)) %>%
+  mutate(fpr = 1-specificity)
+
+# plot all levels together
+avg_sens %>%
+  ggplot(aes(x=fpr,y=mean_sensitivity,
+             ymin=lower_sens,ymax=upper_sens)) +
+  geom_line(aes(color=algorithm),alpha=0.7,lwd=1) +
+  geom_ribbon(aes(fill=algorithm),alpha=0.3) +
+  coord_equal() +
+  geom_abline(intercept = 0,lty="dashed",color="grey50") +
+  scale_x_continuous(expand = c(0,0)) +
+  scale_y_continuous(expand = c(0,0)) +
+  xlab("False Positive Rate") +
+  ylab("Average True Positive Rate") +
+  #facet_wrap(~level,ncol=3) +
+  #scale_x_reverse(expand = c(0,0),labels = scales::percent_format(accuracy = 1)) +
+  theme_classic() +
+  theme(legend.position = "top",
+        axis.text = element_text(size=11),
+        axis.title = element_text(size=12),
+        axis.text.x = element_text(vjust = -0.5),
+        axis.title.x = element_text(vjust= -0.5)) +
+  guides(color = guide_legend(nrow = 1,title=""),
+         fill = guide_legend(nrow = 1,title="")) +
+  scale_color_manual(values=c("brown3","deepskyblue3"))
+
+allAUCs %>% 
+  group_by(algorithm) %>% 
+  summarize(median_AUC = round(median(AUC),3),
+            mean_AUC = round(mean(AUC),3)) %>% 
+  kableExtra::kable(col.names = c("Algorithm","Median AUC","Mean AUC"))
+```
+
+### Prediction Probabilities
+
+```{r pred_prob,fig.width=7,fig.height=7}
+# library(ggforce)
+# 
+# mergedProb <- read_csv("../data/learning/summary/merged_predictions.csv",
+#                        col_types = cols(cancer=col_double(),
+#                                         normal=col_double(),
+#                                         .default=col_character())) %>% 
+#   mutate(GroupDX = paste0(Group," ",dx))
+# 
+# ids = c((mergedProb %>% filter(dx == "normal") %>% pull(Group))[1:5],
+#         (mergedProb %>% filter(dx == "cancer") %>% pull(Group))[1:5])
+# 
+# ### NORMAL
+# p <- mergedProb %>%
+#   #filter(Group %in% ids) %>%
+#   filter(dx == "normal") %>%
+#   ggplot(aes(x=algorithm,y=cancer,color=algorithm)) +
+#     geom_jitter(width = 0.2,height=0,alpha=0.5,size=1.8) +
+#     facet_wrap_paginate(~GroupDX,ncol=5,nrow=5) +
+#     theme_bw() +
+#     geom_hline(yintercept = 0.5,lty="dashed",color="grey") +
+#     ylab("Probability of SRN") + xlab("") +
+#     #scale_y_continuous(limits=c(0,1),breaks=seq(0,1,0.2)) +
+#     theme(axis.text.x = element_blank(),
+#           axis.ticks.x = element_blank(),
+#           strip.background = element_rect(fill="deepskyblue3", size=1, color="black"),
+#           strip.text.x = element_text(face="bold", color="white"),
+#           legend.position = "top")
+# 
+# ###
+#   # plot <- mergedProb %>%
+#   # filter(dx == "normal") %>%
+#   # ggplot(aes(x=algorithm,y=cancer,color=algorithm)) +
+#   #   geom_jitter(width = 0.2,height=0,alpha=0.5,size=1.8) +
+#   #   facet_wrap_paginate(~GroupDX,ncol=5,nrow=5,page=1) +
+#   #   theme_bw() +
+#   #   geom_hline(yintercept = 0.5,lty="dashed",color="grey") +
+#   #   ylab("Probability of SRN") + xlab("") +
+#   #   #scale_y_continuous(limits=c(0,1),breaks=seq(0,1,0.2)) +
+#   #   theme(axis.text.x = element_blank(),
+#   #         axis.ticks.x = element_blank(),
+#   #         strip.background = element_rect(fill="deepskyblue3", size=1, color="black"),
+#   #         strip.text.x = element_text(face="bold", color="white"),
+#   #         legend.position = "top")
+#   # print(plot)
+# ###
+# 
+# for( i in 1:n_pages(p)){
+#   plot <- mergedProb %>%
+#   #filter(Group %in% ids) %>%
+#   filter(dx == "normal") %>%
+#   ggplot(aes(x=algorithm,y=cancer,color=algorithm)) +
+#     geom_jitter(width = 0.2,height=0,alpha=0.5,size=1.8) +
+#     facet_wrap_paginate(~GroupDX,ncol=5,nrow=5,page=i) +
+#     theme_bw() +
+#     geom_hline(yintercept = 0.5,lty="dashed",color="grey") +
+#     ylab("Probability of SRN") + xlab("") +
+#     #scale_y_continuous(limits=c(0,1),breaks=seq(0,1,0.2)) +
+#     theme(axis.text.x = element_blank(),
+#           axis.ticks.x = element_blank(),
+#           strip.background = element_rect(fill="deepskyblue3", size=1, color="black"),
+#           strip.text.x = element_text(face="bold", color="white"),
+#           legend.position = "top")
+#   print(plot)
+# }
+# 
+# ### CANCER
+# p <- mergedProb %>%
+#   #filter(Group %in% ids) %>%
+#   filter(dx == "cancer") %>%
+#   ggplot(aes(x=algorithm,y=cancer,color=algorithm)) +
+#     geom_jitter(width = 0.2,height=0,alpha=0.5,size=1.8) +
+#     facet_wrap_paginate(~GroupDX,ncol=6,nrow=5) +
+#     theme_bw() +
+#     geom_hline(yintercept = 0.5,lty="dashed",color="grey") +
+#     ylab("Probability of SRN") + xlab("") +
+#     #scale_y_continuous(limits=c(0,1),breaks=seq(0,1,0.2)) +
+#     theme(axis.text.x = element_blank(),
+#           axis.ticks.x = element_blank(),
+#           strip.background = element_rect(fill="brown3", size=1, color="black"),
+#           strip.text.x = element_text(face="bold", color="white"),
+#           legend.position = "top")
+# 
+# for( i in 1:n_pages(p)){
+#   plot <- mergedProb %>%
+#   #filter(Group %in% ids) %>%
+#   filter(dx == "cancer") %>%
+#   ggplot(aes(x=algorithm,y=cancer,color=algorithm)) +
+#     geom_jitter(width = 0.2,height=0,alpha=0.5,size=1.8) +
+#     facet_wrap_paginate(~GroupDX,ncol=6,nrow=5,page=i) +
+#     theme_bw() +
+#     geom_hline(yintercept = 0.5,lty="dashed",color="grey") +
+#     ylab("Probability of SRN") + xlab("") +
+#     #scale_y_continuous(limits=c(0,1),breaks=seq(0,1,0.2)) +
+#     theme(axis.text.x = element_blank(),
+#           axis.ticks.x = element_blank(),
+#           strip.background = element_rect(fill="brown3", size=1, color="black"),
+#           strip.text.x = element_text(face="bold", color="white"),
+#           legend.position = "top")
+#   print(plot)
+# }
+# 
+#
+
+# mergedProb %>% 
+#   #filter(Group %in% ids) %>% 
+#   filter(dx == "cancer") %>% 
+#   ggplot(aes(x=algorithm,y=cancer,color=algorithm)) +
+#     geom_jitter(width = 0.2,height=0,alpha=0.7,size=1.8) +
+#     facet_wrap(~GroupDX,ncol=5) +
+#     theme_bw() +
+#     geom_hline(yintercept = 0.5,lty="dashed",color="grey") +
+#     ylab("Probability of SRN") +
+#     theme(axis.text.x = element_text(angle=45,hjust=1),
+#           strip.background = element_rect(fill="brown3", size=1, color="black"),
+#           strip.text.x = element_text(face="bold", color="white"))
+# 
+```
+
+```{r prob_hist,include=F}
+# mergedProb %>% 
+#   #filter(Group %in% ids) %>% 
+#   filter(dx == "normal") %>% 
+#   mutate(GroupDX = paste0(Group," ",dx)) %>% 
+#   ggplot(aes(x=cancer,fill=algorithm)) +
+#     geom_histogram(binwidth = 0.05,color="black") +
+#     geom_vline(xintercept=0.5,lty="dashed",color="grey") +
+#     facet_wrap(~GroupDX,ncol=5) +
+#     xlab("Probability of SRN") +
+#     ylab("Number of Splits") +
+#     theme_bw() +
+#     scale_y_continuous(expand=c(0,0)) +
+#     scale_x_continuous(limits=c(0,1),labels = scales::percent_format(accuracy=1)) +
+#     theme(axis.text.x = element_text(angle=45),
+#           strip.background = element_rect(fill="deepskyblue3", size=1, color="black"),
+#           strip.text.x = element_text(face="bold", color="white"))
+# mergedProb %>% 
+#   #filter(Group %in% ids) %>% 
+#   filter(dx == "cancer") %>% 
+#   mutate(GroupDX = paste0(Group," ",dx)) %>% 
+#   ggplot(aes(x=cancer,fill=algorithm)) +
+#     geom_histogram(binwidth = 0.05,color="black") +
+#     geom_vline(xintercept=0.5,lty="dashed",color="grey") +
+#     facet_wrap(~GroupDX,ncol=5) +
+#     xlab("Probability of SRN") +
+#     ylab("Number of Splits") +
+#     theme_bw() +
+#     scale_y_continuous(expand=c(0,0)) +
+#     scale_x_continuous(limits=c(0,1),labels = scales::percent_format(accuracy=1)) +
+#     theme(axis.text.x = element_text(angle=45),
+#           strip.background = element_rect(fill="brown3", size=1, color="black"),
+#           strip.text.x = element_text(face="bold", color="white"))
 
@@
 
-### Model performance during training
-```{r data}
+```{r prob_diff,include=F}
 
-# optifit_summary <- read_tsv("../data/learning/summary/optifit/model_results.tsv",
-#                             col_types = cols(sample = col_character(),
-#                                              cv_auc = col_double(),
-#                                              cancer = col_double(),
-#                                              normal = col_double(),
-#                                              dx_diff = col_double(),
-#                                              dx_flag = col_factor(),
-#                                              pred_class = col_number(),
-#                                              meta_class = col_number()
-#                                              ))  %>% 
-#   mutate(pred_class = case_when(pred_class == 0 ~ "normal",
-#                                 TRUE ~ "cancer")) %>% 
-#   mutate(meta_class = case_when(meta_class == 0 ~ "normal",
-#                                 TRUE ~ "cancer"))
-
-# opticlust_summary <- read_tsv("../data/learning/summary/opticlust/model_results.tsv",
-#                               col_types = cols(sample = col_character(),
-#                                                cv_auc = col_double(),
-#                                                cancer = col_double(),
-#                                                normal = col_double(),
-#                                                dx_diff = col_double(),
-#                                                dx_flag = col_factor(),
-#                                                pred_class = col_factor(),
-#                                                meta_class = col_factor()
-#                                                )) %>% 
-#   mutate(pred_class = case_when(pred_class == 0 ~ "normal",
-#                                 TRUE ~ "cancer")) %>% 
-#   mutate(meta_class = case_when(meta_class == 0 ~ "normal",
-#                                 TRUE ~ "cancer"))
-
-# opticlust_summary %>% mutate(method="OptiClust") %>% 
-#   bind_rows(optifit_summary %>% mutate(method="OptiFit")) %>% 
-#   ggplot(aes(x=method,y=cv_auc,fill=method)) +
-#     geom_boxplot() +
-#     #geom_jitter(height=0,alpha = 0.5) +
-#     labs(title = "Cross Validation Performance") +
+# mergedProb %>% 
+#   #group_by(Group,dx,algorithm,split) %>%   
+#   #filter(Group %in% ids) %>% 
+#   pivot_wider(id_cols=c("Group","dx","split"),names_from = algorithm,values_from = cancer) %>% 
+#   mutate(diff = opticlust - optifit) %>% 
+#   ggplot(aes(x=diff,fill=dx,color=dx)) +
+#     geom_histogram(alpha=0.7) +
+#     facet_wrap(~Group) +
+#     geom_vline(xintercept = 0,lty="dashed",color="grey") +
+#     xlab("Cancer Probability Difference (OptiClust - OptiFit)") +
+#     scale_y_continuous(expand=c(0,0)) +
 #     theme_bw() +
-#     theme(legend.position = "none") 
-# ggsave("./figures/cv_auc_performance.png")
+#     scale_fill_manual(values=c("brown3","deepskyblue3"),name="Actual Diagnosis") +
+#     scale_color_manual(values=c("brown3","deepskyblue3"),name="Actual Diagnosis")
 
@@
 
-### Prediction Probability 
+### Correct Classification Frequency {#class_correct_freq}
+
+```{r pct_correct}
+pct_correct <- read_csv("../results/tables/pct_class_correct.csv")
+
+pct_correct %>%
+  ggplot(aes(x=factor(algorithm,levels=c("opticlust","optifit"),labels=c("OptiClust","OptiFit")),
+             y=pct_correct,group=Group,color=dx)) +
+    geom_jitter(height=0,width=0.2,alpha=0.7,size=1.8) +
+    theme_bw() + xlab("") +
+    ylab("Percent of splits sample is correctly classified (TP or TN)") +
+    scale_y_continuous(labels = scales::percent_format(accuracy=1),
+                       breaks = seq(0,1,0.1)) +
+    theme(panel.grid.major.y = element_line(color="grey85")) +
+    scale_color_manual(values=c("brown3","deepskyblue3"),name="Actual Diagnosis")
+
+```
+```{r test}
+# 
+# get_categories <- function(filename,threshold){
+#   read_csv(filename,col_types = cols(Group = col_character(),
+#                                     dx = col_character(),
+#                                     .default = col_double())) %>% 
+#     mutate(algorithm = str_replace(filename,
+#                                    "../data/learning/results/(.*)/prediction_results_split_(\\d*).csv", "\\1"),
+#            split = str_replace(filename,
+#                                "../data/learning/results/(.*)/prediction_results_split_(\\d*).csv", "\\2"),
+#            pred_dx = case_when(cancer > threshold ~ "cancer",
+#                                TRUE ~ "normal")) %>% 
+#     filter(Group %in% c("2005650","2009650","2025653","2671660")) %>% 
+#     select(Group,dx,pred_dx,algorithm,split) %>% 
+#     mutate(category = case_when(dx == "normal" & pred_dx == "normal" ~ "TN",
+#                                 dx == "normal" & pred_dx == "cancer" ~ "FP",
+#                                 dx == "cancer" & pred_dx == "cancer" ~ "TP",
+#                                 dx == "cancer" & pred_dx == "normal" ~ "FN",
+#                                 TRUE ~ "NA"))
+# }
+# 
+# get_pct_correct <- function(algorithm, threshold){
+#   #get_pred_file_list(algorithm) %>% 
+#   filelist <- c("../data/learning/results/optifit/prediction_results_split_1.csv",
+#                 "../data/learning/results/optifit/prediction_results_split_2.csv",
+#                 "../data/learning/results/optifit/prediction_results_split_3.csv",
+#                 "../data/learning/results/optifit/prediction_results_split_4.csv",
+#                 "../data/learning/results/optifit/prediction_results_split_5.csv")
+  # filelist <- c("../data/learning/results/optifit/prediction_results_split_1.csv",
+  #               "../data/learning/results/optifit/prediction_results_split_2.csv",
+  #               "../data/learning/results/optifit/prediction_results_split_3.csv",
+  #               "../data/learning/results/optifit/prediction_results_split_4.csv",
+  #               "../data/learning/results/optifit/prediction_results_split_5.csv")
+#   filelist %>% 
+#     map_dfr(get_categories,threshold) %>% 
+#     select(Group,category,dx) %>%
+#     group_by(Group,dx) %>% 
+#     count(category) %>% 
+#     pivot_wider(id_cols=c("Group","dx"),names_from=category,values_from=n,values_fill=0) %>% 
+#     mutate(correct = TN + TP,
+#            incorrect = FN + FP,
+#            total = TN + TP + FN + FP) %>% 
+#     mutate(pct_correct = correct/total) %>% 
+#     select(Group,dx,pct_correct) %>% 
+#     mutate(algorithm = algorithm)
+# }
+# 
+# pct_correct <- map_dfr(algorithms, get_pct_correct, threshold=0.5) %>% 
+#   mutate(Group=as.character(Group))
+# 
+# pct_correct %>% 
+#   ggplot(aes(x=algorithm,y=pct_correct,group=Group,color=dx)) +
+#     geom_jitter(height=0,width=0.2) +
+#     theme_bw() +
+#     ylab("Percent of splits sample is correctly classified (TP or TN)")
+
+```
+
 ```{r }
 
+### Prediction Probability 
 # opticlust_summary %>% 
@@
 
+```{r }
 ### Difference in prediction probablity  
-```{r hist}
-
 # ggplot(opticlust_summary,aes(x=dx_diff)) +
@@
 
+```{r }
 ### Summary of performance with a threshold of 0.5
-```{r confusionMatrix}
-
 # opticlust_cm <- confusionMatrix(data=factor(opticlust_summary$pred_class),reference = factor(opticlust_summary$meta_class))
@@
 
+```{r }
 ### ROC plot  
-
-```{r roc}
-
 #opticlust

### 2022-01-25 code/R/merge_opticlust_20_mcc.R
optifit 20 mcc

@@ -1,0 +1,37 @@
+################################
+# Courtney R Armour
+# January 2022
+# Schloss Lab
+# University of Michigan
+################################
+
+### SETUP -------------------
+library(tidyverse)
+
+outDir <- "results/tables/"
+if(!dir.exists(outDir)){dir.create(outDir)}
+
+input <- snakemake@input[["sensspec"]]
+#file = "data/process/opticlust/split_1/sub/glne.precluster.opti_mcc.0.03.pick.sensspec"
+
+### FUNCTIONS -------------------
+read_sensspec <- function(file){
+    file_parts <- unlist(str_split(file,"/|\\."))
+    
+    split <- file_parts[grepl(pattern="split",file_parts)]
+    split <- gsub("prediction_results_","",split)
+    
+    algorithm <- file_parts[grepl(pattern="^opticlust$|^optifit$",file_parts)]
+    
+    sensspec <- read_tsv(file,na = c("","NA","NaN")) %>% #some of the Pos/Neg Pred Values were NaN 
+        select(mcc)  %>% 
+        mutate(split=split,algorithm=algorithm)
+  
+  return(sensspec)
+}
+
+### MAIN -------------------
+
+mergedMCC <- map_dfr(input,read_sensspec)
+
+write_csv(mergedMCC,paste0(outDir,"opticlust_20_mcc.csv"))

### 2022-01-25 code/bash/get_opticlust_20_mcc.sh

@@ -1,0 +1,51 @@
+#! /bin/bash
+
+LIST=${1:?ERROR: Need to define LIST.}
+DIST=${2:?ERROR: Need to define DIST.}
+COUNT=${3:?ERROR: Need to define COUNT.}
+SPLIT=${4:?ERROR: Need to define SPLIT.} 
+NPROC=${5:?ERROR: Need to define NPROC.} 
+
+NUM=`basename $SPLIT .csv`
+
+OUTDIR=data/process/opticlust/$NUM/
+
+###################
+### GROUP SETUP ###
+###################
+
+#initialize train and test arrays
+trainIDS=()
+testIDS=()
+
+{
+    read
+    while IFS="," read -a line; do    
+        if [[ "${line[1]}" == "train" ]]; then
+            trainIDS+="${line[0]}-"
+        else
+            testIDS+="${line[0]}-"
+        fi
+    done 
+}< "$SPLIT"
+
+echo "training: ${trainIDS%?}" # %? removes last character (extra "-")
+echo "testing: ${testIDS%?}"
+
+##############
+### MOTHUR ###
+##############
+
+# Make output dirs if they don't exist
+mkdir -p "${OUTDIR}"/sub/
+
+### TESTING DATA
+# Removing old files if they exist
+if [ -n "$(ls -A "${OUTDIR}"/sub)" ]; then
+	rm "${OUTDIR}"/sub/*
+fi
+
+# Create OptiClust shared file for training data
+ mothur "#set.current(outputdir="${OUTDIR}"/sub/,processors="${NPROC}");
+ 	get.groups(list="${LIST}", column="${DIST}", count="${COUNT}", groups=${testIDS%?});
+    sens.spec(list=current,column=current,count=current)"

### 2022-01-27 code/bash/get_opticlust_20_mcc.sh
mcc for opticlust 20pct

@@ -1,14 +1,14 @@
 #! /bin/bash
 
-LIST=${1:?ERROR: Need to define LIST.}
-DIST=${2:?ERROR: Need to define DIST.}
-COUNT=${3:?ERROR: Need to define COUNT.}
-SPLIT=${4:?ERROR: Need to define SPLIT.} 
-NPROC=${5:?ERROR: Need to define NPROC.} 
+DIST=${1:?ERROR: Need to define DIST.}
+COUNT=${2:?ERROR: Need to define COUNT.}
+SPLIT=${3:?ERROR: Need to define SPLIT.} 
+NPROC=${4:?ERROR: Need to define NPROC.} 
 
 NUM=`basename $SPLIT .csv`
 
 OUTDIR=data/process/opticlust/$NUM/
+SUBSIZE=10000
 
 ###################
 ### GROUP SETUP ###
@@ -29,7 +29,7 @@ testIDS=()
     done 
 }< "$SPLIT"
 
-echo "training: ${trainIDS%?}" # %? removes last character (extra "-")
+#echo "training: ${trainIDS%?}" # %? removes last character (extra "-")
 echo "testing: ${testIDS%?}"
 
 ##############
@@ -47,5 +47,7 @@ fi
 
 # Create OptiClust shared file for training data
  mothur "#set.current(outputdir="${OUTDIR}"/sub/,processors="${NPROC}");
- 	get.groups(list="${LIST}", column="${DIST}", count="${COUNT}", groups=${testIDS%?});
-    sens.spec(list=current,column=current,count=current)"
+ 	get.groups(column="${DIST}", count="${COUNT}", groups=${testIDS%?});
+    cluster(column=current, count=current);
+    make.shared(list=current, count=current, label=0.03);
+	sub.sample(shared=current, label=0.03, size="${SUBSIZE}")"

### 2022-01-28 code/slurm/snakemake.sh
fix plot rules

@@ -31,4 +31,4 @@ conda activate glne
 mkdir -p logs/slurm/
 
 # Initiating snakemake and running workflow in cluster mode
-snakemake --profile config/slurm/ --latency-wait 90 --rerun-incomplete
+snakemake --profile config/slurm/ --latency-wait 90

### 2022-01-28 exploratory/exploratory.Rmd
updated exploratory doc

@@
                            TRUE ~ paste0(algorithm,"_",subset))) %>% 
-  mutate(color=case_when(algorithm == "opticlust" ~ "OptiClust",
+  mutate(Group=case_when(algorithm == "opticlust" ~ "OptiClust",
                          subset == "train" ~ "OptiFit Reference",
@@
 merged_mcc %>% 
-  ggplot(aes(x=label,y=mcc,color=color)) +
+  ggplot(aes(x=label,y=mcc,color=Group)) +
     geom_jitter(height=0,width=0.2,alpha=0.5,size=3) +
@@
     # geom_text(aes(label,mcc,label=split),data=subset(merged_mcc, split == highest_pct_mapped),color="black",hjust=0)
+
+```
+```{r opticlust_20_mcc}
+
+opticlust_20_mcc <- read_csv("../results/tables/opticlust_20_mcc.csv") %>% 
+  mutate(Group="OptiClust 20%")
+
+merged_mcc %>% 
+  filter(algorithm=="optifit",subset=="test") %>% 
+  select(mcc,split,algorithm,Group) %>% 
+  bind_rows(opticlust_20_mcc) %>% 
+  ggplot(aes(x=algorithm,y=mcc,color=Group)) +
+    geom_jitter(height=0,width=0.2,alpha=0.5,size=3) +
+    theme_bw() +
+    xlab("") + ylab("MCC") +
+    theme(panel.grid.major.y = element_line(color="grey85"),
+          panel.grid.minor.y = element_line(color="grey85"),
+          panel.grid.major.x = element_blank())
 

### 2022-02-07 code/R/preprocess.R
fix subsample optifit

@@ -19,10 +19,10 @@ ncores   <- as.numeric(input[4])
 
 #for testing the pipeline
 # metadata <- "data/metadata/metadata.tsv"
-# training <- "data/process/opticlust/split_1/train/glne.precluster.opti_mcc.0.03.subsample.0.03.pick.shared"
-# #training <- "data/process/optifit/split_5/train/glne.precluster.pick.opti_mcc.0.03.subsample.shared"
-# testing  <- "data/process/opticlust/split_1/test/glne.precluster.opti_mcc.0.03.subsample.0.03.pick.shared"
-# #testing <- "data/process/optifit/split_5/test/glne.precluster.pick.subsample.renamed.fit.optifit_mcc.shared"
+# #training <- "data/process/opticlust/split_1/train/glne.precluster.opti_mcc.0.03.subsample.0.03.pick.shared"
+# training <- "data/process/optifit/split_1/train/glne.precluster.pick.opti_mcc.0.03.subsample.shared"
+# #testing  <- "data/process/opticlust/split_1/test/glne.precluster.opti_mcc.0.03.subsample.0.03.pick.shared"
+# testing <- "data/process/optifit/split_1/test/glne.precluster.pick.renamed.fit.optifit_mcc.shared"
 # ncores=12
 
 split <- unlist(str_split(training,"/"))[4]
@@ -57,7 +57,8 @@ if (str_detect(testing, "optifit")) {
   test_shared <- read_tsv(testing, col_types = cols(Group=col_character(),
                                                   .default = col_double())) %>%
     select(-label, -numRefOtus) %>%
-    rename_all(str_replace, "Ref_", "")
+    select(Group,starts_with("Ref_")) %>% #select only OTUs in reference
+    rename_all(str_replace, "Ref_", "") #remove Ref_ label to match train data
 } else {
   test_shared <- read_tsv(testing, col_types = cols(Group=col_character(),
                                                   .default = col_double())) %>%

### 2022-02-07 code/bash/mothurFitOptiFit.sh

@@ -62,10 +62,18 @@ fi
 
 # Selecting test data to be fit, subsampling input files to account for 'closed' clustering method, and fitting sample to previous shared file
 mothur "#set.current(outputdir="${OUTDIR}"/, processors="${NPROC}");
-	get.groups(fasta="${FASTA}", count="${COUNT}", taxonomy="${TAXONOMY}",  groups=${testIDS%?}, column="${DIST}");
-	sub.sample(fasta=current, count=current, taxonomy=current, size="${SUBSIZE}");
+	get.groups(fasta="${FASTA}", count="${COUNT}", taxonomy="${TAXONOMY}", groups=${testIDS%?});
 	rename.seqs(fasta=current,count=current);
-	cluster.fit(fasta=current, count=current, reffasta="${REFFASTA}", refcolumn="${REFDIST}", reflist="${REFLIST}", method=closed, printref=f);
-	remove.seqs(count=current, accnos=current);
+	cluster.fit(fasta=current, count=current, reffasta="${REFFASTA}", refcolumn="${REFDIST}", reflist="${REFLIST}", method=open, printref=f);
 	make.shared(list=current, count=current, label=0.03);
-	list.seqs(list=current)"  # creates accnos file of seqs in OTUs
+	sub.sample(shared=current, size="${SUBSIZE}")"
+	
+# # Selecting test data to be fit, subsampling input files to account for 'closed' clustering method, and fitting sample to previous shared file
+# mothur "#set.current(outputdir="${OUTDIR}"/, processors="${NPROC}");
+# 	get.groups(fasta="${FASTA}", count="${COUNT}", taxonomy="${TAXONOMY}",  groups=${testIDS%?}, column="${DIST}");
+# 	sub.sample(fasta=current, count=current, taxonomy=current, size="${SUBSIZE}");
+# 	rename.seqs(fasta=current,count=current);
+# 	cluster.fit(fasta=current, count=current, reffasta="${REFFASTA}", refcolumn="${REFDIST}", reflist="${REFLIST}", method=closed, printref=f);
+# 	remove.seqs(count=current, accnos=current);
+# 	make.shared(list=current, count=current, label=0.03);
+# 	list.seqs(list=current)"  # creates accnos file of seqs in OTUs

### 2022-02-07 code/slurm/snakemake.sh
change account

@@ -6,7 +6,7 @@
 #SBATCH --cpus-per-task=1
 #SBATCH --mem-per-cpu=6GB
 #SBATCH --time=5-00:00:00
-#SBATCH --account=pschloss99
+#SBATCH --account=pschloss1
 #SBATCH --partition=standard
 #SBATCH --mail-user=armourc@umich.edu
 #SBATCH --mail-type=FAIL

### 2022-02-07 exploratory/exploratory.Rmd
update exploratory

@@
 ```{r pct_mapped}
-pct_mapped <- read_tsv("../results/tables/fraction_reads_mapped.tsv") %>% 
-  rename(split=sample) %>% 
-  mutate(algorithm="optifit") %>% 
-  mutate(split=paste0("split_",split))
-
-# ggplot(pct_mapped,aes(x=algorithm,y=fraction_mapped)) + 
-#   geom_jitter(width=0.2,color="blue",size=3,alpha = 0.5) +
-#   stat_summary(fun.data=median_hilow, fun.args=0.5, geom="crossbar", width=0.5, color="black") +
+# pct_mapped <- read_tsv("../results/tables/fraction_reads_mapped.tsv") %>% 
+#   rename(split=sample) %>% 
+#   mutate(algorithm="optifit") %>% 
+#   mutate(split=paste0("split_",split))
+# 
+# # ggplot(pct_mapped,aes(x=algorithm,y=fraction_mapped)) + 
+# #   geom_jitter(width=0.2,color="blue",size=3,alpha = 0.5) +
+# #   stat_summary(fun.data=median_hilow, fun.args=0.5, geom="crossbar", width=0.5, color="black") +
+# #   theme_bw() +
+# #   xlab("") + ylab("Fraction of Reads Mapped with OptiFit") +
+# #   scale_y_continuous(limits=c(.98,1),labels = scales::percent_format(accuracy=0.1))
+# 
+# 
+# highest_pct_mapped <- pct_mapped %>% 
+#   filter(fraction_mapped == max(fraction_mapped)) %>% 
+#   pull(split)
+# lowest_pct_mapped <- pct_mapped %>% 
+#   filter(fraction_mapped == min(fraction_mapped)) %>% 
+#   pull(split)
+# 
+# ggplot(pct_mapped,aes(x=fraction_mapped)) + 
+#   geom_histogram(color="black",fill="grey") +
 #   theme_bw() +
-#   xlab("") + ylab("Fraction of Reads Mapped with OptiFit") +
-#   scale_y_continuous(limits=c(.98,1),labels = scales::percent_format(accuracy=0.1))
-
-highest_pct_mapped <- pct_mapped %>% 
-  filter(fraction_mapped == max(fraction_mapped)) %>% 
-  pull(split)
-lowest_pct_mapped <- pct_mapped %>% 
-  filter(fraction_mapped == min(fraction_mapped)) %>% 
-  pull(split)
-
-ggplot(pct_mapped,aes(x=fraction_mapped)) + 
-  geom_histogram(color="black",fill="grey") +
-  theme_bw() +
-  xlab("Percent of Reads Mapped with OptiFit") +
-  scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
-  scale_y_continuous(expand=c(0,0)) +
-  geom_text(x=max(pct_mapped$fraction_mapped),y=1.5,label=highest_pct_mapped,hjust=0.25,size=3) +
-  geom_text(x=min(pct_mapped$fraction_mapped),y=1.5,label=lowest_pct_mapped,hjust=0.5,size=3)
+#   xlab("Percent of Reads Mapped with OptiFit") +
+#   scale_x_continuous(labels = scales::percent_format(accuracy = 0.1)) +
+#   scale_y_continuous(expand=c(0,0)) +
+#   geom_text(x=max(pct_mapped$fraction_mapped),y=1.5,label=highest_pct_mapped,hjust=0.25,size=3) +
+#   geom_text(x=min(pct_mapped$fraction_mapped),y=1.5,label=lowest_pct_mapped,hjust=0.5,size=3)
 
@@
     xlab("") + ylab("MCC") +
+    #scale_y_continuous(expand=c(0,0),limits=c(0.85,0.91)) +
     theme(panel.grid.major.y = element_line(color="grey85"),
@@
 ```{r pct_map_v_mcc_ref}
-
-sub_pct <- pct_mapped %>% 
-  select(split,fraction_mapped)
-
-mcc_pct <- merged_mcc %>% 
-  filter(algorithm == "optifit" & subset == "test",state == "fit") %>% 
-  select(split,mcc) %>% 
-  inner_join(., sub_pct ,by="split")
-
-mcc_pct %>% 
-  ggplot(aes(x=fraction_mapped,y=mcc)) +
-    geom_point(size=2,alpha = 0.5) +
-    theme_bw() +
-    xlab("Percent of Reads Mapped with OptiFit") +
-    ylab("OptiFit MCC (printref=f)") +
-    geom_text(aes(fraction_mapped,mcc,label=split),
-              data=subset(mcc_pct, (fraction_mapped > 0.997 | fraction_mapped < 0.98)),
-              color="black",vjust=-1,size=3) +
-    geom_text(aes(fraction_mapped,mcc,label=split),
-              data=subset(mcc_pct, (mcc > 0.775 | mcc < 0.7)),
-              color="black",vjust=-1,size=3)
+# 
+# sub_pct <- pct_mapped %>% 
+#   select(split,fraction_mapped)
+# 
+# mcc_pct <- merged_mcc %>% 
+#   filter(algorithm == "optifit" & subset == "test",state == "fit") %>% 
+#   select(split,mcc) %>% 
+#   inner_join(., sub_pct ,by="split") 
+# 
+# mcc_pct %>% 
+#   ggplot(aes(x=fraction_mapped,y=mcc)) +
+#     geom_point(size=2,alpha = 0.5) +
+#     theme_bw() +
+#     xlab("Percent of Reads Mapped with OptiFit") +
+#     ylab("OptiFit MCC (printref=f)") +
+#     geom_text(aes(fraction_mapped,mcc,label=split),
+#               data=subset(mcc_pct, (fraction_mapped > 0.997 | fraction_mapped < 0.98)),
+#               color="black",vjust=-1,size=3) +
+#     geom_text(aes(fraction_mapped,mcc,label=split),
+#               data=subset(mcc_pct, (mcc > 0.775 | mcc < 0.7)),
+#               color="black",vjust=-1,size=3)
 ```
@@
 
-sub_pct <- pct_mapped %>% 
-  select(split,fraction_mapped)
-
-mcc_pct <- merged_mcc %>% 
-  filter(algorithm == "optifit" & subset == "test",state == "combo") %>% 
-  select(split,mcc) %>% 
-  inner_join(., sub_pct ,by="split")
-
-mcc_pct %>% 
-  ggplot(aes(x=fraction_mapped,y=mcc)) +
-    geom_point(size=2,alpha = 0.5) +
-    theme_bw() +
-    xlab("Percent of Reads Mapped with OptiFit") +
-    ylab("OptiFit MCC (printref=t)") +
-    geom_text(aes(fraction_mapped,mcc,label=split),
-              data=subset(mcc_pct, (fraction_mapped > 0.997 | fraction_mapped < 0.98)),
-              color="black",vjust=-1,size=3) +
-    geom_text(aes(fraction_mapped,mcc,label=split),
-              data=subset(mcc_pct, (mcc > 0.8961 | mcc < 0.8905)),
-              color="black",vjust=-1,size=3)
+# sub_pct <- pct_mapped %>% 
+#   select(split,fraction_mapped)
+# 
+# mcc_pct <- merged_mcc %>% 
+#   filter(algorithm == "optifit" & subset == "test",state == "combo") %>% 
+#   select(split,mcc) %>% 
+#   inner_join(., sub_pct ,by="split") 
+# 
+# mcc_pct %>% 
+#   ggplot(aes(x=fraction_mapped,y=mcc)) +
+#     geom_point(size=2,alpha = 0.5) +
+#     theme_bw() +
+#     xlab("Percent of Reads Mapped with OptiFit") +
+#     ylab("OptiFit MCC (printref=t)") +
+#     geom_text(aes(fraction_mapped,mcc,label=split),
+#               data=subset(mcc_pct, (fraction_mapped > 0.997 | fraction_mapped < 0.98)),
+#               color="black",vjust=-1,size=3) +
+#     geom_text(aes(fraction_mapped,mcc,label=split),
+#               data=subset(mcc_pct, (mcc > 0.8961 | mcc < 0.8905)),
+#               color="black",vjust=-1,size=3)
 ```
@@
 
-```{r pct_correct}
+For each split, a cutoff of 0.5 was used to determine whether each sample was correctly or incorrectly classified.
+This data was merged for all splits to quantify the percent of splits each sample was correctly classified in.
+For normal samples it is the frequency that sample was a true negative and for cancer samples its the frequency that sample was a true positive.
+For example, if a cancer sample had a probabily of being cancer greater than 0.5 for 10 of the 20 splits it was in the test set for, it would have a percent correct value of 50%.
+I'm trying to figure out how to visualize this data...
+
+```{r pct_correct_hist, out.width="50%"}
 pct_correct <- read_csv("../results/tables/pct_class_correct.csv")
 
+# pct_correct %>%
+#   ggplot(aes(x=factor(algorithm,levels=c("opticlust","optifit"),labels=c("OptiClust","OptiFit")),
+#              y=pct_correct,group=Group,color=dx)) +
+#     geom_jitter(height=0,width=0.2,alpha=0.7,size=1.8) +
+#     theme_bw() + xlab("") +
+#     ylab("Percent of splits sample is correctly classified (TP or TN)") +
+#     scale_y_continuous(labels = scales::percent_format(accuracy=1),
+#                        breaks = seq(0,1,0.1)) +
+#     theme(panel.grid.major.y = element_line(color="grey85")) +
+#     scale_color_manual(values=c("brown3","deepskyblue3"),name="Actual Diagnosis")
+#
+
 pct_correct %>%
-  ggplot(aes(x=factor(algorithm,levels=c("opticlust","optifit"),labels=c("OptiClust","OptiFit")),
-             y=pct_correct,group=Group,color=dx)) +
-    geom_jitter(height=0,width=0.2,alpha=0.7,size=1.8) +
-    theme_bw() + xlab("") +
-    ylab("Percent of splits sample is correctly classified (TP or TN)") +
-    scale_y_continuous(labels = scales::percent_format(accuracy=1),
-                       breaks = seq(0,1,0.1)) +
-    theme(panel.grid.major.y = element_line(color="grey85")) +
-    scale_color_manual(values=c("brown3","deepskyblue3"),name="Actual Diagnosis")
+  ggplot(aes(x=pct_correct,fill=algorithm)) +
+  geom_histogram(binwidth=0.1,color="black",alpha=0.5,position = "identity") +
+  facet_grid(.~dx) +
+  theme_bw() +
+  xlab("Percent of splits sample is correctly classified (TP or TN)") +
+  theme(legend.position = "top")
+
+pct_correct %>%
+  ggplot(aes(x=pct_correct,fill=algorithm)) +
+  geom_histogram(binwidth=0.1,color="black",alpha=0.5,position = "dodge") +
+  facet_grid(.~dx) +
+  theme_bw() +
+  xlab("Percent of splits sample is correctly classified (TP or TN)") +
+  theme(legend.position = "top")
+
+pct_correct %>%
+  ggplot(aes(x=pct_correct,fill=algorithm)) +
+  #geom_histogram(binwidth=0.1,color="black",alpha=0.8) +
+  geom_density(alpha=0.5) +
+  facet_grid(.~dx) +
+  theme_bw() +
+  xlab("Percent of splits sample is correctly classified (TP or TN)") +
+  theme(legend.position = "top")
+
+pct_correct %>%
+  ggplot(aes(x=pct_correct,color=algorithm)) +
+  #geom_histogram(binwidth=0.1,color="black",alpha=0.8) +
+  geom_density(alpha=0.5) +
+  facet_grid(.~dx) +
+  theme_bw() +
+  xlab("Percent of splits sample is correctly classified (TP or TN)") +
+  theme(legend.position = "top")
 
 ```
+
+The scatter plot below has the samples along the x-axis without labels because they are not readable.
+The y-axis is the percent correct.
+The color of the points is the algorithm.
+There is a separate plot for cancer samples and normal samples.
+
+```{r pct_correct_point}
+#order samples by difference in pct_correct between algorithms
+samp_order <- pct_correct %>% 
+  pivot_wider(id_cols = c("Group","dx"),
+              names_from = algorithm,
+              values_from = pct_correct) %>% 
+  mutate(difference = abs(opticlust - optifit)) %>% 
+  arrange(difference) %>% 
+  pull(Group)
+
+pct_correct %>% 
+  ggplot(aes(x=factor(Group,levels = samp_order),y=pct_correct,color=algorithm)) +
+  geom_point(alpha=0.7) +
+  facet_wrap(.~dx,nrow=2) +
+  theme_bw() + 
+  ylab("Percent of splits sample is correctly classified (TP or TN)") +
+  theme(axis.text.x = element_blank(),
+        axis.ticks.x = element_blank()) +
+  xlab("Sample")
+```
+
+To eliminate some noise, I removed any samples where there was no difference in the percent correct between the two algorithms.
+
+```{r mod_pct_correct_point}
+#get rid of samples that are perfectly classified with both algorithms
+#and samples that have the samp pct_correct
+#order samples by difference in pct_correct between algorithms
+samp_order <- pct_correct %>% 
+  pivot_wider(id_cols = c("Group","dx"),
+              names_from = algorithm,
+              values_from = pct_correct) %>% 
+  filter(optifit != 1 & opticlust != 1) %>% 
+  mutate(difference = abs(opticlust - optifit)) %>% 
+  filter(difference > 0) %>% 
+  arrange(difference) %>% 
+  pull(Group)
+
+pct_correct %>% 
+  filter(Group %in% samp_order) %>% 
+  ggplot(aes(x=factor(Group,levels = samp_order),y=pct_correct,color=algorithm)) +
+  geom_point(alpha=0.7) +
+  facet_wrap(.~dx,nrow=2) +
+  theme_bw() + 
+  ylab("Percent of splits sample is correctly classified (TP or TN)") +
+  theme(axis.text.x = element_blank(),
+        axis.ticks.x = element_blank()) +
+  xlab("Sample")
+
+```
+
 ```{r test}

### 2022-02-21 exploratory/exploratory.Rmd
updated exploratory

@@
 -   [Overview of data splits](#splits)
--   [Percent of reads mapped with optifit](#pct_mapped)
 -   [MCC Optifit vs OptiClust](#mcc)
--   [Percent of reads mapped vs MCC](#pct_v_mcc_ref)
 -   [Hyperparameter Performance](#hp_performance)
@@
 -   [Average ROC curves](#avg_roc)
+-   [Sensitivity at Specificity Values](#sens_xspec)
 -   [Correct Classification Frequency](#class_correct_freq)
+-   [Prediction Probabilities](#pred_prob)
+-   [MCC vs Sensitivity/Specificity](#mcc_senspec)
 
@@
 knitr::include_graphics("../results/figures/view_splits.png")
+```
+
+How often do samples appear in the test set together?
+
+```{r split_together_freq}
+
+split_count <- read_csv("../results/tables/splitTogetherFrequency.csv",col_types = cols(Group=col_character(),
+                                                                                        .default = col_double()))
+quant_overlap <- function(table,group1,group2){
+  counts <- table %>% 
+    filter(Group == c(group1,group2)) %>% 
+    select(-Group) %>% 
+    colSums()
+
+  bothTest = length(which(counts == 2))
+  bothTrain = length(which(counts == 0))
+
+  tibble(group1 = group1,
+         group2 = group2,
+         bothTest = bothTest,
+         bothTrain = bothTrain)
+}
+
+samps <- split_count %>% pull(Group)
+
+# togetherFreq <- expand_grid(x=1:length(samps),
+#             y=1:length(samps)) %>% 
+#   filter(x < y) %>% 
+#   mutate(group1 = samps[x],
+#          group2 = samps[y]) %>% 
+#   select(-x,-y) %>% 
+#   group_by(group1,group2) %>% 
+#   summarize(quant_overlap(split_count,group1,group2),
+#             .groups="drop")
+# write_csv(togetherFreq,"./data/togetherFreq.csv")
+togetherFreq <- read_csv("./data/togetherFreq.csv",col_types = cols(group1=col_character(),
+                                                                    group2=col_character(),
+                                                                    .default=col_double()))
+
+meanBothTest = round(mean(togetherFreq$bothTest))
+meanBothTrain = round(mean(togetherFreq$bothTrain))
+
+neverBothTest = togetherFreq %>% 
+  filter(bothTest == 0) %>% 
+  nrow()
+neverBothTestFreq = round(neverBothTest / nrow(togetherFreq),digits=2)*100
+
+alwaysBothTrain = togetherFreq %>% 
+  filter(bothTrain == 100) %>% 
+  nrow()
+alwaysBothTrainFreq = round(alwaysBothTrain / nrow(togetherFreq),digits=2)*100
+
+maxBothTest = togetherFreq %>% 
+  filter(bothTest == max(bothTest))
+```
+
+For each pair of samples, I quantified how many splits both samples are in the test set together.
+On average, samples are both in the test set for `r meanBothTest` split.
+A majority of sample pairs are never in the test set together (`r neverBothTestFreq`%).
+At most there are `r nrow(maxBothTest)` sample pairs that are both in the test set for `r maxBothTest %>% pull(bothTest) %>% unique()` splits.
+
+```{r together_freq_plot}
+# togetherFreq %>% 
+#   pivot_longer(!c(group1,group2),names_to="type",values_to = "freq") %>% 
+#   ggplot(aes(x=freq,fill=type)) +
+#     geom_histogram(binwidth = 1) +
+#     facet_wrap(.~type,nrow=2,scales="free_y") +
+#     theme_bw() +
+#     scale_y_continuous(expand=c(0,0)) + 
+#     scale_x_continuous(breaks = seq(0,100,10),expand=c(0,0))
+# 
+# meanBothTest = round(mean(togetherFreq$bothTest))
+# meanBothTrain = round(mean(togetherFreq$bothTrain))
+
+# togetherFreq %>% 
+#   pivot_longer(!c(group1,group2),names_to="type",values_to = "freq") %>% 
+#   ggplot(aes(x=freq,fill=type)) +
+#     geom_density() +
+#     facet_wrap(.~type,nrow=2,scales = "free_y") +
+#     theme_bw() +
+#     scale_y_continuous(expand=c(0,0)) + 
+#     scale_x_continuous(breaks = seq(0,100,10),expand=c(0,0))
+
+togetherFreq %>% 
+  ggplot(aes(x=bothTest)) +
+    geom_histogram(binwidth=1,fill="#F8766D",color="black",alpha=0.8) +
+    theme_bw() +
+    scale_y_continuous(expand=c(0,0)) +
+    scale_x_continuous(expand=c(0,0)) +
+    xlab("Number of splits where sample pair are both in the test set") +
+    ylab("Number of Sample Pairs")
+
+```
+
+I also split the pairs by diagnosis (both cancer, both normal, or mixed) to make sure there was no difference by dx.
+
+```{r together_freq_plot2}
+sample_dx <- read_csv("../data/learning/summary/merged_predictions.csv",
+                      col_types = cols(cancer=col_double(),                              
+                                       normal=col_double(),
+                                       .default=col_character())) %>% 
+  select(Group,dx) %>% 
+  distinct()
+
+togetherFreq %>% 
+  inner_join(.,sample_dx,by=c("group1" = "Group")) %>% 
+  rename(group1_dx = dx) %>% 
+  inner_join(.,sample_dx,by=c("group2" = "Group")) %>% 
+  rename(group2_dx = dx) %>% 
+  select(group1,group1_dx,group2,group2_dx,bothTest,bothTrain) %>% 
+  mutate(dx=case_when(group1_dx == "normal" & group2_dx == "normal" ~ "normal",
+                      group1_dx == "cancer" & group2_dx == "cancer" ~ "cancer",
+                      TRUE ~ "mix")) %>% 
+  ggplot(aes(x=bothTest,fill=dx)) +
+    geom_histogram(binwidth=1,color="black",alpha=0.8) +
+    #geom_density(color="black",alpha=0.5) +
+    theme_bw() +
+    scale_y_continuous(expand=c(0,0)) +
+    scale_x_continuous(expand=c(0,0)) +
+    xlab("Number of splits where sample pair are both in the test set") +
+    ylab("Number of Sample Pairs") +
+    facet_wrap(.~dx,nrow=3,scales="free_y")
+
+# togetherFreq %>% 
+#   inner_join(.,sample_dx,by=c("group1" = "Group")) %>% 
+#   rename(group1_dx = dx) %>% 
+#   inner_join(.,sample_dx,by=c("group2" = "Group")) %>% 
+#   rename(group2_dx = dx) %>% 
+#   select(group1,group1_dx,group2,group2_dx,bothTest,bothTrain) %>% 
+#   mutate(dx=case_when(group1_dx == "normal" & group2_dx == "normal" ~ "normal",
+#                       group1_dx == "cancer" & group2_dx == "cancer" ~ "cancer",
+#                       TRUE ~ "mix")) %>% 
+#   ggplot(aes(x=dx,y=bothTest,color=dx)) +
+#     geom_jitter(width=0.3,height=0,alpha=0.7)
 ```
@@
 
+```{r pct_map_v_mcc_ref}
 ### Percent Mapped VS MCC (without reference) {#pct_v_mcc_ref}
-```{r pct_map_v_mcc_ref}
 # 
@@
 
+```{r pct_map_v_mcc}
 ### Percent Mapped VS MCC (WITH reference) {#pct_v_mcc_noref}
-```{r pct_map_v_mcc}
 
@@
 
+```{r allAUC}
+
+allAUCs <- read_csv("../data/learning/summary/merged_performance.csv",
+                    col_types = cols(algorithm=col_character(),
+                                     method=col_character(),
+                                     split=col_character(),
+                                     .default=col_double())) 
+meds <- allAUCs %>% 
+  select(cv_metric_AUC,AUC,algorithm) %>% 
+  rename(CV=cv_metric_AUC,
+         Test=AUC) %>% 
+  pivot_longer(!algorithm,names_to="type",values_to = "AUC") %>% 
+  group_by(algorithm,type) %>% 
+  summarise(median_AUC = round(median(AUC),digits=3))
+
+allAUCs %>% 
+  select(cv_metric_AUC,AUC,algorithm) %>% 
+  rename(CV=cv_metric_AUC,
+         Test=AUC) %>% 
+  pivot_longer(!algorithm,names_to="type",values_to = "AUC") %>% 
+  ggplot(aes(x=type,y=AUC,color=algorithm)) +
+    geom_jitter(height=0,width=0.2,alpha = 0.6,size=2) + 
+    stat_summary(fun.data=median_hilow, fun.args=0.5, geom="crossbar", width=0.5, color="black") +
+    theme_bw() +
+    ylab("AUC") +
+    xlab("") +
+    facet_grid(.~algorithm) +
+    theme(legend.position = "top") +
+    geom_text(data=meds,aes(x=type,y=median_AUC,label=median_AUC),
+              color="black",hjust=-1.2)
+
+```
+
 ### Averaged ROC Curve {#avg_roc}
@@
   mutate(fpr = 1-specificity)
+
+#calculate mean/median
+mean_med <- allAUCs %>% 
+  group_by(algorithm) %>% 
+  summarize(median_AUC = round(median(AUC),3),
+            mean_AUC = round(mean(AUC),3))
+
+of_mean <- mean_med %>% filter(algorithm=="optifit") %>% pull(mean_AUC)
+oc_mean <- mean_med %>% filter(algorithm=="opticlust") %>% pull(mean_AUC)
 
@@
          fill = guide_legend(nrow = 1,title="")) +
-  scale_color_manual(values=c("brown3","deepskyblue3"))
+  scale_color_manual(values=c("brown3","deepskyblue3")) +
+  geom_text(x=0.7,y=0.2,label=paste0("OptiClust Mean AUC: ",oc_mean),size=3.5) +
+  geom_text(x=0.7,y=0.15,label=paste0("OptiFit Mean AUC: ",of_mean),size=3.5)
 
-allAUCs %>% 
-  group_by(algorithm) %>% 
-  summarize(median_AUC = round(median(AUC),3),
-            mean_AUC = round(mean(AUC),3)) %>% 
-  kableExtra::kable(col.names = c("Algorithm","Median AUC","Mean AUC"))
 ```
 
-### Prediction Probabilities
+### Sensitivity at Specificity Thresholds {#sens_xspec}
+```{r sens_xspec}
+
+fun_median <- function(x){
+  return(data.frame(y=round(median(x),digits=3),label=round(median(x,na.rm=T),digits=3)))
+}
+
+spec = 0.9
+all_sens_spec %>% 
+  filter(specificity == spec) %>% 
+  ggplot(aes(x=algorithm,y=sensitivity,color=algorithm)) +
+    geom_jitter(height=0,width=0.2,alpha = 0.7,size=2) +
+    stat_summary(fun.data=median_hilow, fun.args=0.5, geom="crossbar", width=0.5, color="black") +
+    theme_bw() +
+    ylab("AUC") +
+    xlab("") +
+    theme(legend.position = "top") +
+    ggtitle(paste0("Sensitivity at ",spec*100,"% Specificity")) +
+    stat_summary(fun.data=fun_median, geom="text",show.legend = FALSE,color="black",hjust=-2.2)
+
+spec = 0.5
+all_sens_spec %>% 
+  filter(specificity == spec) %>% 
+  ggplot(aes(x=algorithm,y=sensitivity,color=algorithm)) +
+    geom_jitter(height=0,width=0.2,alpha = 0.7,size=2) +
+    stat_summary(fun.data=median_hilow, fun.args=0.5, geom="crossbar", width=0.5, color="black") +
+    theme_bw() +
+    ylab("AUC") +
+    xlab("") +
+    theme(legend.position = "top") +
+    ggtitle(paste0("Sensitivity at ",spec*100,"% Specificity")) +
+    stat_summary(fun.data=fun_median, geom="text",show.legend = FALSE,color="black",hjust=-2.2)
+
+spec = 0.1
+all_sens_spec %>% 
+  filter(specificity == spec) %>% 
+  ggplot(aes(x=algorithm,y=sensitivity,color=algorithm)) +
+    geom_jitter(height=0,width=0.2,alpha = 0.7,size=2) +
+    stat_summary(fun.data=median_hilow, fun.args=0.5, geom="crossbar", width=0.5, color="black") +
+    theme_bw() +
+    ylab("AUC") +
+    xlab("") +
+    theme(legend.position = "top") +
+    ggtitle(paste0("Sensitivity at ",spec*100,"% Specificity")) +
+    stat_summary(fun.data=fun_median, geom="text",show.legend = FALSE,color="black",hjust=-2.2)
+```
+
+### Correct Classification Frequency {#class_correct_freq}
+
+For each split, a cutoff of 0.5 was used to determine whether each sample was correctly or incorrectly classified.
+This data was merged for all splits to quantify the percent of splits each sample was correctly classified in.
+For normal samples it is the frequency that sample was a true negative and for cancer samples its the frequency that sample was a true positive.
+For example, if a cancer sample had a probabily of being cancer greater than 0.5 for 10 of the 20 splits it was in the test set for, it would have a percent correct value of 50%.
+I'm trying to figure out how to visualize this data...
+
+```{r pct_correct_hist, out.width="50%"}
+pct_correct <- read_csv("../results/tables/pct_class_correct.csv")
+
+# pct_correct %>%
+#   ggplot(aes(x=factor(algorithm,levels=c("opticlust","optifit"),labels=c("OptiClust","OptiFit")),
+#              y=pct_correct,group=Group,color=dx)) +
+#     geom_jitter(height=0,width=0.2,alpha=0.7,size=1.8) +
+#     theme_bw() + xlab("") +
+#     ylab("Percent of splits sample is correctly classified (TP or TN)") +
+#     scale_y_continuous(labels = scales::percent_format(accuracy=1),
+#                        breaks = seq(0,1,0.1)) +
+#     theme(panel.grid.major.y = element_line(color="grey85")) +
+#     scale_color_manual(values=c("brown3","deepskyblue3"),name="Actual Diagnosis")
+#
+
+pct_correct %>%
+  ggplot(aes(x=pct_correct,fill=algorithm)) +
+  geom_histogram(binwidth=0.1,color="black",alpha=0.5,position = "identity") +
+  facet_grid(.~dx) +
+  theme_bw() +
+  xlab("Percent of splits sample is correctly classified (TP or TN)") +
+  theme(legend.position = "top")
+
+pct_correct %>%
+  ggplot(aes(x=pct_correct,fill=algorithm)) +
+  geom_histogram(binwidth=0.1,color="black",alpha=0.5,position = "dodge") +
+  facet_grid(.~dx) +
+  theme_bw() +
+  xlab("Percent of splits sample is correctly classified (TP or TN)") +
+  theme(legend.position = "top")
+
+pct_correct %>%
+  ggplot(aes(x=pct_correct,fill=algorithm)) +
+  #geom_histogram(binwidth=0.1,color="black",alpha=0.8) +
+  geom_density(alpha=0.5) +
+  facet_grid(.~dx) +
+  theme_bw() +
+  xlab("Percent of splits sample is correctly classified (TP or TN)") +
+  theme(legend.position = "top")
+
+pct_correct %>%
+  ggplot(aes(x=pct_correct,color=algorithm)) +
+  #geom_histogram(binwidth=0.1,color="black",alpha=0.8) +
+  geom_density(alpha=0.5) +
+  facet_grid(.~dx) +
+  theme_bw() +
+  xlab("Percent of splits sample is correctly classified (TP or TN)") +
+  theme(legend.position = "top")
+
+```
+
+The scatter plot below has the samples along the x-axis without labels because they are not readable.
+The y-axis is the percent correct.
+The color of the points is the algorithm.
+There is a separate plot for cancer samples and normal samples.
+
+```{r pct_correct_point,fig.width=11}
+#order samples by difference in pct_correct between algorithms
+samp_order <- pct_correct %>% 
+  pivot_wider(id_cols = c("Group","dx"),
+              names_from = algorithm,
+              values_from = pct_correct) %>% 
+  mutate(difference = abs(opticlust - optifit)) %>% 
+  arrange(difference) %>% 
+  pull(Group)
+
+pct_correct %>% 
+  ggplot(aes(x=factor(Group,levels = samp_order),y=pct_correct,color=algorithm)) +
+  geom_point(alpha=0.7) +
+  facet_wrap(.~dx,nrow=2,scales="free_x") +
+  theme_bw() + 
+  ylab("Percent of splits sample is correctly classified (TP or TN)") +
+  theme(axis.text.x = element_blank(),
+        axis.ticks.x = element_blank()) +
+  xlab("Sample") +
+  theme(legend.position = "top")
+```
+
+To eliminate some noise, I removed any samples where there was no difference in the percent correct between the two algorithms.
+
+```{r mod_pct_correct_point,fig.width=9}
+#get rid of samples that are perfectly classified with both algorithms
+#and samples that have the samp pct_correct
+#order samples by difference in pct_correct between algorithms
+samp_order <- pct_correct %>% 
+  pivot_wider(id_cols = c("Group","dx"),
+              names_from = algorithm,
+              values_from = pct_correct) %>% 
+  #filter(optifit != 1 & opticlust != 1) %>% 
+  mutate(difference = abs(opticlust - optifit)) %>% 
+  filter(difference > 0) %>% 
+  arrange(difference,optifit) %>% 
+  pull(Group)
+
+pct_correct %>% 
+  filter(Group %in% samp_order) %>% 
+  ggplot(aes(x=factor(Group,levels = samp_order),y=pct_correct,color=algorithm)) +
+  geom_point(alpha=0.7) +
+  facet_wrap(.~dx,nrow=2,scales="free_x") +
+  theme_bw() + 
+  ylab("Percent of splits sample is correctly classified (TP or TN)") +
+  theme(axis.text.x = element_blank(),
+        axis.ticks.x = element_blank()) +
+  xlab("Sample") +
+  theme(legend.position = "top")
+
+```
+
+### Prediction Probabilities {#pred_prob}
+I selected the samples where the difference in the correct classification frequency between optifit and opticlust was greater than 0.5 as shown below.
+
+```{r predprob_subset}
+
+samp_order <- pct_correct %>% 
+  pivot_wider(id_cols = c("Group","dx"),
+              names_from = algorithm,
+              values_from = pct_correct) %>% 
+  #filter(optifit != 1 & opticlust != 1) %>% 
+  mutate(difference = abs(opticlust - optifit)) %>% 
+  filter(difference > 0.5) %>% 
+  arrange(difference,optifit) %>% 
+  pull(Group)
+
+pct_correct %>% 
+  filter(Group %in% samp_order) %>% 
+  ggplot(aes(x=factor(Group,levels = samp_order),y=pct_correct,color=algorithm)) +
+  geom_point(alpha=0.7) +
+  facet_wrap(.~dx,nrow=2,scales="free_x") +
+  theme_bw() + 
+  ylab("Percent of splits sample is correctly classified (TP or TN)") +
+  theme(axis.text.x = element_blank(),
+        axis.ticks.x = element_blank()) +
+  xlab("Sample")
+
+```
+
+For each of the `r length(samp_order)` samples in this set, I plotted the probability of cancer from each split this sample was in the test set for.
+The dashed line is the 0.5 probability line.
+Using the 0.5 threshold, a point above the line means the sample would be classified as cancer in that split, below the line normal.
+For the cancer samples, a point above the line is a true positive while for the normal samples a point above the line is a false positive
+
+```{r predprob, fig.height = 10, fig.width = 9}
+mergedProb <- read_csv("../data/learning/summary/merged_predictions.csv",
+                       col_types = cols(cancer=col_double(),
+                                        normal=col_double(),
+                                        .default=col_character())) %>%
+  mutate(GroupDX = paste0(Group," ",dx))
+
+ids = as.character(samp_order)
+
+## NORMAL
+mergedProb %>%
+  filter(Group %in% ids) %>%
+  filter(dx == "normal") %>% 
+  ggplot(aes(x=algorithm,y=cancer,color=algorithm)) +
+    geom_jitter(width = 0.2,height=0,alpha=0.5,size=1.8) +
+    facet_wrap(.~GroupDX) +
+    theme_bw() +
+    geom_hline(yintercept = 0.5,lty="dashed",color="grey") +
+    ylab("Probability of SRN") + xlab("") +
+    scale_y_continuous(limits=c(0,1),breaks=seq(0.2,0.8,0.2)) +
+    theme(axis.text.x = element_blank(),
+          axis.ticks.x = element_blank(),
+          strip.background = element_rect(fill="deepskyblue3", size=1, color="black"),
+          strip.text.x = element_text(face="bold", color="white"),
+          legend.position = "top")
+## CANCER
+mergedProb %>%
+  filter(Group %in% ids) %>%
+  mutate(Group=factor(Group,levels=ids)) %>% 
+  filter(dx == "cancer") %>% 
+  ggplot(aes(x=algorithm,y=cancer,color=algorithm)) +
+    geom_jitter(width = 0.2,height=0,alpha=0.5,size=1.8) +
+    facet_wrap(.~GroupDX) +
+    theme_bw() +
+    geom_hline(yintercept = 0.5,lty="dashed",color="grey") +
+    ylab("Probability of SRN") + xlab("") +
+    scale_y_continuous(limits=c(0,1),breaks=seq(0.2,0.8,0.2)) +
+    theme(axis.text.x = element_blank(),
+          axis.ticks.x = element_blank(),
+          strip.background = element_rect(fill="brown3", size=1, color="black"),
+          strip.text.x = element_text(face="bold", color="white"),
+          legend.position = "top")
+
+```
+
+For comparison, here are a subset of samples where there is not a difference in the percent correctly classified between optifit and opticlust
+
+```{r predprob_subset2}
+
+samp_order1 <- pct_correct %>% 
+  pivot_wider(id_cols = c("Group","dx"),
+              names_from = algorithm,
+              values_from = pct_correct) %>% 
+  filter(optifit == 1 & opticlust == 1) %>% 
+  mutate(difference = abs(opticlust - optifit)) %>% 
+  arrange(difference,optifit) %>% 
+  sample_n(10) %>% #randomly select 5
+  pull(Group)
+
+samp_order2 <- pct_correct %>% 
+  pivot_wider(id_cols = c("Group","dx"),
+              names_from = algorithm,
+              values_from = pct_correct) %>% 
+  filter(optifit < 0.9 & opticlust < 0.9) %>% 
+  filter(optifit > 0.1 & opticlust > 0.1) %>% 
+  mutate(difference = abs(opticlust - optifit)) %>% 
+  filter(difference == 0) %>% 
+  sample_n(min(10,nrow(.))) %>% #randomly select 5
+  arrange(optifit) %>% 
+  pull(Group)
+
+samp_order3 <- pct_correct %>% 
+  pivot_wider(id_cols = c("Group","dx"),
+              names_from = algorithm,
+              values_from = pct_correct) %>% 
+  filter(optifit == 0 & opticlust == 0) %>% 
+  mutate(difference = abs(opticlust - optifit)) %>% 
+  arrange(difference,optifit) %>% 
+  sample_n(10) %>% #randomly select 5
+  pull(Group)
+
+samp_order <- c(samp_order1,samp_order2,samp_order3)
+
+pct_correct %>% 
+  filter(Group %in% samp_order) %>% 
+  ggplot(aes(x=factor(Group,levels = samp_order),y=pct_correct,color=algorithm)) +
+  geom_point(alpha=0.7) +
+  facet_wrap(.~dx,nrow=2,scales="free_x") +
+  theme_bw() + 
+  ylab("Percent of splits sample is correctly classified (TP or TN)") +
+  theme(axis.text.x = element_blank(),
+        axis.ticks.x = element_blank()) +
+  xlab("Sample")
+
+```
+
+```{r predprob2, fig.height = 10, fig.width = 9}
+mergedProb <- read_csv("../data/learning/summary/merged_predictions.csv",
+                       col_types = cols(cancer=col_double(),
+                                        normal=col_double(),
+                                        .default=col_character())) %>%
+  mutate(GroupDX = paste0(Group," ",dx))
+
+ids = as.character(samp_order)
+
+## NORMAL
+mergedProb %>%
+  filter(Group %in% ids) %>%
+  filter(dx == "normal") %>% 
+  ggplot(aes(x=algorithm,y=cancer,color=algorithm)) +
+    geom_jitter(width = 0.2,height=0,alpha=0.5,size=1.8) +
+    facet_wrap(.~GroupDX) +
+    theme_bw() +
+    geom_hline(yintercept = 0.5,lty="dashed",color="grey") +
+    ylab("Probability of SRN") + xlab("") +
+    scale_y_continuous(limits=c(0,1),breaks=seq(0.2,0.8,0.2)) +
+    theme(axis.text.x = element_blank(),
+          axis.ticks.x = element_blank(),
+          strip.background = element_rect(fill="deepskyblue3", size=1, color="black"),
+          strip.text.x = element_text(face="bold", color="white"),
+          legend.position = "top")
+## CANCER
+mergedProb %>%
+  filter(Group %in% ids) %>%
+  mutate(Group=factor(Group,levels=ids)) %>% 
+  filter(dx == "cancer") %>% 
+  ggplot(aes(x=algorithm,y=cancer,color=algorithm)) +
+    geom_jitter(width = 0.2,height=0,alpha=0.5,size=1.8) +
+    facet_wrap(.~GroupDX) +
+    theme_bw() +
+    geom_hline(yintercept = 0.5,lty="dashed",color="grey") +
+    ylab("Probability of SRN") + xlab("") +
+    scale_y_continuous(limits=c(0,1),breaks=seq(0.2,0.8,0.2)) +
+    theme(axis.text.x = element_blank(),
+          axis.ticks.x = element_blank(),
+          strip.background = element_rect(fill="brown3", size=1, color="black"),
+          strip.text.x = element_text(face="bold", color="white"),
+          legend.position = "top")
+
+```
+
+## Does mcc correlate with sensitivity/specificity? {#mcc_senspec}
+
+OptiFit MCC values from just the fit data compared to sensitivity/specificity (threshold = 0.5)
+
+```{r mcc_sensspec}
+library(caret)
+library(assertthat)
+
+get_sensspec_thrsh <- function(senspec,threshold){
+  assert_that(is.numeric(threshold))
+
+  senspec %>%
+    mutate(prediction = case_when(cancer >= threshold ~ "cancer",
+                                  TRUE ~ "normal")) %>%
+    select(-cancer,-normal) %>% 
+    group_by(split,algorithm) %>% 
+    summarize()
+
+  senspec %>%
+    mutate(prediction = case_when(cancer >= 0.5 ~ "cancer",
+                                  TRUE ~ "normal")) %>%
+    mutate(dx = factor(dx,levels=c("normal","cancer")),
+           prediction = factor(prediction,levels=c("normal","cancer"))) %>% 
+    select(-cancer,-normal) %>% 
+    mutate(category = case_when(dx == "normal" & prediction == "normal" ~ "TN",
+                                dx == "cancer" & prediction == "cancer" ~ "TP",
+                                dx == "cancer" & prediction == "normal" ~ "FN",
+                                dx == "normal" & prediction == "cancer" ~ "FP")) %>% 
+    group_by(split,algorithm) %>% 
+    count(category) %>% 
+    pivot_wider(id_cols=c("split","algorithm"),names_from = category,values_from = n) %>% 
+    group_by(split,algorithm) %>% 
+    summarise(sensitivity = TP / (TP+FN),
+              specificity = TN / (TN+FP))
+
+}
+
+senspec_50 <- get_sensspec_thrsh(read_csv("../data/learning/summary/merged_predictions.csv"),0.5)
+
+# senspec_50 %>% 
+#   filter(algorithm == "optifit")
+
+merged_mcc %>% 
+  filter(label=="OptiFit Test" & state=="fit") %>% 
+  select(-subset,-Group) %>% 
+  inner_join(senspec_50 %>% filter(algorithm == "optifit"),by=c("split","algorithm")) %>% 
+  ggplot(aes(x=mcc,y=sensitivity)) +
+    geom_point() +
+    theme_bw()
+
+merged_mcc %>% 
+  filter(label=="OptiFit Test" & state=="fit") %>% 
+  select(-subset,-Group) %>% 
+  inner_join(senspec_50 %>% filter(algorithm == "optifit"),by=c("split","algorithm")) %>% 
+  ggplot(aes(x=mcc,y=specificity)) +
+    geom_point() +
+    theme_bw()
+
+```
+
+```{r test}
+# 
+# get_categories <- function(filename,threshold){
+#   read_csv(filename,col_types = cols(Group = col_character(),
+#                                     dx = col_character(),
+#                                     .default = col_double())) %>% 
+#     mutate(algorithm = str_replace(filename,
+#                                    "../data/learning/results/(.*)/prediction_results_split_(\\d*).csv", "\\1"),
+#            split = str_replace(filename,
+#                                "../data/learning/results/(.*)/prediction_results_split_(\\d*).csv", "\\2"),
+#            pred_dx = case_when(cancer > threshold ~ "cancer",
+#                                TRUE ~ "normal")) %>% 
+#     filter(Group %in% c("2005650","2009650","2025653","2671660")) %>% 
+#     select(Group,dx,pred_dx,algorithm,split) %>% 
+#     mutate(category = case_when(dx == "normal" & pred_dx == "normal" ~ "TN",
+#                                 dx == "normal" & pred_dx == "cancer" ~ "FP",
+#                                 dx == "cancer" & pred_dx == "cancer" ~ "TP",
+#                                 dx == "cancer" & pred_dx == "normal" ~ "FN",
+#                                 TRUE ~ "NA"))
+# }
+# 
+# get_pct_correct <- function(algorithm, threshold){
+#   #get_pred_file_list(algorithm) %>% 
+#   filelist <- c("../data/learning/results/optifit/prediction_results_split_1.csv",
+#                 "../data/learning/results/optifit/prediction_results_split_2.csv",
+#                 "../data/learning/results/optifit/prediction_results_split_3.csv",
+#                 "../data/learning/results/optifit/prediction_results_split_4.csv",
+#                 "../data/learning/results/optifit/prediction_results_split_5.csv")
+  # filelist <- c("../data/learning/results/optifit/prediction_results_split_1.csv",
+  #               "../data/learning/results/optifit/prediction_results_split_2.csv",
+  #               "../data/learning/results/optifit/prediction_results_split_3.csv",
+  #               "../data/learning/results/optifit/prediction_results_split_4.csv",
+  #               "../data/learning/results/optifit/prediction_results_split_5.csv")
+#   filelist %>% 
+#     map_dfr(get_categories,threshold) %>% 
+#     select(Group,category,dx) %>%
+#     group_by(Group,dx) %>% 
+#     count(category) %>% 
+#     pivot_wider(id_cols=c("Group","dx"),names_from=category,values_from=n,values_fill=0) %>% 
+#     mutate(correct = TN + TP,
+#            incorrect = FN + FP,
+#            total = TN + TP + FN + FP) %>% 
+#     mutate(pct_correct = correct/total) %>% 
+#     select(Group,dx,pct_correct) %>% 
+#     mutate(algorithm = algorithm)
+# }
+# 
+# pct_correct <- map_dfr(algorithms, get_pct_correct, threshold=0.5) %>% 
+#   mutate(Group=as.character(Group))
+# 
+# pct_correct %>% 
+#   ggplot(aes(x=algorithm,y=pct_correct,group=Group,color=dx)) +
+#     geom_jitter(height=0,width=0.2) +
+#     theme_bw() +
+#     ylab("Percent of splits sample is correctly classified (TP or TN)")
+
+```
+
+```{r }
+
+### Prediction Probability 
+# opticlust_summary %>% 
+#   ggplot(aes(x=cancer,fill=meta_class)) +
+#     geom_histogram(color="black",alpha=0.8,binwidth = 0.05) + theme_bw() +
+#     labs(xlab = "Probability of Cancer") + 
+#     facet_wrap(~meta_class) +
+#     xlab("probability of cancer") +
+#     geom_vline(xintercept = 0.5,lty="dashed") +
+#     scale_y_continuous(expand=c(0,0)) +
+#     scale_fill_manual(values=c("red","blue"),guide=NULL) +
+#     ggtitle("OptiClust")
+
+# optifit_summary %>% 
+#   ggplot(aes(x=cancer,fill=meta_class)) +
+#     geom_histogram(color="black",alpha=0.8,binwidth = 0.05) + theme_bw() +
+#     labs(xlab = "Probability of Cancer") + 
+#     facet_wrap(~meta_class) +
+#     xlab("probability of cancer") +
+#     geom_vline(xintercept = 0.5,lty="dashed") +
+#     scale_y_continuous(expand=c(0,0)) +
+#     scale_fill_manual(values=c("red","blue"),guide=NULL) +
+#     ggtitle("OptiFit")
+
+```
+
+```{r }
+### Difference in prediction probablity  
+# ggplot(opticlust_summary,aes(x=dx_diff)) +
+#   geom_histogram(binwidth = 0.05,color="black",fill="grey") +
+#   labs(title = "OptiClust: Difference in DX Prediction",
+#        x="Difference in Probability") +
+#   theme_bw() +
+#   scale_x_continuous(expand=c(0,0)) +
+#   scale_y_continuous(expand=c(0,0))
+# 
+# ggplot(optifit_summary,aes(x=dx_diff)) +
+#   geom_histogram(binwidth = 0.05,color="black",fill="grey") +
+#   labs(title = "OptiFit: Difference in DX Prediction",
+#        x="Difference in Probability") +
+#   theme_bw() +
+#   scale_x_continuous(expand=c(0,0)) +
+#   scale_y_continuous(expand=c(0,0))
+
+# opticlust_summary %>% mutate(method="OptiClust") %>% 
+#   bind_rows(optifit_summary %>% mutate(method="OptiFit")) %>% 
+#   ggplot(aes(x=dx_diff)) +
+#     geom_histogram(binwidth = 0.05,color="black",fill="grey") +
+#     labs(title = "Difference in DX Prediction",
+#          x="Difference in Probability") +
+#     theme_bw() +
+#     scale_x_continuous(expand=c(0,0)) +
+#     scale_y_continuous(expand=c(0,0)) +
+#     facet_wrap(~method)
+
+```
+
+```{r }
+### Summary of performance with a threshold of 0.5
+# opticlust_cm <- confusionMatrix(data=factor(opticlust_summary$pred_class),reference = factor(opticlust_summary$meta_class))
+# optifit_cm <- confusionMatrix(data=factor(optifit_summary$pred_class),reference = factor(optifit_summary$meta_class))
+
+# kable(opticlust_cm$table,caption="OptiClust Confusion Matrix") %>% 
+#   kable_styling(full_width = F,position = "float_left")
+# kable(optifit_cm$table,caption="OptiFit Confusion Matrix") %>% 
+#   kable_styling(full_width = F,position = "left")
+
+# cm_summary <- round(opticlust_cm$byClass,digits=3) %>% 
+#   bind_rows(round(optifit_cm$byClass,digits=3)) %>% 
+#   mutate(Method=c("OptiClust","OptiFit")) %>% 
+#   select(Method,everything())
+
+# kable(cm_summary,align="c",booktabs=TRUE) %>% 
+#   kable_classic()
+```
+
+```{r }
+### ROC plot  
+#opticlust
+# response <- ifelse(opticlust_summary$meta_class == "cancer",1,0)
+# predictor <- opticlust_summary$cancer
+
+# clust_roc <- roc(response,predictor)
+# opticlust_thr <- coords(clust_roc, "best", ret = "threshold",transpose = TRUE)
+
+# #optifit
+# response <- ifelse(optifit_summary$meta_class == "normal",0,1)
+# predictor <- optifit_summary$normal
+
+# fit_roc <- roc(response,predictor)
+# optifit_thr <- coords(fit_roc, "best", ret = "threshold",transpose = TRUE)
+
+# #plot
+# ggroc(list(OptiClust=clust_roc,OptiFit=fit_roc),legacy.axes=TRUE) +
+#   theme_bw() + 
+#   geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1),
+#                  color="darkgrey", linetype="dashed") +
+#   scale_y_continuous(expand=c(0,0)) +
+#   scale_x_continuous(expand=c(0,0))
+
+```
+
+```{r}
+# OptiClust AUC: `r round(clust_roc$auc,digits=3)`  
+# OptiFit AUC: `r round(fit_roc$auc,digits=3)`
+
+# Optimal threshold for OptiClust: `r round(opticlust_thr,digits=3)`  
+# Optimal threshold for OptiFit: `r round(optifit_thr,digits=3)`  
+```
 
@@
 ```
-
-### Correct Classification Frequency {#class_correct_freq}
-
-For each split, a cutoff of 0.5 was used to determine whether each sample was correctly or incorrectly classified.
-This data was merged for all splits to quantify the percent of splits each sample was correctly classified in.
-For normal samples it is the frequency that sample was a true negative and for cancer samples its the frequency that sample was a true positive.
-For example, if a cancer sample had a probabily of being cancer greater than 0.5 for 10 of the 20 splits it was in the test set for, it would have a percent correct value of 50%.
-I'm trying to figure out how to visualize this data...
-
-```{r pct_correct_hist, out.width="50%"}
-pct_correct <- read_csv("../results/tables/pct_class_correct.csv")
-
-# pct_correct %>%
-#   ggplot(aes(x=factor(algorithm,levels=c("opticlust","optifit"),labels=c("OptiClust","OptiFit")),
-#              y=pct_correct,group=Group,color=dx)) +
-#     geom_jitter(height=0,width=0.2,alpha=0.7,size=1.8) +
-#     theme_bw() + xlab("") +
-#     ylab("Percent of splits sample is correctly classified (TP or TN)") +
-#     scale_y_continuous(labels = scales::percent_format(accuracy=1),
-#                        breaks = seq(0,1,0.1)) +
-#     theme(panel.grid.major.y = element_line(color="grey85")) +
-#     scale_color_manual(values=c("brown3","deepskyblue3"),name="Actual Diagnosis")
-#
-
-pct_correct %>%
-  ggplot(aes(x=pct_correct,fill=algorithm)) +
-  geom_histogram(binwidth=0.1,color="black",alpha=0.5,position = "identity") +
-  facet_grid(.~dx) +
-  theme_bw() +
-  xlab("Percent of splits sample is correctly classified (TP or TN)") +
-  theme(legend.position = "top")
-
-pct_correct %>%
-  ggplot(aes(x=pct_correct,fill=algorithm)) +
-  geom_histogram(binwidth=0.1,color="black",alpha=0.5,position = "dodge") +
-  facet_grid(.~dx) +
-  theme_bw() +
-  xlab("Percent of splits sample is correctly classified (TP or TN)") +
-  theme(legend.position = "top")
-
-pct_correct %>%
-  ggplot(aes(x=pct_correct,fill=algorithm)) +
-  #geom_histogram(binwidth=0.1,color="black",alpha=0.8) +
-  geom_density(alpha=0.5) +
-  facet_grid(.~dx) +
-  theme_bw() +
-  xlab("Percent of splits sample is correctly classified (TP or TN)") +
-  theme(legend.position = "top")
-
-pct_correct %>%
-  ggplot(aes(x=pct_correct,color=algorithm)) +
-  #geom_histogram(binwidth=0.1,color="black",alpha=0.8) +
-  geom_density(alpha=0.5) +
-  facet_grid(.~dx) +
-  theme_bw() +
-  xlab("Percent of splits sample is correctly classified (TP or TN)") +
-  theme(legend.position = "top")
-
-```
-
-The scatter plot below has the samples along the x-axis without labels because they are not readable.
-The y-axis is the percent correct.
-The color of the points is the algorithm.
-There is a separate plot for cancer samples and normal samples.
-
-```{r pct_correct_point}
-#order samples by difference in pct_correct between algorithms
-samp_order <- pct_correct %>% 
-  pivot_wider(id_cols = c("Group","dx"),
-              names_from = algorithm,
-              values_from = pct_correct) %>% 
-  mutate(difference = abs(opticlust - optifit)) %>% 
-  arrange(difference) %>% 
-  pull(Group)
-
-pct_correct %>% 
-  ggplot(aes(x=factor(Group,levels = samp_order),y=pct_correct,color=algorithm)) +
-  geom_point(alpha=0.7) +
-  facet_wrap(.~dx,nrow=2) +
-  theme_bw() + 
-  ylab("Percent of splits sample is correctly classified (TP or TN)") +
-  theme(axis.text.x = element_blank(),
-        axis.ticks.x = element_blank()) +
-  xlab("Sample")
-```
-
-To eliminate some noise, I removed any samples where there was no difference in the percent correct between the two algorithms.
-
-```{r mod_pct_correct_point}
-#get rid of samples that are perfectly classified with both algorithms
-#and samples that have the samp pct_correct
-#order samples by difference in pct_correct between algorithms
-samp_order <- pct_correct %>% 
-  pivot_wider(id_cols = c("Group","dx"),
-              names_from = algorithm,
-              values_from = pct_correct) %>% 
-  filter(optifit != 1 & opticlust != 1) %>% 
-  mutate(difference = abs(opticlust - optifit)) %>% 
-  filter(difference > 0) %>% 
-  arrange(difference) %>% 
-  pull(Group)
-
-pct_correct %>% 
-  filter(Group %in% samp_order) %>% 
-  ggplot(aes(x=factor(Group,levels = samp_order),y=pct_correct,color=algorithm)) +
-  geom_point(alpha=0.7) +
-  facet_wrap(.~dx,nrow=2) +
-  theme_bw() + 
-  ylab("Percent of splits sample is correctly classified (TP or TN)") +
-  theme(axis.text.x = element_blank(),
-        axis.ticks.x = element_blank()) +
-  xlab("Sample")
-
-```
-
-```{r test}
-# 
-# get_categories <- function(filename,threshold){
-#   read_csv(filename,col_types = cols(Group = col_character(),
-#                                     dx = col_character(),
-#                                     .default = col_double())) %>% 
-#     mutate(algorithm = str_replace(filename,
-#                                    "../data/learning/results/(.*)/prediction_results_split_(\\d*).csv", "\\1"),
-#            split = str_replace(filename,
-#                                "../data/learning/results/(.*)/prediction_results_split_(\\d*).csv", "\\2"),
-#            pred_dx = case_when(cancer > threshold ~ "cancer",
-#                                TRUE ~ "normal")) %>% 
-#     filter(Group %in% c("2005650","2009650","2025653","2671660")) %>% 
-#     select(Group,dx,pred_dx,algorithm,split) %>% 
-#     mutate(category = case_when(dx == "normal" & pred_dx == "normal" ~ "TN",
-#                                 dx == "normal" & pred_dx == "cancer" ~ "FP",
-#                                 dx == "cancer" & pred_dx == "cancer" ~ "TP",
-#                                 dx == "cancer" & pred_dx == "normal" ~ "FN",
-#                                 TRUE ~ "NA"))
-# }
-# 
-# get_pct_correct <- function(algorithm, threshold){
-#   #get_pred_file_list(algorithm) %>% 
-#   filelist <- c("../data/learning/results/optifit/prediction_results_split_1.csv",
-#                 "../data/learning/results/optifit/prediction_results_split_2.csv",
-#                 "../data/learning/results/optifit/prediction_results_split_3.csv",
-#                 "../data/learning/results/optifit/prediction_results_split_4.csv",
-#                 "../data/learning/results/optifit/prediction_results_split_5.csv")
-  # filelist <- c("../data/learning/results/optifit/prediction_results_split_1.csv",
-  #               "../data/learning/results/optifit/prediction_results_split_2.csv",
-  #               "../data/learning/results/optifit/prediction_results_split_3.csv",
-  #               "../data/learning/results/optifit/prediction_results_split_4.csv",
-  #               "../data/learning/results/optifit/prediction_results_split_5.csv")
-#   filelist %>% 
-#     map_dfr(get_categories,threshold) %>% 
-#     select(Group,category,dx) %>%
-#     group_by(Group,dx) %>% 
-#     count(category) %>% 
-#     pivot_wider(id_cols=c("Group","dx"),names_from=category,values_from=n,values_fill=0) %>% 
-#     mutate(correct = TN + TP,
-#            incorrect = FN + FP,
-#            total = TN + TP + FN + FP) %>% 
-#     mutate(pct_correct = correct/total) %>% 
-#     select(Group,dx,pct_correct) %>% 
-#     mutate(algorithm = algorithm)
-# }
-# 
-# pct_correct <- map_dfr(algorithms, get_pct_correct, threshold=0.5) %>% 
-#   mutate(Group=as.character(Group))
-# 
-# pct_correct %>% 
-#   ggplot(aes(x=algorithm,y=pct_correct,group=Group,color=dx)) +
-#     geom_jitter(height=0,width=0.2) +
-#     theme_bw() +
-#     ylab("Percent of splits sample is correctly classified (TP or TN)")
-
-```
-
-```{r }
-
-### Prediction Probability 
-# opticlust_summary %>% 
-#   ggplot(aes(x=cancer,fill=meta_class)) +
-#     geom_histogram(color="black",alpha=0.8,binwidth = 0.05) + theme_bw() +
-#     labs(xlab = "Probability of Cancer") + 
-#     facet_wrap(~meta_class) +
-#     xlab("probability of cancer") +
-#     geom_vline(xintercept = 0.5,lty="dashed") +
-#     scale_y_continuous(expand=c(0,0)) +
-#     scale_fill_manual(values=c("red","blue"),guide=NULL) +
-#     ggtitle("OptiClust")
-
-# optifit_summary %>% 
-#   ggplot(aes(x=cancer,fill=meta_class)) +
-#     geom_histogram(color="black",alpha=0.8,binwidth = 0.05) + theme_bw() +
-#     labs(xlab = "Probability of Cancer") + 
-#     facet_wrap(~meta_class) +
-#     xlab("probability of cancer") +
-#     geom_vline(xintercept = 0.5,lty="dashed") +
-#     scale_y_continuous(expand=c(0,0)) +
-#     scale_fill_manual(values=c("red","blue"),guide=NULL) +
-#     ggtitle("OptiFit")
-
-```
-
-```{r }
-### Difference in prediction probablity  
-# ggplot(opticlust_summary,aes(x=dx_diff)) +
-#   geom_histogram(binwidth = 0.05,color="black",fill="grey") +
-#   labs(title = "OptiClust: Difference in DX Prediction",
-#        x="Difference in Probability") +
-#   theme_bw() +
-#   scale_x_continuous(expand=c(0,0)) +
-#   scale_y_continuous(expand=c(0,0))
-# 
-# ggplot(optifit_summary,aes(x=dx_diff)) +
-#   geom_histogram(binwidth = 0.05,color="black",fill="grey") +
-#   labs(title = "OptiFit: Difference in DX Prediction",
-#        x="Difference in Probability") +
-#   theme_bw() +
-#   scale_x_continuous(expand=c(0,0)) +
-#   scale_y_continuous(expand=c(0,0))
-
-# opticlust_summary %>% mutate(method="OptiClust") %>% 
-#   bind_rows(optifit_summary %>% mutate(method="OptiFit")) %>% 
-#   ggplot(aes(x=dx_diff)) +
-#     geom_histogram(binwidth = 0.05,color="black",fill="grey") +
-#     labs(title = "Difference in DX Prediction",
-#          x="Difference in Probability") +
-#     theme_bw() +
-#     scale_x_continuous(expand=c(0,0)) +
-#     scale_y_continuous(expand=c(0,0)) +
-#     facet_wrap(~method)
-
-```
-
-```{r }
-### Summary of performance with a threshold of 0.5
-# opticlust_cm <- confusionMatrix(data=factor(opticlust_summary$pred_class),reference = factor(opticlust_summary$meta_class))
-# optifit_cm <- confusionMatrix(data=factor(optifit_summary$pred_class),reference = factor(optifit_summary$meta_class))
-
-# kable(opticlust_cm$table,caption="OptiClust Confusion Matrix") %>% 
-#   kable_styling(full_width = F,position = "float_left")
-# kable(optifit_cm$table,caption="OptiFit Confusion Matrix") %>% 
-#   kable_styling(full_width = F,position = "left")
-
-# cm_summary <- round(opticlust_cm$byClass,digits=3) %>% 
-#   bind_rows(round(optifit_cm$byClass,digits=3)) %>% 
-#   mutate(Method=c("OptiClust","OptiFit")) %>% 
-#   select(Method,everything())
-
-# kable(cm_summary,align="c",booktabs=TRUE) %>% 
-#   kable_classic()
-```
-
-```{r }
-### ROC plot  
-#opticlust
-# response <- ifelse(opticlust_summary$meta_class == "cancer",1,0)
-# predictor <- opticlust_summary$cancer
-
-# clust_roc <- roc(response,predictor)
-# opticlust_thr <- coords(clust_roc, "best", ret = "threshold",transpose = TRUE)
-
-# #optifit
-# response <- ifelse(optifit_summary$meta_class == "normal",0,1)
-# predictor <- optifit_summary$normal
-
-# fit_roc <- roc(response,predictor)
-# optifit_thr <- coords(fit_roc, "best", ret = "threshold",transpose = TRUE)
-
-# #plot
-# ggroc(list(OptiClust=clust_roc,OptiFit=fit_roc),legacy.axes=TRUE) +
-#   theme_bw() + 
-#   geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1),
-#                  color="darkgrey", linetype="dashed") +
-#   scale_y_continuous(expand=c(0,0)) +
-#   scale_x_continuous(expand=c(0,0))
-
-```
-
-```{r}
-# OptiClust AUC: `r round(clust_roc$auc,digits=3)`  
-# OptiFit AUC: `r round(fit_roc$auc,digits=3)`
-
-# Optimal threshold for OptiClust: `r round(opticlust_thr,digits=3)`  
-# Optimal threshold for OptiFit: `r round(optifit_thr,digits=3)`  
-```

### 2022-02-23 submission/manuscript.tex
manuscript outline

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  11pt,
]{article}
\title{\textbf{TITLE}}
\author{}
\date{\vspace{-2.5em}}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={TITLE},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1.0in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

\usepackage[none]{hyphenat}

\usepackage{setspace}
\doublespacing
\setlength{\parskip}{1em}

\usepackage{lineno}

\usepackage{pdfpages}
\usepackage{helvet}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\begin{document}
\maketitle

\vspace{20mm}

Running title: INSERT RUNNING TITLE HERE

\vspace{20mm}

Courtney R. Armour\({^1}\), William L. Close\(^{1,*}\), Begüm D.
Topçuoğlu\(^{1,\#}\), Patrick D. Schloss \(^{1,\dagger}\)

\vspace{5mm}

\({^1}\) Department of Microbiology and Immunology, University of
Michigan, Ann Arbor MI.

\({^\#}\) Current Affiliation: Bristol Myers Squibb, Summit, New Jersey,
USA~

\({^*}\) Current Affiliation:

\vspace{20mm}

\(\dagger\) To whom correspondence should be addressed:
\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}

\vspace{20mm}

\textbf{observation format} (max 1200 words, 2 figures, 25 ref)

\newpage

\linenumbers

\hypertarget{abstract-250-word-max}{%
\subsection{Abstract (250 word max)}\label{abstract-250-word-max}}

\hypertarget{importance-150-word-max}{%
\subsection{Importance (150 word max)}\label{importance-150-word-max}}

\newpage

\hypertarget{introduction-250-words}{%
\subsection{Introduction (\textasciitilde250
words)}\label{introduction-250-words}}

\begin{itemize}
\tightlist
\item
  Gut microbiome community composition has proven useful as a resource
  for machine learning prediction of various diseases \{examples\}.
\item
  Amplicon sequencing of the 16S rRNA gene is a reliable tool for
  assessing the taxonomic composition of microbial communities.
\item
  Analysis of 16S rRNA sequence data generally relies on clustering of
  sequences based on similarity into operational taxonomic units (OTUs).
\item
  However OTU clustering depends on the data in the dataset and the
  addition of new data may change the overall OTU clusters.
\item
  The unstable nature of OTU clustering complicates machine learning.
  When building a classification model, changes to the OTU clusters
  means you have to re-create the model. This can change the underlying
  model and be time consuming and resource intensive.
\item
  The ability to integrate new data into an existing model could allow
  for deployment of a single model that new data can be continually
  added to and predicted on.
\item
  Recently Sovacool \emph{et al} described a new method for fitting new
  data into existing OTU clusters \{Kelly optifit 2022\}.
\item
  While OptiFit works well to fit new sequence data and provide high
  quality OTU clusters, it is unknown if the use of OptiFit will have an
  impact on machine learning predictions.
\item
  Here, we use OptiFit with a 16S rRNA sequence dataset consisting of
  normal and SRN samples to test how well new data integrated with
  OptiFit performs for prediction of SRN.
\end{itemize}

\hypertarget{results-700-words}{%
\subsection{Results (\textasciitilde700
words)}\label{results-700-words}}

\begin{itemize}
\tightlist
\item
  Utilize public dataset with normal and SRN samples to compare
  prediction between OptiFit and OptiClust
\item
  randomly split data into 80\% training 20\% test sets 100 times
\item
  Processed the data with both algorithms - Figure 1

  \begin{itemize}
  \item
    \begin{enumerate}
    \def\labelenumi{\arabic{enumi}.}
    \tightlist
    \item
      Used traditional OptiClust method to cluster all data, then split
      into training and test set
    \end{enumerate}
  \item
    \begin{enumerate}
    \def\labelenumi{\arabic{enumi}.}
    \setcounter{enumi}{1}
    \tightlist
    \item
      Used OptiClust on the 80\% training set, then used OptiFit to fit
      the remaining 20\%
    \end{enumerate}
  \end{itemize}
\item
  OptiFit produces similar quality OTU clusters based on MCC
  (supplement?)
\item
  Used mikropml to train a model on the 80\% training set for each data
  split, then predicted dx on the 20\% test set
\item
  Training performance almost identical (OptiClust median CV AUC 0.694,
  OptiFit median CV AUC 0.693) - Figure 2
\item
  Performance on the test set comparable (OptiClust median CV AUC 0.694,
  OptiFit median CV AUC 0.693) - Figure 2
\item
  Maybe expand on where optifit did better/worse?
\end{itemize}

\hypertarget{discussionconclusions-250-words}{%
\subsection{Discussion/Conclusions (\textasciitilde250
words)}\label{discussionconclusions-250-words}}

\begin{itemize}
\tightlist
\item
  OptiFit works!
\item
  future questions:

  \begin{itemize}
  \tightlist
  \item
    how much reference do you need?
  \item
    does it work well for other situations?

    \begin{itemize}
    \tightlist
    \item
      what if we used a new dataset for test set instead of a subset of
      the full dataset?
    \item
      other diseases?
    \end{itemize}
  \end{itemize}
\end{itemize}

\hypertarget{materials-and-methods}{%
\subsection{Materials and Methods}\label{materials-and-methods}}

\begin{itemize}
\tightlist
\item
  16S rRNA amplicon sequence data from 490 subjects \{baxter\}

  \begin{itemize}
  \tightlist
  \item
    261 controls
  \item
    229 SRN
  \end{itemize}
\item
  created 100 random splits of the data (80\% training, 20\% test)
\item
  preprocessed data with mothur v1.45
\item
  two pathways\\
  -opticlust:

  \begin{itemize}
  \tightlist
  \item
    clustered all data together
  \item
    split the shared file based on the random splits
  \end{itemize}

  -optifit:

  \begin{itemize}
  \tightlist
  \item
    split the data
  \item
    opticlust on the 80\% training set
  \item
    optifit to fit the remaining 20\% to the training set OTUs
  \end{itemize}
\item
  ML with mikropml package (version)

  \begin{itemize}
  \tightlist
  \item
    preprocessed training set and applied preprocessing to test set

    \begin{itemize}
    \tightlist
    \item
      correlated collapsed, removed nzv
    \end{itemize}
  \end{itemize}
\end{itemize}

\hypertarget{acknowledgements}{%
\subsection{Acknowledgements}\label{acknowledgements}}

\begin{itemize}
\tightlist
\item
  funding
\end{itemize}

\newpage

\hypertarget{figures}{%
\subsection{Figures}\label{figures}}

\textbf{Figure 1. Workflow} description.

\textbf{Figure 2. Model Performance.} \textbf{A)} Mean AUC \textbf{B)}
Averaged ROC curves

\newpage

\hypertarget{references}{%
\subsection{References}\label{references}}

\end{document}

### 2022-02-23 submission/manuscript.Rmd

---
title: "**TITLE**"
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: header.tex
csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
fontsize: 11pt
geometry: margin=1.0in
header-includes:
  - \usepackage{helvet}
---

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
library(knitr)
opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("cache" = FALSE)

inline_hook <- function(x){
	print(x)

	if(is.list(x)){
		x <- unlist(x)
	}

	if(is.numeric(x)){
		if(abs(x - round(x)) < .Machine$double.eps^0.5){
			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
		} else {
			paste(format(x,big.mark=',', digits=1, nsmall=1, scientific=FALSE))
		}
	} else {
    	paste(x)      
	}
}
knitr::knit_hooks$set(inline=inline_hook)
```

\vspace{20mm}

Running title: INSERT RUNNING TITLE HERE

\vspace{20mm}

Courtney R. Armour${^1}$, William L. Close$^{1,*}$, Begüm D. Topçuoğlu$^{1,\#}$, Patrick D. Schloss $^{1,\dagger}$

\vspace{5mm}

${^1}$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor MI.

${^\#}$ Current Affiliation: Bristol Myers Squibb, Summit, New Jersey, USA

${^*}$ Current Affiliation:

\vspace{20mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

\vspace{20mm}

**observation format** (max 1200 words, 2 figures, 25 ref)

\newpage

\linenumbers

## Abstract (250 word max)

## Importance (150 word max)

\newpage

## Introduction (~250 words)  
- Gut microbiome community composition has proven useful as a resource for machine learning prediction of various diseases {examples}. 
- Amplicon sequencing of the 16S rRNA gene is a reliable tool for assessing the taxonomic composition of microbial communities. 
- Analysis of 16S rRNA sequence data generally relies on clustering of sequences based on similarity into operational taxonomic units (OTUs).
- However OTU clustering depends on the data in the dataset and the addition of new data may change the overall OTU clusters. 
- The unstable nature of OTU clustering complicates machine learning. When building a classification model, changes to the OTU clusters means you have to re-create the model. This can change the underlying model and be time consuming and resource intensive. 
- The ability to integrate new data into an existing model could allow for deployment of a single model that new data can be continually added to and predicted on.
- Recently Sovacool *et al* described a new method for fitting new data into existing OTU clusters {Kelly optifit 2022}. 
- While OptiFit works well to fit new sequence data and provide high quality OTU clusters, it is unknown if the use of OptiFit will have an impact on machine learning predictions.
- Here, we use OptiFit with a 16S rRNA sequence dataset consisting of normal and SRN samples to test how well new data integrated with OptiFit performs for prediction of SRN.

## Results (~700 words)
- Utilize public dataset with normal and SRN samples to compare prediction between OptiFit and OptiClust
- randomly split data into 80% training 20% test sets 100 times
- Processed the data with both algorithms - Figure 1
  - 1. Used traditional OptiClust method to cluster all data, then split into training and test set
  - 2. Used OptiClust on the 80% training set, then used OptiFit to fit the remaining 20%
- OptiFit produces similar quality OTU clusters based on MCC (supplement?)
- Used mikropml to train a model on the 80% training set for each data split, then predicted dx on the 20% test set
- Training performance almost identical (OptiClust median CV AUC 0.694, OptiFit median CV AUC 0.693) - Figure 2
- Performance on the test set comparable (OptiClust median CV AUC 0.694, OptiFit median CV AUC 0.693) - Figure 2
- Maybe expand on where optifit did better/worse?

## Discussion/Conclusions (~250 words)
- OptiFit works!
- future questions:
  - how much reference do you need?
  - does it work well for other situations?
    - what if we used a new dataset for test set instead of a subset of the full dataset?
    - other diseases?

## Materials and Methods
- 16S rRNA amplicon sequence data from 490 subjects {baxter}
  - 261 controls
  - 229 SRN
- created 100 random splits of the data (80% training, 20% test)
- preprocessed data with mothur v1.45
- two pathways  
  -opticlust:  
    - clustered all data together
    - split the shared file based on the random splits

  -optifit:  
    - split the data
    - opticlust on the 80% training set
    - optifit to fit the remaining 20% to the training set OTUs
- ML with mikropml package (version)
  - preprocessed training set and applied preprocessing to test set
    - correlated collapsed, removed nzv

## Acknowledgements
- funding

\newpage

## Figures

**Figure 1. Workflow** description.

**Figure 2. Model Performance.** **A)** Mean AUC **B)** Averaged ROC curves

\newpage

## References

### 2022-02-23 exploratory/testing.Rmd
latest exploratory

---
title: "Testing OptiFit"
author: "Courtney R Armour"
date: "1/31/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,message=FALSE)

library(tidyverse)

```
### MCC OptiFit vs OptiClust

We've been running OptiFit using 80% of samples as the reference. I wanted to see if there would be any change if a higher % of samples was used in the reference. Here I am comparing the MCC scores from 80% (8020) of samples as reference to 90% (9010) of samples as reference. Additionally, when chatting with Kelly about OptiFit she mentioned that she did not do any subsampling in her OptiFit analysis. I ran an addition set where I removed all the subsampling steps and the MCC is plotted below.

```{r mcc}
presubsamp_merged_mcc <- read_csv("data/presubsamp_merged_MCC.csv") %>% 
  mutate(label = case_when(algorithm == "opticlust" ~ algorithm,
                           TRUE ~ paste0(algorithm,"_",subset))) %>% 
  mutate(Group=case_when(algorithm == "opticlust" ~ "OptiClust",
                         subset == "train" ~ "OptiFit Reference",
                         state == "fit" ~ "Optifit Fit w/o Reference",
                         TRUE ~ "Optifit Fit w/ Reference")) %>% 
  mutate(label = factor(label,levels=c("opticlust","optifit_train","optifit_test"),
                        labels = c("OptiClust","OptiFit Train","OptiFit Test"))) %>% 
  mutate(data_split = "8020 - presubsamp")
```

```{r mcc}
postsubsamp_merged_mcc <- read_csv("data/postsubsamp_merged_MCC.csv") %>% 
  mutate(label = case_when(algorithm == "opticlust" ~ algorithm,
                           TRUE ~ paste0(algorithm,"_",subset))) %>% 
  mutate(Group=case_when(algorithm == "opticlust" ~ "OptiClust",
                         subset == "train" ~ "OptiFit Reference",
                         state == "fit" ~ "Optifit Fit w/o Reference",
                         TRUE ~ "Optifit Fit w/ Reference")) %>% 
  mutate(label = factor(label,levels=c("opticlust","optifit_train","optifit_test"),
                        labels = c("OptiClust","OptiFit Train","OptiFit Test"))) %>% 
  mutate(data_split = "8020 - postsubsamp")
```

```{r mcc_9010}
merged_mcc_9010 <- read_csv("data/9010split_merged_MCC.csv") %>% 
  mutate(label = case_when(algorithm == "opticlust" ~ algorithm,
                           TRUE ~ paste0(algorithm,"_",subset))) %>% 
  mutate(Group=case_when(algorithm == "opticlust" ~ "OptiClust",
                         subset == "train" ~ "OptiFit Reference",
                         state == "fit" ~ "Optifit Fit w/o Reference",
                         TRUE ~ "Optifit Fit w/ Reference")) %>% 
  mutate(label = factor(label,levels=c("opticlust","optifit_train","optifit_test"),
                        labels = c("OptiClust","OptiFit Train","OptiFit Test"))) %>% 
  mutate(data_split = "9010 - presubsamp")
```

```{r nosubsamp}
merged_mcc_nosubsamp <- read_csv("data/nosubsamp_merged_MCC.csv") %>% 
  mutate(label = case_when(algorithm == "opticlust" ~ algorithm,
                           TRUE ~ paste0(algorithm,"_",subset))) %>% 
  mutate(Group=case_when(algorithm == "opticlust" ~ "OptiClust",
                         subset == "train" ~ "OptiFit Reference",
                         state == "fit" ~ "Optifit Fit w/o Reference",
                         TRUE ~ "Optifit Fit w/ Reference")) %>% 
  mutate(label = factor(label,levels=c("opticlust","optifit_train","optifit_test"),
                        labels = c("OptiClust","OptiFit Train","OptiFit Test"))) %>% 
  mutate(data_split = "No Subsample")

```

```{r plot}

bind_rows(presubsamp_merged_mcc,merged_mcc_9010) %>% 
  bind_rows(merged_mcc_nosubsamp) %>% 
  bind_rows(postsubsamp_merged_mcc) %>% 
  ggplot(aes(x=label,y=mcc,color=Group)) +
    geom_jitter(height=0,width=0.2,alpha=0.5,size=3) +
    theme_bw() +
    xlab("") + ylab("MCC") +
    theme(panel.grid.major.y = element_line(color="grey85"),
          panel.grid.minor.y = element_line(color="grey85"),
          panel.grid.major.x = element_blank(),
          legend.position = "top",
          axis.text.x = element_text(angle=45,hjust=1)) +
    #geom_text(aes(label,mcc,label=split),data=subset(merged_mcc, mcc < 0.7),color="black",hjust=0) +
    facet_grid(.~data_split)

```

### 2022-04-01 submission/manuscript.Rmd
manscript draft and figs

@@
 
-```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
+```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE, include=FALSE}
+library(tidyverse)
 library(knitr)
@@
 
-inline_hook <- function(x){
-	print(x)
-
-	if(is.list(x)){
-		x <- unlist(x)
-	}
-
-	if(is.numeric(x)){
-		if(abs(x - round(x)) < .Machine$double.eps^0.5){
-			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
-		} else {
-			paste(format(x,big.mark=',', digits=1, nsmall=1, scientific=FALSE))
-		}
-	} else {
-    	paste(x)      
-	}
-}
-knitr::knit_hooks$set(inline=inline_hook)
 ```
 
-\vspace{20mm}
+**Title Options:**  
+Applying OptiFit: An Improved Method for Fitting Amplicon Sequences to Existing OTUs to Predict Colorectal Cancer  
+Predicting colorectal cancer with an improved method for fitting amplicon sequences to existing OTUs  
+Efficient microbiome-based machine learning prediction of colorectal cancer using OptiFit  
+OptiFit streamlines microbiome-based machine learning prediction of colorectal cancer  
+OptiFit enables integration of new amplicon sequence data to existing OTUs for prediction of colorectal cancer
+
+\vspace{10mm}
 
@@
 
-\vspace{20mm}
+\vspace{10mm}
 
@@
 
-${^\#}$ Current Affiliation: Bristol Myers Squibb, Summit, New Jersey, USA
-
 ${^*}$ Current Affiliation:
 
-\vspace{20mm}
+${^\#}$ Current Affiliation: Bristol Myers Squibb, Summit, New Jersey, USA
 
@@
 
-## Introduction (~250 words)  
-- Gut microbiome community composition has proven useful as a resource for machine learning prediction of various diseases {examples}. 
-- Amplicon sequencing of the 16S rRNA gene is a reliable tool for assessing the taxonomic composition of microbial communities. 
-- Analysis of 16S rRNA sequence data generally relies on clustering of sequences based on similarity into operational taxonomic units (OTUs).
-- However OTU clustering depends on the data in the dataset and the addition of new data may change the overall OTU clusters. 
-- The unstable nature of OTU clustering complicates machine learning.
-When building a classification model, changes to the OTU clusters means you have to re-create the model.
-This can change the underlying model and be time consuming and resource intensive.
-- The ability to integrate new data into an existing model could allow for deployment of a single model that new data can be continually added to and predicted on.
-- Recently Sovacool *et al* described a new method for fitting new data into existing OTU clusters {Kelly optifit 2022}. 
-- While OptiFit works well to fit new sequence data and provide high quality OTU clusters, it is unknown if the use of OptiFit will have an impact on machine learning predictions.
-- Here, we use OptiFit with a 16S rRNA sequence dataset consisting of normal and SRN samples to test how well new data integrated with OptiFit performs for prediction of SRN.
+Gut microbiome community composition is useful as a resource for machine learning prediction of various diseases {}.
+Amplicon sequencing of the 16S rRNA gene is a reliable tool for assessing the taxonomic composition of microbial communities.
+Analysis of 16S rRNA sequence data generally relies on clustering of sequences based on similarity into operational taxonomic units (OTUs).
+However, OTU clustering depends on the data in the dataset and the addition of new data could change the overall OTU clusters.
+The unstable nature of OTU clustering complicates deployment of machine learning models since integration of additional data requires re-clustering all the data and re-training of the model.
+The ability to integrate new data into an existing model without re-clustering and re-training could allow for deployment of a single model that new data can be continually added to and predicted on.
+Recently Sovacool *et al* described a new method for fitting new data into existing OTU clusters {Kelly optifit 2022}.
+While OptiFit works well to fit new sequence data and provide high quality OTU clusters, it is unknown if the use of OptiFit will have an impact on machine learning predictions.
+Here, we use OptiFit with a 16S rRNA sequence dataset consisting of normal and SRN samples to test how well new data integrated with OptiFit performs for prediction of SRN.
 
-## Results (~700 words)
-- Utilize public dataset with normal and SRN samples to compare prediction between OptiFit and OptiClust
-- randomly split data into 80% training 20% test sets 100 times
-- Processed the data with both algorithms - Figure 1
-  - 1. Used traditional OptiClust method to cluster all data, then split into training and test set
-  - 2. Used OptiClust on the 80% training set, then used OptiFit to fit the remaining 20%
-- OptiFit produces similar quality OTU clusters based on MCC (supplement?)
-- Used mikropml to train a model on the 80% training set for each data split, then predicted dx on the 20% test set
-- Training performance almost identical (OptiClust median CV AUC 0.694, OptiFit median CV AUC 0.693) - Figure 2
-- Performance on the test set comparable (OptiClust median CV AUC 0.694, OptiFit median CV AUC 0.693) - Figure 2
-- Maybe expand on where optifit did better/worse?
+To assess machine learning prediction performance using OptiFit, we downloaded a publicly available dataset of 16S rRNA amplicon sequences from stool samples.
+The data set includes samples from healthy subjects as well as subjects with screen-relevant neoplasia (SRN) consisting of advanced adenoma and carcinoma.
+To simulate a scenario where we build a model with a set of data and then add additional data, we randomly split the dataset where 80% of the data was clustered into OTUs and the remaining 20% was used integrated into the existing OTUs with the OptiFit algorithm (Figure 1A).
+To account for variation depending on the split of the data, the data was randomly split 100 times and the process repeated for each of the 100 data splits.
+For comparison, the standard process using OptiClust was conducted where all of the data was clustered into OTUs using the OptiClust algorithm and the resulting OTU abundance table was split into the testing and training set (Figure 1B).
 
-## Discussion/Conclusions (~250 words)
-- OptiFit works!
-- future questions:
-  - how much reference do you need?
-  - does it work well for other situations?
-    - what if we used a new dataset for test set instead of a subset of the full dataset?
-    - other diseases?
+```{r mcc,include=F}
+merged_mcc <- read_csv("../data/learning/summary/merged_MCC.csv")
+
+opticlust_mcc <- merged_mcc %>% 
+  filter(algorithm == "opticlust") %>% 
+  pull(mcc)
+
+optifit_avg_mcc <- merged_mcc %>% 
+  filter(algorithm == "optifit" & state=="combo") %>% 
+  summarise(mean_mcc = mean(mcc))
+```
+One way to examine the quality of the OTU clusters is with the Matthews correlation coefficient (MCC) which looks at the similarity between all pairs of sequences and assess whether they are appropriately in the same OTU or not {Westcott 2017 mSphere OptiClust}.
+An MCC score of zero indicates low quality OTU clusters while a score of 1 indicates high quality OTU clusters.
+Since the data is only clustered once in the OptiClust pathway there is only one MCC score while the OptiFit method produces an MCC score for the OTU clusters from each data split.
+Overall the MCC scores were similar between OptiClust (MCC = `r round(opticlust_mcc,digits=3)`) and OptiFit (average MCC = `r round(optifit_avg_mcc,digits=3)`) indicating that OptiFit does a good job of appropriately integrating new sequences into the existing OTUs.
+
+A potential problem to using OptiFit for machine learning prediction is that any sequences in the new data that do not map to the existing OTU clusters will be discarded resulting in a possible loss of information.
+To quantify whether this impacts model performance we used the taxonomic abundances of the training data from the OptiClust and OptiFit pathways to train a model to predict SRNs based on OTU abundances.
+This model was then deployed to predict the diagnosis classification of the held out test data.
+To compare model performance we calculated the area under the receiver operating characteristic curve (AUROC) for each data split on both the training data and the testing data.
+The model performance during training was equivalent between the two algorithms (OptiClust median CV AUC 0.694, OptiFit median CV AUC 0.693, Figure 2A).
+Additionally, the performance on the test data was equivalent between the two algorithms (OptiClust median CV AUC 0.694, OptiFit median CV AUC 0.693, Figure 2A,B) indicating that new data can be integrated into existing OTU clusters without impacting model performance.
+
+Overall this analysis demonstrated that OptiFit can be used to fit new sequence data into existing OTU clusters and perform equally in predicting SRNs compared to clustering all of the sequence data together.
+The ability to integrate new data into existing OTUs enables the deployment of a single machine learning model based on microbiome composition that new data can be predicted on.
+These results are based on a single dataset and disease.
+Further analysis is needed to determine if the process works well for data collected or processed with different methodology or prediction of other conditions.
 
 ## Materials and Methods
-- 16S rRNA amplicon sequence data from 490 subjects {baxter}
-  - 261 controls
-  - 229 SRN
-- created 100 random splits of the data (80% training, 20% test)
-- preprocessed data with mothur v1.45
-- two pathways  
-  -opticlust:  
-    - clustered all data together
-    - split the shared file based on the random splits
+***Data Set.*** Raw 16S rRNA amplicon sequence data isolated from human stool samples was downloaded from NCBI Sequence Read Archive (accession no.
+SRP062005) {baxter}.
+This data set contains stool samples from a total of N subjects, however after preprocessing to screen for sequence quality and subsample to 10,000 reads per sample, 490 samples remained.
+For this analysis, samples from subjects identified in the metadata as normal, high risk normal, or adenoma were categorized as "normal" while samples from subjects identified as advanced adenoma or carcinoma were categorized as "screen relevant neoplasia" (SRN).
+The resulting data set consisted of 261 normal samples and 229 SRN samples.
 
-  -optifit:  
-    - split the data
-    - opticlust on the 80% training set
-    - optifit to fit the remaining 20% to the training set OTUs
-- ML with mikropml package (version)
-  - preprocessed training set and applied preprocessing to test set
-    - correlated collapsed, removed nzv
+***Data Processing.*** The full dataset was pre-processed with mothur (v1.45) using the SILVA reference database (v132) {silva} to join forward and reverse reads, merge duplicate reads, align to the reference, pre-cluster, remove chimeras, assign taxonomy, and remove non-bacterial reads following the Schloss Lab MiSeq standard operating procedure described on the mothur website (https://mothur.org/wiki/miseq_sop/).
+100 splits of the 490 samples were generated where 80% of the samples (392 samples) were randomly assigned to the training set and the remaining 20% (98 samples) were assigned to the test set.
+Using 100 splits of the data accounts for the variation that may be observed depending on the samples that are in the training or test sets.
+Each samples was in the training set an average of N times and the test set and average of N times.
+
+The data was processed through two pathways.
+First, the standard pathway using the OptiClust algorithm.
+In this pathway, all of the data was clustered together to generate OTUs and the resulting abundance tables were split into the training and testing sets.
+In the second pathway, the pre-processed data was split into the training and testing sets.
+The training set was clustered into OTUs, then the test set was fit to the OTUs of the training set using the OptiFit algorithm.
+The OptiFit algorithm was run with method open so that any sequences that didn't map to the existing OTU clusters would form new OTUs.
+Any OTUs that were not in the training set were removed prior to machine learning.
+For both pathways all data was sub-sampled to 10,000 reads per sample.
+
+***Machine Learning.*** Machine learning using Random Forest was conducted with the R package mikrompl (v XXXX) {} to predict the diagnosis (SRN or normal) for the samples in the test set for each data split.
+The training set was preprocessed to normalize values (scale/center), collapse colinear features, and remove features with zero-variance.
+The preprocessing from the training set was then applied to the test set.
+P values comparing model performance were calculated as previously described {}.
+The averaged ROC curves were plotted by taking the average and standard deviation of the sensitivity at each specificity value.
+
+***Code Availability.*** all code used in this analysis is available at https://github.com/SchlossLab/Armour_OptiFitGLNE_XXXX_2021
 
 ## Acknowledgements
-- funding
+(funding)
 
@@
 
+![](../exploratory/figures/figure1B.png)
+![](../exploratory/figures/figure1Apng.png)
+
 **Figure 1. Workflow** description.
+
+![](../exploratory/figures/Figure2.png)
 

### 2022-04-01 submission/manuscript.tex
manscript draft and figs

@@
 ]{article}
-\title{\textbf{TITLE}}
-\author{}
-\date{\vspace{-2.5em}}
-
 \usepackage{amsmath,amssymb}
@@
 
+\title{\textbf{TITLE}}
+\author{}
+\date{\vspace{-2.5em}}
+
 \begin{document}
@@
 
-\vspace{20mm}
+\textbf{Title Options:}\\
+Applying OptiFit: An Improved Method for Fitting Amplicon Sequences to
+Existing OTUs to Predict Colorectal Cancer\\
+Predicting colorectal cancer with an improved method for fitting
+amplicon sequences to existing OTUs\\
+Efficient microbiome-based machine learning prediction of colorectal
+cancer using OptiFit\\
+OptiFit streamlines microbiome-based machine learning prediction of
+colorectal cancer\\
+OptiFit enables integration of new amplicon sequence data to existing
+OTUs for prediction of colorectal cancer
+
+\vspace{10mm}
 
@@
 
+\({^*}\) Current Affiliation:
+
 \({^\#}\) Current Affiliation: Bristol Myers Squibb, Summit, New Jersey,
 USA~
-
-\({^*}\) Current Affiliation:
-
-\vspace{20mm}
 
@@
 
-\hypertarget{introduction-250-words}{%
-\subsection{Introduction (\textasciitilde250
-words)}\label{introduction-250-words}}
+Gut microbiome community composition is useful as a resource for machine
+learning prediction of various diseases \{\}. Amplicon sequencing of the
+16S rRNA gene is a reliable tool for assessing the taxonomic composition
+of microbial communities. Analysis of 16S rRNA sequence data generally
+relies on clustering of sequences based on similarity into operational
+taxonomic units (OTUs). However, OTU clustering depends on the data in
+the dataset and the addition of new data could change the overall OTU
+clusters. The unstable nature of OTU clustering complicates deployment
+of machine learning models since integration of additional data requires
+re-clustering all the data and re-training of the model. The ability to
+integrate new data into an existing model without re-clustering and
+re-training could allow for deployment of a single model that new data
+can be continually added to and predicted on. Recently Sovacool \emph{et
+al} described a new method for fitting new data into existing OTU
+clusters \{Kelly optifit 2022\}. While OptiFit works well to fit new
+sequence data and provide high quality OTU clusters, it is unknown if
+the use of OptiFit will have an impact on machine learning predictions.
+Here, we use OptiFit with a 16S rRNA sequence dataset consisting of
+normal and SRN samples to test how well new data integrated with OptiFit
+performs for prediction of SRN.
 
-\begin{itemize}
-\tightlist
-\item
-  Gut microbiome community composition has proven useful as a resource
-  for machine learning prediction of various diseases \{examples\}.
-\item
-  Amplicon sequencing of the 16S rRNA gene is a reliable tool for
-  assessing the taxonomic composition of microbial communities.
-\item
-  Analysis of 16S rRNA sequence data generally relies on clustering of
-  sequences based on similarity into operational taxonomic units (OTUs).
-\item
-  However OTU clustering depends on the data in the dataset and the
-  addition of new data may change the overall OTU clusters.
-\item
-  The unstable nature of OTU clustering complicates machine learning.
-  When building a classification model, changes to the OTU clusters
-  means you have to re-create the model. This can change the underlying
-  model and be time consuming and resource intensive.
-\item
-  The ability to integrate new data into an existing model could allow
-  for deployment of a single model that new data can be continually
-  added to and predicted on.
-\item
-  Recently Sovacool \emph{et al} described a new method for fitting new
-  data into existing OTU clusters \{Kelly optifit 2022\}.
-\item
-  While OptiFit works well to fit new sequence data and provide high
-  quality OTU clusters, it is unknown if the use of OptiFit will have an
-  impact on machine learning predictions.
-\item
-  Here, we use OptiFit with a 16S rRNA sequence dataset consisting of
-  normal and SRN samples to test how well new data integrated with
-  OptiFit performs for prediction of SRN.
-\end{itemize}
+To assess machine learning prediction performance using OptiFit, we
+downloaded a publicly available dataset of 16S rRNA amplicon sequences
+from stool samples. The data set includes samples from healthy subjects
+as well as subjects with screen-relevant neoplasia (SRN) consisting of
+advanced adenoma and carcinoma. To simulate a scenario where we build a
+model with a set of data and then add additional data, we randomly split
+the dataset where 80\% of the data was clustered into OTUs and the
+remaining 20\% was used integrated into the existing OTUs with the
+OptiFit algorithm (Figure 1A). To account for variation depending on the
+split of the data, the data was randomly split 100 times and the process
+repeated for each of the 100 data splits. For comparison, the standard
+process using OptiClust was conducted where all of the data was
+clustered into OTUs using the OptiClust algorithm and the resulting OTU
+abundance table was split into the testing and training set (Figure 1B).
 
-\hypertarget{results-700-words}{%
-\subsection{Results (\textasciitilde700
-words)}\label{results-700-words}}
+One way to examine the quality of the OTU clusters is with the Matthews
+correlation coefficient (MCC) which looks at the similarity between all
+pairs of sequences and assess whether they are appropriately in the same
+OTU or not \{Westcott 2017 mSphere OptiClust\}. An MCC score of zero
+indicates low quality OTU clusters while a score of 1 indicates high
+quality OTU clusters. Since the data is only clustered once in the
+OptiClust pathway there is only one MCC score while the OptiFit method
+produces an MCC score for the OTU clusters from each data split. Overall
+the MCC scores were similar between OptiClust (MCC = 0.895) and OptiFit
+(average MCC = 0.892) indicating that OptiFit does a good job of
+appropriately integrating new sequences into the existing OTUs.
 
-\begin{itemize}
-\tightlist
-\item
-  Utilize public dataset with normal and SRN samples to compare
-  prediction between OptiFit and OptiClust
-\item
-  randomly split data into 80\% training 20\% test sets 100 times
-\item
-  Processed the data with both algorithms - Figure 1
+A potential problem to using OptiFit for machine learning prediction is
+that any sequences in the new data that do not map to the existing OTU
+clusters will be discarded resulting in a possible loss of information.
+To quantify whether this impacts model performance we used the taxonomic
+abundances of the training data from the OptiClust and OptiFit pathways
+to train a model to predict SRNs based on OTU abundances. This model was
+then deployed to predict the diagnosis classification of the held out
+test data. To compare model performance we calculated the area under the
+receiver operating characteristic curve (AUROC) for each data split on
+both the training data and the testing data. The model performance
+during training was equivalent between the two algorithms (OptiClust
+median CV AUC 0.694, OptiFit median CV AUC 0.693, Figure 2A).
+Additionally, the performance on the test data was equivalent between
+the two algorithms (OptiClust median CV AUC 0.694, OptiFit median CV AUC
+0.693, Figure 2A,B) indicating that new data can be integrated into
+existing OTU clusters without impacting model performance.
 
-  \begin{itemize}
-  \item
-    \begin{enumerate}
-    \def\labelenumi{\arabic{enumi}.}
-    \tightlist
-    \item
-      Used traditional OptiClust method to cluster all data, then split
-      into training and test set
-    \end{enumerate}
-  \item
-    \begin{enumerate}
-    \def\labelenumi{\arabic{enumi}.}
-    \setcounter{enumi}{1}
-    \tightlist
-    \item
-      Used OptiClust on the 80\% training set, then used OptiFit to fit
-      the remaining 20\%
-    \end{enumerate}
-  \end{itemize}
-\item
-  OptiFit produces similar quality OTU clusters based on MCC
-  (supplement?)
-\item
-  Used mikropml to train a model on the 80\% training set for each data
-  split, then predicted dx on the 20\% test set
-\item
-  Training performance almost identical (OptiClust median CV AUC 0.694,
-  OptiFit median CV AUC 0.693) - Figure 2
-\item
-  Performance on the test set comparable (OptiClust median CV AUC 0.694,
-  OptiFit median CV AUC 0.693) - Figure 2
-\item
-  Maybe expand on where optifit did better/worse?
-\end{itemize}
-
-\hypertarget{discussionconclusions-250-words}{%
-\subsection{Discussion/Conclusions (\textasciitilde250
-words)}\label{discussionconclusions-250-words}}
-
-\begin{itemize}
-\tightlist
-\item
-  OptiFit works!
-\item
-  future questions:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    how much reference do you need?
-  \item
-    does it work well for other situations?
-
-    \begin{itemize}
-    \tightlist
-    \item
-      what if we used a new dataset for test set instead of a subset of
-      the full dataset?
-    \item
-      other diseases?
-    \end{itemize}
-  \end{itemize}
-\end{itemize}
+Overall this analysis demonstrated that OptiFit can be used to fit new
+sequence data into existing OTU clusters and perform equally in
+predicting SRNs compared to clustering all of the sequence data
+together. The ability to integrate new data into existing OTUs enables
+the deployment of a single machine learning model based on microbiome
+composition that new data can be predicted on. These results are based
+on a single dataset and disease. Further analysis is needed to determine
+if the process works well for data collected or processed with different
+methodology or prediction of other conditions.
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  16S rRNA amplicon sequence data from 490 subjects \{baxter\}
+\textbf{\emph{Data Set.}} Raw 16S rRNA amplicon sequence data isolated
+from human stool samples was downloaded from NCBI Sequence Read Archive
+(accession no. SRP062005) \{baxter\}. This data set contains stool
+samples from a total of N subjects, however after preprocessing to
+screen for sequence quality and subsample to 10,000 reads per sample,
+490 samples remained. For this analysis, samples from subjects
+identified in the metadata as normal, high risk normal, or adenoma were
+categorized as ``normal'' while samples from subjects identified as
+advanced adenoma or carcinoma were categorized as ``screen relevant
+neoplasia'' (SRN). The resulting data set consisted of 261 normal
+samples and 229 SRN samples.
 
-  \begin{itemize}
-  \tightlist
-  \item
-    261 controls
-  \item
-    229 SRN
-  \end{itemize}
-\item
-  created 100 random splits of the data (80\% training, 20\% test)
-\item
-  preprocessed data with mothur v1.45
-\item
-  two pathways\\
-  -opticlust:
+\textbf{\emph{Data Processing.}} The full dataset was pre-processed with
+mothur (v1.45) using the SILVA reference database (v132) \{silva\} to
+join forward and reverse reads, merge duplicate reads, align to the
+reference, pre-cluster, remove chimeras, assign taxonomy, and remove
+non-bacterial reads following the Schloss Lab MiSeq standard operating
+procedure described on the mothur website
+(\url{https://mothur.org/wiki/miseq_sop/}). 100 splits of the 490
+samples were generated where 80\% of the samples (392 samples) were
+randomly assigned to the training set and the remaining 20\% (98
+samples) were assigned to the test set. Using 100 splits of the data
+accounts for the variation that may be observed depending on the samples
+that are in the training or test sets. Each samples was in the training
+set an average of N times and the test set and average of N times.
 
-  \begin{itemize}
-  \tightlist
-  \item
-    clustered all data together
-  \item
-    split the shared file based on the random splits
-  \end{itemize}
+The data was processed through two pathways. First, the standard pathway
+using the OptiClust algorithm. In this pathway, all of the data was
+clustered together to generate OTUs and the resulting abundance tables
+were split into the training and testing sets. In the second pathway,
+the pre-processed data was split into the training and testing sets. The
+training set was clustered into OTUs, then the test set was fit to the
+OTUs of the training set using the OptiFit algorithm. The OptiFit
+algorithm was run with method open so that any sequences that didn't map
+to the existing OTU clusters would form new OTUs. Any OTUs that were not
+in the training set were removed prior to machine learning. For both
+pathways all data was sub-sampled to 10,000 reads per sample.
 
-  -optifit:
+\textbf{\emph{Machine Learning.}} Machine learning using Random Forest
+was conducted with the R package mikrompl (v XXXX) \{\} to predict the
+diagnosis (SRN or normal) for the samples in the test set for each data
+split. The training set was preprocessed to normalize values
+(scale/center), collapse colinear features, and remove features with
+zero-variance. The preprocessing from the training set was then applied
+to the test set. P values comparing model performance were calculated as
+previously described \{\}. The averaged ROC curves were plotted by
+taking the average and standard deviation of the sensitivity at each
+specificity value.
 
-  \begin{itemize}
-  \tightlist
-  \item
-    split the data
-  \item
-    opticlust on the 80\% training set
-  \item
-    optifit to fit the remaining 20\% to the training set OTUs
-  \end{itemize}
-\item
-  ML with mikropml package (version)
-
-  \begin{itemize}
-  \tightlist
-  \item
-    preprocessed training set and applied preprocessing to test set
-
-    \begin{itemize}
-    \tightlist
-    \item
-      correlated collapsed, removed nzv
-    \end{itemize}
-  \end{itemize}
-\end{itemize}
+\textbf{\emph{Code Availability.}} all code used in this analysis is
+available at
+\url{https://github.com/SchlossLab/Armour_OptiFitGLNE_XXXX_2021}
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  funding
-\end{itemize}
+(funding)
 
@@
 
+\includegraphics{../exploratory/figures/figure1B.png}
+\includegraphics{../exploratory/figures/figure1Apng.png}
+
 \textbf{Figure 1. Workflow} description.
+
+\includegraphics{../exploratory/figures/Figure2.png}
 

### 2022-04-01 code/bash/mothurReferences.sh
update references

@@ -35,44 +35,29 @@ mkdir -p "${TMP}"/
 echo PROGRESS: Preparing SILVA database v4 sequence alignment files.
 
 # Downloading the prepared SILVA database from the mothur website
-# This version is from v123 and described at http://blog.mothur.org/2015/12/03/SILVA-v123-reference-files/
-# This version is used to maintain methods from Baxter NT, et al. Microbiota-based model improves the sensitivity
-# of fecal immunochemical test for detecting colonic lesions. Genome Med. 2016;8(1):37.
-# wget -N -P "${TMP}"/ http://mothur.org/w/images/1/15/Silva.seed_v123.tgz
-# updated location of silva reference v123
-wget -N -P "${TMP}"/ https://mothur.s3.us-east-2.amazonaws.com/wiki/silva.seed_v123.tgz
+wget -N -P "${TMP}"/ https://mothur.s3.us-east-2.amazonaws.com/wiki/silva.seed_v132.tgz
 
 # Decompressing the database
-# tar xvzf "${TMP}"/Silva.seed_v123.tgz -C "${TMP}"/
-# change capitalization
-tar xvzf "${TMP}"/silva.seed_v123.tgz -C "${TMP}"/
+tar xvzf "${TMP}"/silva.seed_v132.tgz -C "${TMP}"/
 
 # Using mothur to pull out the v4 region from bacterial sequences
 mothur "#set.current(outputdir="${TMP}"/, processors="${NPROC}");
-	get.lineage(fasta="${TMP}"/silva.seed_v123.align, taxonomy="${TMP}"/silva.seed_v123.tax, taxon=Bacteria);
+	get.lineage(fasta="${TMP}"/silva.seed_v132.align, taxonomy="${TMP}"/silva.seed_v132.tax, taxon=Bacteria);
 	pcr.seqs(fasta=current, start=11894, end=25319, keepdots=F)"
 
 # Renaming the output file and moving it from the tmp dir to the output dir
-mv "${TMP}"/silva.seed_v123.pick.align "${OUTDIR}"/silva.seed.align
-mv "${TMP}"/silva.seed_v123.pick.pcr.align "${OUTDIR}"/silva.v4.align
-
-
+mv "${TMP}"/silva.seed_v132.pick.pcr.align "${OUTDIR}"/silva.v4.align
 
 echo PROGRESS: Preparing Ribosomal Database Project taxonomy files.
 
[…]

### 2022-06-15 submission/manuscript.Rmd
full manuscript draft

--- 2022-02-23 submission/manuscript.Rmd
@@
 ---
-title: "**TITLE**"
-bibliography: references.bib
+title: "**Reference-based OTU clustering for machine learning classification**"
 output:
-  pdf_document:
-    keep_tex: true
-    includes:
-      in_header: header.tex
-csl: mbio.csl #Get themes at https://github.com/citation-style-language/styles
-fontsize: 11pt
+  rmarkdown::pdf_document:
+    keep_tex: yes
 geometry: margin=1.0in
+font-size: 11pt
 header-includes:
   - \usepackage{helvet}
+  - \renewcommand*\familydefault{\sfdefault}
+  - \usepackage{setspace}
+  - \doublespacing
+  - \usepackage[left]{lineno}
+  - \usepackage{multirow}
+  - \usepackage[none]{hyphenat}
+editor_options: 
+  markdown: 
+    wrap: 72
+bibliography: references.bib
+csl: mbio.csl
+link-citations: true
 ---
 
-```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
+```{r settings, eval=TRUE, echo=FALSE, cache=FALSE, include=FALSE}
+library(tidyverse)
 library(knitr)
@@
 
-inline_hook <- function(x){
-	print(x)
-
-	if(is.list(x)){
-		x <- unlist(x)
+format_decimal <- function(number,digits){
+  return(round(number,digits=digits))
 }
 
-	if(is.numeric(x)){
-		if(abs(x - round(x)) < .Machine$double.eps^0.5){
-			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
-		} else {
-			paste(format(x,big.mark=',', digits=1, nsmall=1, scientific=FALSE))
-		}
-	} else {
-    	paste(x)      
-	}
-}
-knitr::knit_hooks$set(inline=inline_hook)
 ```
 
-\vspace{20mm}
+Other Title Options:\
+A solution to the problem of inconsistent OTU clustering for machine
+learning classification
+
+Optimized reference-based OTU clustering for machine learning
+classification
+
+\vspace{5mm}
 
@@
 
-\vspace{20mm}
+\vspace{5mm}
 
-Courtney R. Armour${^1}$, William L. Close$^{1,*}$, Begüm D. Topçuoğlu$^{1,\#}$, Patrick D. Schloss $^{1,\dagger}$
+Courtney R. Armour${^1}$, Kelly L. Sovacool${^2}$, William L.
+Close$^{1,*}$, Begüm D. Topçuoğlu$^{1,\#}$, Jenna Wiens${^3}$, Patrick
+D. Schloss $^{1,\dagger}$
 
@@
 
-${^1}$ Department of Microbiology and Immunology, University of Michigan, Ann Arbor MI.
+${^1}$ Department of Microbiology and Immunology, University of
+Michigan, Ann Arbor, Michigan, USA
 
-${^\#}$ Current Affiliation: Bristol Myers Squibb, Summit, New Jersey, USA
+${^2}$ Department of Computational Medicine and Bioinformatics,
+University of Michigan, Ann Arbor, Michigan, USA
+
+${^3}$ Department of Electrical Engineering and Computer Science,
+University of Michigan, Ann Arbor, Michigan, USA
 
@@
 
-\vspace{20mm}
+${^\#}$ Current Affiliation: Bristol Myers Squibb, Summit, New Jersey,
+USA
 
-$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu
+$\dagger$ To whom correspondence should be addressed:
+[pschloss\@umich.edu](mailto:pschloss@umich.edu)
 
-\vspace{20mm}
+\vspace{10mm}
 
@@
 
+Machine learning classification of disease based on the gut microbiome
+often relies on operational taxonomic units (OTUs) to quantify microbial
+composition. The standard methodology for quantifying taxonomic
+composition is clustering 16S rRNA amplicon sequences *de novo* into
+OTUs and using their abundances to train a classification model.
+However, OTU clusters depend on the sequences in the data set and
+therefore can change if new data is added. This inconsistency
+complicates machine learning classification because in order to use the
+model to classify additional samples, the OTUs have to be re-clustered
+to include the new sequences and the model retrained with the new OTU
+clusters. A new algorithm in mothur, called OptiFit, addresses this
+issue by clustering new sequences into the existing OTU clusters. While
+OptiFit is proven to produce high quality OTU clusters, it is unclear
+whether this method for clustering new sequence data into existing OTUs
+will have any impact on machine learning classification using the OTU
+abundances. To evaluate the potential for use in machine learning
+classification, we used OptiFit to cluster additional data into existing
+OTU clusters and quantified model performance in classifying a data set
+containing samples from controls and people with advanced adenoma and
+carcinoma (SRN). We compared the model performance using OptiFit to the
+standard procedure of clustering all the data together. We found that
+OptiFit can be used to fit new sequence data to existing OTUs and
+perform equally in classifying SRNs. Moving forward, when OTUs are used
+in machine learning classification problems, OptiFit can be used to
+avoid the need to re-cluster sequences into OTUs when classifying
+additional samples that were not part of the data set used to build the
+model.
+
 ## Importance (150 word max)
+
+There are instances when OTU abundance data is optimal for machine
+learning prediction of disease based on microbiome composition. The
+current methodology for OTU clustering complicates machine learning
+since *de novo* OTU clusters can change depending on the data in the
+data set. Attempting to classify additional samples that were not part
+of the original model build requires re-clustering OTUs and re-training
+the model. OptiFit provides an elegant solution to the problem of OTU
+inconsistency by enabling new data to be fit to the existing OTUs while
+maintaining machine learning classification performance. The use of
+OptiFit will enable a single model based on OTU abundance data to be
+used in classifying additional samples that were not part of the
+original OTU clustering.
 
@@
 
-## Introduction (~250 words)  
-- Gut microbiome community composition has proven useful as a resource for machine learning prediction of various diseases {examples}. 
-- Amplicon sequencing of the 16S rRNA gene is a reliable tool for assessing the taxonomic composition of microbial communities. 
-- Analysis of 16S rRNA sequence data generally relies on clustering of sequences based on similarity into operational taxonomic units (OTUs).
-- However OTU clustering depends on the data in the dataset and the addition of new data may change the overall OTU clusters. 
-- The unstable nature of OTU clustering complicates machine learning.
-When building a classification model, changes to the OTU clusters means you have to re-create the model.
-This can change the underlying model and be time consuming and resource intensive.
-- The ability to integrate new data into an existing model could allow for deployment of a single model that new data can be continually added to and predicted on.
-- Recently Sovacool *et al* described a new method for fitting new data into existing OTU clusters {Kelly optifit 2022}. 
-- While OptiFit works well to fit new sequence data and provide high quality OTU clusters, it is unknown if the use of OptiFit will have an impact on machine learning predictions.
-- Here, we use OptiFit with a 16S rRNA sequence dataset consisting of normal and SRN samples to test how well new data integrated with OptiFit performs for prediction of SRN.
+Gut microbiome community composition is useful as a resource for machine
+learning prediction of diseases, including colorectal cancer
+[@baxter2016; @zackular2014]. Amplicon sequencing of the 16S rRNA gene
+is a reliable tool for assessing the taxonomic composition of microbial
+communities. Analysis of 16S rRNA amplicon sequence (16S) data generally
+relies on clustering of sequences based on similarity into operational
+taxonomic units (OTUs). The process of OTU clustering can either be
+reference-based or *de novo*. The quality of OTUs generated with
+reference-based clustering is dependent on the quality of the reference
+used, while OTUs generated with *de novo* clustering are optimized to
+the data set @westcott2015. While *de novo* clustering can produce
+high-quality OTU clusters where sequences are accurately grouped based
+on similarity thresholds, the resulting OTU clusters depend on the data
+in the data set and the addition of new data could change the overall
+OTU clusters. The inconsistent nature of OTU clustering complicates
+deployment of machine learning models since integration of additional
+data requires re-clustering all the data and re-training of the model.
+The ability to integrate new data into an existing model without
+re-clustering and re-training could allow for deployment of a single
+model that new data can be continually added to. Recently Sovacool *et
+al* introduced OptiFit: a method for fitting new data into existing OTU
+clusters @sovacool2022. While OptiFit is proven to effectively fit new
+sequence data to existing OTU clusters, it is unknown if the use of
+OptiFit will have an impact on machine learning classification. Here we
+test the ability of OptiFit to cluster new sequence data into existing
+OTU clusters for the purpose of machine learning classification of
+disease based on gut microbiome composition.
 
-## Results (~700 words)
-- Utilize public dataset with normal and SRN samples to compare prediction between OptiFit and OptiClust
-- randomly split data into 80% training 20% test sets 100 times
-- Processed the data with both algorithms - Figure 1
-  - 1. Used traditional OptiClust method to cluster all data, then split into training and test set
-  - 2. Used OptiClust on the 80% training set, then used OptiFit to fit the remaining 20%
-- OptiFit produces similar quality OTU clusters based on MCC (supplement?)
-- Used mikropml to train a model on the 80% training set for each data split, then predicted dx on the 20% test set
-- Training performance almost identical (OptiClust median CV AUC 0.694, OptiFit median CV AUC 0.693) - Figure 2
-- Performance on the test set comparable (OptiClust median CV AUC 0.694, OptiFit median CV AUC 0.693) - Figure 2
-- Maybe expand on where optifit did better/worse?
+When using OTU abundances for machine learning classification, the
+current methodology is to *de novo* cluster all of the sequence data
+into OTUs with the OptiClust algorithm in mothur @westcott2017. The
+resulting abundance data is then split into training and testing sets,
+where the training set is used to tune hyperparameters and ultimately
+train the model. The testing set is then classified with the model and
+the performance of the model can be quantified (Figure 1A). However,
+with this methodology we would have to re-generate the OTU clusters and
+re-train the model if we wanted to classify additional samples. The
+OptiFit algorithm @sovacool2022 addresses this problem by enabling new
+sequences to be clustered into existing OTUs. The OptiFit workflow is
+similar to the OptiClust workflow where the data is clustered into OTUs
+and used to tune hyperparameters and ultimately train the model. Then,
+we can use OptiFit to fit sequence data of samples not part of the
+original data set into the existing OTUs and use the same model to
+classify the samples (Figure 1B). To test how the model performance
+compares between these two methodologies, we used a publicly available
+data set of 16S sequences from stool samples of healthy subjects as well
+as subjects with SRN consisting of advanced adenoma and carcinoma
+@baxter2016. The data set was randomly split into an 80% train set and
+20% test set. For the standard OptiClust workflow, the training and test
+sets were *de novo* clustered together into OTUs then the resulting
+abundance table was split into the training and testing set. For the
+OptiFit workflow, the train set was clustered *de novo* into OTUs and
+the remaining test set was fit to the OTU clusters using the OptiFit
+algorithm. For both workflows, the abundance table of the train set was
+used to tune hyperparameters and train a random forest model to classify
+SRN. The test set was classified as either control or SRN using the
+trained models.To account for variation depending on the split of the
+data, the data set was randomly split 100 times and the process repeated
+for each of the 100 data splits. By comparing the model performance of
+classifying the samples in the test data set between the OptiFit and
+OptiClust algorithms, we can quantify the impact of using OptiFit on
+model classification performance.
 
-## Discussion/Conclusions (~250 words)
-- OptiFit works!
-- future questions:
-  - how much reference do you need?
-  - does it work well for other situations?
-    - what if we used a new dataset for test set instead of a subset of the full dataset?
-    - other diseases?
+```{r mcc,include=F}
+merged_mcc <- read_csv("../data/learning/summary/merged_MCC.csv")
+
+opticlust_mcc <- merged_mcc %>% 
+  filter(algorithm == "opticlust") %>% 
+  pull(mcc)
+
+optifit_avg_mcc <- merged_mcc %>% 
+  filter(algorithm == "optifit" & state=="combo") %>% 
+  summarise(mean_mcc = mean(mcc))
+```
+
+We first examined the quality of the resulting OTU clusters from the two
+algorithms using the Matthews correlation coefficient (MCC). The MCC
+score is quantified by examining all pairs of sequences and assessing
+whether they belong together in an OTU or not based on sequence
+similarity @westcott2017. MCC scores range between negative one and one.
+A score of negative one means none of the sequences in an OTU are within
+the similarity threshold and any sequences within the similarity
+threshold are not in an OTU together. An MCC score of zero essentially
+means the sequences are randomly clustered. An MCC score of 1 means all
+sequences in an OTU are within the similarity threshold and all sequence
+pairs within the similarity threshold are in the same OTU. To ensure
+that OptiFit is appropriately integrating new sequence data into the
+existing OTUs, we would expect that the MCC scores produced by the
+OptiClust and OptiFit workflows are similar. Since the data is only
+clustered once in the OptiClust workflow there is only one MCC score
+while the OptiFit workflow produces an MCC score for the OTU clusters
+from each data split. Overall the MCC scores were similar between
+OptiClust (MCC = `r round(opticlust_mcc,digits=3)`) and OptiFit (average
+MCC = `r round(optifit_avg_mcc,digits=3)`) indicating that OptiFit
+performs as well as OptiClust when integrating new sequences into the
+existing OTUs.
+
+```{r performance,include=F}
+performance <- read_csv("../data/learning/summary/merged_performance.csv",
+                    col_types = cols(algorithm=col_character(),
+                                     method=col_character(),
+                                     split=col_character(),
+                                     .default=col_double()))
+
+performance_summary <- performance %>% 
+  group_by(algorithm) %>% 
+  summarise(median_cv_AUC = round(median(cv_metric_AUC),digits=3),
+            mean_cv_AUC = round(mean(cv_metric_AUC),digits=3),
+            median_AUC = round(median(AUC),digits=3),
+            mean_AUC = round(mean(AUC),digits=3))
+
+pvals <- read_csv("../results/tables/pvalues.csv",col_types = cols(p_value = col_double(),
+                                                                   .default = col_character())) %>% 
+  select(metric,p_value)
+```
+
+After verifying that the quality of the OTUs was consistent between
+OptiClust and OptiFit, we next examined the model performance for
+classifying samples in the test data set as control or SRN. To quantify
+model performance we used the taxonomic abundances of the training data
+from the OptiClust and OptiFit workflows to train a model to predict
+SRNs. Using the predicted and actual diagnosis classification, we
+calculated the area under the receiver operating characteristic curve
+(AUROC) for each data split to quantify model performance. During
+cross-validation (CV) training, the model performance was equivalent
+between the two algorithms (p-value =
+`r format_decimal(pvals %>% filter(metric == "cv_metric_AUC") %>% select(p_value),2)`,
+OptiClust mean CV AUROC =
+`r format_decimal(performance_summary %>% filter(algorithm == "opticlust") %>% select(mean_cv_AUC),3)`,
+OptiFit mean CV AUROC =
+`r format_decimal(performance_summary %>% filter(algorithm == "optifit") %>% select(mean_cv_AUC),3)`,
+Figure 2A). The trained model was then deployed to classify the samples
+of the test data as control or SRN. The performance on the test data was
+equivalent between the two algorithms (p-value =
+`r format_decimal(pvals %>% filter(metric == "AUC") %>% select(p_value),2)`,
+OptiClust mean test AUROC =
+`r format_decimal(performance_summary %>% filter(algorithm == "opticlust") %>% select(mean_AUC),3)`,
+OptiFit mean test AUROC =
+`r format_decimal(performance_summary %>% filter(algorithm == "optifit") %>% select(mean_AUC),3)`,
+Figure 2B,C) indicating that new data can be fit to existing OTU
+clusters without impacting model performance.
+
+We tested the ability of OptiFit to integrate new data into existing
+OTUs for the purpose of machine learning classification of disease based
+on microbiome abundance. A potential problem to using OptiFit for
+machine learning prediction is that any sequences in the new data that
+do not map to the existing OTU clusters will be discarded resulting in a
+possible loss of information. However, we demonstrated that OptiFit can
+be used to fit new sequence data into existing OTU clusters and perform
+equally in predicting SRN compared to clustering all of the sequence
+data together. The ability to integrate new data into existing OTUs
+enables the deployment of a single machine learning model based on
+microbiome composition that new data can be classified with. These
+results are based on a single data set and disease. Further analysis is
+needed to determine the data size necessary to build a robust model
+capable of classifying diverse data sets. A robust machine learning
+model could be implemented as part of a non-invasive and low-cost aid in
+diagnosing SRN.
 
 ## Materials and Methods
-- 16S rRNA amplicon sequence data from 490 subjects {baxter}
-  - 261 controls
-  - 229 SRN
-- created 100 random splits of the data (80% training, 20% test)
-- preprocessed data with mothur v1.45
-- two pathways  
-  -opticlust:  
-    - clustered all data together
-    - split the shared file based on the random splits
 
-  -optifit:  
-    - split the data
-    - opticlust on the 80% training set
-    - optifit to fit the remaining 20% to the training set OTUs
-- ML with mikropml package (version)
-  - preprocessed training set and applied preprocessing to test set
-    - correlated collapsed, removed nzv
+***Data Set.*** Raw 16S rRNA amplicon sequence data isolated from human
+stool samples was downloaded from NCBI Sequence Read Archive (accession
+no. SRP062005) @baxter2016.; @edgar2011 This data set contains stool
+samples from a total of N subjects, however after preprocessing to
+screen for sequence quality and subsample to 10,000 reads per sample,
+490 samples remained. For this analysis, samples from subjects
+identified in the metadata as normal, high risk normal, or adenoma were
+categorized as "normal" while samples from subjects identified as
+advanced adenoma or carcinoma were categorized as "screen relevant
+neoplasia" (SRN). The resulting data set consisted of 261 normal samples
+and 229 SRN samples.
+
+```{r n_test_train,include=F}
+n_test_train <- read_csv("../results/tables/splitTogetherFrequency.csv") %>% 
+  mutate(n_test = rowSums(across(starts_with("split_")))) %>% 
+  select(Group,n_test) %>% 
+  mutate(n_train = 100 - n_test) %>% 
+  summarize(avg_test = mean(n_test),
+            sd_test = sd(n_test),
+            avg_train = mean(n_train),
+            sd_train = sd(n_train))
+```
+
+***Data Processing.*** The full dataset was pre-processed with mothur
+(v1.45) @schloss2009 using the SILVA reference database (v132)
+@quast2013 to join forward and reverse reads, merge duplicate reads,
+align to the reference, pre-cluster, remove chimeras with UCHIME
+@edgar2011, assign taxonomy, and remove non-bacterial reads following
+the Schloss Lab MiSeq standard operating procedure described on the
+mothur website (<https://mothur.org/wiki/miseq_sop/>). 100 splits of the
+490 samples were generated where 80% of the samples (392 samples) were
+randomly assigned to the training set and the remaining 20% (98 samples)
+were assigned to the test set. Using 100 splits of the data accounts for
+the variation that may be observed depending on the samples that are in
+the training or test sets. Each samples was in the training set an
+average of `r n_test_train %>% pull(avg_train)` times
+(SD=`r format_decimal(n_test_train %>% pull(sd_train),digits = 1)`) and
+the test set an average of `r n_test_train %>% pull(avg_test)` times
+(SD=`r format_decimal(n_test_train %>% pull(sd_test),digits = 1)`).
+
+The data was processed through two workflows. First, the standard
+workflow using the OptiClust algorithm @westcott2017. In this pathway,
+all of the data was clustered together with OptiClust to generate OTUs
+and the resulting abundance tables were split into the training and
+testing sets. In the second workflow, the pre-processed data was split
+into the training and testing sets. The training set was clustered into
+OTUs, then the test set was fit to the OTUs of the training set using
+the OptiFit algorithm @sovacool2022. The OptiFit algorithm was run with
+method open so that any sequences that didn't map to the existing OTU
+clusters would form new OTUs. Any OTUs that were not in the training set
+were removed prior to machine learning. For both pathways, the shared
+files were sub-sampled to 10,000 reads per sample.
+
+***Machine Learning.*** Machine learning using Random Forest was
+conducted with the R package mikrompl (v XXXX) @topçuoglu2021 to predict
+the diagnosis (SRN or normal) for the samples in the test set for each
+data split. The training set was preprocessed to normalize values
+(scale/center), collapse correlated features, and remove features with
+zero-variance. The preprocessing from the training set was then applied
+to the test set. P values comparing model performance were calculated as
+previously described {}. The averaged ROC curves were plotted by taking
+the average and standard deviation of the sensitivity at each
+specificity value.
+
+***Code Availability.*** The analysis workflow was implemented in
+Snakemake @koster2012 . Scripts for analysis were written in R @R2020
+and GNU bash @GNUbash. The software used includes mothur v1.47.0
+@schloss2009, RStudio @RStudio2019, the Tidyverse metapackage
+@wickham2019, R Markdown @xie_r_2018, the SRA toolkit
+@noauthor_sra-tools_nodate, and conda @noauthor_anaconda_2016. The
+complete workflow and supporting files required to reproduce this study
+are available at:
+<https://github.com/SchlossLab/Armour_OptiFitGLNE_XXXX_2021>
 
 ## Acknowledgements
-- funding
 
+(funding)
+
+\newpage
+
+## References
+
+```{=tex}
+\setlength{\parindent}{-0.25in}
+\setlength{\leftskip}{0.25in}
+```
+\noindent
+
+::: {#refs}
+:::
+
+```{=tex}
+\setlength{\parindent}{0in}
+\setlength{\leftskip}{0in}
+```
 \newpage
@@
 
-**Figure 1. Workflow** description.
+![](../exploratory/figures/figure1_a.pdf)
 
-**Figure 2. Model Performance.** **A)** Mean AUC **B)** Averaged ROC curves
+![](../exploratory/figures/figure1_b.pdf)
+
+**Figure 1: Workflows.** **A)** OptiClust workflow: The full data set
+was clustered into OTUs using the OptiClust algorithm in mothur. The
+data was then split into two sets where 80% of the samples were assigned
+to the training set and 20% to the testing set. The training set was
+preprocessed with mikropml to normalize values (scale/center), collapse
+correlated features, and remove features with zero-variance. Using
+mikropml, the training set was split into train and validate sets to
+compare results using different hyperparameter settings. The highest
+performing hyperparameter setting was then used to train the model with
+the full training set. The preprocessing scale from the training set was
+applied to the test data set, then the trained model was used to
+classify the samples in the test set. Based on the actual classification
+and predicted classification, the area under the receiver operating
+characteristic curve (AUROC) was calculated to summarize model
+performance. The entire process was repeated 100 times to account for
+variability depending on the split of the data resulting in a total of
+100 AUROC values summarizing the performance of the standard OptiClust
+workflow. **B)** OptiFit workflow: The data set was first split into two
+sets where 80% of the samples were assigned to the training set and 20%
+to the testing set. The training set was then clustered into OTUs using
+the OptiClust algorithm in mothur. The resulting abundance data was
+preprocessed with mikropml to normalize values (scale/center), collapse
+correlated features, and remove features with zero-variance. Using
+mikropml, the training set was split into train and validate sets to
+compare results using different hyperparameter settings. The highest
+performing hyperparameter setting was then used to train the model with
+the full training set. The OptiFit algorithm in mothur was used to
+cluster the left out testing data set using the OTUs of the training set
+as a reference. The preprocessing scale from the training set was
+applied to the test data set, then the trained model was used to
+classify the samples in the test set. Based on the actual classification
+and predicted classification, the area under the receiver operating
+characteristic curve (AUROC) was calculated to summarize model
+performance. The entire process was repeated 100 times to account for
+variability depending on the split of the data resulting in a total of
+100 AUROC values summarizing the performance of the new OptiFit
+workflow.
 
@@
 
-## References
+![](../exploratory/figures/figure2.pdf)
+
+**Figure 2: Model Performance.** **A)** Area under the receiver
+operating characteristic (AUROC) curve during cross-validation for the
+OptiClust and OptiFit workflows. Mean and standard deviation of the
+AUROC is represented by the black dot and whiskers. Mean AUROC is
+printed to the right of the points. **B)** AUROC on the test data for
+the OptiClust and OptiFit workflows. Mean and standard deviation of the
+AUROC is represented by the black dot and whiskers. Mean AUROC is
+printed to the right of the points. **C)** Receiver operating
+characteristic (ROC)Averaged ROC curves

### 2022-06-15 submission/manuscript.tex
full manuscript draft

--- 2022-02-23 submission/manuscript.tex
@@
 \documentclass[
-  11pt,
 ]{article}
-\title{\textbf{TITLE}}
-\author{}
-\date{\vspace{-2.5em}}
-
 \usepackage{amsmath,amssymb}
@@
 \hypersetup{
-  pdftitle={TITLE},
+  pdftitle={Reference-based OTU clustering for machine learning classification},
   hidelinks,
@@
 \setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-\usepackage{helvet} % Helvetica font
-\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
-\usepackage[T1]{fontenc}
-
-\usepackage[none]{hyphenat}
-
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+\usepackage{helvet}
+\renewcommand*\familydefault{\sfdefault}
 \usepackage{setspace}
 \doublespacing
-\setlength{\parskip}{1em}
-
-\usepackage{lineno}
-
-\usepackage{pdfpages}
-\usepackage{helvet}
+\usepackage[left]{lineno}
+\usepackage{multirow}
+\usepackage[none]{hyphenat}
 \ifLuaTeX
@@
 
+\title{\textbf{Reference-based OTU clustering for machine learning
+classification}}
+\author{}
+\date{\vspace{-2.5em}}
+
 \begin{document}
@@
 
-\vspace{20mm}
+Other Title Options:\\
+A solution to the problem of inconsistent OTU clustering for machine
+learning classification
+
+Optimized reference-based OTU clustering for machine learning
+classification
+
+\vspace{5mm}
 
@@
 
-\vspace{20mm}
+\vspace{5mm}
 
-Courtney R. Armour\({^1}\), William L. Close\(^{1,*}\), Begüm D.
-Topçuoğlu\(^{1,\#}\), Patrick D. Schloss \(^{1,\dagger}\)
+Courtney R. Armour\({^1}\), Kelly L. Sovacool\({^2}\), William L.
+Close\(^{1,*}\), Begüm D. Topçuoğlu\(^{1,\#}\), Jenna Wiens\({^3}\),
+Patrick D. Schloss \(^{1,\dagger}\)
 
@@
 \({^1}\) Department of Microbiology and Immunology, University of
-Michigan, Ann Arbor MI.
+Michigan, Ann Arbor, Michigan, USA
+
+\({^2}\) Department of Computational Medicine and Bioinformatics,
+University of Michigan, Ann Arbor, Michigan, USA
+
+\({^3}\) Department of Electrical Engineering and Computer Science,
+University of Michigan, Ann Arbor, Michigan, USA
+
+\({^*}\) Current Affiliation:
 
@@
 
-\({^*}\) Current Affiliation:
-
-\vspace{20mm}
-
 \(\dagger\) To whom correspondence should be addressed:
@@
 
+Machine learning classification of disease based on the gut microbiome
+often relies on operational taxonomic units (OTUs) to quantify microbial
+composition. The standard methodology for quantifying taxonomic
+composition is clustering 16S rRNA amplicon sequences \emph{de novo}
+into OTUs and using their abundances to train a classification model.
+However, OTU clusters depend on the sequences in the data set and
+therefore can change if new data is added. This inconsistency
+complicates machine learning classification because in order to use the
+model to classify additional samples, the OTUs have to be re-clustered
+to include the new sequences and the model retrained with the new OTU
+clusters. A new algorithm in mothur, called OptiFit, addresses this
+issue by clustering new sequences into the existing OTU clusters. While
+OptiFit is proven to produce high quality OTU clusters, it is unclear
+whether this method for clustering new sequence data into existing OTUs
+will have any impact on machine learning classification using the OTU
+abundances. To evaluate the potential for use in machine learning
+classification, we used OptiFit to cluster additional data into existing
+OTU clusters and quantified model performance in classifying a data set
+containing samples from controls and people with advanced adenoma and
+carcinoma (SRN). We compared the model performance using OptiFit to the
+standard procedure of clustering all the data together. We found that
+OptiFit can be used to fit new sequence data to existing OTUs and
+perform equally in classifying SRNs. Moving forward, when OTUs are used
+in machine learning classification problems, OptiFit can be used to
+avoid the need to re-cluster sequences into OTUs when classifying
+additional samples that were not part of the data set used to build the
+model.
+
 \hypertarget{importance-150-word-max}{%
@@
 
+There are instances when OTU abundance data is optimal for machine
+learning prediction of disease based on microbiome composition. The
+current methodology for OTU clustering complicates machine learning
+since \emph{de novo} OTU clusters can change depending on the data in
+the data set. Attempting to classify additional samples that were not
+part of the original model build requires re-clustering OTUs and
+re-training the model. OptiFit provides an elegant solution to the
+problem of OTU inconsistency by enabling new data to be fit to the
+existing OTUs while maintaining machine learning classification
+performance. The use of OptiFit will enable a single model based on OTU
+abundance data to be used in classifying additional samples that were
+not part of the original OTU clustering.
+
 \newpage
 
-\hypertarget{introduction-250-words}{%
-\subsection{Introduction (\textasciitilde250
-words)}\label{introduction-250-words}}
+Gut microbiome community composition is useful as a resource for machine
+learning prediction of diseases, including colorectal cancer
+(\protect\hyperlink{ref-baxter2016}{1},
+\protect\hyperlink{ref-zackular2014}{2}). Amplicon sequencing of the 16S
+rRNA gene is a reliable tool for assessing the taxonomic composition of
+microbial communities. Analysis of 16S rRNA amplicon sequence (16S) data
+generally relies on clustering of sequences based on similarity into
+operational taxonomic units (OTUs). The process of OTU clustering can
+either be reference-based or \emph{de novo}. The quality of OTUs
+generated with reference-based clustering is dependent on the quality of
+the reference used, while OTUs generated with \emph{de novo} clustering
+are optimized to the data set (\protect\hyperlink{ref-westcott2015}{3}).
+While \emph{de novo} clustering can produce high-quality OTU clusters
+where sequences are accurately grouped based on similarity thresholds,
+the resulting OTU clusters depend on the data in the data set and the
+addition of new data could change the overall OTU clusters. The
+inconsistent nature of OTU clustering complicates deployment of machine
+learning models since integration of additional data requires
+re-clustering all the data and re-training of the model. The ability to
+integrate new data into an existing model without re-clustering and
+re-training could allow for deployment of a single model that new data
+can be continually added to. Recently Sovacool \emph{et al} introduced
+OptiFit: a method for fitting new data into existing OTU clusters
+(\protect\hyperlink{ref-sovacool2022}{4}). While OptiFit is proven to
+effectively fit new sequence data to existing OTU clusters, it is
+unknown if the use of OptiFit will have an impact on machine learning
+classification. Here we test the ability of OptiFit to cluster new
+sequence data into existing OTU clusters for the purpose of machine
+learning classification of disease based on gut microbiome composition.
 
-\begin{itemize}
-\tightlist
-\item
-  Gut microbiome community composition has proven useful as a resource
-  for machine learning prediction of various diseases \{examples\}.
-\item
-  Amplicon sequencing of the 16S rRNA gene is a reliable tool for
-  assessing the taxonomic composition of microbial communities.
-\item
-  Analysis of 16S rRNA sequence data generally relies on clustering of
-  sequences based on similarity into operational taxonomic units (OTUs).
-\item
-  However OTU clustering depends on the data in the dataset and the
-  addition of new data may change the overall OTU clusters.
-\item
-  The unstable nature of OTU clustering complicates machine learning.
-  When building a classification model, changes to the OTU clusters
-  means you have to re-create the model. This can change the underlying
-  model and be time consuming and resource intensive.
-\item
-  The ability to integrate new data into an existing model could allow
-  for deployment of a single model that new data can be continually
-  added to and predicted on.
-\item
-  Recently Sovacool \emph{et al} described a new method for fitting new
-  data into existing OTU clusters \{Kelly optifit 2022\}.
-\item
-  While OptiFit works well to fit new sequence data and provide high
-  quality OTU clusters, it is unknown if the use of OptiFit will have an
-  impact on machine learning predictions.
-\item
-  Here, we use OptiFit with a 16S rRNA sequence dataset consisting of
-  normal and SRN samples to test how well new data integrated with
-  OptiFit performs for prediction of SRN.
-\end{itemize}
+When using OTU abundances for machine learning classification, the
+current methodology is to \emph{de novo} cluster all of the sequence
+data into OTUs with the OptiClust algorithm in mothur
+(\protect\hyperlink{ref-westcott2017}{5}). The resulting abundance data
+is then split into training and testing sets, where the training set is
+used to tune hyperparameters and ultimately train the model. The testing
+set is then classified with the model and the performance of the model
+can be quantified (Figure 1A). However, with this methodology we would
+have to re-generate the OTU clusters and re-train the model if we wanted
+to classify additional samples. The OptiFit algorithm
+(\protect\hyperlink{ref-sovacool2022}{4}) addresses this problem by
+enabling new sequences to be clustered into existing OTUs. The OptiFit
+workflow is similar to the OptiClust workflow where the data is
+clustered into OTUs and used to tune hyperparameters and ultimately
+train the model. Then, we can use OptiFit to fit sequence data of
+samples not part of the original data set into the existing OTUs and use
+the same model to classify the samples (Figure 1B). To test how the
+model performance compares between these two methodologies, we used a
+publicly available data set of 16S sequences from stool samples of
+healthy subjects as well as subjects with SRN consisting of advanced
+adenoma and carcinoma (\protect\hyperlink{ref-baxter2016}{1}). The data
+set was randomly split into an 80\% train set and 20\% test set. For the
+standard OptiClust workflow, the training and test sets were \emph{de
+novo} clustered together into OTUs then the resulting abundance table
+was split into the training and testing set. For the OptiFit workflow,
+the train set was clustered \emph{de novo} into OTUs and the remaining
+test set was fit to the OTU clusters using the OptiFit algorithm. For
+both workflows, the abundance table of the train set was used to tune
+hyperparameters and train a random forest model to classify SRN. The
+test set was classified as either control or SRN using the trained
+models.To account for variation depending on the split of the data, the
+data set was randomly split 100 times and the process repeated for each
+of the 100 data splits. By comparing the model performance of
+classifying the samples in the test data set between the OptiFit and
+OptiClust algorithms, we can quantify the impact of using OptiFit on
+model classification performance.
 
-\hypertarget{results-700-words}{%
-\subsection{Results (\textasciitilde700
-words)}\label{results-700-words}}
+We first examined the quality of the resulting OTU clusters from the two
+algorithms using the Matthews correlation coefficient (MCC). The MCC
+score is quantified by examining all pairs of sequences and assessing
+whether they belong together in an OTU or not based on sequence
+similarity (\protect\hyperlink{ref-westcott2017}{5}). MCC scores range
+between negative one and one. A score of negative one means none of the
+sequences in an OTU are within the similarity threshold and any
+sequences within the similarity threshold are not in an OTU together. An
+MCC score of zero essentially means the sequences are randomly
+clustered. An MCC score of 1 means all sequences in an OTU are within
+the similarity threshold and all sequence pairs within the similarity
+threshold are in the same OTU. To ensure that OptiFit is appropriately
+integrating new sequence data into the existing OTUs, we would expect
+that the MCC scores produced by the OptiClust and OptiFit workflows are
+similar. Since the data is only clustered once in the OptiClust workflow
+there is only one MCC score while the OptiFit workflow produces an MCC
+score for the OTU clusters from each data split. Overall the MCC scores
+were similar between OptiClust (MCC = 0.884) and OptiFit (average MCC =
+0.879) indicating that OptiFit performs as well as OptiClust when
+integrating new sequences into the existing OTUs.
 
-\begin{itemize}
-\tightlist
-\item
-  Utilize public dataset with normal and SRN samples to compare
-  prediction between OptiFit and OptiClust
-\item
-  randomly split data into 80\% training 20\% test sets 100 times
-\item
-  Processed the data with both algorithms - Figure 1
+After verifying that the quality of the OTUs was consistent between
+OptiClust and OptiFit, we next examined the model performance for
+classifying samples in the test data set as control or SRN. To quantify
+model performance we used the taxonomic abundances of the training data
+from the OptiClust and OptiFit workflows to train a model to predict
+SRNs. Using the predicted and actual diagnosis classification, we
+calculated the area under the receiver operating characteristic curve
+(AUROC) for each data split to quantify model performance. During
+cross-validation (CV) training, the model performance was equivalent
+between the two algorithms (p-value = 0.13, OptiClust mean CV AUROC =
+0.694, OptiFit mean CV AUROC = 0.697, Figure 2A). The trained model was
+then deployed to classify the samples of the test data as control or
+SRN. The performance on the test data was equivalent between the two
+algorithms (p-value = 0.63, OptiClust mean test AUROC = 0.709, OptiFit
+mean test AUROC = 0.712, Figure 2B,C) indicating that new data can be
+fit to existing OTU clusters without impacting model performance.
 
-  \begin{itemize}
-  \item
-    \begin{enumerate}
-    \def\labelenumi{\arabic{enumi}.}
-    \tightlist
-    \item
-      Used traditional OptiClust method to cluster all data, then split
-      into training and test set
-    \end{enumerate}
-  \item
-    \begin{enumerate}
-    \def\labelenumi{\arabic{enumi}.}
-    \setcounter{enumi}{1}
-    \tightlist
-    \item
-      Used OptiClust on the 80\% training set, then used OptiFit to fit
-      the remaining 20\%
-    \end{enumerate}
-  \end{itemize}
-\item
-  OptiFit produces similar quality OTU clusters based on MCC
-  (supplement?)
-\item
-  Used mikropml to train a model on the 80\% training set for each data
-  split, then predicted dx on the 20\% test set
-\item
-  Training performance almost identical (OptiClust median CV AUC 0.694,
-  OptiFit median CV AUC 0.693) - Figure 2
-\item
-  Performance on the test set comparable (OptiClust median CV AUC 0.694,
-  OptiFit median CV AUC 0.693) - Figure 2
-\item
-  Maybe expand on where optifit did better/worse?
-\end{itemize}
-
-\hypertarget{discussionconclusions-250-words}{%
-\subsection{Discussion/Conclusions (\textasciitilde250
-words)}\label{discussionconclusions-250-words}}
-
-\begin{itemize}
-\tightlist
-\item
-  OptiFit works!
-\item
-  future questions:
-
-  \begin{itemize}
-  \tightlist
-  \item
-    how much reference do you need?
-  \item
-    does it work well for other situations?
-
-    \begin{itemize}
-    \tightlist
-    \item
-      what if we used a new dataset for test set instead of a subset of
-      the full dataset?
-    \item
-      other diseases?
-    \end{itemize}
-  \end{itemize}
-\end{itemize}
+We tested the ability of OptiFit to integrate new data into existing
+OTUs for the purpose of machine learning classification of disease based
+on microbiome abundance. A potential problem to using OptiFit for
+machine learning prediction is that any sequences in the new data that
+do not map to the existing OTU clusters will be discarded resulting in a
+possible loss of information. However, we demonstrated that OptiFit can
+be used to fit new sequence data into existing OTU clusters and perform
+equally in predicting SRN compared to clustering all of the sequence
+data together. The ability to integrate new data into existing OTUs
+enables the deployment of a single machine learning model based on
+microbiome composition that new data can be classified with. These
+results are based on a single data set and disease. Further analysis is
+needed to determine the data size necessary to build a robust model
+capable of classifying diverse data sets. A robust machine learning
+model could be implemented as part of a non-invasive and low-cost aid in
+diagnosing SRN.
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  16S rRNA amplicon sequence data from 490 subjects \{baxter\}
+\textbf{\emph{Data Set.}} Raw 16S rRNA amplicon sequence data isolated
+from human stool samples was downloaded from NCBI Sequence Read Archive
+(accession no. SRP062005) (\protect\hyperlink{ref-baxter2016}{1}).;
+(\protect\hyperlink{ref-edgar2011}{6}) This data set contains stool
+samples from a total of N subjects, however after preprocessing to
+screen for sequence quality and subsample to 10,000 reads per sample,
+490 samples remained. For this analysis, samples from subjects
+identified in the metadata as normal, high risk normal, or adenoma were
+categorized as ``normal'' while samples from subjects identified as
+advanced adenoma or carcinoma were categorized as ``screen relevant
+neoplasia'' (SRN). The resulting data set consisted of 261 normal
+samples and 229 SRN samples.
 
-  \begin{itemize}
-  \tightlist
-  \item
-    261 controls
-  \item
-    229 SRN
-  \end{itemize}
-\item
-  created 100 random splits of the data (80\% training, 20\% test)
-\item
-  preprocessed data with mothur v1.45
-\item
-  two pathways\\
-  -opticlust:
+\textbf{\emph{Data Processing.}} The full dataset was pre-processed with
+mothur (v1.45) (\protect\hyperlink{ref-schloss2009}{7}) using the SILVA
+reference database (v132) (\protect\hyperlink{ref-quast2013}{8}) to join
+forward and reverse reads, merge duplicate reads, align to the
+reference, pre-cluster, remove chimeras with UCHIME
+(\protect\hyperlink{ref-edgar2011}{6}), assign taxonomy, and remove
+non-bacterial reads following the Schloss Lab MiSeq standard operating
+procedure described on the mothur website
+(\url{https://mothur.org/wiki/miseq_sop/}). 100 splits of the 490
+samples were generated where 80\% of the samples (392 samples) were
+randomly assigned to the training set and the remaining 20\% (98
+samples) were assigned to the test set. Using 100 splits of the data
+accounts for the variation that may be observed depending on the samples
+that are in the training or test sets. Each samples was in the training
+set an average of 80 times (SD=4.1) and the test set an average of 20
+times (SD=4.1).
 
-  \begin{itemize}
-  \tightlist
-  \item
-    clustered all data together
-  \item
-    split the shared file based on the random splits
-  \end{itemize}
+The data was processed through two workflows. First, the standard
+workflow using the OptiClust algorithm
+(\protect\hyperlink{ref-westcott2017}{5}). In this pathway, all of the
+data was clustered together with OptiClust to generate OTUs and the
+resulting abundance tables were split into the training and testing
+sets. In the second workflow, the pre-processed data was split into the
+training and testing sets. The training set was clustered into OTUs,
+then the test set was fit to the OTUs of the training set using the
+OptiFit algorithm (\protect\hyperlink{ref-sovacool2022}{4}). The OptiFit
+algorithm was run with method open so that any sequences that didn't map
+to the existing OTU clusters would form new OTUs. Any OTUs that were not
+in the training set were removed prior to machine learning. For both
+pathways, the shared files were sub-sampled to 10,000 reads per sample.
 
-  -optifit:
+\textbf{\emph{Machine Learning.}} Machine learning using Random Forest
+was conducted with the R package mikrompl (v XXXX)
+(\protect\hyperlink{ref-topuxe7uoglu2021}{9}) to predict the diagnosis
+(SRN or normal) for the samples in the test set for each data split. The
+training set was preprocessed to normalize values (scale/center),
+collapse correlated features, and remove features with zero-variance.
+The preprocessing from the training set was then applied to the test
+set. P values comparing model performance were calculated as previously
+described \{\}. The averaged ROC curves were plotted by taking the
+average and standard deviation of the sensitivity at each specificity
+value.
 
-  \begin{itemize}
-  \tightlist
-  \item
-    split the data
-  \item
-    opticlust on the 80\% training set
-  \item
-    optifit to fit the remaining 20\% to the training set OTUs
-  \end{itemize}
-\item
-  ML with mikropml package (version)
-
-  \begin{itemize}
-  \tightlist
-  \item
-    preprocessed training set and applied preprocessing to test set
-
-    \begin{itemize}
-    \tightlist
-    \item
-      correlated collapsed, removed nzv
-    \end{itemize}
-  \end{itemize}
-\end{itemize}
+\textbf{\emph{Code Availability.}} The analysis workflow was implemented
+in Snakemake (\protect\hyperlink{ref-koster2012}{10}) . Scripts for
+analysis were written in R (\protect\hyperlink{ref-R2020}{11}) and GNU
+bash (\protect\hyperlink{ref-GNUbash}{12}). The software used includes
+mothur v1.47.0 (\protect\hyperlink{ref-schloss2009}{7}), RStudio
+(\protect\hyperlink{ref-RStudio2019}{13}), the Tidyverse metapackage
+(\protect\hyperlink{ref-wickham2019}{14}), R Markdown
+(\protect\hyperlink{ref-xie_r_2018}{15}), the SRA toolkit
+(\protect\hyperlink{ref-noauthor_sra-tools_nodate}{16}), and conda
+(\protect\hyperlink{ref-noauthor_anaconda_2016}{17}). The complete
+workflow and supporting files required to reproduce this study are
+available at:
+\url{https://github.com/SchlossLab/Armour_OptiFitGLNE_XXXX_2021}
 
@@
 
-\begin{itemize}
-\tightlist
-\item
-  funding
-\end{itemize}
+(funding)
+
+\newpage
+
+\hypertarget{references}{%
+\subsection{References}\label{references}}
+
+\setlength{\parindent}{-0.25in}
+\setlength{\leftskip}{0.25in}
+
+\noindent
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{0}{1}
+\leavevmode\vadjust pre{\hypertarget{ref-baxter2016}{}}%
+\CSLLeftMargin{1. }
+\CSLRightInline{\textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers
+MAM}, \textbf{Schloss PD}. 2016. Microbiota-based model improves the
+sensitivity of fecal immunochemical test for detecting colonic lesions.
+Genome Medicine \textbf{8}:37.
+doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-zackular2014}{}}%
+\CSLLeftMargin{2. }
+\CSLRightInline{\textbf{Zackular JP}, \textbf{Rogers MAM},
+\textbf{Ruffin MT,I}, \textbf{Schloss PD}. 2014. The human gut
+microbiome as a screening tool for colorectal cancer. Cancer Prevention
+Research \textbf{7}:1112--1121.
+doi:\href{https://doi.org/10.1158/1940-6207.CAPR-14-0129}{10.1158/1940-6207.CAPR-14-0129}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-westcott2015}{}}%
+\CSLLeftMargin{3. }
+\CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo
+clustering methods outperform reference-based methods for assigning 16S
+rRNA gene sequences to operational taxonomic units. PeerJ
+\textbf{3}:e1487.
+doi:\href{https://doi.org/10.7717/peerj.1487}{10.7717/peerj.1487}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-sovacool2022}{}}%
+\CSLLeftMargin{4. }
+\CSLRightInline{\textbf{Sovacool KL}, \textbf{Westcott SL},
+\textbf{Mumphrey MB}, \textbf{Dotson GA}, \textbf{Schloss PD}. 2022.
+OptiFit: An improved method for fitting amplicon sequences to existing
+OTUs. mSphere \textbf{7}:e00916--21.
+doi:\href{https://doi.org/10.1128/msphere.00916-21}{10.1128/msphere.00916-21}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-westcott2017}{}}%
+\CSLLeftMargin{5. }
+\CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2017.
+OptiClust, an improved method for assigning amplicon-based sequence data
+to operational taxonomic units. mSphere \textbf{2}:e00073--17.
+doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-edgar2011}{}}%
+\CSLLeftMargin{6. }
+\CSLRightInline{\textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente
+JC}, \textbf{Quince C}, \textbf{Knight R}. 2011. UCHIME improves
+sensitivity and speed of chimera detection. Bioinformatics
+\textbf{27}:2194--2200.
+doi:\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-schloss2009}{}}%
+\CSLLeftMargin{7. }
+\CSLRightInline{\textbf{Schloss PD}, \textbf{Westcott SL},
+\textbf{Ryabin T}, \textbf{Hall JR}, \textbf{Hartmann M},
+\textbf{Hollister EB}, \textbf{Lesniewski RA}, \textbf{Oakley BB},
+\textbf{Parks DH}, \textbf{Robinson CJ}, \textbf{Sahl JW}, \textbf{Stres
+B}, \textbf{Thallinger GG}, \textbf{Van Horn DJ}, \textbf{Weber CF}.
+2009. Introducing mothur: Open-source, platform-independent,
+community-supported software for describing and comparing microbial
+communities. Applied and Environmental Microbiology
+\textbf{75}:7537--7541.
+doi:\href{https://doi.org/10.1128/AEM.01541-09}{10.1128/AEM.01541-09}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-quast2013}{}}%
+\CSLLeftMargin{8. }
+\CSLRightInline{\textbf{Quast C}, \textbf{Pruesse E}, \textbf{Yilmaz P},
+\textbf{Gerken J}, \textbf{Schweer T}, \textbf{Yarza P}, \textbf{Peplies
+J}, \textbf{Glöckner FO}. 2013. The SILVA ribosomal RNA gene database
+project: Improved data processing and web-based tools. Nucleic Acids
+Research \textbf{41}:D590--D596.
+doi:\href{https://doi.org/10.1093/nar/gks1219}{10.1093/nar/gks1219}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-topuxe7uoglu2021}{}}%
+\CSLLeftMargin{9. }
+\CSLRightInline{\textbf{Topçuoğlu BD}, \textbf{Lapp Z}, \textbf{Sovacool
+KL}, \textbf{Snitkin E}, \textbf{Wiens J}, \textbf{Schloss PD}. 2021.
+mikropml: User-Friendly R Package for Supervised Machine Learning
+Pipelines. Journal of Open Source Software \textbf{6}:3073.
+doi:\href{https://doi.org/10.21105/joss.03073}{10.21105/joss.03073}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-koster2012}{}}%
+\CSLLeftMargin{10. }
+\CSLRightInline{\textbf{Koster J}, \textbf{Rahmann S}. 2012.
+Snakemake--a scalable bioinformatics workflow engine. Bioinformatics
+\textbf{28}:2520--2522.
+doi:\href{https://doi.org/10.1093/bioinformatics/bts480}{10.1093/bioinformatics/bts480}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-R2020}{}}%
+\CSLLeftMargin{11. }
+\CSLRightInline{\textbf{R Core Team}. 2020.
+\href{https://www.R-project.org/}{R: A language and environment for
+statistical computing}. R Foundation for Statistical Computing, Vienna,
+Austria.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-GNUbash}{}}%
+\CSLLeftMargin{12. }
+\CSLRightInline{\textbf{GNU Project}.
+\href{https://www.gnu.org/software/bash/\%20manual/bash.html/}{Bash
+reference manual}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-RStudio2019}{}}%
+\CSLLeftMargin{13. }
+\CSLRightInline{\textbf{RStudio Team}. 2019.
+\href{http://www.rstudio.com/}{RStudio: Integrated development
+environment for r}. RStudio, Inc., Boston, MA.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-wickham2019}{}}%
+\CSLLeftMargin{14. }
+\CSLRightInline{\textbf{Wickham H}, \textbf{Averick M}, \textbf{Bryan
+J}, \textbf{Chang W}, \textbf{McGowan LD}, \textbf{François R},
+\textbf{Grolemund G}, \textbf{Hayes A}, \textbf{Henry L}, \textbf{Hester
+J}, \textbf{Kuhn M}, \textbf{Pedersen TL}, \textbf{Miller E},
+\textbf{Bache SM}, \textbf{Müller K}, \textbf{Ooms J}, \textbf{Robinson
+D}, \textbf{Seidel DP}, \textbf{Spinu V}, \textbf{Takahashi K},
+\textbf{Vaughan D}, \textbf{Wilke C}, \textbf{Woo K}, \textbf{Yutani H}.
+2019. Welcome to the Tidyverse. Journal of Open Source Software
+\textbf{4}:1686.
+doi:\href{https://doi.org/10.21105/joss.01686}{10.21105/joss.01686}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-xie_r_2018}{}}%
+\CSLLeftMargin{15. }
+\CSLRightInline{\textbf{Xie Y}, \textbf{Allaire JJ}, \textbf{Grolemund
+G}. 2018. R {Markdown}: {The Definitive Guide}. {Taylor \& Francis, CRC
+Press}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-noauthor_sra-tools_nodate}{}}%
+\CSLLeftMargin{16. }
+\CSLRightInline{{SRA}-{Tools} - {NCBI}.
+http://ncbi.github.io/sra-tools/.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-noauthor_anaconda_2016}{}}%
+\CSLLeftMargin{17. }
+\CSLRightInline{2016. Anaconda {Software Distribution}. Anaconda
+Documentation. Anaconda Inc.}
+
+\end{CSLReferences}
+
+\setlength{\parindent}{0in}
+\setlength{\leftskip}{0in}
 
@@
 
-\textbf{Figure 1. Workflow} description.
+\includegraphics{../exploratory/figures/figure1_a.pdf}
 
-\textbf{Figure 2. Model Performance.} \textbf{A)} Mean AUC \textbf{B)}
-Averaged ROC curves
+\includegraphics{../exploratory/figures/figure1_b.pdf}
+
+\textbf{Figure 1: Workflows.} \textbf{A)} OptiClust workflow: The full
+data set was clustered into OTUs using the OptiClust algorithm in
+mothur. The data was then split into two sets where 80\% of the samples
+were assigned to the training set and 20\% to the testing set. The
+training set was preprocessed with mikropml to normalize values
+(scale/center), collapse correlated features, and remove features with
+zero-variance. Using mikropml, the training set was split into train and
+validate sets to compare results using different hyperparameter
+settings. The highest performing hyperparameter setting was then used to
+train the model with the full training set. The preprocessing scale from
+the training set was applied to the test data set, then the trained
+model was used to classify the samples in the test set. Based on the
+actual classification and predicted classification, the area under the
+receiver operating characteristic curve (AUROC) was calculated to
+summarize model performance. The entire process was repeated 100 times
+to account for variability depending on the split of the data resulting
+in a total of 100 AUROC values summarizing the performance of the
+standard OptiClust workflow. \textbf{B)} OptiFit workflow: The data set
+was first split into two sets where 80\% of the samples were assigned to
+the training set and 20\% to the testing set. The training set was then
+clustered into OTUs using the OptiClust algorithm in mothur. The
+resulting abundance data was preprocessed with mikropml to normalize
+values (scale/center), collapse correlated features, and remove features
+with zero-variance. Using mikropml, the training set was split into
+train and validate sets to compare results using different
+hyperparameter settings. The highest performing hyperparameter setting
+was then used to train the model with the full training set. The OptiFit
+algorithm in mothur was used to cluster the left out testing data set
+using the OTUs of the training set as a reference. The preprocessing
+scale from the training set was applied to the test data set, then the
+trained model was used to classify the samples in the test set. Based on
+the actual classification and predicted classification, the area under
+the receiver operating characteristic curve (AUROC) was calculated to
+summarize model performance. The entire process was repeated 100 times
+to account for variability depending on the split of the data resulting
+in a total of 100 AUROC values summarizing the performance of the new
+OptiFit workflow.
 
@@
 
-\hypertarget{references}{%
-\subsection{References}\label{references}}
+\includegraphics{../exploratory/figures/figure2.pdf}
+
+\textbf{Figure 2: Model Performance.} \textbf{A)} Area under the
+receiver operating characteristic (AUROC) curve during cross-validation
+for the OptiClust and OptiFit workflows. Mean and standard deviation of
+the AUROC is represented by the black dot and whiskers. Mean AUROC is
+printed to the right of the points. \textbf{B)} AUROC on the test data
+for the OptiClust and OptiFit workflows. Mean and standard deviation of
+the AUROC is represented by the black dot and whiskers. Mean AUROC is
+printed to the right of the points. \textbf{C)} Receiver operating
+characteristic (ROC)Averaged ROC curves
 

### 2022-06-20 submission/manuscript.Rmd
revised ms

@@
 
-Other Title Options:\
-A solution to the problem of inconsistent OTU clustering for machine
-learning classification
-
-Optimized reference-based OTU clustering for machine learning
-classification
-
 \vspace{5mm}
 
-Running title: INSERT RUNNING TITLE HERE
+Running title: Reference-based OTU clustering for ML classification
 
-\vspace{5mm}
+\vspace{10mm}
 
@@
 
-\vspace{5mm}
+\vspace{10mm}
 
@@
 
-## Abstract (250 word max)
+## Abstract
 
 Machine learning classification of disease based on the gut microbiome
-often relies on operational taxonomic units (OTUs) to quantify microbial
-composition. The standard methodology for quantifying taxonomic
-composition is clustering 16S rRNA amplicon sequences *de novo* into
-OTUs and using their abundances to train a classification model.
-However, OTU clusters depend on the sequences in the data set and
-therefore can change if new data is added. This inconsistency
-complicates machine learning classification because in order to use the
-model to classify additional samples, the OTUs have to be re-clustered
-to include the new sequences and the model retrained with the new OTU
-clusters. A new algorithm in mothur, called OptiFit, addresses this
-issue by clustering new sequences into the existing OTU clusters. While
-OptiFit is proven to produce high quality OTU clusters, it is unclear
-whether this method for clustering new sequence data into existing OTUs
-will have any impact on machine learning classification using the OTU
-abundances. To evaluate the potential for use in machine learning
-classification, we used OptiFit to cluster additional data into existing
-OTU clusters and quantified model performance in classifying a data set
-containing samples from controls and people with advanced adenoma and
-carcinoma (SRN). We compared the model performance using OptiFit to the
-standard procedure of clustering all the data together. We found that
-OptiFit can be used to fit new sequence data to existing OTUs and
-perform equally in classifying SRNs. Moving forward, when OTUs are used
-in machine learning classification problems, OptiFit can be used to
-avoid the need to re-cluster sequences into OTUs when classifying
-additional samples that were not part of the data set used to build the
-model.
+often relies on clustering 16S rRNA gene sequences into operational
+taxonomic units (OTUs) to quantify microbial composition. The standard
+approach to clustering sequences into OTUs leverages the similarity of
+the sequences to each other rather than to a reference database. The
+abundance of each OTU is used to train a classification model. However,
+OTU assignments depend on the sequences in the data set and therefore
+can change if new data are added. This lack of stability complicates
+classification because in order to use the model to classify additional
+samples, the OTUs have to be reclustered to include the new sequences
+and the model retrained with the new OTU clusters. A new reference-based
+clustering algorithm, called OptiFit, addresses this issue by fitting
+new sequences into existing OTUs. While OptiFit is proven to produce
+high quality OTU clusters, it is unclear whether this method for fitting
+new sequence data into existing OTUs will impact the performance of
+classification models. We used OptiFit to cluster additional data into
+existing OTU clusters and quantified model performance in classifying a
+data set containing samples from patients with and without colonic
+screen relevant neoplasias (SRN). We compared the performance of this
+model to the standard procedure of clustering all the data together. We
+found that both approaches performed equally well in classifying SRNs.
+Moving forward, when OTUs are used in classification problems, OptiFit
+can be used to avoid the need to retrain models using reclustered
+sequences when classifying new samples.
 
-## Importance (150 word max)
+## Importance
 
-There are instances when OTU abundance data is optimal for machine
-learning prediction of disease based on microbiome composition. The
-current methodology for OTU clustering complicates machine learning
-since *de novo* OTU clusters can change depending on the data in the
-data set. Attempting to classify additional samples that were not part
-of the original model build requires re-clustering OTUs and re-training
-the model. OptiFit provides an elegant solution to the problem of OTU
-inconsistency by enabling new data to be fit to the existing OTUs while
-maintaining machine learning classification performance. The use of
-OptiFit will enable a single model based on OTU abundance data to be
-used in classifying additional samples that were not part of the
-original OTU clustering.
+There is great potential for using microbiome data to non-invasively
+diagnose people. One of the challenges with using classification models
+generated using the relative abundance of operational taxonomic unit
+(OTU) data is that 16S rRNA gene sequences are assigned to OTUs based on
+their similarity to other sequences in the dataset. If data are
+generated from new patients seeking a diagnosis, then it would be
+necessary to reassign sequences to OTUs and retrain the classification
+model. Yet there is a desire to have a single, validated model that can
+be deployed. To overcome this obstacle, we applied the OptiFit
+clustering algorithm which fits new sequence data to existing OTUs
+allowing the reuse of a consistent model. A random forest machine
+learning model deployed using OptiFit performed as well as the
+traditional reassignment and retraining approach. This result indicates
+there is potential for developing machine learning classification models
+based on OTU relative abundance data.
 
@@
 
-Gut microbiome community composition is useful as a resource for machine
-learning prediction of diseases, including colorectal cancer
-[@baxter2016; @zackular2014]. Amplicon sequencing of the 16S rRNA gene
-is a reliable tool for assessing the taxonomic composition of microbial
-communities. Analysis of 16S rRNA amplicon sequence (16S) data generally
-relies on clustering of sequences based on similarity into operational
-taxonomic units (OTUs). The process of OTU clustering can either be
-reference-based or *de novo*. The quality of OTUs generated with
-reference-based clustering is dependent on the quality of the reference
-used, while OTUs generated with *de novo* clustering are optimized to
-the data set @westcott2015. While *de novo* clustering can produce
-high-quality OTU clusters where sequences are accurately grouped based
-on similarity thresholds, the resulting OTU clusters depend on the data
-in the data set and the addition of new data could change the overall
-OTU clusters. The inconsistent nature of OTU clustering complicates
-deployment of machine learning models since integration of additional
-data requires re-clustering all the data and re-training of the model.
-The ability to integrate new data into an existing model without
-re-clustering and re-training could allow for deployment of a single
-model that new data can be continually added to. Recently Sovacool *et
-al* introduced OptiFit: a method for fitting new data into existing OTU
-clusters @sovacool2022. While OptiFit is proven to effectively fit new
-sequence data to existing OTU clusters, it is unknown if the use of
-OptiFit will have an impact on machine learning classification. Here we
-test the ability of OptiFit to cluster new sequence data into existing
-OTU clusters for the purpose of machine learning classification of
-disease based on gut microbiome composition.
+Gut community composition is useful as a resource for machine learning
+classification of diseases, including colorectal cancer [@baxter2016;
+@zackular2014]. Amplicon sequencing of the 16S rRNA gene is a reliable
+tool for assessing the taxonomic composition of microbial communities,
+which is the input to these models. Analysis of 16S rRNA gene sequence
+data generally relies on clustering of sequences based on similarity
+into operational taxonomic units (OTUs). The process of OTU clustering
+can either be reference-based or *de novo*. The quality of OTUs
+generated with reference-based clustering is generally poor compared to
+those generated with *de novo* clustering @westcott2015. While *de novo*
+clustering produces high-quality OTU clusters where sequences are
+accurately grouped based on similarity thresholds, the resulting OTU
+clusters depend on the data in the data set and the addition of new data
+could change the overall OTU clusters. The unstable nature of OTU
+clustering complicates deployment of machine learning models since
+integration of additional data requires reclustering all the data and
+retraining of the model. The ability to integrate new data into a
+validated model without reclustering and retraining could allow for
+deployment of a single model that new data can be continually added to.
+Recently Sovacool *et al* introduced OptiFit: a method for fitting new
+sequence data into existing OTUs @sovacool2022. While OptiFit is proven
+to effectively fit new sequence data to existing OTU clusters, it is
+unknown if the use of OptiFit will have an impact on classification.
+Here we tested the ability of OptiFit to cluster new sequence data into
+existing OTU clusters for the purpose of classification of disease based
+on gut microbiome composition.
 
-When using OTU abundances for machine learning classification, the
-current methodology is to *de novo* cluster all of the sequence data
-into OTUs with the OptiClust algorithm in mothur @westcott2017. The
-resulting abundance data is then split into training and testing sets,
-where the training set is used to tune hyperparameters and ultimately
-train the model. The testing set is then classified with the model and
-the performance of the model can be quantified (Figure 1A). However,
-with this methodology we would have to re-generate the OTU clusters and
-re-train the model if we wanted to classify additional samples. The
-OptiFit algorithm @sovacool2022 addresses this problem by enabling new
-sequences to be clustered into existing OTUs. The OptiFit workflow is
-similar to the OptiClust workflow where the data is clustered into OTUs
-and used to tune hyperparameters and ultimately train the model. Then,
-we can use OptiFit to fit sequence data of samples not part of the
-original data set into the existing OTUs and use the same model to
-classify the samples (Figure 1B). To test how the model performance
-compares between these two methodologies, we used a publicly available
-data set of 16S sequences from stool samples of healthy subjects as well
-as subjects with SRN consisting of advanced adenoma and carcinoma
-@baxter2016. The data set was randomly split into an 80% train set and
-20% test set. For the standard OptiClust workflow, the training and test
-sets were *de novo* clustered together into OTUs then the resulting
-abundance table was split into the training and testing set. For the
-OptiFit workflow, the train set was clustered *de novo* into OTUs and
-the remaining test set was fit to the OTU clusters using the OptiFit
-algorithm. For both workflows, the abundance table of the train set was
-used to tune hyperparameters and train a random forest model to classify
-SRN. The test set was classified as either control or SRN using the
-trained models.To account for variation depending on the split of the
-data, the data set was randomly split 100 times and the process repeated
-for each of the 100 data splits. By comparing the model performance of
-classifying the samples in the test data set between the OptiFit and
-OptiClust algorithms, we can quantify the impact of using OptiFit on
-model classification performance.
+We compared two approaches, one using all of the data to generate OTU
+clusters and the other generating OTU clusters with a portion of the
+data and then fitting the remaining sequence data to the existing OTUs
+using OptiFit. In the first approach, all of the 16S rRNA sequence data
+was *de novo* clustered into OTUs with the OptiClust algorithm in mothur
+@westcott2017. The resulting abundance data was then split into training
+and testing sets, where the training set was used to tune
+hyperparameters and ultimately train the model. The testing set was then
+classified with the model and the performance of the model was
+quantified (Figure 1A). However, with this methodology we would have to
+regenerate the OTU clusters and retrain the model if we wanted to
+classify additional samples. The OptiFit algorithm @sovacool2022
+addresses this problem by enabling new sequences to be clustered into
+existing OTUs. The OptiFit workflow is similar to the OptiClust workflow
+where the data was clustered into OTUs and used to tune hyperparameters
+and ultimately train the model. Then, we used OptiFit to fit sequence
+data of samples not part of the original data set into the existing OTUs
+and used the same model to classify the samples (Figure 1B). To test how
+the model performance compares between these two methodologies, we used
+a publicly available data set of 16S rRNA gene sequences from stool
+samples of healthy subjects as well as subjects with SRN consisting of
+advanced adenoma and carcinoma @baxter2016. The data set was randomly
+split into an 80% train set and 20% test set. For the standard OptiClust
+workflow, the training and test sets were *de novo* clustered together
+into OTUs then the resulting abundance table was split into the training
+and testing set. For the OptiFit workflow, the train set was clustered
+*de novo* into OTUs and the remaining test set was fit to the OTU
+clusters using the OptiFit algorithm. For both workflows, the abundance
+table of the train set was used to tune hyperparameters and train a
+random forest model to classify SRN. The test set was classified as
+either control or SRN using the trained models.To account for variation
+depending on the split of the data, the data set was randomly split 100
+times and the process repeated for each of the 100 data splits. By
+comparing the model performance of classifying the samples in the test
+data set between the OptiFit and OptiClust algorithms, we quantified the
+impact of using OptiFit on model classification performance.
 
@@
 algorithms using the Matthews correlation coefficient (MCC). The MCC
-score is quantified by examining all pairs of sequences and assessing
-whether they belong together in an OTU or not based on sequence
-similarity @westcott2017. MCC scores range between negative one and one.
-A score of negative one means none of the sequences in an OTU are within
-the similarity threshold and any sequences within the similarity
-threshold are not in an OTU together. An MCC score of zero essentially
-means the sequences are randomly clustered. An MCC score of 1 means all
-sequences in an OTU are within the similarity threshold and all sequence
-pairs within the similarity threshold are in the same OTU. To ensure
-that OptiFit is appropriately integrating new sequence data into the
-existing OTUs, we would expect that the MCC scores produced by the
-OptiClust and OptiFit workflows are similar. Since the data is only
-clustered once in the OptiClust workflow there is only one MCC score
-while the OptiFit workflow produces an MCC score for the OTU clusters
-from each data split. Overall the MCC scores were similar between
-OptiClust (MCC = `r round(opticlust_mcc,digits=3)`) and OptiFit (average
-MCC = `r round(optifit_avg_mcc,digits=3)`) indicating that OptiFit
-performs as well as OptiClust when integrating new sequences into the
+score was quantified by examining all pairs of sequences and assessing
+whether they belonged together in an OTU based on their similarity
+@westcott2017. MCC scores range between negative one and one. A score of
+negative one means none of the sequences in an OTU are within the
+similarity threshold and any sequences within the similarity threshold
+are not in an OTU together. An MCC score of zero essentially means the
+sequences are randomly clustered. An MCC score of 1 means all sequences
+in an OTU are within the similarity threshold and all sequence pairs
+within the similarity threshold are in the same OTU. To ensure that
+OptiFit is appropriately integrating new sequence data into the existing
+OTUs, we expected the MCC scores produced by the OptiClust and OptiFit
+workflows to be similar. Since the data was only clustered once in the
+OptiClust workflow there was only one MCC score while the OptiFit
+workflow produced an MCC score for the OTU clusters from each data
+split. Overall the MCC scores were similar between OptiClust (MCC =
+`r round(opticlust_mcc,digits=3)`) and OptiFit (average MCC =
+`r round(optifit_avg_mcc,digits=3)`). This indicated that OptiFit
+performed as well as OptiClust when integrating new sequences into the
 existing OTUs.
@@
 After verifying that the quality of the OTUs was consistent between
-OptiClust and OptiFit, we next examined the model performance for
-classifying samples in the test data set as control or SRN. To quantify
-model performance we used the taxonomic abundances of the training data
-from the OptiClust and OptiFit workflows to train a model to predict
-SRNs. Using the predicted and actual diagnosis classification, we
-calculated the area under the receiver operating characteristic curve
-(AUROC) for each data split to quantify model performance. During
-cross-validation (CV) training, the model performance was equivalent
-between the two algorithms (p-value =
+OptiClust and OptiFit, we examined the model performance for classifying
+samples in the held out test data set. To quantify model performance we
+used the OTU relative abundances from the training data from the
+OptiClust and OptiFit workflows to train a model to predict SRNs. Using
+the predicted and actual diagnosis classification, we calculated the
+area under the receiver operating characteristic curve (AUROC) for each
+data split to quantify model performance. During cross-validation (CV)
+training, the model performance was equivalent between the two
+algorithms (p-value =
 `r format_decimal(pvals %>% filter(metric == "cv_metric_AUC") %>% select(p_value),2)`,
@@
 `r format_decimal(performance_summary %>% filter(algorithm == "optifit") %>% select(mean_AUC),3)`,
-Figure 2B,C) indicating that new data can be fit to existing OTU
+Figure 2B,C) indicating that new data could be fit to existing OTU
 clusters without impacting model performance.
@@
 We tested the ability of OptiFit to integrate new data into existing
-OTUs for the purpose of machine learning classification of disease based
-on microbiome abundance. A potential problem to using OptiFit for
-machine learning prediction is that any sequences in the new data that
-do not map to the existing OTU clusters will be discarded resulting in a
-possible loss of information. However, we demonstrated that OptiFit can
-be used to fit new sequence data into existing OTU clusters and perform
-equally in predicting SRN compared to clustering all of the sequence
-data together. The ability to integrate new data into existing OTUs
-enables the deployment of a single machine learning model based on
-microbiome composition that new data can be classified with. These
-results are based on a single data set and disease. Further analysis is
-needed to determine the data size necessary to build a robust model
-capable of classifying diverse data sets. A robust machine learning
+OTUs for the purpose of machine learning classification using OTU
+relative abundance. A potential problem with using OptiFit is that any
+sequences in the new data that do not map to the existing OTU clusters
+will be discarded resulting in a possible loss of information. However,
+we demonstrated that OptiFit can be used to fit new sequence data into
+existing OTU clusters and perform equally well in predicting SRN
+compared to clustering all of the sequence data together. The ability to
+integrate data from new samples into existing OTUs enables the
+deployment of a single machine learning model. These results are based
+on a single data set and disease. Further analysis is needed to
+determine the number of samples that are necessary to build a robust
+model capable of classifying diverse samples. A robust machine learning
 model could be implemented as part of a non-invasive and low-cost aid in
-diagnosing SRN.
+diagnosing SRN and other diseases.
 
@@
 
-***Data Set.*** Raw 16S rRNA amplicon sequence data isolated from human
+***Data Set.*** Raw 16S rRNA gene sequence data isolated from human
 stool samples was downloaded from NCBI Sequence Read Archive (accession
-no. SRP062005) @baxter2016.; @edgar2011 This data set contains stool
-samples from a total of N subjects, however after preprocessing to
-screen for sequence quality and subsample to 10,000 reads per sample,
-490 samples remained. For this analysis, samples from subjects
-identified in the metadata as normal, high risk normal, or adenoma were
-categorized as "normal" while samples from subjects identified as
-advanced adenoma or carcinoma were categorized as "screen relevant
-neoplasia" (SRN). The resulting data set consisted of 261 normal samples
-and 229 SRN samples.
+no. SRP062005) [@edgar2011; @baxter2016]. This data set contains stool
+samples from a total of 490 subjects. For this analysis, samples from
+subjects identified in the metadata as normal, high risk normal, or
+adenoma were categorized as "normal" while samples from subjects
+identified as advanced adenoma or carcinoma were categorized as "screen
+relevant neoplasia" (SRN). The resulting data set consisted of 261
+normal samples and 229 SRN samples.
 
@@
 
-***Data Processing.*** The full dataset was pre-processed with mothur
-(v1.45) @schloss2009 using the SILVA reference database (v132)
-@quast2013 to join forward and reverse reads, merge duplicate reads,
-align to the reference, pre-cluster, remove chimeras with UCHIME
-@edgar2011, assign taxonomy, and remove non-bacterial reads following
-the Schloss Lab MiSeq standard operating procedure described on the
-mothur website (<https://mothur.org/wiki/miseq_sop/>). 100 splits of the
-490 samples were generated where 80% of the samples (392 samples) were
-randomly assigned to the training set and the remaining 20% (98 samples)
-were assigned to the test set. Using 100 splits of the data accounts for
-the variation that may be observed depending on the samples that are in
-the training or test sets. Each samples was in the training set an
-average of `r n_test_train %>% pull(avg_train)` times
+***Data Processing.*** The full dataset was preprocessed with mothur
+(v1.47) @schloss2009 to join forward and reverse reads, merge duplicate
+reads, align to the SILVA reference database (v132) @quast2013,
+precluster, remove chimeras with UCHIME @edgar2011, assign taxonomy, and
+remove non-bacterial reads following the Schloss Lab MiSeq standard
+operating procedure described on the mothur website
+(<https://mothur.org/wiki/miseq_sop/>). 100 splits of the 490 samples
+were generated where 80% of the samples (392 samples) were randomly
+assigned to the training set and the remaining 20% (98 samples) were
+assigned to the test set. Using 100 splits of the data accounts for the
+variation that may be observed depending on the samples that are in the
+training or test sets. Each sample was in the training set an average of
+`r n_test_train %>% pull(avg_train)` times
 (SD=`r format_decimal(n_test_train %>% pull(sd_train),digits = 1)`) and
@@
 and the resulting abundance tables were split into the training and
-testing sets. In the second workflow, the pre-processed data was split
+testing sets. In the second workflow, the preprocessed data was split
 into the training and testing sets. The training set was clustered into
@@
 the OptiFit algorithm @sovacool2022. The OptiFit algorithm was run with
-method open so that any sequences that didn't map to the existing OTU
-clusters would form new OTUs. Any OTUs that were not in the training set
-were removed prior to machine learning. For both pathways, the shared
-files were sub-sampled to 10,000 reads per sample.
+method open so that any sequences that did not map to the existing OTU
+clusters would form new OTUs. For both pathways, the shared files were
+sub-sampled to 10,000 reads per sample.
 
 ***Machine Learning.*** Machine learning using Random Forest was
-conducted with the R package mikrompl (v XXXX) @topçuoglu2021 to predict
-the diagnosis (SRN or normal) for the samples in the test set for each
-data split. The training set was preprocessed to normalize values
-(scale/center), collapse correlated features, and remove features with
+conducted with the R package mikrompl (v 1.2.0) @topçuoglu2021 to
+predict the diagnosis (SRN or normal) for the samples in the test set
+for each data split. The training set was preprocessed to normalize OTU
+counts (scale/center), collapse correlated OTUs, and remove OTUs with
 zero-variance. The preprocessing from the training set was then applied
-to the test set. P values comparing model performance were calculated as
-previously described {}. The averaged ROC curves were plotted by taking
-the average and standard deviation of the sensitivity at each
-specificity value.
+to the test set. Any OTUs in the test set that were not in the training
+set were removed. P values comparing model performance were calculated
+as previously described @topçuoglu2020. The averaged ROC curves were
+plotted by taking the average and standard deviation of the sensitivity
+at each specificity value.
 
@@
 
-(funding)
+This work was supported through a grant from the NIH (R01CA215574).
 

### 2022-06-20 submission/manuscript.tex
revised ms

@@
 
-Other Title Options:\\
-A solution to the problem of inconsistent OTU clustering for machine
-learning classification
-
-Optimized reference-based OTU clustering for machine learning
-classification
-
 \vspace{5mm}
 
-Running title: INSERT RUNNING TITLE HERE
+Running title: Reference-based OTU clustering for ML classification
 
-\vspace{5mm}
+\vspace{10mm}
 
@@
 
-\hypertarget{abstract-250-word-max}{%
-\subsection{Abstract (250 word max)}\label{abstract-250-word-max}}
+\hypertarget{abstract}{%
+\subsection{Abstract}\label{abstract}}
 
 Machine learning classification of disease based on the gut microbiome
-often relies on operational taxonomic units (OTUs) to quantify microbial
-composition. The standard methodology for quantifying taxonomic
-composition is clustering 16S rRNA amplicon sequences \emph{de novo}
-into OTUs and using their abundances to train a classification model.
-However, OTU clusters depend on the sequences in the data set and
-therefore can change if new data is added. This inconsistency
-complicates machine learning classification because in order to use the
-model to classify additional samples, the OTUs have to be re-clustered
-to include the new sequences and the model retrained with the new OTU
-clusters. A new algorithm in mothur, called OptiFit, addresses this
-issue by clustering new sequences into the existing OTU clusters. While
-OptiFit is proven to produce high quality OTU clusters, it is unclear
-whether this method for clustering new sequence data into existing OTUs
-will have any impact on machine learning classification using the OTU
-abundances. To evaluate the potential for use in machine learning
-classification, we used OptiFit to cluster additional data into existing
-OTU clusters and quantified model performance in classifying a data set
-containing samples from controls and people with advanced adenoma and
-carcinoma (SRN). We compared the model performance using OptiFit to the
-standard procedure of clustering all the data together. We found that
-OptiFit can be used to fit new sequence data to existing OTUs and
-perform equally in classifying SRNs. Moving forward, when OTUs are used
-in machine learning classification problems, OptiFit can be used to
-avoid the need to re-cluster sequences into OTUs when classifying
-additional samples that were not part of the data set used to build the
-model.
+often relies on clustering 16S rRNA gene sequences into operational
+taxonomic units (OTUs) to quantify microbial composition. The standard
+approach to clustering sequences into OTUs leverages the similarity of
+the sequences to each other rather than to a reference database. The
+abundance of each OTU is used to train a classification model. However,
+OTU assignments depend on the sequences in the data set and therefore
+can change if new data are added. This lack of stability complicates
+classification because in order to use the model to classify additional
+samples, the OTUs have to be reclustered to include the new sequences
+and the model retrained with the new OTU clusters. A new reference-based
+clustering algorithm, called OptiFit, addresses this issue by fitting
+new sequences into existing OTUs. While OptiFit is proven to produce
+high quality OTU clusters, it is unclear whether this method for fitting
+new sequence data into existing OTUs will impact the performance of
+classification models. We used OptiFit to cluster additional data into
+existing OTU clusters and quantified model performance in classifying a
+data set containing samples from patients with and without colonic
+screen relevant neoplasias (SRN). We compared the performance of this
+model to the standard procedure of clustering all the data together. We
+found that both approaches performed equally well in classifying SRNs.
+Moving forward, when OTUs are used in classification problems, OptiFit
+can be used to avoid the need to retrain models using reclustered
+sequences when classifying new samples.
 
-\hypertarget{importance-150-word-max}{%
-\subsection{Importance (150 word max)}\label{importance-150-word-max}}
+\hypertarget{importance}{%
+\subsection{Importance}\label{importance}}
 
-There are instances when OTU abundance data is optimal for machine
-learning prediction of disease based on microbiome composition. The
-current methodology for OTU clustering complicates machine learning
-since \emph{de novo} OTU clusters can change depending on the data in
-the data set. Attempting to classify additional samples that were not
-part of the original model build requires re-clustering OTUs and
-re-training the model. OptiFit provides an elegant solution to the
-problem of OTU inconsistency by enabling new data to be fit to the
-existing OTUs while maintaining machine learning classification
-performance. The use of OptiFit will enable a single model based on OTU
-abundance data to be used in classifying additional samples that were
-not part of the original OTU clustering.
+There is great potential for using microbiome data to non-invasively
+diagnose people. One of the challenges with using classification models
+generated using the relative abundance of operational taxonomic unit
+(OTU) data is that 16S rRNA gene sequences are assigned to OTUs based on
+their similarity to other sequences in the dataset. If data are
+generated from new patients seeking a diagnosis, then it would be
+necessary to reassign sequences to OTUs and retrain the classification
+model. Yet there is a desire to have a single, validated model that can
+be deployed. To overcome this obstacle, we applied the OptiFit
+clustering algorithm which fits new sequence data to existing OTUs
+allowing the reuse of a consistent model. A random forest machine
+learning model deployed using OptiFit performed as well as the
+traditional reassignment and retraining approach. This result indicates
+there is potential for developing machine learning classification models
+based on OTU relative abundance data.
 
@@
 
-Gut microbiome community composition is useful as a resource for machine
-learning prediction of diseases, including colorectal cancer
+Gut community composition is useful as a resource for machine learning
+classification of diseases, including colorectal cancer
 (\protect\hyperlink{ref-baxter2016}{1},
@@
 rRNA gene is a reliable tool for assessing the taxonomic composition of
-microbial communities. Analysis of 16S rRNA amplicon sequence (16S) data
-generally relies on clustering of sequences based on similarity into
-operational taxonomic units (OTUs). The process of OTU clustering can
-either be reference-based or \emph{de novo}. The quality of OTUs
-generated with reference-based clustering is dependent on the quality of
-the reference used, while OTUs generated with \emph{de novo} clustering
-are optimized to the data set (\protect\hyperlink{ref-westcott2015}{3}).
-While \emph{de novo} clustering can produce high-quality OTU clusters
-where sequences are accurately grouped based on similarity thresholds,
-the resulting OTU clusters depend on the data in the data set and the
-addition of new data could change the overall OTU clusters. The
-inconsistent nature of OTU clustering complicates deployment of machine
-learning models since integration of additional data requires
-re-clustering all the data and re-training of the model. The ability to
-integrate new data into an existing model without re-clustering and
-re-training could allow for deployment of a single model that new data
-can be continually added to. Recently Sovacool \emph{et al} introduced
-OptiFit: a method for fitting new data into existing OTU clusters
+microbial communities, which is the input to these models. Analysis of
+16S rRNA gene sequence data generally relies on clustering of sequences
+based on similarity into operational taxonomic units (OTUs). The process
+of OTU clustering can either be reference-based or \emph{de novo}. The
+quality of OTUs generated with reference-based clustering is generally
+poor compared to those generated with \emph{de novo} clustering
+(\protect\hyperlink{ref-westcott2015}{3}). While \emph{de novo}
+clustering produces high-quality OTU clusters where sequences are
+accurately grouped based on similarity thresholds, the resulting OTU
+clusters depend on the data in the data set and the addition of new data
+could change the overall OTU clusters. The unstable nature of OTU
+clustering complicates deployment of machine learning models since
+integration of additional data requires reclustering all the data and
+retraining of the model. The ability to integrate new data into a
+validated model without reclustering and retraining could allow for
+deployment of a single model that new data can be continually added to.
+Recently Sovacool \emph{et al} introduced OptiFit: a method for fitting
+new sequence data into existing OTUs
 (\protect\hyperlink{ref-sovacool2022}{4}). While OptiFit is proven to
 effectively fit new sequence data to existing OTU clusters, it is
-unknown if the use of OptiFit will have an impact on machine learning
-classification. Here we test the ability of OptiFit to cluster new
-sequence data into existing OTU clusters for the purpose of machine
-learning classification of disease based on gut microbiome composition.
+unknown if the use of OptiFit will have an impact on classification.
+Here we tested the ability of OptiFit to cluster new sequence data into
+existing OTU clusters for the purpose of classification of disease based
+on gut microbiome composition.
 
-When using OTU abundances for machine learning classification, the
-current methodology is to \emph{de novo} cluster all of the sequence
-data into OTUs with the OptiClust algorithm in mothur
-(\protect\hyperlink{ref-westcott2017}{5}). The resulting abundance data
-is then split into training and testing sets, where the training set is
-used to tune hyperparameters and ultimately train the model. The testing
-set is then classified with the model and the performance of the model
-can be quantified (Figure 1A). However, with this methodology we would
-have to re-generate the OTU clusters and re-train the model if we wanted
-to classify additional samples. The OptiFit algorithm
+We compared two approaches, one using all of the data to generate OTU
+clusters and the other generating OTU clusters with a portion of the
+data and then fitting the remaining sequence data to the existing OTUs
+using OptiFit. In the first approach, all of the 16S rRNA sequence data
+was \emph{de novo} clustered into OTUs with the OptiClust algorithm in
+mothur (\protect\hyperlink{ref-westcott2017}{5}). The resulting
+abundance data was then split into training and testing sets, where the
+training set was used to tune hyperparameters and ultimately train the
+model. The testing set was then classified with the model and the
+performance of the model was quantified (Figure 1A). However, with this
+methodology we would have to regenerate the OTU clusters and retrain the
+model if we wanted to classify additional samples. The OptiFit algorithm
 (\protect\hyperlink{ref-sovacool2022}{4}) addresses this problem by
 enabling new sequences to be clustered into existing OTUs. The OptiFit
-workflow is similar to the OptiClust workflow where the data is
+workflow is similar to the OptiClust workflow where the data was
 clustered into OTUs and used to tune hyperparameters and ultimately
-train the model. Then, we can use OptiFit to fit sequence data of
-samples not part of the original data set into the existing OTUs and use
-the same model to classify the samples (Figure 1B). To test how the
-model performance compares between these two methodologies, we used a
-publicly available data set of 16S sequences from stool samples of
+train the model. Then, we used OptiFit to fit sequence data of samples
+not part of the original data set into the existing OTUs and used the
+same model to classify the samples (Figure 1B). To test how the model
+performance compares between these two methodologies, we used a publicly
+available data set of 16S rRNA gene sequences from stool samples of
 healthy subjects as well as subjects with SRN consisting of advanced
@@
 classifying the samples in the test data set between the OptiFit and
-OptiClust algorithms, we can quantify the impact of using OptiFit on
-model classification performance.
+OptiClust algorithms, we quantified the impact of using OptiFit on model
+classification performance.
 
@@
 algorithms using the Matthews correlation coefficient (MCC). The MCC
-score is quantified by examining all pairs of sequences and assessing
-whether they belong together in an OTU or not based on sequence
-similarity (\protect\hyperlink{ref-westcott2017}{5}). MCC scores range
-between negative one and one. A score of negative one means none of the
+score was quantified by examining all pairs of sequences and assessing
+whether they belonged together in an OTU based on their similarity
+(\protect\hyperlink{ref-westcott2017}{5}). MCC scores range between
+negative one and one. A score of negative one means none of the
 sequences in an OTU are within the similarity threshold and any
@@
 threshold are in the same OTU. To ensure that OptiFit is appropriately
-integrating new sequence data into the existing OTUs, we would expect
-that the MCC scores produced by the OptiClust and OptiFit workflows are
-similar. Since the data is only clustered once in the OptiClust workflow
-there is only one MCC score while the OptiFit workflow produces an MCC
-score for the OTU clusters from each data split. Overall the MCC scores
-were similar between OptiClust (MCC = 0.884) and OptiFit (average MCC =
-0.879) indicating that OptiFit performs as well as OptiClust when
-integrating new sequences into the existing OTUs.
+integrating new sequence data into the existing OTUs, we expected the
+MCC scores produced by the OptiClust and OptiFit workflows to be
+similar. Since the data was only clustered once in the OptiClust
+workflow there was only one MCC score while the OptiFit workflow
+produced an MCC score for the OTU clusters from each data split. Overall
+the MCC scores were similar between OptiClust (MCC = 0.884) and OptiFit
+(average MCC = 0.879). This indicated that OptiFit performed as well as
+OptiClust when integrating new sequences into the existing OTUs.
 
 After verifying that the quality of the OTUs was consistent between
-OptiClust and OptiFit, we next examined the model performance for
-classifying samples in the test data set as control or SRN. To quantify
-model performance we used the taxonomic abundances of the training data
-from the OptiClust and OptiFit workflows to train a model to predict
-SRNs. Using the predicted and actual diagnosis classification, we
-calculated the area under the receiver operating characteristic curve
-(AUROC) for each data split to quantify model performance. During
-cross-validation (CV) training, the model performance was equivalent
-between the two algorithms (p-value = 0.13, OptiClust mean CV AUROC =
-0.694, OptiFit mean CV AUROC = 0.697, Figure 2A). The trained model was
-then deployed to classify the samples of the test data as control or
-SRN. The performance on the test data was equivalent between the two
-algorithms (p-value = 0.63, OptiClust mean test AUROC = 0.709, OptiFit
-mean test AUROC = 0.712, Figure 2B,C) indicating that new data can be
-fit to existing OTU clusters without impacting model performance.
+OptiClust and OptiFit, we examined the model performance for classifying
+samples in the held out test data set. To quantify model performance we
+used the OTU relative abundances from the training data from the
+OptiClust and OptiFit workflows to train a model to predict SRNs. Using
+the predicted and actual diagnosis classification, we calculated the
+area under the receiver operating characteristic curve (AUROC) for each
+data split to quantify model performance. During cross-validation (CV)
+training, the model performance was equivalent between the two
+algorithms (p-value = 0.13, OptiClust mean CV AUROC = 0.694, OptiFit
+mean CV AUROC = 0.697, Figure 2A). The trained model was then deployed
+to classify the samples of the test data as control or SRN. The
+performance on the test data was equivalent between the two algorithms
+(p-value = 0.63, OptiClust mean test AUROC = 0.709, OptiFit mean test
+AUROC = 0.712, Figure 2B,C) indicating that new data could be fit to
+existing OTU clusters without impacting model performance.
 
 We tested the ability of OptiFit to integrate new data into existing
-OTUs for the purpose of machine learning classification of disease based
-on microbiome abundance. A potential problem to using OptiFit for
-machine learning prediction is that any sequences in the new data that
-do not map to the existing OTU clusters will be discarded resulting in a
-possible loss of information. However, we demonstrated that OptiFit can
-be used to fit new sequence data into existing OTU clusters and perform
-equally in predicting SRN compared to clustering all of the sequence
-data together. The ability to integrate new data into existing OTUs
-enables the deployment of a single machine learning model based on
-microbiome composition that new data can be classified with. These
-results are based on a single data set and disease. Further analysis is
-needed to determine the data size necessary to build a robust model
-capable of classifying diverse data sets. A robust machine learning
+OTUs for the purpose of machine learning classification using OTU
+relative abundance. A potential problem with using OptiFit is that any
+sequences in the new data that do not map to the existing OTU clusters
+will be discarded resulting in a possible loss of information. However,
+we demonstrated that OptiFit can be used to fit new sequence data into
+existing OTU clusters and perform equally well in predicting SRN
+compared to clustering all of the sequence data together. The ability to
+integrate data from new samples into existing OTUs enables the
+deployment of a single machine learning model. These results are based
+on a single data set and disease. Further analysis is needed to
+determine the number of samples that are necessary to build a robust
+model capable of classifying diverse samples. A robust machine learning
 model could be implemented as part of a non-invasive and low-cost aid in
-diagnosing SRN.
+diagnosing SRN and other diseases.
 
@@
 
-\textbf{\emph{Data Set.}} Raw 16S rRNA amplicon sequence data isolated
-from human stool samples was downloaded from NCBI Sequence Read Archive
-(accession no. SRP062005) (\protect\hyperlink{ref-baxter2016}{1}).;
-(\protect\hyperlink{ref-edgar2011}{6}) This data set contains stool
-samples from a total of N subjects, however after preprocessing to
-screen for sequence quality and subsample to 10,000 reads per sample,
-490 samples remained. For this analysis, samples from subjects
-identified in the metadata as normal, high risk normal, or adenoma were
-categorized as ``normal'' while samples from subjects identified as
-advanced adenoma or carcinoma were categorized as ``screen relevant
-neoplasia'' (SRN). The resulting data set consisted of 261 normal
-samples and 229 SRN samples.
+\textbf{\emph{Data Set.}} Raw 16S rRNA gene sequence data isolated from
+human stool samples was downloaded from NCBI Sequence Read Archive
+(accession no. SRP062005) (\protect\hyperlink{ref-baxter2016}{1},
+\protect\hyperlink{ref-edgar2011}{6}). This data set contains stool
+samples from a total of 490 subjects. For this analysis, samples from
+subjects identified in the metadata as normal, high risk normal, or
+adenoma were categorized as ``normal'' while samples from subjects
+identified as advanced adenoma or carcinoma were categorized as ``screen
+relevant neoplasia'' (SRN). The resulting data set consisted of 261
+normal samples and 229 SRN samples.
 
-\textbf{\emph{Data Processing.}} The full dataset was pre-processed with
-mothur (v1.45) (\protect\hyperlink{ref-schloss2009}{7}) using the SILVA
-reference database (v132) (\protect\hyperlink{ref-quast2013}{8}) to join
-forward and reverse reads, merge duplicate reads, align to the
-reference, pre-cluster, remove chimeras with UCHIME
-(\protect\hyperlink{ref-edgar2011}{6}), assign taxonomy, and remove
-non-bacterial reads following the Schloss Lab MiSeq standard operating
-procedure described on the mothur website
+\textbf{\emph{Data Processing.}} The full dataset was preprocessed with
+mothur (v1.47) (\protect\hyperlink{ref-schloss2009}{7}) to join forward
+and reverse reads, merge duplicate reads, align to the SILVA reference
+database (v132) (\protect\hyperlink{ref-quast2013}{8}), precluster,
+remove chimeras with UCHIME (\protect\hyperlink{ref-edgar2011}{6}),
+assign taxonomy, and remove non-bacterial reads following the Schloss
+Lab MiSeq standard operating procedure described on the mothur website
 (\url{https://mothur.org/wiki/miseq_sop/}). 100 splits of the 490
@@
 accounts for the variation that may be observed depending on the samples
-that are in the training or test sets. Each samples was in the training
+that are in the training or test sets. Each sample was in the training
 set an average of 80 times (SD=4.1) and the test set an average of 20
@@
 resulting abundance tables were split into the training and testing
-sets. In the second workflow, the pre-processed data was split into the
+sets. In the second workflow, the preprocessed data was split into the
 training and testing sets. The training set was clustered into OTUs,
@@
 OptiFit algorithm (\protect\hyperlink{ref-sovacool2022}{4}). The OptiFit
-algorithm was run with method open so that any sequences that didn't map
-to the existing OTU clusters would form new OTUs. Any OTUs that were not
-in the training set were removed prior to machine learning. For both
-pathways, the shared files were sub-sampled to 10,000 reads per sample.
+algorithm was run with method open so that any sequences that did not
+map to the existing OTU clusters would form new OTUs. For both pathways,
+the shared files were sub-sampled to 10,000 reads per sample.
 
 \textbf{\emph{Machine Learning.}} Machine learning using Random Forest
-was conducted with the R package mikrompl (v XXXX)
+was conducted with the R package mikrompl (v 1.2.0)
 (\protect\hyperlink{ref-topuxe7uoglu2021}{9}) to predict the diagnosis
 (SRN or normal) for the samples in the test set for each data split. The
-training set was preprocessed to normalize values (scale/center),
-collapse correlated features, and remove features with zero-variance.
-The preprocessing from the training set was then applied to the test
-set. P values comparing model performance were calculated as previously
-described \{\}. The averaged ROC curves were plotted by taking the
-average and standard deviation of the sensitivity at each specificity
-value.
+training set was preprocessed to normalize OTU counts (scale/center),
+collapse correlated OTUs, and remove OTUs with zero-variance. The
+preprocessing from the training set was then applied to the test set.
+Any OTUs in the test set that were not in the training set were removed.
+P values comparing model performance were calculated as previously
+described (\protect\hyperlink{ref-topuxe7uoglu2020}{10}). The averaged
+ROC curves were plotted by taking the average and standard deviation of
+the sensitivity at each specificity value.
 
 \textbf{\emph{Code Availability.}} The analysis workflow was implemented
-in Snakemake (\protect\hyperlink{ref-koster2012}{10}) . Scripts for
-analysis were written in R (\protect\hyperlink{ref-R2020}{11}) and GNU
-bash (\protect\hyperlink{ref-GNUbash}{12}). The software used includes
+in Snakemake (\protect\hyperlink{ref-koster2012}{11}) . Scripts for
+analysis were written in R (\protect\hyperlink{ref-R2020}{12}) and GNU
+bash (\protect\hyperlink{ref-GNUbash}{13}). The software used includes
 mothur v1.47.0 (\protect\hyperlink{ref-schloss2009}{7}), RStudio
-(\protect\hyperlink{ref-RStudio2019}{13}), the Tidyverse metapackage
-(\protect\hyperlink{ref-wickham2019}{14}), R Markdown
-(\protect\hyperlink{ref-xie_r_2018}{15}), the SRA toolkit
-(\protect\hyperlink{ref-noauthor_sra-tools_nodate}{16}), and conda
-(\protect\hyperlink{ref-noauthor_anaconda_2016}{17}). The complete
+(\protect\hyperlink{ref-RStudio2019}{14}), the Tidyverse metapackage
+(\protect\hyperlink{ref-wickham2019}{15}), R Markdown
+(\protect\hyperlink{ref-xie_r_2018}{16}), the SRA toolkit
+(\protect\hyperlink{ref-noauthor_sra-tools_nodate}{17}), and conda
+(\protect\hyperlink{ref-noauthor_anaconda_2016}{18}). The complete
 workflow and supporting files required to reproduce this study are
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-topuxe7uoglu2020}{}}%
+\CSLLeftMargin{10. }
+\CSLRightInline{\textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
+\textbf{Ruffin MT}, \textbf{Wiens J}, \textbf{Schloss PD}. 2020. A
+framework for effective application of machine learning to
+microbiome-based classification problems. mBio \textbf{11}:e00434--20.
+doi:\href{https://doi.org/10.1128/mBio.00434-20}{10.1128/mBio.00434-20}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-koster2012}{}}%
-\CSLLeftMargin{10. }
+\CSLLeftMargin{11. }
 \CSLRightInline{\textbf{Koster J}, \textbf{Rahmann S}. 2012.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-R2020}{}}%
-\CSLLeftMargin{11. }
+\CSLLeftMargin{12. }
 \CSLRightInline{\textbf{R Core Team}. 2020.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-GNUbash}{}}%
-\CSLLeftMargin{12. }
+\CSLLeftMargin{13. }
 \CSLRightInline{\textbf{GNU Project}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-RStudio2019}{}}%
-\CSLLeftMargin{13. }
+\CSLLeftMargin{14. }
 \CSLRightInline{\textbf{RStudio Team}. 2019.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-wickham2019}{}}%
-\CSLLeftMargin{14. }
+\CSLLeftMargin{15. }
 \CSLRightInline{\textbf{Wickham H}, \textbf{Averick M}, \textbf{Bryan
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xie_r_2018}{}}%
-\CSLLeftMargin{15. }
+\CSLLeftMargin{16. }
 \CSLRightInline{\textbf{Xie Y}, \textbf{Allaire JJ}, \textbf{Grolemund
@@
 \leavevmode\vadjust pre{\hypertarget{ref-noauthor_sra-tools_nodate}{}}%
-\CSLLeftMargin{16. }
+\CSLLeftMargin{17. }
 \CSLRightInline{{SRA}-{Tools} - {NCBI}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-noauthor_anaconda_2016}{}}%
-\CSLLeftMargin{17. }
+\CSLLeftMargin{18. }
 \CSLRightInline{2016. Anaconda {Software Distribution}. Anaconda

### 2022-06-20 submission/manuscript.Rmd

@@
 taxonomic units (OTUs) to quantify microbial composition. The standard
-approach to clustering sequences into OTUs leverages the similarity of
+*de novo* approach to clustering sequences into OTUs leverages the similarity of
 the sequences to each other rather than to a reference database. The
@@
 OTU assignments depend on the sequences in the data set and therefore
-can change if new data are added. This lack of stability complicates
+can change if new data are added. This lack of consistency complicates
 classification because in order to use the model to classify additional
@@
 clustering algorithm, called OptiFit, addresses this issue by fitting
-new sequences into existing OTUs. While OptiFit is proven to produce
-high quality OTU clusters, it is unclear whether this method for fitting
-new sequence data into existing OTUs will impact the performance of
+new sequences into existing OTUs. While OptiFit has been shown to produce
+high quality OTU clusters, it is unknown whether this method will impact 
+the performance of
 classification models. We used OptiFit to cluster additional data into
@@
 screen relevant neoplasias (SRN). We compared the performance of this
-model to the standard procedure of clustering all the data together. We
+model to the standard procedure of *de novo* clustering all the data together. We
 found that both approaches performed equally well in classifying SRNs.
 Moving forward, when OTUs are used in classification problems, OptiFit
-can be used to avoid the need to retrain models using reclustered
-sequences when classifying new samples.
+can streamline the process of classifying new samples by avoiding the need to 
+retrain models using reclustered sequences.
 
@@
 be deployed. To overcome this obstacle, we applied the OptiFit
-clustering algorithm which fits new sequence data to existing OTUs
+clustering algorithm which fits new sequence data to existing OTUs,
 allowing the reuse of a consistent model. A random forest machine
@@
 which is the input to these models. Analysis of 16S rRNA gene sequence
-data generally relies on clustering of sequences based on similarity
+data generally relies on clustering sequences based on similarity
 into operational taxonomic units (OTUs). The process of OTU clustering
@@
 clusters depend on the data in the data set and the addition of new data
-could change the overall OTU clusters. The unstable nature of OTU
+could change the overall OTU clusters. The inconsistent nature of *de novo* OTU
 clustering complicates deployment of machine learning models since
 integration of additional data requires reclustering all the data and
-retraining of the model. The ability to integrate new data into a
+retraining the model. The ability to integrate new data into a
 validated model without reclustering and retraining could allow for
-deployment of a single model that new data can be continually added to.
-Recently Sovacool *et al* introduced OptiFit: a method for fitting new
+deployment of a single model that new data can be continually tested against.
+Recently, Sovacool *et al* introduced OptiFit: a method for fitting new
 sequence data into existing OTUs @sovacool2022. While OptiFit is proven
 to effectively fit new sequence data to existing OTU clusters, it is
-unknown if the use of OptiFit will have an impact on classification.
-Here we tested the ability of OptiFit to cluster new sequence data into
+unknown if the use of OptiFit will have an impact on classification performance.
+Here, we tested the ability of OptiFit to cluster new sequence data into
 existing OTU clusters for the purpose of classification of disease based
@@
 
-We compared two approaches, one using all of the data to generate OTU
-clusters and the other generating OTU clusters with a portion of the
+We compared two approaches, one using all of the data to generate *de novo* OTU
+clusters, and the other generating *de novo* OTU clusters with a portion of the
 data and then fitting the remaining sequence data to the existing OTUs
@@
 classified with the model and the performance of the model was
-quantified (Figure 1A). However, with this methodology we would have to
+quantified (Figure 1A). However, with this methodology, we would have to
 regenerate the OTU clusters and retrain the model if we wanted to
@@
 addresses this problem by enabling new sequences to be clustered into
-existing OTUs. The OptiFit workflow is similar to the OptiClust workflow
+existing OTUs. The OptiFit workflow is similar to the OptiClust workflow,
 where the data was clustered into OTUs and used to tune hyperparameters
 and ultimately train the model. Then, we used OptiFit to fit sequence
-data of samples not part of the original data set into the existing OTUs
+data of samples not part of the original data set into the existing OTUs,
 and used the same model to classify the samples (Figure 1B). To test how
@@
 workflow, the training and test sets were *de novo* clustered together
-into OTUs then the resulting abundance table was split into the training
+into OTUs, then the resulting abundance table was split into the training
 and testing set. For the OptiFit workflow, the train set was clustered
-*de novo* into OTUs and the remaining test set was fit to the OTU
+*de novo* into OTUs, and the remaining test set was fit to the OTU
 clusters using the OptiFit algorithm. For both workflows, the abundance
@@
 random forest model to classify SRN. The test set was classified as
-either control or SRN using the trained models.To account for variation
+either control or SRN using the trained models. To account for variation
 depending on the split of the data, the data set was randomly split 100
@@
 workflow produced an MCC score for the OTU clusters from each data
-split. Overall the MCC scores were similar between OptiClust (MCC =
+split. Overall, the MCC scores were similar between OptiClust (MCC =
 `r round(opticlust_mcc,digits=3)`) and OptiFit (average MCC =
@@
 OptiClust and OptiFit, we examined the model performance for classifying
-samples in the held out test data set. To quantify model performance we
+samples in the held out test data set. To quantify model performance, we
 used the OTU relative abundances from the training data from the
@@
 sequences in the new data that do not map to the existing OTU clusters
-will be discarded resulting in a possible loss of information. However,
+will be discarded, resulting in a possible loss of information. However,
 we demonstrated that OptiFit can be used to fit new sequence data into
-existing OTU clusters and perform equally well in predicting SRN
-compared to clustering all of the sequence data together. The ability to
+existing OTU clusters and performs equally well in predicting SRN
+compared to *de novo* clustering all of the sequence data together. The ability to
 integrate data from new samples into existing OTUs enables the
@@
 subjects identified in the metadata as normal, high risk normal, or
-adenoma were categorized as "normal" while samples from subjects
+adenoma were categorized as "normal", while samples from subjects
 identified as advanced adenoma or carcinoma were categorized as "screen
@@
 The data was processed through two workflows. First, the standard
-workflow using the OptiClust algorithm @westcott2017. In this pathway,
+workflow using the OptiClust algorithm @westcott2017. In this workflow,
 all of the data was clustered together with OptiClust to generate OTUs
@@
 OTUs, then the test set was fit to the OTUs of the training set using
-the OptiFit algorithm @sovacool2022. The OptiFit algorithm was run with
-method open so that any sequences that did not map to the existing OTU
-clusters would form new OTUs. For both pathways, the shared files were
+the OptiFit algorithm @sovacool2022. The OptiFit algorithm was run with the open
+method so that any sequences that did not map to the existing OTU
+clusters would form new OTUs. 
+<!-- Did you then remove the columns corresponding to the additional OTUs? 
+Typically you'd want to run the closed method for this use-case, right? -->
+For both workflows, the shared files were
 sub-sampled to 10,000 reads per sample.
@@
 for each data split. The training set was preprocessed to normalize OTU
-counts (scale/center), collapse correlated OTUs, and remove OTUs with
-zero-variance. The preprocessing from the training set was then applied
+counts (scale and center), collapse correlated OTUs, and remove OTUs with
+zero variance. The preprocessing from the training set was then applied
 to the test set. Any OTUs in the test set that were not in the training
@@
 to the training set and 20% to the testing set. The training set was
-preprocessed with mikropml to normalize values (scale/center), collapse
-correlated features, and remove features with zero-variance. Using
+preprocessed with mikropml to normalize values (scale and center), collapse
+correlated features, and remove features with zero variance. Using
 mikropml, the training set was split into train and validate sets to
@@
 preprocessed with mikropml to normalize values (scale/center), collapse
-correlated features, and remove features with zero-variance. Using
+correlated features, and remove features with zero variance. Using
 mikropml, the training set was split into train and validate sets to
@@
 the full training set. The OptiFit algorithm in mothur was used to
-cluster the left out testing data set using the OTUs of the training set
+cluster the held out testing data set using the OTUs of the training set
 as a reference. The preprocessing scale from the training set was

### 2022-07-11 submission/manuscript.tex
will and pat comment edits

@@
 
-\({^*}\) Current Affiliation:
+\({^*}\) Current Affiliation: Bio-Rad Laboratories, Hercules,
+California, USA
 
 \({^\#}\) Current Affiliation: Bristol Myers Squibb, Summit, New Jersey,
-USA~
+USA
 
@@
 often relies on clustering 16S rRNA gene sequences into operational
-taxonomic units (OTUs) to quantify microbial composition. The standard
-approach to clustering sequences into OTUs leverages the similarity of
-the sequences to each other rather than to a reference database. The
-abundance of each OTU is used to train a classification model. However,
-OTU assignments depend on the sequences in the data set and therefore
-can change if new data are added. This lack of stability complicates
-classification because in order to use the model to classify additional
-samples, the OTUs have to be reclustered to include the new sequences
-and the model retrained with the new OTU clusters. A new reference-based
-clustering algorithm, called OptiFit, addresses this issue by fitting
-new sequences into existing OTUs. While OptiFit is proven to produce
-high quality OTU clusters, it is unclear whether this method for fitting
-new sequence data into existing OTUs will impact the performance of
-classification models. We used OptiFit to cluster additional data into
-existing OTU clusters and quantified model performance in classifying a
-data set containing samples from patients with and without colonic
-screen relevant neoplasias (SRN). We compared the performance of this
-model to the standard procedure of clustering all the data together. We
-found that both approaches performed equally well in classifying SRNs.
-Moving forward, when OTUs are used in classification problems, OptiFit
-can be used to avoid the need to retrain models using reclustered
-sequences when classifying new samples.
+taxonomic units (OTUs) to quantify microbial composition. The abundance
+of each OTU is then used to train a classification model. The standard
+\emph{de novo} approach to clustering sequences into OTUs leverages the
+similarity of the sequences to each other rather than to a reference
+database. However, OTU assignments depend on the sequences in the data
+set and therefore can change if new data are added. This lack of
+stability complicates classification because in order to use the model
+to classify additional samples, the new sequences must be reclustered
+with the old data and the model must be retrained with the new OTU
+assignments. The new reference-based clustering algorithm, OptiFit,
+addresses this issue by fitting new sequences into existing OTUs. While
+OptiFit can produce high quality OTU clusters, it is unclear whether
+this method for fitting new sequence data into existing OTUs will impact
+the performance of classification models. We used OptiFit to cluster
+additional data into existing OTU clusters and then quantified model
+performance in classifying a data set containing samples from patients
+with and without colonic screen relevant neoplasia (SRN). We compared
+the performance of this model to the standard procedure of \emph{de
+novo} clustering all the data together. We found that both approaches
+performed equally well in classifying SRNs. Moving forward, when OTUs
+are used in classification problems, OptiFit can streamline the process
+of classifying new samples by avoiding the need to retrain models using
+reclustered sequences.
 
@@
 diagnose people. One of the challenges with using classification models
-generated using the relative abundance of operational taxonomic unit
-(OTU) data is that 16S rRNA gene sequences are assigned to OTUs based on
-their similarity to other sequences in the dataset. If data are
-generated from new patients seeking a diagnosis, then it would be
-necessary to reassign sequences to OTUs and retrain the classification
-model. Yet there is a desire to have a single, validated model that can
-be deployed. To overcome this obstacle, we applied the OptiFit
-clustering algorithm which fits new sequence data to existing OTUs
-allowing the reuse of a consistent model. A random forest machine
-learning model deployed using OptiFit performed as well as the
-traditional reassignment and retraining approach. This result indicates
-there is potential for developing machine learning classification models
-based on OTU relative abundance data.
+based on the relative abundance of operational taxonomic units (OTUs) is
+that 16S rRNA gene sequences are assigned to OTUs based on their
+similarity to other sequences in the data set. If data are generated
+from new patients seeking a diagnosis, then it would be necessary to
+reassign sequences to OTUs and retrain the classification model. Yet
+there is a desire to have a single, validated model that can be widely
+deployed. To overcome this obstacle, we applied the OptiFit clustering
+algorithm which fits new sequence data to existing OTUs and allows for
+the reuse of a consistent model. A random forest machine learning model
+deployed using OptiFit performed as well as the traditional reassignment
+and retrain approach. This result indicates that there is potential for
+developing deployable machine learning classification models based on
+OTU relative abundance data.
 
@@
 Gut community composition is useful as a resource for machine learning
-classification of diseases, including colorectal cancer
+classification of diseases, such as colorectal cancer
 (\protect\hyperlink{ref-baxter2016}{1},
-\protect\hyperlink{ref-zackular2014}{2}). Amplicon sequencing of the 16S
-rRNA gene is a reliable tool for assessing the taxonomic composition of
-microbial communities, which is the input to these models. Analysis of
-16S rRNA gene sequence data generally relies on clustering of sequences
-based on similarity into operational taxonomic units (OTUs). The process
-of OTU clustering can either be reference-based or \emph{de novo}. The
-quality of OTUs generated with reference-based clustering is generally
-poor compared to those generated with \emph{de novo} clustering
-(\protect\hyperlink{ref-westcott2015}{3}). While \emph{de novo}
-clustering produces high-quality OTU clusters where sequences are
+\protect\hyperlink{ref-zackular2014}{2}). Taxonomic composition of
+microbial communities can be assessed using amplicon sequencing of the
+16S rRNA gene, which is the input to classification models. Analysis of
+16S rRNA gene sequence data generally relies on similarity-based
+clustering of sequences into operational taxonomic units (OTUs). The
+process of OTU clustering can either be reference-based or \emph{de
+novo}. The quality of OTUs generated with reference-based clustering is
+generally poor compared to those generated with \emph{de novo}
+clustering (\protect\hyperlink{ref-westcott2015}{3}). While \emph{de
+novo} clustering produces high-quality OTU clusters where sequences are
 accurately grouped based on similarity thresholds, the resulting OTU
-clusters depend on the data in the data set and the addition of new data
-could change the overall OTU clusters. The unstable nature of OTU
-clustering complicates deployment of machine learning models since
-integration of additional data requires reclustering all the data and
-retraining of the model. The ability to integrate new data into a
-validated model without reclustering and retraining could allow for
-deployment of a single model that new data can be continually added to.
-Recently Sovacool \emph{et al} introduced OptiFit: a method for fitting
-new sequence data into existing OTUs
-(\protect\hyperlink{ref-sovacool2022}{4}). While OptiFit is proven to
+clusters depend on the sequences within the data set and the addition of
+new data has the potential to redefine OTU cluster composition. The
+unstable nature of \emph{de novo} OTU clustering complicates deployment
+of machine learning models since integration of additional data requires
+reclustering all the data and retraining the model. The ability to
+integrate new data into a validated model without reclustering and
+retraining could allow for deployment of a single model that new data
+can be continually tested against. Recently, Sovacool \emph{et al}
+introduced OptiFit, a method for fitting new sequence data into existing
+OTUs (\protect\hyperlink{ref-sovacool2022}{4}). While OptiFit can
 effectively fit new sequence data to existing OTU clusters, it is
-unknown if the use of OptiFit will have an impact on classification.
-Here we tested the ability of OptiFit to cluster new sequence data into
-existing OTU clusters for the purpose of classification of disease based
-on gut microbiome composition.
+unknown if the use of OptiFit will have an impact on classification
+performance. Here, we tested the ability of OptiFit to cluster new
+sequence data into existing OTU clusters for the purpose of classifying
+disease based on gut microbiome composition.
 
-We compared two approaches, one using all of the data to generate OTU
-clusters and the other generating OTU clusters with a portion of the
-data and then fitting the remaining sequence data to the existing OTUs
-using OptiFit. In the first approach, all of the 16S rRNA sequence data
-was \emph{de novo} clustered into OTUs with the OptiClust algorithm in
-mothur (\protect\hyperlink{ref-westcott2017}{5}). The resulting
-abundance data was then split into training and testing sets, where the
-training set was used to tune hyperparameters and ultimately train the
-model. The testing set was then classified with the model and the
-performance of the model was quantified (Figure 1A). However, with this
-methodology we would have to regenerate the OTU clusters and retrain the
-model if we wanted to classify additional samples. The OptiFit algorithm
-(\protect\hyperlink{ref-sovacool2022}{4}) addresses this problem by
-enabling new sequences to be clustered into existing OTUs. The OptiFit
-workflow is similar to the OptiClust workflow where the data was
-clustered into OTUs and used to tune hyperparameters and ultimately
-train the model. Then, we used OptiFit to fit sequence data of samples
-not part of the original data set into the existing OTUs and used the
-same model to classify the samples (Figure 1B). To test how the model
-performance compares between these two methodologies, we used a publicly
-available data set of 16S rRNA gene sequences from stool samples of
-healthy subjects as well as subjects with SRN consisting of advanced
-adenoma and carcinoma (\protect\hyperlink{ref-baxter2016}{1}). The data
-set was randomly split into an 80\% train set and 20\% test set. For the
-standard OptiClust workflow, the training and test sets were \emph{de
-novo} clustered together into OTUs then the resulting abundance table
-was split into the training and testing set. For the OptiFit workflow,
-the train set was clustered \emph{de novo} into OTUs and the remaining
-test set was fit to the OTU clusters using the OptiFit algorithm. For
-both workflows, the abundance table of the train set was used to tune
-hyperparameters and train a random forest model to classify SRN. The
-test set was classified as either control or SRN using the trained
-models.To account for variation depending on the split of the data, the
-data set was randomly split 100 times and the process repeated for each
-of the 100 data splits. By comparing the model performance of
-classifying the samples in the test data set between the OptiFit and
-OptiClust algorithms, we quantified the impact of using OptiFit on model
-classification performance.
+We compared two approaches, one using all the data to generate OTU
+clusters, and the other generating \emph{de novo} OTU clusters with a
+portion of the data and then fitting the remaining sequence data to the
+existing OTUs using OptiFit. In the first approach, all the 16S rRNA
+sequence data was \emph{de novo} clustered into OTUs with the OptiClust
+algorithm in mothur (\protect\hyperlink{ref-westcott2017}{5}). The
+resulting abundance data was then split into training and testing sets,
+where the training set was used to tune hyperparameters and ultimately
+train the model. The testing set was then classified with the model and
+the performance of the model was quantified (Figure 1A). However, with
+this methodology, we would have to regenerate the OTU clusters and
+retrain the model if we wanted to classify additional samples. The
+OptiFit algorithm (\protect\hyperlink{ref-sovacool2022}{4}) addresses
+this problem by enabling new sequences to be clustered into existing
+OTUs. The OptiFit workflow is similar to the OptiClust workflow, where
+the data was clustered into OTUs and used to tune hyperparameters and
+ultimately train the model. Then, we used OptiFit to fit sequence data
+of samples not part of the original data set into the existing OTUs, and
+used the same model to classify the samples (Figure 1B). To test how the
+model performance compared between these two approaches, we used a
+publicly available data set of 16S rRNA gene sequences from stool
+samples of healthy subjects (n = 261) as well as subjects with SRN
+consisting of advanced adenoma and carcinoma (n = 229)
+(\protect\hyperlink{ref-baxter2016}{1}). The data set was randomly split
+into an 80\% train set and 20\% test set. For the standard OptiClust
+workflow, the training and test sets were \emph{de novo} clustered
+together into OTUs, then the resulting abundance table was split into
+the training and testing set. For the OptiFit workflow, the train set
+was clustered \emph{de novo} into OTUs, and the remaining test set was
+fit to the OTU clusters using the OptiFit algorithm. For both workflows,
+the abundance table of the train set was used to tune hyperparameters
+and train a random forest model to classify SRN. The test set was
+classified as either control or SRN using the trained models. To account
+for variation, the data set was randomly split 100 times and the process
+repeated for each of the 100 data splits. By comparing the model
+performance of classifying the samples in the test data set between the
+OptiFit and OptiClust algorithms, we quantified the impact of using
+OptiFit on model classification performance.
 
@@
 sequences within the similarity threshold are not in an OTU together. An
-MCC score of zero essentially means the sequences are randomly
-clustered. An MCC score of 1 means all sequences in an OTU are within
-the similarity threshold and all sequence pairs within the similarity
-threshold are in the same OTU. To ensure that OptiFit is appropriately
-integrating new sequence data into the existing OTUs, we expected the
-MCC scores produced by the OptiClust and OptiFit workflows to be
-similar. Since the data was only clustered once in the OptiClust
-workflow there was only one MCC score while the OptiFit workflow
-produced an MCC score for the OTU clusters from each data split. Overall
-the MCC scores were similar between OptiClust (MCC = 0.884) and OptiFit
-(average MCC = 0.879). This indicated that OptiFit performed as well as
-OptiClust when integrating new sequences into the existing OTUs.
+MCC score of zero means the sequences are randomly clustered. An MCC
+score of 1 means all sequences in an OTU are within the similarity
+threshold and all sequence pairs within the similarity threshold are in
+the same OTU. To ensure that OptiFit is appropriately integrating new
+sequence data into the existing OTUs, we expected the MCC scores
+produced by the OptiClust and OptiFit workflows to be similar. Since the
+data was only clustered once in the OptiClust workflow there was only
+one MCC score while the OptiFit workflow produced an MCC score for the
+OTU clusters from each data split. Overall, the MCC scores were similar
+between OptiClust (MCC = 0.884) and OptiFit (average MCC = 0.879,
+standard deviation = 0.002). This indicated that OptiFit performed as
+well as OptiClust when integrating new sequences into the existing OTUs.
 
@@
 algorithms (p-value = 0.13, OptiClust mean CV AUROC = 0.694, OptiFit
-mean CV AUROC = 0.697, Figure 2A). The trained model was then deployed
+mean CV AUROC = 0.697; Figure 2A). The trained model was then deployed
 to classify the samples of the test data as control or SRN. The
@@
 (p-value = 0.63, OptiClust mean test AUROC = 0.709, OptiFit mean test
-AUROC = 0.712, Figure 2B,C) indicating that new data could be fit to
-existing OTU clusters without impacting model performance.
+AUROC = 0.712; Figures 2B and 2C) indicating that new data could be fit
+to existing OTU clusters without impacting model performance.
 
@@
 relative abundance. A potential problem with using OptiFit is that any
-sequences in the new data that do not map to the existing OTU clusters
-will be discarded resulting in a possible loss of information. However,
-we demonstrated that OptiFit can be used to fit new sequence data into
-existing OTU clusters and perform equally well in predicting SRN
-compared to clustering all of the sequence data together. The ability to
-integrate data from new samples into existing OTUs enables the
-deployment of a single machine learning model. These results are based
-on a single data set and disease. Further analysis is needed to
-determine the number of samples that are necessary to build a robust
-model capable of classifying diverse samples. A robust machine learning
-model could be implemented as part of a non-invasive and low-cost aid in
-diagnosing SRN and other diseases.
+sequences from the new samples that do not map to the existing OTU
+clusters will be discarded, resulting in a possible loss of information.
+Regardless, we demonstrated that OptiFit can be used to fit new sequence
+data into existing OTU clusters and performs equally well in predicting
+SRN compared to \emph{de novo} clustering all the sequence data
+together. The ability to integrate data from new samples into existing
+OTUs enables the deployment of a single machine learning model. Further
+analysis is needed to determine the number of samples that are necessary
+to build a robust model capable of classifying diverse samples. A robust
+machine learning model could be implemented as part of a non-invasive
+and low-cost aid in diagnosing SRN and other diseases.
 
@@
 subjects identified in the metadata as normal, high risk normal, or
-adenoma were categorized as ``normal'' while samples from subjects
+adenoma were categorized as ``normal'', while samples from subjects
 identified as advanced adenoma or carcinoma were categorized as ``screen
@@
 
-\textbf{\emph{Data Processing.}} The full dataset was preprocessed with
+\textbf{\emph{Data Processing.}} The full data set was preprocessed with
 mothur (v1.47) (\protect\hyperlink{ref-schloss2009}{7}) to join forward
@@
 that are in the training or test sets. Each sample was in the training
-set an average of 80 times (SD=4.1) and the test set an average of 20
-times (SD=4.1).
+set an average of 80 times (standard deviation = 4.1) and the test set
+an average of 20 times (standard deviation = 4.1).
 
-The data was processed through two workflows. First, the standard
-workflow using the OptiClust algorithm
-(\protect\hyperlink{ref-westcott2017}{5}). In this pathway, all of the
-data was clustered together with OptiClust to generate OTUs and the
+The data was processed through two workflows. In the control workflow,
+all the data was clustered together with OptiClust
+(\protect\hyperlink{ref-westcott2017}{5}) to generate OTUs and the
 resulting abundance tables were split into the training and testing
-sets. In the second workflow, the preprocessed data was split into the
-training and testing sets. The training set was clustered into OTUs,
-then the test set was fit to the OTUs of the training set using the
-OptiFit algorithm (\protect\hyperlink{ref-sovacool2022}{4}). The OptiFit
-algorithm was run with method open so that any sequences that did not
-map to the existing OTU clusters would form new OTUs. For both pathways,
-the shared files were sub-sampled to 10,000 reads per sample.
+sets. In the experimental workflow, the preprocessed data was split into
+the training and testing sets. The training set was clustered into OTUs
+using OptiClust, then the test set was fit to the OTUs of the training
+set using the OptiFit algorithm
+(\protect\hyperlink{ref-sovacool2022}{4}). The OptiFit algorithm was run
+with method open so that any sequences that did not map to the existing
+OTU clusters would form new OTUs. For both pathways, the shared files
+were sub-sampled to 10,000 reads per sample.
 
-\textbf{\emph{Machine Learning.}} Machine learning using Random Forest
+\textbf{\emph{Machine Learning.}} Machine learning using random forest
 was conducted with the R package mikrompl (v 1.2.0)
@@
 (SRN or normal) for the samples in the test set for each data split. The
-training set was preprocessed to normalize OTU counts (scale/center),
-collapse correlated OTUs, and remove OTUs with zero-variance. The
-preprocessing from the training set was then applied to the test set.
-Any OTUs in the test set that were not in the training set were removed.
-P values comparing model performance were calculated as previously
-described (\protect\hyperlink{ref-topuxe7uoglu2020}{10}). The averaged
-ROC curves were plotted by taking the average and standard deviation of
-the sensitivity at each specificity value.
+training set was preprocessed to normalize OTU counts (scale and
+center), collapse correlated OTUs, and remove OTUs with zero variance.
+The preprocessing from the training set was then applied to the test
+set. Any OTUs in the test set that were not in the training set were
+removed. P values comparing model performance were calculated as
+previously described (\protect\hyperlink{ref-topuxe7uoglu2020}{10}). The
+averaged ROC curves were plotted by taking the average and standard
+deviation of the sensitivity at each specificity value.
 
@@
 were assigned to the training set and 20\% to the testing set. The
-training set was preprocessed with mikropml to normalize values
-(scale/center), collapse correlated features, and remove features with
-zero-variance. Using mikropml, the training set was split into train and
+training set was preprocessed with mikropml to normalize values (scale
+and center), collapse correlated features, and remove features with zero
+variance. Using mikropml, the training set was split into train and
 validate sets to compare results using different hyperparameter
@@
 was then used to train the model with the full training set. The OptiFit
-algorithm in mothur was used to cluster the left out testing data set
+algorithm in mothur was used to cluster the held out testing data set
 using the OTUs of the training set as a reference. The preprocessing
@@
 for the OptiClust and OptiFit workflows. Mean and standard deviation of
-the AUROC is represented by the black dot and whiskers. Mean AUROC is
-printed to the right of the points. \textbf{C)} Receiver operating
-characteristic (ROC)Averaged ROC curves
+the AUROC is represented by the black dot and whiskers. The mean AUROC
+is printed to the right of the points. \textbf{C)} Averaged receiver
+operating characteristic (ROC) curves.
 

### 2022-08-18 code/R/calc_nonmapped.R
use subsampled file for optifit modeling, calc fraction mapped

@@ -1,0 +1,37 @@
+################################
+# Courtney R Armour
+# August 2022
+# Schloss Lab
+# University of Michigan
+################################
+
+### SETUP -------------------
+library(tidyverse)
+
+outDir <- "data/learning/summary/"
+if(!dir.exists(outDir)){dir.create(outDir)}
+
+input <- commandArgs(trailingOnly = TRUE)
+
+# input <- c("data/process/optifit/split_1/test/glne.precluster.pick.renamed.fit.optifit_mcc.0.03.subsample.shared",
+#            "data/process/optifit/split_2/test/glne.precluster.pick.renamed.fit.optifit_mcc.0.03.subsample.shared")
+
+### FUNCTIONS -----------------
+
+get_frac_nonmapped <- function(file){
+    file_parts <- unlist(str_split(file,"/"))  
+    split <- file_parts[grepl(pattern="split",file_parts)]
+      
+    read_tsv(file) %>% 
+        select(-label,-numOtus) %>% 
+        mutate(RefSum=rowSums(select(.,starts_with("Ref_"))),
+               OtherSum=rowSums(select(.,starts_with("Otu")))) %>% 
+        select(Group,RefSum,OtherSum) %>% 
+        mutate(total = RefSum+OtherSum) %>% 
+        mutate(frac_nonmapped = OtherSum/total,
+               split = split)
+    
+}
+
+table <- map_dfr(input,get_frac_nonmapped)
+write_csv(table,"results/tables/fracNonMapped.csv")

### 2022-08-18 code/R/preprocess.R

@@ -23,6 +23,8 @@ ncores   <- as.numeric(input[4])
 # training <- "data/process/optifit/split_1/train/glne.precluster.pick.opti_mcc.0.03.subsample.shared"
 # #testing  <- "data/process/opticlust/split_1/test/glne.precluster.opti_mcc.0.03.subsample.0.03.pick.shared"
 # testing <- "data/process/optifit/split_1/test/glne.precluster.pick.renamed.fit.optifit_mcc.shared"
+# training <- "data/process/optifit/split_8/train/glne.precluster.pick.opti_mcc.0.03.subsample.shared"
+# testing <- "data/process/optifit/split_8/test/glne.precluster.pick.renamed.fit.optifit_mcc.0.03.subsample.shared"
 # ncores=12
 
 split <- unlist(str_split(training,"/"))[4]
@@ -56,7 +58,7 @@ train_shared <- read_tsv(training, col_types = cols(Group=col_character(),
 if (str_detect(testing, "optifit")) { 
   test_shared <- read_tsv(testing, col_types = cols(Group=col_character(),
                                                   .default = col_double())) %>%
-    select(-label, -numRefOtus) %>%
+    select(-label, -numOtus) %>%
     select(Group,starts_with("Ref_")) %>% #select only OTUs in reference
     rename_all(str_replace, "Ref_", "") #remove Ref_ label to match train data
 } else {
@@ -82,7 +84,7 @@ missing_cols <- setdiff(train_labels, test_labels)
 
 # Only keeping otus that are in both files and assigning as test data for ML prediction
 test <- test_shared %>%
-  select(common_labels)
+  select(all_of(common_labels))
 
 # Adding in missing otus all coded as 0 so test has the same cols as the loo shared
 test[missing_cols] <- 0

### 2022-08-18 submission/manuscript.Rmd

@@
 
-${^*}$ Current Affiliation:
+${^*}$ Current Affiliation: Bio-Rad Laboratories, Hercules, California,
+USA
 
@@
 often relies on clustering 16S rRNA gene sequences into operational
-taxonomic units (OTUs) to quantify microbial composition. The standard
-*de novo* approach to clustering sequences into OTUs leverages the similarity of
-the sequences to each other rather than to a reference database. The
-abundance of each OTU is used to train a classification model. However,
-OTU assignments depend on the sequences in the data set and therefore
-can change if new data are added. This lack of consistency complicates
-classification because in order to use the model to classify additional
-samples, the OTUs have to be reclustered to include the new sequences
-and the model retrained with the new OTU clusters. A new reference-based
-clustering algorithm, called OptiFit, addresses this issue by fitting
-new sequences into existing OTUs. While OptiFit has been shown to produce
-high quality OTU clusters, it is unknown whether this method will impact 
-the performance of
-classification models. We used OptiFit to cluster additional data into
-existing OTU clusters and quantified model performance in classifying a
-data set containing samples from patients with and without colonic
-screen relevant neoplasias (SRN). We compared the performance of this
-model to the standard procedure of *de novo* clustering all the data together. We
-found that both approaches performed equally well in classifying SRNs.
-Moving forward, when OTUs are used in classification problems, OptiFit
-can streamline the process of classifying new samples by avoiding the need to 
-retrain models using reclustered sequences.
+taxonomic units (OTUs) to quantify microbial composition. The abundance
+of each OTU is then used to train a classification model. The standard
+*de novo* approach to clustering sequences into OTUs leverages the
+similarity of the sequences to each other rather than to a reference
+database. However, OTU assignments depend on the sequences in the data
+set and therefore can change if new data are added. This lack of
+stability complicates classification because in order to use the model
+to classify additional samples, the new sequences must be reclustered
+with the old data and the model must be retrained with the new OTU
+assignments. The new reference-based clustering algorithm, OptiFit,
+addresses this issue by fitting new sequences into existing OTUs. While
+OptiFit can produce high quality OTU clusters, it is unclear whether
+this method for fitting new sequence data into existing OTUs will impact
+the performance of classification models. We used OptiFit to cluster
+additional data into existing OTU clusters and then quantified model
+performance in classifying a data set containing samples from patients
+with and without colonic screen relevant neoplasia (SRN). We compared
+the performance of this model to the standard procedure of *de novo*
+clustering all the data together. We found that both approaches
+performed equally well in classifying SRNs. Moving forward, when OTUs
+are used in classification problems, OptiFit can streamline the process
+of classifying new samples by avoiding the need to retrain models using
+reclustered sequences.
 
@@
 diagnose people. One of the challenges with using classification models
-generated using the relative abundance of operational taxonomic unit
-(OTU) data is that 16S rRNA gene sequences are assigned to OTUs based on
-their similarity to other sequences in the dataset. If data are
-generated from new patients seeking a diagnosis, then it would be
-necessary to reassign sequences to OTUs and retrain the classification
-model. Yet there is a desire to have a single, validated model that can
-be deployed. To overcome this obstacle, we applied the OptiFit
-clustering algorithm which fits new sequence data to existing OTUs,
-allowing the reuse of a consistent model. A random forest machine
-learning model deployed using OptiFit performed as well as the
-traditional reassignment and retraining approach. This result indicates
-there is potential for developing machine learning classification models
-based on OTU relative abundance data.
+based on the relative abundance of operational taxonomic units (OTUs) is
+that 16S rRNA gene sequences are assigned to OTUs based on their
+similarity to other sequences in the data set. If data are generated
+from new patients seeking a diagnosis, then it would be necessary to
+reassign sequences to OTUs and retrain the classification model. Yet
+there is a desire to have a single, validated model that can be widely
+deployed. To overcome this obstacle, we applied the OptiFit clustering
+algorithm which fits new sequence data to existing OTUs and allows for
+the reuse of a consistent model. A random forest machine learning model
+deployed using OptiFit performed as well as the traditional reassignment
+and retrain approach. This result indicates that there is potential for
+developing deployable machine learning classification models based on
+OTU relative abundance data.
 
@@
 Gut community composition is useful as a resource for machine learning
-classification of diseases, including colorectal cancer [@baxter2016;
-@zackular2014]. Amplicon sequencing of the 16S rRNA gene is a reliable
-tool for assessing the taxonomic composition of microbial communities,
-which is the input to these models. Analysis of 16S rRNA gene sequence
-data generally relies on clustering sequences based on similarity
-into operational taxonomic units (OTUs). The process of OTU clustering
-can either be reference-based or *de novo*. The quality of OTUs
-generated with reference-based clustering is generally poor compared to
-those generated with *de novo* clustering @westcott2015. While *de novo*
+classification of diseases, such as colorectal cancer [@baxter2016;
+@zackular2014]. Taxonomic composition of microbial communities can be
+assessed using amplicon sequencing of the 16S rRNA gene, which is the
+input to classification models. Analysis of 16S rRNA gene sequence data
+generally relies on similarity-based clustering of sequences into
+operational taxonomic units (OTUs). The process of OTU clustering can
+either be reference-based or *de novo*. The quality of OTUs generated
+with reference-based clustering is generally poor compared to those
+generated with *de novo* clustering @westcott2015. While *de novo*
 clustering produces high-quality OTU clusters where sequences are
 accurately grouped based on similarity thresholds, the resulting OTU
-clusters depend on the data in the data set and the addition of new data
-could change the overall OTU clusters. The inconsistent nature of *de novo* OTU
-clustering complicates deployment of machine learning models since
-integration of additional data requires reclustering all the data and
-retraining the model. The ability to integrate new data into a
-validated model without reclustering and retraining could allow for
-deployment of a single model that new data can be continually tested against.
-Recently, Sovacool *et al* introduced OptiFit: a method for fitting new
-sequence data into existing OTUs @sovacool2022. While OptiFit is proven
-to effectively fit new sequence data to existing OTU clusters, it is
-unknown if the use of OptiFit will have an impact on classification performance.
-Here, we tested the ability of OptiFit to cluster new sequence data into
-existing OTU clusters for the purpose of classification of disease based
-on gut microbiome composition.
+clusters depend on the sequences within the data set and the addition of
+new data has the potential to redefine OTU cluster composition. The
+unstable nature of *de novo* OTU clustering complicates deployment of
+machine learning models since integration of additional data requires
+reclustering all the data and retraining the model. The ability to
+integrate new data into a validated model without reclustering and
+retraining could allow for deployment of a single model that new data
+can be continually tested against. Recently, Sovacool *et al* introduced
+OptiFit, a method for fitting new sequence data into existing OTUs
+@sovacool2022. While OptiFit can effectively fit new sequence data to
+existing OTU clusters, it is unknown if the use of OptiFit will have an
+impact on classification performance. Here, we tested the ability of
+OptiFit to cluster new sequence data into existing OTU clusters for the
+purpose of classifying disease based on gut microbiome composition.
 
-We compared two approaches, one using all of the data to generate *de novo* OTU
-clusters, and the other generating *de novo* OTU clusters with a portion of the
-data and then fitting the remaining sequence data to the existing OTUs
-using OptiFit. In the first approach, all of the 16S rRNA sequence data
-was *de novo* clustered into OTUs with the OptiClust algorithm in mothur
-@westcott2017. The resulting abundance data was then split into training
-and testing sets, where the training set was used to tune
+We compared two approaches, one using all the data to generate OTU
+clusters, and the other generating *de novo* OTU clusters with a portion
+of the data and then fitting the remaining sequence data to the existing
+OTUs using OptiFit. In the first approach, all the 16S rRNA sequence
+data was *de novo* clustered into OTUs with the OptiClust algorithm in
+mothur @westcott2017. The resulting abundance data was then split into
+training and testing sets, where the training set was used to tune
 hyperparameters and ultimately train the model. The testing set was then
@@
 addresses this problem by enabling new sequences to be clustered into
-existing OTUs. The OptiFit workflow is similar to the OptiClust workflow,
-where the data was clustered into OTUs and used to tune hyperparameters
-and ultimately train the model. Then, we used OptiFit to fit sequence
-data of samples not part of the original data set into the existing OTUs,
-and used the same model to classify the samples (Figure 1B). To test how
-the model performance compares between these two methodologies, we used
-a publicly available data set of 16S rRNA gene sequences from stool
-samples of healthy subjects as well as subjects with SRN consisting of
-advanced adenoma and carcinoma @baxter2016. The data set was randomly
-split into an 80% train set and 20% test set. For the standard OptiClust
-workflow, the training and test sets were *de novo* clustered together
-into OTUs, then the resulting abundance table was split into the training
-and testing set. For the OptiFit workflow, the train set was clustered
-*de novo* into OTUs, and the remaining test set was fit to the OTU
-clusters using the OptiFit algorithm. For both workflows, the abundance
-table of the train set was used to tune hyperparameters and train a
-random forest model to classify SRN. The test set was classified as
-either control or SRN using the trained models. To account for variation
-depending on the split of the data, the data set was randomly split 100
-times and the process repeated for each of the 100 data splits. By
-comparing the model performance of classifying the samples in the test
-data set between the OptiFit and OptiClust algorithms, we quantified the
-impact of using OptiFit on model classification performance.
+existing OTUs. The OptiFit workflow is similar to the OptiClust
+workflow, where the data was clustered into OTUs and used to tune
+hyperparameters and ultimately train the model. Then, we used OptiFit to
+fit sequence data of samples not part of the original data set into the
+existing OTUs, and used the same model to classify the samples (Figure
+1B). To test how the model performance compared between these two
+approaches, we used a publicly available data set of 16S rRNA gene
+sequences from stool samples of healthy subjects (n = 261) as well as
+subjects with SRN consisting of advanced adenoma and carcinoma (n = 229)
+@baxter2016. The data set was randomly split into an 80% train set and
+20% test set. For the standard OptiClust workflow, the training and test
+sets were *de novo* clustered together into OTUs, then the resulting
+abundance table was split into the training and testing set. For the
+OptiFit workflow, the train set was clustered *de novo* into OTUs, and
+the remaining test set was fit to the OTU clusters using the OptiFit
+algorithm. For both workflows, the abundance table of the train set was
+used to tune hyperparameters and train a random forest model to classify
+SRN. The test set was classified as either control or SRN using the
+trained models. To account for variation, the data set was randomly
+split 100 times and the process repeated for each of the 100 data
+splits. By comparing the model performance of classifying the samples in
+the test data set between the OptiFit and OptiClust algorithms, we
+quantified the impact of using OptiFit on model classification
+performance.
 
@@
   filter(algorithm == "optifit" & state=="combo") %>% 
-  summarise(mean_mcc = mean(mcc))
+  summarise(mean_mcc = mean(mcc),
+            sd_mcc = sd(mcc))
 ```
@@
 similarity threshold and any sequences within the similarity threshold
-are not in an OTU together. An MCC score of zero essentially means the
-sequences are randomly clustered. An MCC score of 1 means all sequences
-in an OTU are within the similarity threshold and all sequence pairs
-within the similarity threshold are in the same OTU. To ensure that
-OptiFit is appropriately integrating new sequence data into the existing
-OTUs, we expected the MCC scores produced by the OptiClust and OptiFit
-workflows to be similar. Since the data was only clustered once in the
-OptiClust workflow there was only one MCC score while the OptiFit
-workflow produced an MCC score for the OTU clusters from each data
-split. Overall, the MCC scores were similar between OptiClust (MCC =
+are not in an OTU together. An MCC score of zero means the sequences are
+randomly clustered. An MCC score of 1 means all sequences in an OTU are
+within the similarity threshold and all sequence pairs within the
+similarity threshold are in the same OTU. To ensure that OptiFit is
+appropriately integrating new sequence data into the existing OTUs, we
+expected the MCC scores produced by the OptiClust and OptiFit workflows
+to be similar. Since the data was only clustered once in the OptiClust
+workflow there was only one MCC score while the OptiFit workflow
+produced an MCC score for the OTU clusters from each data split.
+Overall, the MCC scores were similar between OptiClust (MCC =
 `r round(opticlust_mcc,digits=3)`) and OptiFit (average MCC =
-`r round(optifit_avg_mcc,digits=3)`). This indicated that OptiFit
-performed as well as OptiClust when integrating new sequences into the
-existing OTUs.
+`r round(optifit_avg_mcc %>% pull(mean_mcc),digits=3)`, standard
+deviation = `r round(optifit_avg_mcc %>% pull(sd_mcc),digits=3)`). This
+indicated that OptiFit performed as well as OptiClust when integrating
+new sequences into the existing OTUs.
 
@@
 OptiFit mean CV AUROC =
-`r format_decimal(performance_summary %>% filter(algorithm == "optifit") %>% select(mean_cv_AUC),3)`,
+`r format_decimal(performance_summary %>% filter(algorithm == "optifit") %>% select(mean_cv_AUC),3)`;
 Figure 2A). The trained model was then deployed to classify the samples
@@
 OptiFit mean test AUROC =
-`r format_decimal(performance_summary %>% filter(algorithm == "optifit") %>% select(mean_AUC),3)`,
-Figure 2B,C) indicating that new data could be fit to existing OTU
+`r format_decimal(performance_summary %>% filter(algorithm == "optifit") %>% select(mean_AUC),3)`;
+Figures 2B and 2C) indicating that new data could be fit to existing OTU
 clusters without impacting model performance.
@@
 relative abundance. A potential problem with using OptiFit is that any
-sequences in the new data that do not map to the existing OTU clusters
-will be discarded, resulting in a possible loss of information. However,
-we demonstrated that OptiFit can be used to fit new sequence data into
-existing OTU clusters and performs equally well in predicting SRN
-compared to *de novo* clustering all of the sequence data together. The ability to
-integrate data from new samples into existing OTUs enables the
-deployment of a single machine learning model. These results are based
-on a single data set and disease. Further analysis is needed to
-determine the number of samples that are necessary to build a robust
-model capable of classifying diverse samples. A robust machine learning
-model could be implemented as part of a non-invasive and low-cost aid in
-diagnosing SRN and other diseases.
+sequences from the new samples that do not map to the existing OTU
+clusters will be discarded, resulting in a possible loss of information.
+Regardless, we demonstrated that OptiFit can be used to fit new sequence
+data into existing OTU clusters and performs equally well in predicting
+SRN compared to *de novo* clustering all the sequence data together. The
+ability to integrate data from new samples into existing OTUs enables
+the deployment of a single machine learning model. Further analysis is
+needed to determine the number of samples that are necessary to build a
+robust model capable of classifying diverse samples. A robust machine
+learning model could be implemented as part of a non-invasive and
+low-cost aid in diagnosing SRN and other diseases.
 
@@
 
-***Data Processing.*** The full dataset was preprocessed with mothur
+***Data Processing.*** The full data set was preprocessed with mothur
 (v1.47) @schloss2009 to join forward and reverse reads, merge duplicate
@@
 training or test sets. Each sample was in the training set an average of
-`r n_test_train %>% pull(avg_train)` times
-(SD=`r format_decimal(n_test_train %>% pull(sd_train),digits = 1)`) and
-the test set an average of `r n_test_train %>% pull(avg_test)` times
-(SD=`r format_decimal(n_test_train %>% pull(sd_test),digits = 1)`).
+`r n_test_train %>% pull(avg_train)` times (standard deviation =
+`r format_decimal(n_test_train %>% pull(sd_train),digits = 1)`) and the
+test set an average of `r n_test_train %>% pull(avg_test)` times
+(standard deviation =
+`r format_decimal(n_test_train %>% pull(sd_test),digits = 1)`).
 
-The data was processed through two workflows. First, the standard
-workflow using the OptiClust algorithm @westcott2017. In this workflow,
-all of the data was clustered together with OptiClust to generate OTUs
-and the resulting abundance tables were split into the training and
-testing sets. In the second workflow, the preprocessed data was split
-into the training and testing sets. The training set was clustered into
-OTUs, then the test set was fit to the OTUs of the training set using
-the OptiFit algorithm @sovacool2022. The OptiFit algorithm was run with the open
-method so that any sequences that did not map to the existing OTU
-clusters would form new OTUs. 
-<!-- Did you then remove the columns corresponding to the additional OTUs? 
-Typically you'd want to run the closed method for this use-case, right? -->
-For both workflows, the shared files were
-sub-sampled to 10,000 reads per sample.
+The data was processed through two workflows. In the control workflow,
+all the data was clustered together with OptiClust @westcott2017 to
+generate OTUs and the resulting abundance tables were split into the
+training and testing sets. In the experimental workflow, the
+preprocessed data was split into the training and testing sets. The
+training set was clustered into OTUs using OptiClust, then the test set
+was fit to the OTUs of the training set using the OptiFit algorithm
+@sovacool2022. The OptiFit algorithm was run with method open so that
+any sequences that did not map to the existing OTU clusters would form
+new OTUs. For both pathways, the shared files were sub-sampled to 10,000
+reads per sample.
 
-***Machine Learning.*** Machine learning using Random Forest was
+***Machine Learning.*** Machine learning using random forest was
 conducted with the R package mikrompl (v 1.2.0) @topçuoglu2021 to
@@
 for each data split. The training set was preprocessed to normalize OTU
-counts (scale and center), collapse correlated OTUs, and remove OTUs with
-zero variance. The preprocessing from the training set was then applied
-to the test set. Any OTUs in the test set that were not in the training
-set were removed. P values comparing model performance were calculated
-as previously described @topçuoglu2020. The averaged ROC curves were
-plotted by taking the average and standard deviation of the sensitivity
-at each specificity value.
+counts (scale and center), collapse correlated OTUs, and remove OTUs
+with zero variance. The preprocessing from the training set was then
+applied to the test set. Any OTUs in the test set that were not in the
+training set were removed. P values comparing model performance were
+calculated as previously described @topçuoglu2020. The averaged ROC
+curves were plotted by taking the average and standard deviation of the
+sensitivity at each specificity value.
 
@@
 to the training set and 20% to the testing set. The training set was
-preprocessed with mikropml to normalize values (scale and center), collapse
-correlated features, and remove features with zero variance. Using
-mikropml, the training set was split into train and validate sets to
-compare results using different hyperparameter settings. The highest
+preprocessed with mikropml to normalize values (scale and center),
+collapse correlated features, and remove features with zero variance.
+Using mikropml, the training set was split into train and validate sets
+to compare results using different hyperparameter settings. The highest
 performing hyperparameter setting was then used to train the model with
@@
 preprocessed with mikropml to normalize values (scale/center), collapse
-correlated features, and remove features with zero variance. Using
+correlated features, and remove features with zero-variance. Using
 mikropml, the training set was split into train and validate sets to
@@
 the OptiClust and OptiFit workflows. Mean and standard deviation of the
-AUROC is represented by the black dot and whiskers. Mean AUROC is
-printed to the right of the points. **C)** Receiver operating
-characteristic (ROC)Averaged ROC curves
+AUROC is represented by the black dot and whiskers. The mean AUROC is
+printed to the right of the points. **C)** Averaged receiver operating
+characteristic (ROC) curves.

### 2022-08-31 submission/manuscript.Rmd
latest paper edits

@@
 ---
-title: "**Reference-based OTU clustering for machine learning classification**"
+title: "**Streamlined implementation of a machine learning model to classify screen relevant neoplasia using reference-based OTU clustering**"
 output:
@@
 
-Running title: Reference-based OTU clustering for ML classification
+Running title: Reference-based OTU clustering for machine learning
+classification
 
@@
 similarity of the sequences to each other rather than to a reference
-database. However, OTU assignments depend on the sequences in the data
-set and therefore can change if new data are added. This lack of
-stability complicates classification because in order to use the model
-to classify additional samples, the new sequences must be reclustered
-with the old data and the model must be retrained with the new OTU
-assignments. The new reference-based clustering algorithm, OptiFit,
-addresses this issue by fitting new sequences into existing OTUs. While
-OptiFit can produce high quality OTU clusters, it is unclear whether
-this method for fitting new sequence data into existing OTUs will impact
-the performance of classification models. We used OptiFit to cluster
-additional data into existing OTU clusters and then quantified model
-performance in classifying a data set containing samples from patients
-with and without colonic screen relevant neoplasia (SRN). We compared
-the performance of this model to the standard procedure of *de novo*
-clustering all the data together. We found that both approaches
+database. However, such an approach depends on the sequences in the
+dataset and therefore OTU assignments can change if new data are added.
+This lack of stability complicates classification because in order to
+use the model to classify additional samples, the new sequences must be
+reclustered with the old data and the model must be retrained with the
+new OTU assignments. The new reference-based clustering algorithm,
+OptiFit, addresses this issue by fitting new sequences into existing
+OTUs. While OptiFit can produce high quality OTU clusters, it is unclear
+whether this method for fitting new sequence data into existing OTUs
+will impact the performance of classification models. We used OptiFit to
+cluster additional data into existing OTU clusters and then evaluated
+model performance in classifying a dataset containing samples from
+patients with and without colonic screen relevant neoplasia (SRN). We
+compared the performance of this model to the standard procedure of *de
+novo* clustering all the data together. We found that both approaches
 performed equally well in classifying SRNs. Moving forward, when OTUs
@@
 that 16S rRNA gene sequences are assigned to OTUs based on their
-similarity to other sequences in the data set. If data are generated
-from new patients seeking a diagnosis, then it would be necessary to
-reassign sequences to OTUs and retrain the classification model. Yet
-there is a desire to have a single, validated model that can be widely
-deployed. To overcome this obstacle, we applied the OptiFit clustering
-algorithm which fits new sequence data to existing OTUs and allows for
-the reuse of a consistent model. A random forest machine learning model
-deployed using OptiFit performed as well as the traditional reassignment
-and retrain approach. This result indicates that there is potential for
-developing deployable machine learning classification models based on
-OTU relative abundance data.
+similarity to other sequences in a dataset. If data are generated from
+new patients seeking a diagnosis, then a standard approach requires
+reassigning sequences to OTUs and retraining the classification model.
+Yet there is a desire to have a single, validated model that can be
+widely deployed. To overcome this obstacle, we applied the OptiFit
+clustering algorithm which fits new sequence data to existing OTUs and
+allows for the reuse of models. A random forest machine learning model
+implemented using OptiFit performed as well as the traditional
+reassignment and retrain approach. This result indicates that there is
+potential for developing and applying machine learning models based on
+OTU relative abundance data that do not require retraining.
 
@@
 accurately grouped based on similarity thresholds, the resulting OTU
-clusters depend on the sequences within the data set and the addition of
+clusters depend on the sequences within the dataset and the addition of
 new data has the potential to redefine OTU cluster composition. The
@@
 integrate new data into a validated model without reclustering and
-retraining could allow for deployment of a single model that new data
-can be continually tested against. Recently, Sovacool *et al* introduced
+retraining could allow for the application of a single model that can
+continually classify new data. Recently, Sovacool *et al.* introduced
 OptiFit, a method for fitting new sequence data into existing OTUs
@@
 OTUs using OptiFit. In the first approach, all the 16S rRNA sequence
-data was *de novo* clustered into OTUs with the OptiClust algorithm in
-mothur @westcott2017. The resulting abundance data was then split into
-training and testing sets, where the training set was used to tune
-hyperparameters and ultimately train the model. The testing set was then
-classified with the model and the performance of the model was
-quantified (Figure 1A). However, with this methodology, we would have to
-regenerate the OTU clusters and retrain the model if we wanted to
-classify additional samples. The OptiFit algorithm @sovacool2022
-addresses this problem by enabling new sequences to be clustered into
-existing OTUs. The OptiFit workflow is similar to the OptiClust
-workflow, where the data was clustered into OTUs and used to tune
-hyperparameters and ultimately train the model. Then, we used OptiFit to
-fit sequence data of samples not part of the original data set into the
-existing OTUs, and used the same model to classify the samples (Figure
-1B). To test how the model performance compared between these two
-approaches, we used a publicly available data set of 16S rRNA gene
+data was clustered into OTUs with the OptiClust algorithm in mothur
+@westcott2017. The resulting abundance data was then split into training
+and testing sets, where the training set was used to tune
+hyperparameters and ultimately train and select the model. The model was
+applied to the testing set and performance evaluated (Figure 1A).
+However, this methodology requires one to regenerate the OTU clusters
+and retrain the model to classify additional samples. The OptiFit
+algorithm @sovacool2022 addresses this problem by enabling new sequences
+to be clustered into existing OTUs. The OptiFit workflow is similar to
+the OptiClust workflow, where the data was clustered into OTUs and used
+to tune hyperparameters and ultimately train the model. Then, we used
+OptiFit to fit sequence data of samples not part of the original dataset
+into the existing OTUs, and used the same model to classify the samples
+(Figure 1B). To test how the model performance compared between these
+two approaches, we used a publicly available dataset of 16S rRNA gene
 sequences from stool samples of healthy subjects (n = 261) as well as
 subjects with SRN consisting of advanced adenoma and carcinoma (n = 229)
-@baxter2016. The data set was randomly split into an 80% train set and
+@baxter2016. The dataset was randomly split into an 80% train set and
 20% test set. For the standard OptiClust workflow, the training and test
@@
 SRN. The test set was classified as either control or SRN using the
-trained models. To account for variation, the data set was randomly
-split 100 times and the process repeated for each of the 100 data
-splits. By comparing the model performance of classifying the samples in
-the test data set between the OptiFit and OptiClust algorithms, we
-quantified the impact of using OptiFit on model classification
-performance.
+trained models. To account for variation, the dataset was randomly split
+100 times and the process repeated for each of the 100 data splits. By
+comparing the model performance of classifying the samples in the test
+dataset between the OptiFit and OptiClust algorithms, we quantified the
+impact of using OptiFit on model classification performance.
 
@@
             sd_mcc = sd(mcc))
+```
+
+```{r frac_mapped,include=F}
+avg_frac_mapped <- read_csv("../results/tables/fracNonMapped.csv",show_col_types = F) %>% 
+  mutate(avg_frac = RefSum/10000) %>% 
+  summarise(mean = mean(avg_frac),
+            sd = sd(avg_frac))
 ```
@@
 to be similar. Since the data was only clustered once in the OptiClust
-workflow there was only one MCC score while the OptiFit workflow
-produced an MCC score for the OTU clusters from each data split.
-Overall, the MCC scores were similar between OptiClust (MCC =
-`r round(opticlust_mcc,digits=3)`) and OptiFit (average MCC =
+workflow there was only one MCC score (MCC =
+`r round(opticlust_mcc,digits=3)`) while the OptiFit workflow produced
+an MCC score for the OTU clusters from each data split (average MCC =
 `r round(optifit_avg_mcc %>% pull(mean_mcc),digits=3)`, standard
-deviation = `r round(optifit_avg_mcc %>% pull(sd_mcc),digits=3)`). This
-indicated that OptiFit performed as well as OptiClust when integrating
+deviation = `r round(optifit_avg_mcc %>% pull(sd_mcc),digits=3)`). As
+expected, the MCC scores were similar between the two workflows. Another
+metric we examined for the OptiFit workflow was the fraction of
+sequences from the test set that mapped to the reference OTUs. Any
+sequences that did not map to reference OTUs were eliminated therefore,
+if a high percentage of reads did not map we might expect this loss of
+data to negatively impact classification performance. We found that loss
+of data was not an issue since on average
+`r format_decimal((avg_frac_mapped %>% pull(mean))*100,digits = 1)`%
+(standard deviation =
+`r format_decimal(avg_frac_mapped %>% pull(sd),digits = 3)`) of
+sequences in the test set mapped to the reference OTUs. These results
+indicate that OptiFit performed as well as OptiClust when integrating
 new sequences into the existing OTUs.
@@
             mean_cv_AUC = round(mean(cv_metric_AUC),digits=3),
+            sd_cv_AUC = round(sd(cv_metric_AUC),digits=3),
             median_AUC = round(median(AUC),digits=3),
-            mean_AUC = round(mean(AUC),digits=3))
+            mean_AUC = round(mean(AUC),digits=3),
+            sd_AUC = round(sd(AUC),digits=3))
 
@@
 OptiClust and OptiFit, we examined the model performance for classifying
-samples in the held out test data set. To quantify model performance, we
+samples in the held out test dataset. To evaluate model performance, we
 used the OTU relative abundances from the training data from the
@@
 data split to quantify model performance. During cross-validation (CV)
-training, the model performance was equivalent between the two
-algorithms (p-value =
-`r format_decimal(pvals %>% filter(metric == "cv_metric_AUC") %>% select(p_value),2)`,
-OptiClust mean CV AUROC =
-`r format_decimal(performance_summary %>% filter(algorithm == "opticlust") %>% select(mean_cv_AUC),3)`,
-OptiFit mean CV AUROC =
-`r format_decimal(performance_summary %>% filter(algorithm == "optifit") %>% select(mean_cv_AUC),3)`;
-Figure 2A). The trained model was then deployed to classify the samples
-of the test data as control or SRN. The performance on the test data was
-equivalent between the two algorithms (p-value =
-`r format_decimal(pvals %>% filter(metric == "AUC") %>% select(p_value),2)`,
-OptiClust mean test AUROC =
-`r format_decimal(performance_summary %>% filter(algorithm == "opticlust") %>% select(mean_AUC),3)`,
-OptiFit mean test AUROC =
-`r format_decimal(performance_summary %>% filter(algorithm == "optifit") %>% select(mean_AUC),3)`;
+training, the model performance was equivalent between the OptiClust and
+OptiFit approaches (p-value =
+`r format_decimal(pvals %>% filter(metric == "cv_metric_AUC") %>% select(p_value),2)`;
+Figure 2A). The trained model was then applied to the test data
+classifying samples as either control or SRN. The performance on the
+test data was equivalent between the OptiClust and OptiFit approaches
+(p-value =
+`r format_decimal(pvals %>% filter(metric == "AUC") %>% select(p_value),2)`;
 Figures 2B and 2C) indicating that new data could be fit to existing OTU
-clusters without impacting model performance.
+clusters using OptiFit without impacting model performance.
 
@@
 clusters will be discarded, resulting in a possible loss of information.
-Regardless, we demonstrated that OptiFit can be used to fit new sequence
-data into existing OTU clusters and performs equally well in predicting
-SRN compared to *de novo* clustering all the sequence data together. The
+However, we demonstrated OptiFit can be used to fit new sequence data
+into existing OTU clusters and performs equally well in predicting SRN
+compared to *de novo* clustering all the sequence data together. The
 ability to integrate data from new samples into existing OTUs enables
-the deployment of a single machine learning model. Further analysis is
-needed to determine the number of samples that are necessary to build a
-robust model capable of classifying diverse samples. A robust machine
-learning model could be implemented as part of a non-invasive and
-low-cost aid in diagnosing SRN and other diseases.
+the implementation of a single machine learning model. This is important
+for model implementation because not all of the data needs to be
+available or known at the time of model generation. Since results may
+depend on the amount of data in the training set, further analysis is
+needed to determine the number of samples that are necessary in training
+to build a robust model capable of classifying diverse samples. A robust
+machine learning model could be implemented as part of a non-invasive
+and low-cost aid in diagnosing SRN and other diseases.
 
@@
 
-***Data Set.*** Raw 16S rRNA gene sequence data isolated from human
-stool samples was downloaded from NCBI Sequence Read Archive (accession
-no. SRP062005) [@edgar2011; @baxter2016]. This data set contains stool
+***Dataset.*** Raw 16S rRNA gene sequence data isolated from human stool
+samples was downloaded from NCBI Sequence Read Archive (accession no.
+SRP062005) [@edgar2011; @baxter2016]. This dataset contains stool
 samples from a total of 490 subjects. For this analysis, samples from
@@
 identified as advanced adenoma or carcinoma were categorized as "screen
-relevant neoplasia" (SRN). The resulting data set consisted of 261
-normal samples and 229 SRN samples.
+relevant neoplasia" (SRN). The resulting dataset consisted of 261 normal
+samples and 229 SRN samples.
 
@@
 
-***Data Processing.*** The full data set was preprocessed with mothur
+***Data Processing.*** The full dataset was preprocessed with mothur
 (v1.47) @schloss2009 to join forward and reverse reads, merge duplicate
@@
 
-***Machine Learning.*** Machine learning using random forest was
-conducted with the R package mikrompl (v 1.2.0) @topçuoglu2021 to
-predict the diagnosis (SRN or normal) for the samples in the test set
-for each data split. The training set was preprocessed to normalize OTU
-counts (scale and center), collapse correlated OTUs, and remove OTUs
-with zero variance. The preprocessing from the training set was then
-applied to the test set. Any OTUs in the test set that were not in the
-training set were removed. P values comparing model performance were
-calculated as previously described @topçuoglu2020. The averaged ROC
-curves were plotted by taking the average and standard deviation of the
-sensitivity at each specificity value.
+***Machine Learning.*** A random forest model was trained with the R
+package mikrompl (v 1.2.0) @topçuoglu2021 to predict the diagnosis (SRN
+or normal) for the samples in the test set for each data split. The
+training set was preprocessed to normalize OTU counts (scale and
+center), collapse correlated OTUs, and remove OTUs with zero variance.
+The preprocessing from the training set was then applied to the test
+set. Any OTUs in the test set that were not in the training set were
+removed. P values comparing model performance were calculated as
+previously described @topçuoglu2020. The averaged ROC curves were
+plotted by taking the average and standard deviation of the sensitivity
+at each specificity value.
 
@@
 
-**Figure 1: Workflows.** **A)** OptiClust workflow: The full data set
-was clustered into OTUs using the OptiClust algorithm in mothur. The
-data was then split into two sets where 80% of the samples were assigned
-to the training set and 20% to the testing set. The training set was
+**Figure 1: Workflows.** **A)** OptiClust workflow: The full dataset was
+clustered into OTUs using the OptiClust algorithm in mothur. The data
+was then split into two sets where 80% of the samples were assigned to
+the training set and 20% to the testing set. The training set was
 preprocessed with mikropml to normalize values (scale and center),
@@
 the full training set. The preprocessing scale from the training set was
-applied to the test data set, then the trained model was used to
-classify the samples in the test set. Based on the actual classification
-and predicted classification, the area under the receiver operating
+applied to the test dataset, then the trained model was used to classify
+the samples in the test set. Based on the actual classification and
+predicted classification, the area under the receiver operating
 characteristic curve (AUROC) was calculated to summarize model
@@
 100 AUROC values summarizing the performance of the standard OptiClust
-workflow. **B)** OptiFit workflow: The data set was first split into two
+workflow. **B)** OptiFit workflow: The dataset was first split into two
 sets where 80% of the samples were assigned to the training set and 20%
@@
 the full training set. The OptiFit algorithm in mothur was used to
-cluster the held out testing data set using the OTUs of the training set
+cluster the held out testing dataset using the OTUs of the training set
 as a reference. The preprocessing scale from the training set was
-applied to the test data set, then the trained model was used to
-classify the samples in the test set. Based on the actual classification
-and predicted classification, the area under the receiver operating
+applied to the test dataset, then the trained model was used to classify
+the samples in the test set. Based on the actual classification and
+predicted classification, the area under the receiver operating
 characteristic curve (AUROC) was calculated to summarize model
@@
 
-![](../exploratory/figures/figure2.pdf)
+![](figures/fig2.png)
 

### 2022-08-31 submission/manuscript.tex
latest paper edits

@@
 \usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Reference-based OTU clustering for machine learning classification},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
 \usepackage[margin=1.0in]{geometry}
@@
 \fi
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
+\hypersetup{
+  pdftitle={Streamlined implementation of a machine learning model to classify screen relevant neoplasia using reference-based OTU clustering},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
 
-\title{\textbf{Reference-based OTU clustering for machine learning
-classification}}
+\title{\textbf{Streamlined implementation of a machine learning model to
+classify screen relevant neoplasia using reference-based OTU
+clustering}}
 \author{}
@@
 similarity of the sequences to each other rather than to a reference
-database. However, OTU assignments depend on the sequences in the data
-set and therefore can change if new data are added. This lack of
-stability complicates classification because in order to use the model
-to classify additional samples, the new sequences must be reclustered
-with the old data and the model must be retrained with the new OTU
-assignments. The new reference-based clustering algorithm, OptiFit,
-addresses this issue by fitting new sequences into existing OTUs. While
-OptiFit can produce high quality OTU clusters, it is unclear whether
-this method for fitting new sequence data into existing OTUs will impact
-the performance of classification models. We used OptiFit to cluster
-additional data into existing OTU clusters and then quantified model
-performance in classifying a data set containing samples from patients
-with and without colonic screen relevant neoplasia (SRN). We compared
-the performance of this model to the standard procedure of \emph{de
-novo} clustering all the data together. We found that both approaches
-performed equally well in classifying SRNs. Moving forward, when OTUs
-are used in classification problems, OptiFit can streamline the process
-of classifying new samples by avoiding the need to retrain models using
-reclustered sequences.
+database. However, such an approach depends on the sequences in the
+dataset and therefore OTU assignments can change if new data are added.
+This lack of stability complicates classification because in order to
+use the model to classify additional samples, the new sequences must be
+reclustered with the old data and the model must be retrained with the
+new OTU assignments. The new reference-based clustering algorithm,
+OptiFit, addresses this issue by fitting new sequences into existing
+OTUs. While OptiFit can produce high quality OTU clusters, it is unclear
+whether this method for fitting new sequence data into existing OTUs
+will impact the performance of classification models. We used OptiFit to
+cluster additional data into existing OTU clusters and then evaluated
+model performance in classifying a dataset containing samples from
+patients with and without colonic screen relevant neoplasia (SRN). We
+compared the performance of this model to the standard procedure of
+\emph{de novo} clustering all the data together. We found that both
+approaches performed equally well in classifying SRNs. Moving forward,
+when OTUs are used in classification problems, OptiFit can streamline
+the process of classifying new samples by avoiding the need to retrain
+models using reclustered sequences.
 
@@
 integrate new data into a validated model without reclustering and
-retraining could allow for deployment of a single model that new data
-can be continually tested against. Recently, Sovacool \emph{et al}
+retraining could allow for the application of a single model that can
+continually classify new data. Recently, Sovacool \emph{et al.}
 introduced OptiFit, a method for fitting new sequence data into existing
@@
 existing OTUs using OptiFit. In the first approach, all the 16S rRNA
-sequence data was \emph{de novo} clustered into OTUs with the OptiClust
-algorithm in mothur (\protect\hyperlink{ref-westcott2017}{5}). The
-resulting abundance data was then split into training and testing sets,
-where the training set was used to tune hyperparameters and ultimately
-train the model. The testing set was then classified with the model and
-the performance of the model was quantified (Figure 1A). However, with
-this methodology, we would have to regenerate the OTU clusters and
-retrain the model if we wanted to classify additional samples. The
-OptiFit algorithm (\protect\hyperlink{ref-sovacool2022}{4}) addresses
-this problem by enabling new sequences to be clustered into existing
-OTUs. The OptiFit workflow is similar to the OptiClust workflow, where
-the data was clustered into OTUs and used to tune hyperparameters and
-ultimately train the model. Then, we used OptiFit to fit sequence data
-of samples not part of the original data set into the existing OTUs, and
-used the same model to classify the samples (Figure 1B). To test how the
-model performance compared between these two approaches, we used a
-publicly available data set of 16S rRNA gene sequences from stool
-samples of healthy subjects (n = 261) as well as subjects with SRN
-consisting of advanced adenoma and carcinoma (n = 229)
-(\protect\hyperlink{ref-baxter2016}{1}). The data set was randomly split
+sequence data was clustered into OTUs with the OptiClust algorithm in
+mothur (\protect\hyperlink{ref-westcott2017}{5}). The resulting
+abundance data was then split into training and testing sets, where the
+training set was used to tune hyperparameters and ultimately train and
+select the model. The model was applied to the testing set and
+performance evaluated (Figure 1A). However, this methodology requires
+one to regenerate the OTU clusters and retrain the model to classify
+additional samples. The OptiFit algorithm
+(\protect\hyperlink{ref-sovacool2022}{4}) addresses this problem by
+enabling new sequences to be clustered into existing OTUs. The OptiFit
+workflow is similar to the OptiClust workflow, where the data was
+clustered into OTUs and used to tune hyperparameters and ultimately
+train the model. Then, we used OptiFit to fit sequence data of samples
+not part of the original dataset into the existing OTUs, and used the
+same model to classify the samples (Figure 1B). To test how the model
+performance compared between these two approaches, we used a publicly
+available dataset of 16S rRNA gene sequences from stool samples of
+healthy subjects (n = 261) as well as subjects with SRN consisting of
+advanced adenoma and carcinoma (n = 229)
+(\protect\hyperlink{ref-baxter2016}{1}). The dataset was randomly split
 into an 80\% train set and 20\% test set. For the standard OptiClust
@@
 classified as either control or SRN using the trained models. To account
-for variation, the data set was randomly split 100 times and the process
+for variation, the dataset was randomly split 100 times and the process
 repeated for each of the 100 data splits. By comparing the model
-performance of classifying the samples in the test data set between the
+performance of classifying the samples in the test dataset between the
 OptiFit and OptiClust algorithms, we quantified the impact of using
@@
 data was only clustered once in the OptiClust workflow there was only
-one MCC score while the OptiFit workflow produced an MCC score for the
-OTU clusters from each data split. Overall, the MCC scores were similar
-between OptiClust (MCC = 0.884) and OptiFit (average MCC = 0.879,
-standard deviation = 0.002). This indicated that OptiFit performed as
-well as OptiClust when integrating new sequences into the existing OTUs.
+one MCC score (MCC = 0.884) while the OptiFit workflow produced an MCC
+score for the OTU clusters from each data split (average MCC = 0.879,
+standard deviation = 0.002). As expected, the MCC scores were similar
+between the two workflows. Another metric we examined for the OptiFit
+workflow was the fraction of sequences from the test set that mapped to
+the reference OTUs. Any sequences that did not map to reference OTUs
+were eliminated therefore, if a high percentage of reads did not map we
+might expect this loss of data to negatively impact classification
+performance. We found that loss of data was not an issue since on
+average 99.9\% (standard deviation = 0.004) of sequences in the test set
+mapped to the reference OTUs. These results indicate that OptiFit
+performed as well as OptiClust when integrating new sequences into the
+existing OTUs.
 
@@
 data split to quantify model performance. During cross-validation (CV)
-training, the model performance was equivalent between the two
-algorithms (p-value = 0.13, OptiClust mean CV AUROC = 0.694, OptiFit
-mean CV AUROC = 0.697; Figure 2A). The trained model was then deployed
-to classify the samples of the test data as control or SRN. The
-performance on the test data was equivalent between the two algorithms
-(p-value = 0.63, OptiClust mean test AUROC = 0.709, OptiFit mean test
-AUROC = 0.712; Figures 2B and 2C) indicating that new data could be fit
-to existing OTU clusters without impacting model performance.
+training, the model performance was equivalent between the OptiClust and
+OptiFit approaches (p-value = 0.13; Figure 2A). The trained model was
+then applied to the test data classifying samples as either control or
+SRN. The performance on the test data was equivalent between the
+OptiClust and OptiFit approaches (p-value = 0.38; Figures 2B and 2C)
+indicating that new data could be fit to existing OTU clusters using
+OptiFit without impacting model performance.
 
@@
 clusters will be discarded, resulting in a possible loss of information.
-Regardless, we demonstrated that OptiFit can be used to fit new sequence
-data into existing OTU clusters and performs equally well in predicting
-SRN compared to \emph{de novo} clustering all the sequence data
-together. The ability to integrate data from new samples into existing
-OTUs enables the deployment of a single machine learning model. Further
-analysis is needed to determine the number of samples that are necessary
+However, we demonstrated OptiFit can be used to fit new sequence data
+into existing OTU clusters and performs equally well in predicting SRN
+compared to \emph{de novo} clustering all the sequence data together.
+The ability to integrate data from new samples into existing OTUs
+enables the implementation of a single machine learning model. This is
+important for model implementation because not all of the data needs to
+be available or known at the time of model generation. Since results may
+depend on the amount of data in the training set, further analysis is
+needed to determine the number of samples that are necessary in training
 to build a robust model capable of classifying diverse samples. A robust
@@
 
-\textbf{\emph{Data Set.}} Raw 16S rRNA gene sequence data isolated from
+\textbf{\emph{Dataset.}} Raw 16S rRNA gene sequence data isolated from
 human stool samples was downloaded from NCBI Sequence Read Archive
 (accession no. SRP062005) (\protect\hyperlink{ref-baxter2016}{1},
-\protect\hyperlink{ref-edgar2011}{6}). This data set contains stool
+\protect\hyperlink{ref-edgar2011}{6}). This dataset contains stool
 samples from a total of 490 subjects. For this analysis, samples from
@@
 identified as advanced adenoma or carcinoma were categorized as ``screen
-relevant neoplasia'' (SRN). The resulting data set consisted of 261
+relevant neoplasia'' (SRN). The resulting dataset consisted of 261
 normal samples and 229 SRN samples.
 
-\textbf{\emph{Data Processing.}} The full data set was preprocessed with
+\textbf{\emph{Data Processing.}} The full dataset was preprocessed with
 mothur (v1.47) (\protect\hyperlink{ref-schloss2009}{7}) to join forward
@@
 
-\textbf{\emph{Machine Learning.}} Machine learning using random forest
-was conducted with the R package mikrompl (v 1.2.0)
+\textbf{\emph{Machine Learning.}} A random forest model was trained with
+the R package mikrompl (v 1.2.0)
 (\protect\hyperlink{ref-topuxe7uoglu2021}{9}) to predict the diagnosis
@@
 \leavevmode\vadjust pre{\hypertarget{ref-baxter2016}{}}%
-\CSLLeftMargin{1. }
+\CSLLeftMargin{1. }%
 \CSLRightInline{\textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers
@@
 \leavevmode\vadjust pre{\hypertarget{ref-zackular2014}{}}%
-\CSLLeftMargin{2. }
+\CSLLeftMargin{2. }%
 \CSLRightInline{\textbf{Zackular JP}, \textbf{Rogers MAM},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-westcott2015}{}}%
-\CSLLeftMargin{3. }
+\CSLLeftMargin{3. }%
 \CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo
@@
 \leavevmode\vadjust pre{\hypertarget{ref-sovacool2022}{}}%
-\CSLLeftMargin{4. }
+\CSLLeftMargin{4. }%
 \CSLRightInline{\textbf{Sovacool KL}, \textbf{Westcott SL},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-westcott2017}{}}%
-\CSLLeftMargin{5. }
+\CSLLeftMargin{5. }%
 \CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2017.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-edgar2011}{}}%
-\CSLLeftMargin{6. }
+\CSLLeftMargin{6. }%
 \CSLRightInline{\textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente
@@
 \leavevmode\vadjust pre{\hypertarget{ref-schloss2009}{}}%
-\CSLLeftMargin{7. }
+\CSLLeftMargin{7. }%
 \CSLRightInline{\textbf{Schloss PD}, \textbf{Westcott SL},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-quast2013}{}}%
-\CSLLeftMargin{8. }
+\CSLLeftMargin{8. }%
 \CSLRightInline{\textbf{Quast C}, \textbf{Pruesse E}, \textbf{Yilmaz P},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-topuxe7uoglu2021}{}}%
-\CSLLeftMargin{9. }
+\CSLLeftMargin{9. }%
 \CSLRightInline{\textbf{Topçuoğlu BD}, \textbf{Lapp Z}, \textbf{Sovacool
@@
 \leavevmode\vadjust pre{\hypertarget{ref-topuxe7uoglu2020}{}}%
-\CSLLeftMargin{10. }
+\CSLLeftMargin{10. }%
 \CSLRightInline{\textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-koster2012}{}}%
-\CSLLeftMargin{11. }
+\CSLLeftMargin{11. }%
 \CSLRightInline{\textbf{Koster J}, \textbf{Rahmann S}. 2012.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-R2020}{}}%
-\CSLLeftMargin{12. }
+\CSLLeftMargin{12. }%
 \CSLRightInline{\textbf{R Core Team}. 2020.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-GNUbash}{}}%
-\CSLLeftMargin{13. }
+\CSLLeftMargin{13. }%
 \CSLRightInline{\textbf{GNU Project}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-RStudio2019}{}}%
-\CSLLeftMargin{14. }
+\CSLLeftMargin{14. }%
 \CSLRightInline{\textbf{RStudio Team}. 2019.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-wickham2019}{}}%
-\CSLLeftMargin{15. }
+\CSLLeftMargin{15. }%
 \CSLRightInline{\textbf{Wickham H}, \textbf{Averick M}, \textbf{Bryan
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xie_r_2018}{}}%
-\CSLLeftMargin{16. }
+\CSLLeftMargin{16. }%
 \CSLRightInline{\textbf{Xie Y}, \textbf{Allaire JJ}, \textbf{Grolemund
@@
 \leavevmode\vadjust pre{\hypertarget{ref-noauthor_sra-tools_nodate}{}}%
-\CSLLeftMargin{17. }
+\CSLLeftMargin{17. }%
 \CSLRightInline{{SRA}-{Tools} - {NCBI}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-noauthor_anaconda_2016}{}}%
-\CSLLeftMargin{18. }
+\CSLLeftMargin{18. }%
 \CSLRightInline{2016. Anaconda {Software Distribution}. Anaconda
@@
 \textbf{Figure 1: Workflows.} \textbf{A)} OptiClust workflow: The full
-data set was clustered into OTUs using the OptiClust algorithm in
-mothur. The data was then split into two sets where 80\% of the samples
-were assigned to the training set and 20\% to the testing set. The
-training set was preprocessed with mikropml to normalize values (scale
-and center), collapse correlated features, and remove features with zero
+dataset was clustered into OTUs using the OptiClust algorithm in mothur.
+The data was then split into two sets where 80\% of the samples were
+assigned to the training set and 20\% to the testing set. The training
+set was preprocessed with mikropml to normalize values (scale and
+center), collapse correlated features, and remove features with zero
 variance. Using mikropml, the training set was split into train and
@@
 train the model with the full training set. The preprocessing scale from
-the training set was applied to the test data set, then the trained
-model was used to classify the samples in the test set. Based on the
-actual classification and predicted classification, the area under the
-receiver operating characteristic curve (AUROC) was calculated to
-summarize model performance. The entire process was repeated 100 times
-to account for variability depending on the split of the data resulting
-in a total of 100 AUROC values summarizing the performance of the
-standard OptiClust workflow. \textbf{B)} OptiFit workflow: The data set
-was first split into two sets where 80\% of the samples were assigned to
-the training set and 20\% to the testing set. The training set was then
-clustered into OTUs using the OptiClust algorithm in mothur. The
-resulting abundance data was preprocessed with mikropml to normalize
-values (scale/center), collapse correlated features, and remove features
-with zero-variance. Using mikropml, the training set was split into
-train and validate sets to compare results using different
-hyperparameter settings. The highest performing hyperparameter setting
-was then used to train the model with the full training set. The OptiFit
-algorithm in mothur was used to cluster the held out testing data set
-using the OTUs of the training set as a reference. The preprocessing
-scale from the training set was applied to the test data set, then the
-trained model was used to classify the samples in the test set. Based on
-the actual classification and predicted classification, the area under
-the receiver operating characteristic curve (AUROC) was calculated to
-summarize model performance. The entire process was repeated 100 times
-to account for variability depending on the split of the data resulting
-in a total of 100 AUROC values summarizing the performance of the new
-OptiFit workflow.
+the training set was applied to the test dataset, then the trained model
+was used to classify the samples in the test set. Based on the actual
+classification and predicted classification, the area under the receiver
+operating characteristic curve (AUROC) was calculated to summarize model
+performance. The entire process was repeated 100 times to account for
+variability depending on the split of the data resulting in a total of
+100 AUROC values summarizing the performance of the standard OptiClust
+workflow. \textbf{B)} OptiFit workflow: The dataset was first split into
+two sets where 80\% of the samples were assigned to the training set and
+20\% to the testing set. The training set was then clustered into OTUs
+using the OptiClust algorithm in mothur. The resulting abundance data
+was preprocessed with mikropml to normalize values (scale/center),
+collapse correlated features, and remove features with zero-variance.
+Using mikropml, the training set was split into train and validate sets
+to compare results using different hyperparameter settings. The highest
+performing hyperparameter setting was then used to train the model with
+the full training set. The OptiFit algorithm in mothur was used to
+cluster the held out testing dataset using the OTUs of the training set
+as a reference. The preprocessing scale from the training set was
+applied to the test dataset, then the trained model was used to classify
+the samples in the test set. Based on the actual classification and
+predicted classification, the area under the receiver operating
+characteristic curve (AUROC) was calculated to summarize model
+performance. The entire process was repeated 100 times to account for
+variability depending on the split of the data resulting in a total of
+100 AUROC values summarizing the performance of the new OptiFit
+workflow.
 
@@
 
-\includegraphics{../exploratory/figures/figure2.pdf}
+\includegraphics{figures/fig2.png}
 

### 2022-08-31 submission/manuscript.Rmd
shorted abstract and running title

@@
 
-Running title: Reference-based OTU clustering for machine learning
-classification
+Running title: Reference-based OTU clustering for ML classification
 
@@
 
-Machine learning classification of disease based on the gut microbiome
+Machine learning classification of disease using the gut microbiome
 often relies on clustering 16S rRNA gene sequences into operational
@@
 of each OTU is then used to train a classification model. The standard
-*de novo* approach to clustering sequences into OTUs leverages the
+de novo approach to clustering sequences into OTUs leverages the
 similarity of the sequences to each other rather than to a reference
@@
 patients with and without colonic screen relevant neoplasia (SRN). We
-compared the performance of this model to the standard procedure of *de
-novo* clustering all the data together. We found that both approaches
-performed equally well in classifying SRNs. Moving forward, when OTUs
-are used in classification problems, OptiFit can streamline the process
-of classifying new samples by avoiding the need to retrain models using
-reclustered sequences.
+compared the performance of this model to the standard procedure of de
+novo clustering all the data together. We found that both approaches
+performed equally well in classifying SRNs. OptiFit can streamline the
+process of classifying new samples by avoiding the need to retrain
+models using reclustered sequences.
 

### 2022-08-31 submission/manuscript.tex
shorted abstract and running title

@@
 of each OTU is then used to train a classification model. The standard
-\emph{de novo} approach to clustering sequences into OTUs leverages the
+de novo approach to clustering sequences into OTUs leverages the
 similarity of the sequences to each other rather than to a reference
@@
 patients with and without colonic screen relevant neoplasia (SRN). We
-compared the performance of this model to the standard procedure of
-\emph{de novo} clustering all the data together. We found that both
-approaches performed equally well in classifying SRNs. Moving forward,
-when OTUs are used in classification problems, OptiFit can streamline
-the process of classifying new samples by avoiding the need to retrain
+compared the performance of this model to the standard procedure of de
+novo clustering all the data together. We found that both approaches
+performed equally well in classifying SRNs. OptiFit can streamline the
+process of classifying new samples by avoiding the need to retrain
 models using reclustered sequences.

### 2022-08-31 submission/manuscript.Rmd
italicize de novo

@@
 of each OTU is then used to train a classification model. The standard
-de novo approach to clustering sequences into OTUs leverages the
+*de novo* approach to clustering sequences into OTUs leverages the
 similarity of the sequences to each other rather than to a reference
@@
 patients with and without colonic screen relevant neoplasia (SRN). We
-compared the performance of this model to the standard procedure of de
-novo clustering all the data together. We found that both approaches
+compared the performance of this model to the standard procedure of *de
+novo* clustering all the data together. We found that both approaches
 performed equally well in classifying SRNs. OptiFit can streamline the

### 2022-10-28 submission/manuscript.tex

@@
 of each OTU is then used to train a classification model. The standard
-de novo approach to clustering sequences into OTUs leverages the
+\emph{de novo} approach to clustering sequences into OTUs leverages the
 similarity of the sequences to each other rather than to a reference
@@
 patients with and without colonic screen relevant neoplasia (SRN). We
-compared the performance of this model to the standard procedure of de
-novo clustering all the data together. We found that both approaches
-performed equally well in classifying SRNs. OptiFit can streamline the
-process of classifying new samples by avoiding the need to retrain
-models using reclustered sequences.
+compared the performance of this model to the standard procedure of
+\emph{de novo} clustering all the data together. We found that both
+approaches performed equally well in classifying SRNs. OptiFit can
+streamline the process of classifying new samples by avoiding the need
+to retrain models using reclustered sequences.
 

### 2023-04-13 submission/response_to_reviewer.Rmd
first round revisions after pat comments

---
output: pdf_document
geometry: margin=1.0in
font-size: 11pt
header-includes:
  - \usepackage{helvet}
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \usepackage{multirow}
  - \usepackage[none]{hyphenat}
---

## Response to Reviewer Comments

Machine learning classification using self-reference-based OTU clustering

Courtney R. Armour, Kelly L. Sovacool, William L. Close, Begüm D. Topçuoğlu, Jenna Wiens, Patrick D. Schloss

**Reviewer #1:**

Armour describe using the closed-reference OTU method OptiRef to predict screen-relevant-neoplasia (SRN) from 16S sequencing data of the fecal microbiome that was originally processed into OTUs by the de novo method OptiClust.

Is this a case study, or something more general? The results here are based on a single dataset, a single set of methods, and a single set of evaluation metrics. Thus, the generalizability of these results is unclear. This concern is exacerbated by the various places in the manuscript that infer generalizable conclusions.

\textcolor{blue}{The purpose of this paper is to provide an example of a scenario when the previously publish OptiFit algorithm could be useful. We are demonstrating here that it is possible to use your own data as a reference for consistent OTU classifications without a needing a reference database. We have clarified this throughout the text.}

The authors need to situate their methods within the universe of methods commonly used in this area. Closed-reference OTUs and ASVs are widely-used alternative approaches to 16S sequencing data analysis that are already used for ML classification. How does the hybrid approach proposed here of de novo OTUs plus closed-reference OTU assignment compare?

\textcolor{blue}{We thank the reviewer for this suggestion and agree that this strengthens the paper. We added the following methods: reference based clustring using OptiFit and the GreenGenes reference, \*de novo\* clustering using VSEARCH, and reference-based clustering using VSEARCH and the GreenGenes reference. See lines 47 - 64 and figure 1 for workflow descriptions and lines 87 - 98 and Figure 2 for the results.}

Is OptiFit only usable with OptiClust as the precursor?

\textcolor{blue}{OptiFit is not dependent on use of OptiClust as the precurson, you could use any algorithm to cluster the reference and then use OptiFit to fit additional data.}

What is the relevance of the MCC evaluation? What if an alternative method had lower MCC but better pred

\textcolor{blue}{The MCC score is a measure of OTU cluster quality based on the similarity of sequences and whether they are appropriately clustered together or not. Since we added additional methods, we can see that some methods do have lower MCC but equal model performance. This likely indicates that the model depends on well clustered OTUs. We've added some discussion on this to the paper (lines: ).}

### 2023-04-13 submission/manuscript.Rmd
first round revisions after pat comments

--- 2022-08-31 submission/manuscript.Rmd
@@
 ---
-title: "**Streamlined implementation of a machine learning model to classify screen relevant neoplasia using reference-based OTU clustering**"
+title: "**Machine learning classification using self-reference-based OTU clustering**"
 output:
@@
 ```
-
-\vspace{5mm}
 
@@
 
-Machine learning classification of disease using the gut microbiome
-often relies on clustering 16S rRNA gene sequences into operational
-taxonomic units (OTUs) to quantify microbial composition. The abundance
-of each OTU is then used to train a classification model. The standard
-de novo approach to clustering sequences into OTUs leverages the
-similarity of the sequences to each other rather than to a reference
-database. However, such an approach depends on the sequences in the
-dataset and therefore OTU assignments can change if new data are added.
-This lack of stability complicates classification because in order to
-use the model to classify additional samples, the new sequences must be
-reclustered with the old data and the model must be retrained with the
-new OTU assignments. The new reference-based clustering algorithm,
-OptiFit, addresses this issue by fitting new sequences into existing
-OTUs. While OptiFit can produce high quality OTU clusters, it is unclear
-whether this method for fitting new sequence data into existing OTUs
-will impact the performance of classification models. We used OptiFit to
-cluster additional data into existing OTU clusters and then evaluated
-model performance in classifying a dataset containing samples from
-patients with and without colonic screen relevant neoplasia (SRN). We
-compared the performance of this model to the standard procedure of de
-novo clustering all the data together. We found that both approaches
-performed equally well in classifying SRNs. OptiFit can streamline the
-process of classifying new samples by avoiding the need to retrain
-models using reclustered sequences.
+Machine learning classification using the gut microbiome relies on
+assigning 16S rRNA gene sequences into operational taxonomic units
+(OTUs) to quantify microbial composition. OTU abundances are then used
+to train a classification model that can be applied to classify new
+samples. The standard approaches to clustering sequences include
+reference-based and *de novo* clustering. Reference-based clustering
+requires a well-curated reference database that may not exist for all
+systems. *De novo* clustering tends to produce higher quality OTU
+assignments than reference-based, but clusters depend on the sequences
+in the dataset and therefore OTU assignments will change when new
+samples are sequenced. This lack of stability complicates machine
+learning classification since new sequences must be reclustered with the
+old data and the model retrained with the new OTU assignments. The
+OptiFit algorithm addresses these issues by fitting new sequences into
+existing OTUs. While OptiFit produces high quality OTU clusters, it is
+unclear whether this method for fitting new sequence data into existing
+OTUs will impact the performance of classification models trained with
+the older data. We used OptiFit to cluster sequences into existing OTUs
+and evaluated model performance in classifying a dataset containing
+samples from patients with and without colonic screen relevant neoplasia
+(SRN). We compared the performance of this model to standard methods
+including *de novo* and database-reference-based clustering. We found
+that using OptiFit performed as well or better in classifying SRNs.
+OptiFit can streamline the process of classifying new samples by
+avoiding the need to retrain models using reclustered sequences.
 
@@
 
-There is great potential for using microbiome data to non-invasively
-diagnose people. One of the challenges with using classification models
-based on the relative abundance of operational taxonomic units (OTUs) is
-that 16S rRNA gene sequences are assigned to OTUs based on their
-similarity to other sequences in a dataset. If data are generated from
-new patients seeking a diagnosis, then a standard approach requires
-reassigning sequences to OTUs and retraining the classification model.
-Yet there is a desire to have a single, validated model that can be
-widely deployed. To overcome this obstacle, we applied the OptiFit
-clustering algorithm which fits new sequence data to existing OTUs and
-allows for the reuse of models. A random forest machine learning model
-implemented using OptiFit performed as well as the traditional
-reassignment and retrain approach. This result indicates that there is
-potential for developing and applying machine learning models based on
-OTU relative abundance data that do not require retraining.
+There is great potential for using microbiome data to aid in diagnosis.
+A challenge with OTU-based classification models is that 16S rRNA gene
+sequences are often assigned to OTUs based on similarity to other
+sequences in the dataset. If data are generated from new patients, the
+old and new sequences must all be reassigned to OTUs and the
+classification model retrained. Yet there is a desire to have a single,
+validated model that can be widely deployed. To overcome this obstacle,
+we applied the OptiFit clustering algorithm to fit new sequence data to
+existing OTUs allowing for reuse of the model. A random forest model
+implemented using OptiFit performed as well as the traditional reassign
+and retrain approach. This result shows that it is possible to train and
+apply machine learning models based on OTU relative abundance data that
+do not require retraining or the use of a reference database.
 
@@
 
-Gut community composition is useful as a resource for machine learning
-classification of diseases, such as colorectal cancer [@baxter2016;
-@zackular2014]. Taxonomic composition of microbial communities can be
-assessed using amplicon sequencing of the 16S rRNA gene, which is the
-input to classification models. Analysis of 16S rRNA gene sequence data
-generally relies on similarity-based clustering of sequences into
+There is increasing evidence for an association between the composition of the gut microbiome and a variety of diseases, such as crohn's disease and colorectal cancer.
+The signatures that associate with disease have shown potential to aid in diagnosis of disease throught gut microbiome profiling and machine learning.
+Taxonomic composition of microbial
+communities can be assessed using amplicon sequencing of the 16S rRNA
+gene, which is the input to classification models. Analysis of 16S rRNA
+gene sequence data generally relies on assigning sequences into
 operational taxonomic units (OTUs). The process of OTU clustering can
@@
 
-We compared two approaches, one using all the data to generate OTU
-clusters, and the other generating *de novo* OTU clusters with a portion
-of the data and then fitting the remaining sequence data to the existing
-OTUs using OptiFit. In the first approach, all the 16S rRNA sequence
-data was clustered into OTUs with the OptiClust algorithm in mothur
-@westcott2017. The resulting abundance data was then split into training
-and testing sets, where the training set was used to tune
-hyperparameters and ultimately train and select the model. The model was
-applied to the testing set and performance evaluated (Figure 1A).
-However, this methodology requires one to regenerate the OTU clusters
-and retrain the model to classify additional samples. The OptiFit
-algorithm @sovacool2022 addresses this problem by enabling new sequences
-to be clustered into existing OTUs. The OptiFit workflow is similar to
-the OptiClust workflow, where the data was clustered into OTUs and used
-to tune hyperparameters and ultimately train the model. Then, we used
-OptiFit to fit sequence data of samples not part of the original dataset
-into the existing OTUs, and used the same model to classify the samples
-(Figure 1B). To test how the model performance compared between these
-two approaches, we used a publicly available dataset of 16S rRNA gene
-sequences from stool samples of healthy subjects (n = 261) as well as
-subjects with SRN consisting of advanced adenoma and carcinoma (n = 229)
-@baxter2016. The dataset was randomly split into an 80% train set and
-20% test set. For the standard OptiClust workflow, the training and test
-sets were *de novo* clustered together into OTUs, then the resulting
-abundance table was split into the training and testing set. For the
-OptiFit workflow, the train set was clustered *de novo* into OTUs, and
-the remaining test set was fit to the OTU clusters using the OptiFit
-algorithm. For both workflows, the abundance table of the train set was
-used to tune hyperparameters and train a random forest model to classify
-SRN. The test set was classified as either control or SRN using the
-trained models. To account for variation, the dataset was randomly split
-100 times and the process repeated for each of the 100 data splits. By
-comparing the model performance of classifying the samples in the test
-dataset between the OptiFit and OptiClust algorithms, we quantified the
-impact of using OptiFit on model classification performance.
+We compared the ability of several approaches for assigning 16S rRNA
+gene seequences to OTUs including, *de novo*, database-reference-based,
+and self-reference-based with OptiFit. To test how the model performance
+compared between these approaches, we used a publicly available dataset
+of 16S rRNA gene sequences from stool samples of healthy subjects (n =
+226) as well as subjects with screen-relevant neoplasia (SRN) consisting
+of advanced adenoma and carcinoma (n = 229) @baxter2016. For the *de
+novo* workflows, all the 16S rRNA sequence data was clustered into OTUs.
+The OTU clustering was conducted using two common algorithms: 1) the
+OptiClust algorithm in mothur @westcott2017 and 2) the VSEARCH algorithm
+used in QIIME2 [@rognes2016; @bolyen2019]. For both algorithms, the
+resulting abundance data was then split into training and testing sets,
+where the training set was used to tune hyperparameters and ultimately
+train and select the model. The model was applied to the testing set and
+performance evaluated (Figure 1A). We also conducted reference-based OTU
+clustering using OptiFit to fit the sequence data into OTUs based on the
+greengenes reference database. To compare with another commonly used
+method, we also used the VSEARCH algorithm to fit the sequence data to
+the greengenes reference (Figure 1B). In the OptiFit self-reference
+workflow, the data was split into a training and a testing set. The
+training set was clustered into OTUs and used to train a classification
+model. The OptiFit algorithm was used to fit sequence data of samples
+not part of the original dataset into the existing OTUs, and used the
+same model to classify the samples (Figure 1C). For each of the
+workflows the process was repeated for 100 random splits of the data to
+account for variation caused by the choice of the random number
+generator seed.
 
-```{r mcc,include=F}
-merged_mcc <- read_csv("../data/learning/summary/merged_MCC.csv")
+```{r mcc, include=F}
+merged_mcc <- read_csv("../results/ml/summary/merged_mcc.csv")
 
-opticlust_mcc <- merged_mcc %>% 
-  filter(algorithm == "opticlust") %>% 
-  pull(mcc)
+opticlust_denovo_mcc <- merged_mcc %>% 
+  filter(algorithm == "opticlust_denovo") %>% 
+  pull(mcc) %>% round(digits=3)
 
-optifit_avg_mcc <- merged_mcc %>% 
-  filter(algorithm == "optifit" & state=="combo") %>% 
+optifit_self_avg_mcc <- merged_mcc %>% 
+  filter(algorithm == "optifit_self" & state=="combo") %>% 
   summarise(mean_mcc = mean(mcc),
-            sd_mcc = sd(mcc))
+            sd_mcc = sd(mcc)) %>% 
+  round(digits=3)
+
+optifit_gg_mcc <- merged_mcc %>% 
+  filter(algorithm == "optifit_gg") %>% 
+  pull(mcc) %>% round(digits=3)
+
+vsearch_denovo_mcc <- merged_mcc %>% 
+  filter(algorithm == "vsearch_denovo") %>% 
+  pull(mcc) %>% round(digits=3)
+
+vsearch_gg_mcc <- merged_mcc %>% 
+  filter(algorithm == "vsearch_gg") %>% 
+  pull(mcc) %>% round(digits=3)
 ```
 
-```{r frac_mapped,include=F}
-avg_frac_mapped <- read_csv("../results/tables/fracNonMapped.csv",show_col_types = F) %>% 
-  mutate(avg_frac = RefSum/10000) %>% 
-  summarise(mean = mean(avg_frac),
-            sd = sd(avg_frac))
+```{r frac_mapped, include=F}
+counts <- read_csv("../results/tables/counts.csv")
+
+avg_frac_mapped <- counts %>%
+  mutate(frac_mapped = count/10000 * 100) %>%
+  group_by(methods,set) %>%
+  summarize(mean_frac = mean(frac_mapped),
+            sd_frac = sd(frac_mapped),
+            .groups="drop") %>% 
+  filter(set == "test") 
 ```
 
-We first examined the quality of the resulting OTU clusters from the two
-algorithms using the Matthews correlation coefficient (MCC). The MCC
-score was quantified by examining all pairs of sequences and assessing
-whether they belonged together in an OTU based on their similarity
-@westcott2017. MCC scores range between negative one and one. A score of
-negative one means none of the sequences in an OTU are within the
-similarity threshold and any sequences within the similarity threshold
-are not in an OTU together. An MCC score of zero means the sequences are
-randomly clustered. An MCC score of 1 means all sequences in an OTU are
-within the similarity threshold and all sequence pairs within the
-similarity threshold are in the same OTU. To ensure that OptiFit is
-appropriately integrating new sequence data into the existing OTUs, we
-expected the MCC scores produced by the OptiClust and OptiFit workflows
-to be similar. Since the data was only clustered once in the OptiClust
-workflow there was only one MCC score (MCC =
-`r round(opticlust_mcc,digits=3)`) while the OptiFit workflow produced
-an MCC score for the OTU clusters from each data split (average MCC =
-`r round(optifit_avg_mcc %>% pull(mean_mcc),digits=3)`, standard
-deviation = `r round(optifit_avg_mcc %>% pull(sd_mcc),digits=3)`). As
-expected, the MCC scores were similar between the two workflows. Another
-metric we examined for the OptiFit workflow was the fraction of
-sequences from the test set that mapped to the reference OTUs. Any
-sequences that did not map to reference OTUs were eliminated therefore,
-if a high percentage of reads did not map we might expect this loss of
-data to negatively impact classification performance. We found that loss
-of data was not an issue since on average
-`r format_decimal((avg_frac_mapped %>% pull(mean))*100,digits = 1)`%
+We first examined the quality of the resulting OTU clusters from each
+method using the Matthews correlation coefficient (MCC). MCC is a metric
+used to measure OTU cluster quality based on the similarity of all pairs
+of sequences and whether they are appropriately clustered or not
+@westcott2015. MCC scores range between negative one and one, and
+measure how well clustering assignment correlates with the distance
+between sequences. To ensure that OptiFit appropriately integrated new
+sequence data into the existing OTUs, we expected the MCC scores
+produced by the OptiFit workflow to be similar to that of *de novo*
+clustering using the OptiClust algorithm. In the OptiFit workflow the
+test data was fit to the clustered training data for each of the 100
+data splits resulting in an MCC score for each split of the data. In the
+remaining workflows, the data was only clustered once and then split
+into the training and testing sets resulting in a single MCC score for
+each method. Indeed, the MCC scores were similar between the OptiClust
+*de novo* (MCC = `r opticlust_denovo_mcc`) and OptiFit self-reference
+workflows (average MCC = `r optifit_self_avg_mcc %>% pull(mean_mcc)`,
+standard deviation = `r optifit_self_avg_mcc %>% pull(sd_mcc)`).
+Previously, we observed that *de novo* clustering tends to produce
+higher MCC scores than reference-based clustering @sovacool2022.
+Consistent with prior findings the reference-based methods produced
+lower MCC scores (OptiFit Greengenes MCC = `r optifit_gg_mcc`; VSEARCH
+Greengenes MCC = `r vsearch_gg_mcc`) than the *de novo* methods with the
+same algorithm (OptiClust *de novo* MCC = `r opticlust_denovo_mcc`;
+VSEARCH *de novo* MCC = `r vsearch_denovo_mcc`). Another metric we
+examined for the OptiFit workflow was the fraction of sequences from the
+test set that mapped to the reference OTUs. Since sequences that did not
+map to reference OTUs were eliminated, if a high percentage of reads did
+not map to an OTU we expected this loss of data to negatively impact
+classification performance. We found that loss of data was not an issue
+since on average
+`r avg_frac_mapped %>% filter(methods == "optifit_self", set == "test") %>% pull(mean_frac) %>% round(digits=1)`%
 (standard deviation =
-`r format_decimal(avg_frac_mapped %>% pull(sd),digits = 3)`) of
-sequences in the test set mapped to the reference OTUs. These results
-indicate that OptiFit performed as well as OptiClust when integrating
-new sequences into the existing OTUs.
+`r avg_frac_mapped %>% filter(methods == "optifit_self", set == "test") %>% pull(sd_frac) %>% round(digits=2)`%)
+of sequences in the test set mapped to the reference OTUs. This number
+is higher than the average fraction of reads mapped in the OptiFit
+Greengenes workflow (
+`r avg_frac_mapped %>% filter(methods == "optifit_gg",set == "test") %>% pull(mean_frac) %>% round(digits=1)`%
++/-
+`r avg_frac_mapped %>% filter(methods == "optifit_gg", set == "test") %>% pull(sd_frac) %>% round(digits=1)`).
+These results indicate that the OptiFit self-reference method performed
+as well as the OptiClust *de novo* method and better than using an
+external database.
 
-```{r performance,include=F}
-performance <- read_csv("../data/learning/summary/merged_performance.csv",
+```{r}
+performance <- read_csv("../results/ml/summary/merged_performance.csv",
                     col_types = cols(algorithm=col_character(),
@@
 pvals <- read_csv("../results/tables/pvalues.csv",col_types = cols(p_value = col_double(),
-                                                                   .default = col_character())) %>% 
-  select(metric,p_value)
+                                                                   .default = col_character()))
+
+get_pval <- function(pvals,met,g1,g2){
+  pvals %>% 
+    filter(metric == met) %>% 
+    filter(group1 == g1 & group2 == g2 | group1 == g2 & group2 == g1) %>% 
+    pull(p_value) %>% round(digits = 3)
+}
 ```
 
-After verifying that the quality of the OTUs was consistent between
-OptiClust and OptiFit, we examined the model performance for classifying
-samples in the held out test dataset. To evaluate model performance, we
-used the OTU relative abundances from the training data from the
-OptiClust and OptiFit workflows to train a model to predict SRNs. Using
-the predicted and actual diagnosis classification, we calculated the
-area under the receiver operating characteristic curve (AUROC) for each
-data split to quantify model performance. During cross-validation (CV)
-training, the model performance was equivalent between the OptiClust and
-OptiFit approaches (p-value =
-`r format_decimal(pvals %>% filter(metric == "cv_metric_AUC") %>% select(p_value),2)`;
-Figure 2A). The trained model was then applied to the test data
-classifying samples as either control or SRN. The performance on the
-test data was equivalent between the OptiClust and OptiFit approaches
-(p-value =
-`r format_decimal(pvals %>% filter(metric == "AUC") %>% select(p_value),2)`;
-Figures 2B and 2C) indicating that new data could be fit to existing OTU
+We next assessed model performance using OTU relative abundances from
+the training data from the workflows to train a model to predict SRNs
+and used the model on the held out data. Using the predicted and actual
+diagnosis classification, we calculated the area under the receiver
+operating characteristic curve (AUROC) for each data split. During
+cross-validation (CV) training, the model performance was equivalent
+between OptFit self-reference and OptiClust *de novo* (p-value =
+`r get_pval(pvals, "cv_metric_AUC", "opticlust_denovo", "optifit_self")`;
+Figure 2A) while while performance for both VSEARCH methods was lower
+than the OptiClust *de novo*, OptiFit self, and OptiFit Greengenes
+methods (p-values \< 0.05). The trained model was then applied to the
+test data classifying samples as either control or SRN. Both VSEARCH
+methods perform slightly worse than the OptiClust \*de novo\* method
+(both p-values \< 0.05). However the performance on the test data was
+equivalent between the OptiClust *de novo*, OptiFit Greengenes, and
+OptiFit self-reference approaches (p-value \> 0.05; Figures 2B and 2C).
+These results indicate that new data could be fit to existing OTU
 clusters using OptiFit without impacting model performance.
@@
 clusters will be discarded, resulting in a possible loss of information.
-However, we demonstrated OptiFit can be used to fit new sequence data
-into existing OTU clusters and performs equally well in predicting SRN
-compared to *de novo* clustering all the sequence data together. The
-ability to integrate data from new samples into existing OTUs enables
-the implementation of a single machine learning model. This is important
-for model implementation because not all of the data needs to be
-available or known at the time of model generation. Since results may
-depend on the amount of data in the training set, further analysis is
-needed to determine the number of samples that are necessary in training
-to build a robust model capable of classifying diverse samples. A robust
-machine learning model could be implemented as part of a non-invasive
-and low-cost aid in diagnosing SRN and other diseases.
+However, we demonstrated that OptiFit can be used to fit new sequence
+data into existing OTU clusters and it could perform as well in
+predicting SRN compared to *de novo* clustering all the sequence data
+together. In this instance, the performance of OptiFit was equivalent to
+using a database-reference-based method despite the lower quality of the
+OTU clusters in the database-reference-based approach. This likely
+indicates that the sequences that are important to the model are well
+characterized by the reference database. However, a less well studied
+system may not be as well characterized by a reference-database which
+would make the ability to utilize one's own data a reference an exciting
+possiblility. The ability to integrate data from new samples into
+existing OTUs enables the implementation of a single machine learning
+model. This is important for model implementation because not all of the
+data needs to be available or known at the time of model generation. A
+robust machine learning model can be implemented as part of a
+non-invasive and low-cost diagnostic for SRN and other diseases.
 
@@
 
-***Dataset.*** Raw 16S rRNA gene sequence data isolated from human stool
-samples was downloaded from NCBI Sequence Read Archive (accession no.
-SRP062005) [@edgar2011; @baxter2016]. This dataset contains stool
-samples from a total of 490 subjects. For this analysis, samples from
-subjects identified in the metadata as normal, high risk normal, or
-adenoma were categorized as "normal", while samples from subjects
-identified as advanced adenoma or carcinoma were categorized as "screen
-relevant neoplasia" (SRN). The resulting dataset consisted of 261 normal
-samples and 229 SRN samples.
+**Dataset.** Raw 16S rRNA gene sequence data from the V4 region were
+previously generated from human stool samples. Sequences were downloaded
+from the NCBI Sequence Read Archive (accession no. SRP062005)
+[@edgar2011; @baxter2016]. This dataset contains stool samples from 490
+subjects. For this analysis, samples from subjects identified in the
+metadata as normal, high risk normal, or adenoma were categorized as
+"normal", while samples from subjects identified as advanced adenoma or
+carcinoma were categorized as "screen relevant neoplasia" (SRN). The
+resulting dataset consisted of 261 normal samples and 229 SRN samples.
 
 ```{r n_test_train,include=F}
-n_test_train <- read_csv("../results/tables/splitTogetherFrequency.csv") %>% 
-  mutate(n_test = rowSums(across(starts_with("split_")))) %>% 
-  select(Group,n_test) %>% 
-  mutate(n_train = 100 - n_test) %>% 
-  summarize(avg_test = mean(n_test),
-            sd_test = sd(n_test),
-            avg_train = mean(n_train),
-            sd_train = sd(n_train))
+n_test_train <- read_csv("../results/tables/train_test_frequency.csv") %>% 
+  pivot_wider(names_from="train_test",values_from = "n") %>% 
+  summarize(avg_test = mean(test),
+            sd_test = sd(test),
+            avg_train = mean(train),
+            sd_train = sd(train))
 ```
 
-***Data Processing.*** The full dataset was preprocessed with mothur
+**Data processing.** The full dataset was preprocessed with mothur
 (v1.47) @schloss2009 to join forward and reverse reads, merge duplicate
@@
 (standard deviation =
-`r format_decimal(n_test_train %>% pull(sd_test),digits = 1)`).
+`r format_decimal(n_test_train %>% pull(sd_test),digits = 1)`\`).
 
-The data was processed through two workflows. In the control workflow,
-all the data was clustered together with OptiClust @westcott2017 to
-generate OTUs and the resulting abundance tables were split into the
-training and testing sets. In the experimental workflow, the
-preprocessed data was split into the training and testing sets. The
-training set was clustered into OTUs using OptiClust, then the test set
-was fit to the OTUs of the training set using the OptiFit algorithm
-@sovacool2022. The OptiFit algorithm was run with method open so that
-any sequences that did not map to the existing OTU clusters would form
-new OTUs. For both pathways, the shared files were sub-sampled to 10,000
-reads per sample.
+***Reference-based workflows.***
+
+1.  OptiFit Self: The preprocess data was split into the training and
+    testing sets. The training set was clustered into OTUs using
+    OptiClust, then the test set was fit to the OTUs of the training set
+    using the OptiFit algorithm @sovacool2022. The OptiFit algorithm
+    was run with method open so that any sequences that did not map to
+    the existing OTU clusters would form new OTUs. The data was then
+    subsampled to 10,000 reads and any novel OTUs from the test set were
+    removed. This process was repeated for each of the 100 splits
+    resulting in 100 training and testing datasets.
+2.  OptiFit Greengenes: Reference sequences from the Greengenes database
+    v13_8\_99 @desantis2006 were downloaded and processed with mothur by
+    trimming to the V4 region and clustered *de novo* with OptiClust
+    @westcott2017. The preprocessed data was fit to the clustered
+    reference data using OptiFit with the method open to allow any
+    sequences that did not map to the existing reference clusters would
+    form new OTUs. The data was then subsampled to 10,000 reads and any
+    novel OTUs from the test set were removed. The data set was then
+    split into two sets where 80% of the samples were assigned to the
+    training set and 20% to the testing set. This process was repeated
+    for each of the 100 splits resulting in 100 training and testing
+    datasets.
+3.  VSEARCH Greengenes: Preprocessed data was clustered using VSEARCH
+    v2.15.2 @rognes2016 directly to unprocessed Greengenes 97% OTU
+    reference alignment consistent with how VSEARCH is typically used by
+    the QIIME2 software for reference-based clustering @bolyen2019. The
+    data was then subsampled to 10,000 reads and any novel OTUs from the
+    test set were removed. The data set was then split into two sets
+    where 80% of the samples were assigned to the training set and 20%
+    to the testing set. This process was repeated for each of the 100
+    splits resulting in 100 training and testing datasets.
+
+***De novo workflows.***
+
+4.  OptiClust *de novo*: All the preprocessed data was clustered
+    together with OptiClust @westcott2017 to generate OTUs. The data was
+    subsampled to 10,000 reads per sample and the resulting abundance
+    tables were split into the training and testing sets. The process
+    was repeated for each of the 100 splits resulting in 100 training
+    and testing datasets.
+
+5.  VSEARCH *de novo*: All the preprocessed data was clustered using
+    VSEARCH v2.15.2 @rognes2016 with 97% identity and then subsampled to
+    10,000 reads per sample. The process was repeated for each of the
+    100 splits resulting in 100 training and testing datasets for both
+    workflows.
 
@@
 
-***Code Availability.*** The analysis workflow was implemented in
-Snakemake @koster2012 . Scripts for analysis were written in R @R2020
-and GNU bash @GNUbash. The software used includes mothur v1.47.0
-@schloss2009, RStudio @RStudio2019, the Tidyverse metapackage
+***Code Availability.***
+
+The analysis workflow was implemented in Snakemake @koster2012. Scripts
+for analysis were written in R @R2020 and GNU bash @GNUbash. The
+software used includes mothur v1.47.0 @schloss2009, VSEARCH v2.15.2
+@rognes2016, RStudio @RStudio2019, the Tidyverse metapackage
 @wickham2019, R Markdown @xie_r_2018, the SRA toolkit
@@
 are available at:
-<https://github.com/SchlossLab/Armour_OptiFitGLNE_XXXX_2021>
+<https://github.com/SchlossLab/Armour_OptiFitGLNE_mBio_2023>
 
-## Acknowledgements
+## Acknowledgments
 
@@
 
-![](../exploratory/figures/figure1_a.pdf)
+![](./figures/Figure1.png)
 
-![](../exploratory/figures/figure1_b.pdf)
-
-**Figure 1: Workflows.** **A)** OptiClust workflow: The full dataset was
-clustered into OTUs using the OptiClust algorithm in mothur. The data
-was then split into two sets where 80% of the samples were assigned to
-the training set and 20% to the testing set. The training set was
-preprocessed with mikropml to normalize values (scale and center),
-collapse correlated features, and remove features with zero variance.
-Using mikropml, the training set was split into train and validate sets
-to compare results using different hyperparameter settings. The highest
-performing hyperparameter setting was then used to train the model with
-the full training set. The preprocessing scale from the training set was
-applied to the test dataset, then the trained model was used to classify
-the samples in the test set. Based on the actual classification and
-predicted classification, the area under the receiver operating
-characteristic curve (AUROC) was calculated to summarize model
-performance. The entire process was repeated 100 times to account for
-variability depending on the split of the data resulting in a total of
-100 AUROC values summarizing the performance of the standard OptiClust
-workflow. **B)** OptiFit workflow: The dataset was first split into two
-sets where 80% of the samples were assigned to the training set and 20%
-to the testing set. The training set was then clustered into OTUs using
-the OptiClust algorithm in mothur. The resulting abundance data was
-preprocessed with mikropml to normalize values (scale/center), collapse
-correlated features, and remove features with zero-variance. Using
-mikropml, the training set was split into train and validate sets to
-compare results using different hyperparameter settings. The highest
-performing hyperparameter setting was then used to train the model with
-the full training set. The OptiFit algorithm in mothur was used to
-cluster the held out testing dataset using the OTUs of the training set
-as a reference. The preprocessing scale from the training set was
-applied to the test dataset, then the trained model was used to classify
-the samples in the test set. Based on the actual classification and
-predicted classification, the area under the receiver operating
-characteristic curve (AUROC) was calculated to summarize model
-performance. The entire process was repeated 100 times to account for
-variability depending on the split of the data resulting in a total of
-100 AUROC values summarizing the performance of the new OptiFit
-workflow.
+**Figure1: Overview of clustering workflows.** The *de novo* and
+database-reference-based workflows were conducted using two approaches:
+OptiClust with mothur and VSEARCH as is used in the QIIME pipeline.
 
@@
 
-**Figure 2: Model Performance.** **A)** Area under the receiver
-operating characteristic (AUROC) curve during cross-validation for the
-OptiClust and OptiFit workflows. Mean and standard deviation of the
-AUROC is represented by the black dot and whiskers. Mean AUROC is
-printed to the right of the points. **B)** AUROC on the test data for
-the OptiClust and OptiFit workflows. Mean and standard deviation of the
-AUROC is represented by the black dot and whiskers. The mean AUROC is
-printed to the right of the points. **C)** Averaged receiver operating
-characteristic (ROC) curves.
+**Figure 2: Model perfomance of OptiFit self-reference workflow is as
+good or better than other methods.** **A)** Area under the receiver
+operating characteristic (AUROC) curve during cross-validation (train)
+for the various workflows. **B)** AUROC on the test data for the various
+workflows. The mean and standard deviation of the AUROC is represented
+by the black dot and whiskers in panels A and B. The mean AUROC is
+printed below the points. **C)** Averaged receiver operating
+characteristic (ROC) curves. Lines represent the average true positive
+rate for the range of false positive rates.

### 2023-04-13 submission/manuscript.tex
first round revisions after pat comments

--- 2022-08-31 submission/manuscript.tex
@@
 \hypersetup{
-  pdftitle={Streamlined implementation of a machine learning model to classify screen relevant neoplasia using reference-based OTU clustering},
+  pdftitle={Machine learning classification using self-reference-based OTU clustering},
   hidelinks,
@@
 
-\title{\textbf{Streamlined implementation of a machine learning model to
-classify screen relevant neoplasia using reference-based OTU
-clustering}}
+\title{\textbf{Machine learning classification using
+self-reference-based OTU clustering}}
 \author{}
@@
 \maketitle
-
-\vspace{5mm}
 
@@
 
-Machine learning classification of disease using the gut microbiome
-often relies on clustering 16S rRNA gene sequences into operational
-taxonomic units (OTUs) to quantify microbial composition. The abundance
-of each OTU is then used to train a classification model. The standard
-de novo approach to clustering sequences into OTUs leverages the
-similarity of the sequences to each other rather than to a reference
-database. However, such an approach depends on the sequences in the
-dataset and therefore OTU assignments can change if new data are added.
-This lack of stability complicates classification because in order to
-use the model to classify additional samples, the new sequences must be
-reclustered with the old data and the model must be retrained with the
-new OTU assignments. The new reference-based clustering algorithm,
-OptiFit, addresses this issue by fitting new sequences into existing
-OTUs. While OptiFit can produce high quality OTU clusters, it is unclear
-whether this method for fitting new sequence data into existing OTUs
-will impact the performance of classification models. We used OptiFit to
-cluster additional data into existing OTU clusters and then evaluated
-model performance in classifying a dataset containing samples from
-patients with and without colonic screen relevant neoplasia (SRN). We
-compared the performance of this model to the standard procedure of de
-novo clustering all the data together. We found that both approaches
-performed equally well in classifying SRNs. OptiFit can streamline the
-process of classifying new samples by avoiding the need to retrain
-models using reclustered sequences.
+Machine learning classification using the gut microbiome relies on
+assigning 16S rRNA gene sequences into operational taxonomic units
+(OTUs) to quantify microbial composition. OTU abundances are then used
+to train a classification model that can be applied to classify new
+samples. The standard approaches to clustering sequences include
+reference-based and \emph{de novo} clustering. Reference-based
+clustering requires a well-curated reference database that may not exist
+for all systems. \emph{De novo} clustering tends to produce higher
+quality OTU assignments than reference-based, but clusters depend on the
+sequences in the dataset and therefore OTU assignments will change when
+new samples are sequenced. This lack of stability complicates machine
+learning classification since new sequences must be reclustered with the
+old data and the model retrained with the new OTU assignments. The
+OptiFit algorithm addresses these issues by fitting new sequences into
+existing OTUs. While OptiFit produces high quality OTU clusters, it is
+unclear whether this method for fitting new sequence data into existing
+OTUs will impact the performance of classification models trained with
+the older data. We used OptiFit to cluster sequences into existing OTUs
+and evaluated model performance in classifying a dataset containing
+samples from patients with and without colonic screen relevant neoplasia
+(SRN). We compared the performance of this model to standard methods
+including \emph{de novo} and database-reference-based clustering. We
+found that using OptiFit performed as well or better in classifying
+SRNs. OptiFit can streamline the process of classifying new samples by
+avoiding the need to retrain models using reclustered sequences.
 
@@
 
-Gut community composition is useful as a resource for machine learning
-classification of diseases, such as colorectal cancer
-(\protect\hyperlink{ref-baxter2016}{1},
-\protect\hyperlink{ref-zackular2014}{2}). Taxonomic composition of
-microbial communities can be assessed using amplicon sequencing of the
-16S rRNA gene, which is the input to classification models. Analysis of
-16S rRNA gene sequence data generally relies on similarity-based
-clustering of sequences into operational taxonomic units (OTUs). The
-process of OTU clustering can either be reference-based or \emph{de
-novo}. The quality of OTUs generated with reference-based clustering is
-generally poor compared to those generated with \emph{de novo}
-clustering (\protect\hyperlink{ref-westcott2015}{3}). While \emph{de
-novo} clustering produces high-quality OTU clusters where sequences are
+There is increasing evidence for an association between the composition
+of the gut microbiome and a variety of diseases, such as crohn's disease
+and colorectal cancer. The signatures that associate with disease have
+shown potential to aid in diagnosis of disease throught gut microbiome
+profiling and machine learning. Taxonomic composition of microbial
+communities can be assessed using amplicon sequencing of the 16S rRNA
+gene, which is the input to classification models. Analysis of 16S rRNA
+gene sequence data generally relies on assigning sequences into
+operational taxonomic units (OTUs). The process of OTU clustering can
+either be reference-based or \emph{de novo}. The quality of OTUs
+generated with reference-based clustering is generally poor compared to
+those generated with \emph{de novo} clustering
+(\protect\hyperlink{ref-westcott2015}{1}). While \emph{de novo}
+clustering produces high-quality OTU clusters where sequences are
 accurately grouped based on similarity thresholds, the resulting OTU
@@
 introduced OptiFit, a method for fitting new sequence data into existing
-OTUs (\protect\hyperlink{ref-sovacool2022}{4}). While OptiFit can
+OTUs (\protect\hyperlink{ref-sovacool2022}{2}). While OptiFit can
 effectively fit new sequence data to existing OTU clusters, it is
@@
 
-We compared two approaches, one using all the data to generate OTU
-clusters, and the other generating \emph{de novo} OTU clusters with a
-portion of the data and then fitting the remaining sequence data to the
-existing OTUs using OptiFit. In the first approach, all the 16S rRNA
-sequence data was clustered into OTUs with the OptiClust algorithm in
-mothur (\protect\hyperlink{ref-westcott2017}{5}). The resulting
-abundance data was then split into training and testing sets, where the
-training set was used to tune hyperparameters and ultimately train and
-select the model. The model was applied to the testing set and
-performance evaluated (Figure 1A). However, this methodology requires
-one to regenerate the OTU clusters and retrain the model to classify
-additional samples. The OptiFit algorithm
-(\protect\hyperlink{ref-sovacool2022}{4}) addresses this problem by
-enabling new sequences to be clustered into existing OTUs. The OptiFit
-workflow is similar to the OptiClust workflow, where the data was
-clustered into OTUs and used to tune hyperparameters and ultimately
-train the model. Then, we used OptiFit to fit sequence data of samples
+We compared the ability of several approaches for assigning 16S rRNA
+gene seequences to OTUs including, \emph{de novo},
+database-reference-based, and self-reference-based with OptiFit. To test
+how the model performance compared between these approaches, we used a
+publicly available dataset of 16S rRNA gene sequences from stool samples
+of healthy subjects (n = 226) as well as subjects with screen-relevant
+neoplasia (SRN) consisting of advanced adenoma and carcinoma (n = 229)
+(\protect\hyperlink{ref-baxter2016}{3}). For the \emph{de novo}
+workflows, all the 16S rRNA sequence data was clustered into OTUs. The
+OTU clustering was conducted using two common algorithms: 1) the
+OptiClust algorithm in mothur 1 and 2) the VSEARCH algorithm used in
+QIIME2 (\protect\hyperlink{ref-rognes2016}{4},
+\protect\hyperlink{ref-bolyen2019}{5}). For both algorithms, the
+resulting abundance data was then split into training and testing sets,
+where the training set was used to tune hyperparameters and ultimately
+train and select the model. The model was applied to the testing set and
+performance evaluated (Figure 1A). We also conducted reference-based OTU
+clustering using OptiFit to fit the sequence data into OTUs based on the
+greengenes reference database. To compare with another commonly used
+method, we also used the VSEARCH algorithm to fit the sequence data to
+the greengenes reference (Figure 1B). In the OptiFit self-reference
+workflow, the data was split into a training and a testing set. The
+training set was clustered into OTUs and used to train a classification
+model. The OptiFit algorithm was used to fit sequence data of samples
 not part of the original dataset into the existing OTUs, and used the
-same model to classify the samples (Figure 1B). To test how the model
-performance compared between these two approaches, we used a publicly
-available dataset of 16S rRNA gene sequences from stool samples of
-healthy subjects (n = 261) as well as subjects with SRN consisting of
-advanced adenoma and carcinoma (n = 229)
-(\protect\hyperlink{ref-baxter2016}{1}). The dataset was randomly split
-into an 80\% train set and 20\% test set. For the standard OptiClust
-workflow, the training and test sets were \emph{de novo} clustered
-together into OTUs, then the resulting abundance table was split into
-the training and testing set. For the OptiFit workflow, the train set
-was clustered \emph{de novo} into OTUs, and the remaining test set was
-fit to the OTU clusters using the OptiFit algorithm. For both workflows,
-the abundance table of the train set was used to tune hyperparameters
-and train a random forest model to classify SRN. The test set was
-classified as either control or SRN using the trained models. To account
-for variation, the dataset was randomly split 100 times and the process
-repeated for each of the 100 data splits. By comparing the model
-performance of classifying the samples in the test dataset between the
-OptiFit and OptiClust algorithms, we quantified the impact of using
-OptiFit on model classification performance.
+same model to classify the samples (Figure 1C). For each of the
+workflows the process was repeated for 100 random splits of the data to
+account for variation caused by the choice of the random number
+generator seed.
 
-We first examined the quality of the resulting OTU clusters from the two
-algorithms using the Matthews correlation coefficient (MCC). The MCC
-score was quantified by examining all pairs of sequences and assessing
-whether they belonged together in an OTU based on their similarity
-(\protect\hyperlink{ref-westcott2017}{5}). MCC scores range between
-negative one and one. A score of negative one means none of the
-sequences in an OTU are within the similarity threshold and any
-sequences within the similarity threshold are not in an OTU together. An
-MCC score of zero means the sequences are randomly clustered. An MCC
-score of 1 means all sequences in an OTU are within the similarity
-threshold and all sequence pairs within the similarity threshold are in
-the same OTU. To ensure that OptiFit is appropriately integrating new
-sequence data into the existing OTUs, we expected the MCC scores
-produced by the OptiClust and OptiFit workflows to be similar. Since the
-data was only clustered once in the OptiClust workflow there was only
-one MCC score (MCC = 0.884) while the OptiFit workflow produced an MCC
-score for the OTU clusters from each data split (average MCC = 0.879,
-standard deviation = 0.002). As expected, the MCC scores were similar
-between the two workflows. Another metric we examined for the OptiFit
-workflow was the fraction of sequences from the test set that mapped to
-the reference OTUs. Any sequences that did not map to reference OTUs
-were eliminated therefore, if a high percentage of reads did not map we
-might expect this loss of data to negatively impact classification
-performance. We found that loss of data was not an issue since on
-average 99.9\% (standard deviation = 0.004) of sequences in the test set
-mapped to the reference OTUs. These results indicate that OptiFit
-performed as well as OptiClust when integrating new sequences into the
-existing OTUs.
+We first examined the quality of the resulting OTU clusters from each
+method using the Matthews correlation coefficient (MCC). MCC is a metric
+used to measure OTU cluster quality based on the similarity of all pairs
+of sequences and whether they are appropriately clustered or not
+(\protect\hyperlink{ref-westcott2015}{1}). MCC scores range between
+negative one and one, and measure how well clustering assignment
+correlates with the distance between sequences. To ensure that OptiFit
+appropriately integrated new sequence data into the existing OTUs, we
+expected the MCC scores produced by the OptiFit workflow to be similar
+to that of \emph{de novo} clustering using the OptiClust algorithm. In
+the OptiFit workflow the test data was fit to the clustered training
+data for each of the 100 data splits resulting in an MCC score for each
+split of the data. In the remaining workflows, the data was only
+clustered once and then split into the training and testing sets
+resulting in a single MCC score for each method. Indeed, the MCC scores
+were similar between the OptiClust \emph{de novo} (MCC = 0.884) and
+OptiFit self-reference workflows (average MCC = 0.879, standard
+deviation = 0.002). Previously, we observed that \emph{de novo}
+clustering tends to produce higher MCC scores than reference-based
+clustering (\protect\hyperlink{ref-sovacool2022}{2}). Consistent with
+prior findings the reference-based methods produced lower MCC scores
+(OptiFit Greengenes MCC = 0.786; VSEARCH Greengenes MCC = 0.531) than
+the \emph{de novo} methods with the same algorithm (OptiClust \emph{de
+novo} MCC = 0.884; VSEARCH \emph{de novo} MCC = 0.641). Another metric
+we examined for the OptiFit workflow was the fraction of sequences from
+the test set that mapped to the reference OTUs. Since sequences that did
+not map to reference OTUs were eliminated, if a high percentage of reads
+did not map to an OTU we expected this loss of data to negatively impact
+classification performance. We found that loss of data was not an issue
+since on average 99.8\% (standard deviation = 0.68\%) of sequences in
+the test set mapped to the reference OTUs. This number is higher than
+the average fraction of reads mapped in the OptiFit Greengenes workflow
+( 96.8\% +/- 3.5). These results indicate that the OptiFit
+self-reference method performed as well as the OptiClust \emph{de novo}
+method and better than using an external database.
 
-After verifying that the quality of the OTUs was consistent between
-OptiClust and OptiFit, we examined the model performance for classifying
-samples in the held out test dataset. To evaluate model performance, we
-used the OTU relative abundances from the training data from the
-OptiClust and OptiFit workflows to train a model to predict SRNs. Using
-the predicted and actual diagnosis classification, we calculated the
-area under the receiver operating characteristic curve (AUROC) for each
-data split to quantify model performance. During cross-validation (CV)
-training, the model performance was equivalent between the OptiClust and
-OptiFit approaches (p-value = 0.13; Figure 2A). The trained model was
+We next assessed model performance using OTU relative abundances from
+the training data from the workflows to train a model to predict SRNs
+and used the model on the held out data. Using the predicted and actual
+diagnosis classification, we calculated the area under the receiver
+operating characteristic curve (AUROC) for each data split. During
+cross-validation (CV) training, the model performance was equivalent
+between OptFit self-reference and OptiClust \emph{de novo} (p-value =
+0.071; Figure 2A) while while performance for both VSEARCH methods was
+lower than the OptiClust \emph{de novo}, OptiFit self, and OptiFit
+Greengenes methods (p-values \textless{} 0.05). The trained model was
 then applied to the test data classifying samples as either control or
-SRN. The performance on the test data was equivalent between the
-OptiClust and OptiFit approaches (p-value = 0.38; Figures 2B and 2C)
-indicating that new data could be fit to existing OTU clusters using
-OptiFit without impacting model performance.
+SRN. Both VSEARCH methods perform slightly worse than the OptiClust *de
+novo* method (both p-values \textless{} 0.05). However the performance
+on the test data was equivalent between the OptiClust \emph{de novo},
+OptiFit Greengenes, and OptiFit self-reference approaches (p-value
+\textgreater{} 0.05; Figures 2B and 2C). These results indicate that new
+data could be fit to existing OTU clusters using OptiFit without
+impacting model performance.
 
@@
 clusters will be discarded, resulting in a possible loss of information.
-However, we demonstrated OptiFit can be used to fit new sequence data
-into existing OTU clusters and performs equally well in predicting SRN
-compared to \emph{de novo} clustering all the sequence data together.
-The ability to integrate data from new samples into existing OTUs
-enables the implementation of a single machine learning model. This is
-important for model implementation because not all of the data needs to
-be available or known at the time of model generation. Since results may
-depend on the amount of data in the training set, further analysis is
-needed to determine the number of samples that are necessary in training
-to build a robust model capable of classifying diverse samples. A robust
-machine learning model could be implemented as part of a non-invasive
-and low-cost aid in diagnosing SRN and other diseases.
+However, we demonstrated that OptiFit can be used to fit new sequence
+data into existing OTU clusters and it could perform as well in
+predicting SRN compared to \emph{de novo} clustering all the sequence
+data together. In this instance, the performance of OptiFit was
+equivalent to using a database-reference-based method despite the lower
+quality of the OTU clusters in the database-reference-based approach.
+This likely indicates that the sequences that are important to the model
+are well characterized by the reference database. However, a less well
+studied system may not be as well characterized by a reference-database
+which would make the ability to utilize one's own data a reference an
+exciting possiblility. The ability to integrate data from new samples
+into existing OTUs enables the implementation of a single machine
+learning model. This is important for model implementation because not
+all of the data needs to be available or known at the time of model
+generation. A robust machine learning model can be implemented as part
+of a non-invasive and low-cost diagnostic for SRN and other diseases.
 
@@
 
-\textbf{\emph{Dataset.}} Raw 16S rRNA gene sequence data isolated from
-human stool samples was downloaded from NCBI Sequence Read Archive
-(accession no. SRP062005) (\protect\hyperlink{ref-baxter2016}{1},
+\textbf{Dataset.} Raw 16S rRNA gene sequence data from the V4 region
+were previously generated from human stool samples. Sequences were
+downloaded from the NCBI Sequence Read Archive (accession no. SRP062005)
+(\protect\hyperlink{ref-baxter2016}{3},
 \protect\hyperlink{ref-edgar2011}{6}). This dataset contains stool
-samples from a total of 490 subjects. For this analysis, samples from
-subjects identified in the metadata as normal, high risk normal, or
-adenoma were categorized as ``normal'', while samples from subjects
-identified as advanced adenoma or carcinoma were categorized as ``screen
-relevant neoplasia'' (SRN). The resulting dataset consisted of 261
-normal samples and 229 SRN samples.
+samples from 490 subjects. For this analysis, samples from subjects
+identified in the metadata as normal, high risk normal, or adenoma were
+categorized as ``normal'', while samples from subjects identified as
+advanced adenoma or carcinoma were categorized as ``screen relevant
+neoplasia'' (SRN). The resulting dataset consisted of 261 normal samples
+and 229 SRN samples.
 
-\textbf{\emph{Data Processing.}} The full dataset was preprocessed with
-mothur (v1.47) (\protect\hyperlink{ref-schloss2009}{7}) to join forward
-and reverse reads, merge duplicate reads, align to the SILVA reference
+\textbf{Data processing.} The full dataset was preprocessed with mothur
+(v1.47) (\protect\hyperlink{ref-schloss2009}{7}) to join forward and
+reverse reads, merge duplicate reads, align to the SILVA reference
 database (v132) (\protect\hyperlink{ref-quast2013}{8}), precluster,
@@
 set an average of 80 times (standard deviation = 4.1) and the test set
-an average of 20 times (standard deviation = 4.1).
+an average of 20 times (standard deviation = 4.1`).
 
-The data was processed through two workflows. In the control workflow,
-all the data was clustered together with OptiClust
-(\protect\hyperlink{ref-westcott2017}{5}) to generate OTUs and the
-resulting abundance tables were split into the training and testing
-sets. In the experimental workflow, the preprocessed data was split into
-the training and testing sets. The training set was clustered into OTUs
-using OptiClust, then the test set was fit to the OTUs of the training
-set using the OptiFit algorithm
-(\protect\hyperlink{ref-sovacool2022}{4}). The OptiFit algorithm was run
-with method open so that any sequences that did not map to the existing
-OTU clusters would form new OTUs. For both pathways, the shared files
-were sub-sampled to 10,000 reads per sample.
+\textbf{\emph{Reference-based workflows.}}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  OptiFit Self: The preprocess data was split into the training and
+  testing sets. The training set was clustered into OTUs using
+  OptiClust, then the test set was fit to the OTUs of the training set
+  using the OptiFit algorithm (\protect\hyperlink{ref-sovacool2022}{2}).
+  The OptiFit algorithm was run with method open so that any sequences
+  that did not map to the existing OTU clusters would form new OTUs. The
+  data was then subsampled to 10,000 reads and any novel OTUs from the
+  test set were removed. This process was repeated for each of the 100
+  splits resulting in 100 training and testing datasets.
+\item
+  OptiFit Greengenes: Reference sequences from the Greengenes database
+  v13\_8\_99 (\protect\hyperlink{ref-desantis2006}{9}) were downloaded
+  and processed with mothur by trimming to the V4 region and clustered
+  \emph{de novo} with OptiClust
+
+  \begin{enumerate}
+  \def\labelenumii{\arabic{enumii}.}
+  \tightlist
+  \item
+    The preprocessed data was fit to the clustered reference data using
+    OptiFit with the method open to allow any sequences that did not map
+    to the existing reference clusters would form new OTUs. The data was
+    then subsampled to 10,000 reads and any novel OTUs from the test set
+    were removed. The data set was then split into two sets where 80\%
+    of the samples were assigned to the training set and 20\% to the
+    testing set. This process was repeated for each of the 100 splits
+    resulting in 100 training and testing datasets.
+  \end{enumerate}
+\item
+  VSEARCH Greengenes: Preprocessed data was clustered using VSEARCH
+  v2.15.2 (\protect\hyperlink{ref-rognes2016}{4}) directly to
+  unprocessed Greengenes 97\% OTU reference alignment consistent with
+  how VSEARCH is typically used by the QIIME2 software for
+  reference-based clustering (\protect\hyperlink{ref-bolyen2019}{5}).
+  The data was then subsampled to 10,000 reads and any novel OTUs from
+  the test set were removed. The data set was then split into two sets
+  where 80\% of the samples were assigned to the training set and 20\%
+  to the testing set. This process was repeated for each of the 100
+  splits resulting in 100 training and testing datasets.
+\end{enumerate}
+
+\textbf{\emph{De novo workflows.}}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{3}
+\item
+  OptiClust \emph{de novo}: All the preprocessed data was clustered
+  together with OptiClust 1 to generate OTUs. The data was subsampled to
+  10,000 reads per sample and the resulting abundance tables were split
+  into the training and testing sets. The process was repeated for each
+  of the 100 splits resulting in 100 training and testing datasets.
+\item
+  VSEARCH \emph{de novo}: All the preprocessed data was clustered using
+  VSEARCH v2.15.2 (\protect\hyperlink{ref-rognes2016}{4}) with 97\%
+  identity and then subsampled to 10,000 reads per sample. The process
+  was repeated for each of the 100 splits resulting in 100 training and
+  testing datasets for both workflows.
+\end{enumerate}
 
@@
 the R package mikrompl (v 1.2.0)
-(\protect\hyperlink{ref-topuxe7uoglu2021}{9}) to predict the diagnosis
+(\protect\hyperlink{ref-topuxe7uoglu2021}{10}) to predict the diagnosis
 (SRN or normal) for the samples in the test set for each data split. The
@@
 removed. P values comparing model performance were calculated as
-previously described (\protect\hyperlink{ref-topuxe7uoglu2020}{10}). The
+previously described (\protect\hyperlink{ref-topuxe7uoglu2020}{11}). The
 averaged ROC curves were plotted by taking the average and standard
@@
 
-\textbf{\emph{Code Availability.}} The analysis workflow was implemented
-in Snakemake (\protect\hyperlink{ref-koster2012}{11}) . Scripts for
-analysis were written in R (\protect\hyperlink{ref-R2020}{12}) and GNU
-bash (\protect\hyperlink{ref-GNUbash}{13}). The software used includes
-mothur v1.47.0 (\protect\hyperlink{ref-schloss2009}{7}), RStudio
-(\protect\hyperlink{ref-RStudio2019}{14}), the Tidyverse metapackage
-(\protect\hyperlink{ref-wickham2019}{15}), R Markdown
-(\protect\hyperlink{ref-xie_r_2018}{16}), the SRA toolkit
-(\protect\hyperlink{ref-noauthor_sra-tools_nodate}{17}), and conda
-(\protect\hyperlink{ref-noauthor_anaconda_2016}{18}). The complete
+\textbf{\emph{Code Availability.}}
+
+The analysis workflow was implemented in Snakemake
+(\protect\hyperlink{ref-koster2012}{12}). Scripts for analysis were
+written in R (\protect\hyperlink{ref-R2020}{13}) and GNU bash
+(\protect\hyperlink{ref-GNUbash}{14}). The software used includes mothur
+v1.47.0 (\protect\hyperlink{ref-schloss2009}{7}), VSEARCH v2.15.2
+(\protect\hyperlink{ref-rognes2016}{4}), RStudio
+(\protect\hyperlink{ref-RStudio2019}{15}), the Tidyverse metapackage
+(\protect\hyperlink{ref-wickham2019}{16}), R Markdown
+(\protect\hyperlink{ref-xie_r_2018}{17}), the SRA toolkit
+(\protect\hyperlink{ref-noauthor_sra-tools_nodate}{18}), and conda
+(\protect\hyperlink{ref-noauthor_anaconda_2016}{19}). The complete
 workflow and supporting files required to reproduce this study are
 available at:
-\url{https://github.com/SchlossLab/Armour_OptiFitGLNE_XXXX_2021}
+\url{https://github.com/SchlossLab/Armour_OptiFitGLNE_mBio_2023}
 
-\hypertarget{acknowledgements}{%
-\subsection{Acknowledgements}\label{acknowledgements}}
+\hypertarget{acknowledgments}{%
+\subsection{Acknowledgments}\label{acknowledgments}}
 
@@
 \begin{CSLReferences}{0}{1}
-\leavevmode\vadjust pre{\hypertarget{ref-baxter2016}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-westcott2015}{}}%
 \CSLLeftMargin{1. }%
-\CSLRightInline{\textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers
-MAM}, \textbf{Schloss PD}. 2016. Microbiota-based model improves the
-sensitivity of fecal immunochemical test for detecting colonic lesions.
-Genome Medicine \textbf{8}:37.
-doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-zackular2014}{}}%
-\CSLLeftMargin{2. }%
-\CSLRightInline{\textbf{Zackular JP}, \textbf{Rogers MAM},
-\textbf{Ruffin MT,I}, \textbf{Schloss PD}. 2014. The human gut
-microbiome as a screening tool for colorectal cancer. Cancer Prevention
-Research \textbf{7}:1112--1121.
-doi:\href{https://doi.org/10.1158/1940-6207.CAPR-14-0129}{10.1158/1940-6207.CAPR-14-0129}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-westcott2015}{}}%
-\CSLLeftMargin{3. }%
 \CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo
@@
 \leavevmode\vadjust pre{\hypertarget{ref-sovacool2022}{}}%
-\CSLLeftMargin{4. }%
+\CSLLeftMargin{2. }%
 \CSLRightInline{\textbf{Sovacool KL}, \textbf{Westcott SL},
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-westcott2017}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-baxter2016}{}}%
+\CSLLeftMargin{3. }%
+\CSLRightInline{\textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers
+MAM}, \textbf{Schloss PD}. 2016. Microbiota-based model improves the
+sensitivity of fecal immunochemical test for detecting colonic lesions.
+Genome Medicine \textbf{8}:37.
+doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-rognes2016}{}}%
+\CSLLeftMargin{4. }%
+\CSLRightInline{\textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols
+B}, \textbf{Quince C}, \textbf{Mahé F}. 2016. VSEARCH: a versatile open
+source tool for metagenomics. PeerJ \textbf{4}:e2584.
+doi:\href{https://doi.org/10.7717/peerj.2584}{10.7717/peerj.2584}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-bolyen2019}{}}%
 \CSLLeftMargin{5. }%
-\CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2017.
-OptiClust, an improved method for assigning amplicon-based sequence data
-to operational taxonomic units. mSphere \textbf{2}:e00073--17.
-doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.}
+\CSLRightInline{\textbf{Bolyen E}, \textbf{Rideout JR}, \textbf{Dillon
+MR}, \textbf{Bokulich NA}, \textbf{Abnet CC}, \textbf{Al-Ghalith GA},
+\textbf{Alexander H}, \textbf{Alm EJ}, \textbf{Arumugam M},
+\textbf{Asnicar F}, \textbf{Bai Y}, \textbf{Bisanz JE},
+\textbf{Bittinger K}, \textbf{Brejnrod A}, \textbf{Brislawn CJ},
+\textbf{Brown CT}, \textbf{Callahan BJ}, \textbf{Caraballo-Rodríguez
+AM}, \textbf{Chase J}, \textbf{Cope EK}, \textbf{Da Silva R},
+\textbf{Diener C}, \textbf{Dorrestein PC}, \textbf{Douglas GM},
+\textbf{Durall DM}, \textbf{Duvallet C}, \textbf{Edwardson CF},
+\textbf{Ernst M}, \textbf{Estaki M}, \textbf{Fouquier J},
+\textbf{Gauglitz JM}, \textbf{Gibbons SM}, \textbf{Gibson DL},
+\textbf{Gonzalez A}, \textbf{Gorlick K}, \textbf{Guo J},
+\textbf{Hillmann B}, \textbf{Holmes S}, \textbf{Holste H},
+\textbf{Huttenhower C}, \textbf{Huttley GA}, \textbf{Janssen S},
+\textbf{Jarmusch AK}, \textbf{Jiang L}, \textbf{Kaehler BD},
+\textbf{Kang KB}, \textbf{Keefe CR}, \textbf{Keim P}, \textbf{Kelley
+ST}, \textbf{Knights D}, \textbf{Koester I}, \textbf{Kosciolek T},
+\textbf{Kreps J}, \textbf{Langille MGI}, \textbf{Lee J}, \textbf{Ley R},
+\textbf{Liu Y-X}, \textbf{Loftfield E}, \textbf{Lozupone C},
+\textbf{Maher M}, \textbf{Marotz C}, \textbf{Martin BD},
+\textbf{McDonald D}, \textbf{McIver LJ}, \textbf{Melnik AV},
+\textbf{Metcalf JL}, \textbf{Morgan SC}, \textbf{Morton JT},
+\textbf{Naimey AT}, \textbf{Navas-Molina JA}, \textbf{Nothias LF},
+\textbf{Orchanian SB}, \textbf{Pearson T}, \textbf{Peoples SL},
+\textbf{Petras D}, \textbf{Preuss ML}, \textbf{Pruesse E},
+\textbf{Rasmussen LB}, \textbf{Rivers A}, \textbf{Robeson MS},
+\textbf{Rosenthal P}, \textbf{Segata N}, \textbf{Shaffer M},
+\textbf{Shiffer A}, \textbf{Sinha R}, \textbf{Song SJ}, \textbf{Spear
+JR}, \textbf{Swafford AD}, \textbf{Thompson LR}, \textbf{Torres PJ},
+\textbf{Trinh P}, \textbf{Tripathi A}, \textbf{Turnbaugh PJ},
+\textbf{Ul-Hasan S}, \textbf{Hooft JJJ van der}, \textbf{Vargas F},
+\textbf{Vázquez-Baeza Y}, \textbf{Vogtmann E}, \textbf{Hippel M von},
+\textbf{Walters W}, \textbf{Wan Y}, \textbf{Wang M}, \textbf{Warren J},
+\textbf{Weber KC}, \textbf{Williamson CHD}, \textbf{Willis AD},
+\textbf{Xu ZZ}, \textbf{Zaneveld JR}, \textbf{Zhang Y}, \textbf{Zhu Q},
+\textbf{Knight R}, \textbf{Caporaso JG}. 2019. Reproducible,
+interactive, scalable and extensible microbiome data science using QIIME
+2. Nature Biotechnology \textbf{37}:852--857.
+doi:\href{https://doi.org/10.1038/s41587-019-0209-9}{10.1038/s41587-019-0209-9}.}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-desantis2006}{}}%
+\CSLLeftMargin{9. }%
+\CSLRightInline{\textbf{DeSantis TZ}, \textbf{Hugenholtz P},
+\textbf{Larsen N}, \textbf{Rojas M}, \textbf{Brodie EL}, \textbf{Keller
+K}, \textbf{Huber T}, \textbf{Dalevi D}, \textbf{Hu P}, \textbf{Andersen
+GL}. 2006. Greengenes, a chimera-checked 16S rRNA gene database and
+workbench compatible with ARB. Applied and Environmental Microbiology
+\textbf{72}:5069--5072.
+doi:\href{https://doi.org/10.1128/AEM.03006-05}{10.1128/AEM.03006-05}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-topuxe7uoglu2021}{}}%
-\CSLLeftMargin{9. }%
+\CSLLeftMargin{10. }%
 \CSLRightInline{\textbf{Topçuoğlu BD}, \textbf{Lapp Z}, \textbf{Sovacool
@@
 \leavevmode\vadjust pre{\hypertarget{ref-topuxe7uoglu2020}{}}%
-\CSLLeftMargin{10. }%
+\CSLLeftMargin{11. }%
 \CSLRightInline{\textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-koster2012}{}}%
-\CSLLeftMargin{11. }%
+\CSLLeftMargin{12. }%
 \CSLRightInline{\textbf{Koster J}, \textbf{Rahmann S}. 2012.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-R2020}{}}%
-\CSLLeftMargin{12. }%
+\CSLLeftMargin{13. }%
 \CSLRightInline{\textbf{R Core Team}. 2020.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-GNUbash}{}}%
-\CSLLeftMargin{13. }%
+\CSLLeftMargin{14. }%
 \CSLRightInline{\textbf{GNU Project}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-RStudio2019}{}}%
-\CSLLeftMargin{14. }%
+\CSLLeftMargin{15. }%
 \CSLRightInline{\textbf{RStudio Team}. 2019.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-wickham2019}{}}%
-\CSLLeftMargin{15. }%
+\CSLLeftMargin{16. }%
 \CSLRightInline{\textbf{Wickham H}, \textbf{Averick M}, \textbf{Bryan
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xie_r_2018}{}}%
-\CSLLeftMargin{16. }%
+\CSLLeftMargin{17. }%
 \CSLRightInline{\textbf{Xie Y}, \textbf{Allaire JJ}, \textbf{Grolemund
@@
 \leavevmode\vadjust pre{\hypertarget{ref-noauthor_sra-tools_nodate}{}}%
-\CSLLeftMargin{17. }%
+\CSLLeftMargin{18. }%
 \CSLRightInline{{SRA}-{Tools} - {NCBI}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-noauthor_anaconda_2016}{}}%
-\CSLLeftMargin{18. }%
+\CSLLeftMargin{19. }%
 \CSLRightInline{2016. Anaconda {Software Distribution}. Anaconda
@@
 
-\includegraphics{../exploratory/figures/figure1_a.pdf}
+\includegraphics{./figures/Figure1.png}
 
-\includegraphics{../exploratory/figures/figure1_b.pdf}
-
-\textbf{Figure 1: Workflows.} \textbf{A)} OptiClust workflow: The full
-dataset was clustered into OTUs using the OptiClust algorithm in mothur.
-The data was then split into two sets where 80\% of the samples were
-assigned to the training set and 20\% to the testing set. The training
-set was preprocessed with mikropml to normalize values (scale and
-center), collapse correlated features, and remove features with zero
-variance. Using mikropml, the training set was split into train and
-validate sets to compare results using different hyperparameter
-settings. The highest performing hyperparameter setting was then used to
-train the model with the full training set. The preprocessing scale from
-the training set was applied to the test dataset, then the trained model
-was used to classify the samples in the test set. Based on the actual
-classification and predicted classification, the area under the receiver
-operating characteristic curve (AUROC) was calculated to summarize model
-performance. The entire process was repeated 100 times to account for
-variability depending on the split of the data resulting in a total of
-100 AUROC values summarizing the performance of the standard OptiClust
-workflow. \textbf{B)} OptiFit workflow: The dataset was first split into
-two sets where 80\% of the samples were assigned to the training set and
-20\% to the testing set. The training set was then clustered into OTUs
-using the OptiClust algorithm in mothur. The resulting abundance data
-was preprocessed with mikropml to normalize values (scale/center),
-collapse correlated features, and remove features with zero-variance.
-Using mikropml, the training set was split into train and validate sets
-to compare results using different hyperparameter settings. The highest
-performing hyperparameter setting was then used to train the model with
-the full training set. The OptiFit algorithm in mothur was used to
-cluster the held out testing dataset using the OTUs of the training set
-as a reference. The preprocessing scale from the training set was
-applied to the test dataset, then the trained model was used to classify
-the samples in the test set. Based on the actual classification and
-predicted classification, the area under the receiver operating
-characteristic curve (AUROC) was calculated to summarize model
-performance. The entire process was repeated 100 times to account for
-variability depending on the split of the data resulting in a total of
-100 AUROC values summarizing the performance of the new OptiFit
-workflow.
+\textbf{Figure1: Overview of clustering workflows.} The \emph{de novo}
+and database-reference-based workflows were conducted using two
+approaches: OptiClust with mothur and VSEARCH as is used in the QIIME
+pipeline.
 
@@
 
-\textbf{Figure 2: Model Performance.} \textbf{A)} Area under the
+\textbf{Figure 2: Model perfomance of OptiFit self-reference workflow is
+as good or better than other methods.} \textbf{A)} Area under the
 receiver operating characteristic (AUROC) curve during cross-validation
-for the OptiClust and OptiFit workflows. Mean and standard deviation of
-the AUROC is represented by the black dot and whiskers. Mean AUROC is
-printed to the right of the points. \textbf{B)} AUROC on the test data
-for the OptiClust and OptiFit workflows. Mean and standard deviation of
-the AUROC is represented by the black dot and whiskers. The mean AUROC
-is printed to the right of the points. \textbf{C)} Averaged receiver
-operating characteristic (ROC) curves.
+(train) for the various workflows. \textbf{B)} AUROC on the test data
+for the various workflows. The mean and standard deviation of the AUROC
+is represented by the black dot and whiskers in panels A and B. The mean
+AUROC is printed below the points. \textbf{C)} Averaged receiver
+operating characteristic (ROC) curves. Lines represent the average true
+positive rate for the range of false positive rates.
 

### 2023-04-13 submission/manuscript.Rmd
fix typos

@@
     form new OTUs. The data was then subsampled to 10,000 reads and any
-    novel OTUs from the test set were removed. The data set was then
+    novel OTUs from the test set were removed. The dataset was then
     split into two sets where 80% of the samples were assigned to the
@@
     data was then subsampled to 10,000 reads and any novel OTUs from the
-    test set were removed. The data set was then split into two sets
+    test set were removed. The dataset was then split into two sets
     where 80% of the samples were assigned to the training set and 20%
@@
 
-**Figure 2: Model perfomance of OptiFit self-reference workflow is as
+**Figure 2: Model performance of OptiFit self-reference workflow is as
 good or better than other methods.** **A)** Area under the receiver

### 2023-04-13 submission/manuscript.tex
fix typos

@@
     then subsampled to 10,000 reads and any novel OTUs from the test set
-    were removed. The data set was then split into two sets where 80\%
-    of the samples were assigned to the training set and 20\% to the
+    were removed. The dataset was then split into two sets where 80\% of
+    the samples were assigned to the training set and 20\% to the
     testing set. This process was repeated for each of the 100 splits
@@
   The data was then subsampled to 10,000 reads and any novel OTUs from
-  the test set were removed. The data set was then split into two sets
+  the test set were removed. The dataset was then split into two sets
   where 80\% of the samples were assigned to the training set and 20\%
@@
 
-\textbf{Figure 2: Model perfomance of OptiFit self-reference workflow is
-as good or better than other methods.} \textbf{A)} Area under the
+\textbf{Figure 2: Model performance of OptiFit self-reference workflow
+is as good or better than other methods.} \textbf{A)} Area under the
 receiver operating characteristic (AUROC) curve during cross-validation

### 2023-04-18 submission/manuscript.Rmd
second round revision with pat

@@
 ---
-title: "**Machine learning classification using self-reference-based OTU clustering**"
+title: "**Machine learning classification by fitting amplicon sequences to existing OTUs**"
 output:
@@
 
-Running title: Reference-based OTU clustering for ML classification
+Running title: self-reference-based OTU clustering for ML classification
 
@@
 
-There is increasing evidence for an association between the composition of the gut microbiome and a variety of diseases, such as crohn's disease and colorectal cancer.
-The signatures that associate with disease have shown potential to aid in diagnosis of disease throught gut microbiome profiling and machine learning.
+There is increasing evidence for an association between the composition of the gut microbiome and a variety of diseases, such as crohn's disease and colorectal cancer [@Morgan2012;@Sobhani2011].
+There is great potential diagnose of disease with gut microbiome sequence data and machine learning.
 Taxonomic composition of microbial
@@
 with reference-based clustering is generally poor compared to those
-generated with *de novo* clustering @westcott2015. While *de novo*
-clustering produces high-quality OTU clusters where sequences are
-accurately grouped based on similarity thresholds, the resulting OTU
-clusters depend on the sequences within the dataset and the addition of
-new data has the potential to redefine OTU cluster composition. The
-unstable nature of *de novo* OTU clustering complicates deployment of
+generated with *de novo* clustering @westcott2015.
+While *de novo* clustering produces high-quality OTU clusters where sequences are accurately grouped based on similarity thresholds, the resulting OTU clusters depend on the sequences within the dataset and the addition of new data has the potential to redefine OTU cluster composition.
+The unstable nature of *de novo* OTU clustering complicates deployment of
 machine learning models since integration of additional data requires
@@
 We compared the ability of several approaches for assigning 16S rRNA
-gene seequences to OTUs including, *de novo*, database-reference-based,
-and self-reference-based with OptiFit. To test how the model performance
+gene sequences to OTUs including, *de novo* and reference-based clustering.
+For reference-based clustering, we used closed-reference clustering to a public database (database-reference-based) and to OTUs generated froma subset of the samples (self-reference-based).
+To test how the model performance
 compared between these approaches, we used a publicly available dataset
@@
 The OTU clustering was conducted using two common algorithms: 1) the
-OptiClust algorithm in mothur @westcott2017 and 2) the VSEARCH algorithm
+OptiClust algorithm in mothur [@westcott2017] and 2) the VSEARCH algorithm
 used in QIIME2 [@rognes2016; @bolyen2019]. For both algorithms, the
@@
 standard deviation = `r optifit_self_avg_mcc %>% pull(sd_mcc)`).
-Previously, we observed that *de novo* clustering tends to produce
-higher MCC scores than reference-based clustering @sovacool2022.
-Consistent with prior findings the reference-based methods produced
+Consistent with prior findings, the reference-based methods produced
 lower MCC scores (OptiFit Greengenes MCC = `r optifit_gg_mcc`; VSEARCH
-Greengenes MCC = `r vsearch_gg_mcc`) than the *de novo* methods with the
-same algorithm (OptiClust *de novo* MCC = `r opticlust_denovo_mcc`;
-VSEARCH *de novo* MCC = `r vsearch_denovo_mcc`). Another metric we
+Greengenes MCC = `r vsearch_gg_mcc`) than the *de novo* methods (OptiClust *de novo* MCC = `r opticlust_denovo_mcc`;
+VSEARCH *de novo* MCC = `r vsearch_denovo_mcc`) @sovacool2022. Another metric we
 examined for the OptiFit workflow was the fraction of sequences from the
@@
 operating characteristic curve (AUROC) for each data split. During
-cross-validation (CV) training, the model performance was equivalent
-between OptFit self-reference and OptiClust *de novo* (p-value =
+cross-validation (CV) training, the performance of the OptFit self-reference and OptiClust *de novo* models were not significantly different (p-value =
 `r get_pval(pvals, "cv_metric_AUC", "opticlust_denovo", "optifit_self")`;
-Figure 2A) while while performance for both VSEARCH methods was lower
+Figure 2A), while performance for both VSEARCH methods was significantly lower
 than the OptiClust *de novo*, OptiFit self, and OptiFit Greengenes
@@
 test data classifying samples as either control or SRN. Both VSEARCH
-methods perform slightly worse than the OptiClust \*de novo\* method
-(both p-values \< 0.05). However the performance on the test data was
-equivalent between the OptiClust *de novo*, OptiFit Greengenes, and
-OptiFit self-reference approaches (p-value \> 0.05; Figures 2B and 2C).
+methods perform slightly worse than the OptiClust *de novo* method
+(both p-values \< 0.05). However the performance on the test data for the OptiClust *de novo*, OptiFit Greengenes, and
+OptiFit self-reference approaches were not significantly different (p-value \> 0.05; Figures 2B and 2C).
 These results indicate that new data could be fit to existing OTU
@@
 (standard deviation =
-`r format_decimal(n_test_train %>% pull(sd_test),digits = 1)`\`).
+`r format_decimal(n_test_train %>% pull(sd_test),digits = 1)`).
 
@@
 2.  OptiFit Greengenes: Reference sequences from the Greengenes database
-    v13_8\_99 @desantis2006 were downloaded and processed with mothur by
-    trimming to the V4 region and clustered *de novo* with OptiClust
-    @westcott2017. The preprocessed data was fit to the clustered
+    v13_8_99 @desantis2006 were downloaded and processed with mothur by
+    trimming to the V4 region and clustered *de novo* with OptiClust [@westcott2017]. The preprocessed data was fit to the clustered
     reference data using OptiFit with the method open to allow any
@@
 set. Any OTUs in the test set that were not in the training set were
-removed. P values comparing model performance were calculated as
+removed. P-values comparing model performance were calculated as
 previously described @topçuoglu2020. The averaged ROC curves were
@@
 
-The analysis workflow was implemented in Snakemake @koster2012. Scripts
+The analysis workflow was implemented in Snakemake [@koster2012]. Scripts
 for analysis were written in R @R2020 and GNU bash @GNUbash. The

### 2023-04-18 submission/manuscript.tex
second round revision with pat

@@
 \hypersetup{
-  pdftitle={Machine learning classification using self-reference-based OTU clustering},
+  pdftitle={Machine learning classification by fitting amplicon sequences to existing OTUs},
   hidelinks,
@@
 
-\title{\textbf{Machine learning classification using
-self-reference-based OTU clustering}}
+\title{\textbf{Machine learning classification by fitting amplicon
+sequences to existing OTUs}}
 \author{}
@@
 of the gut microbiome and a variety of diseases, such as crohn's disease
-and colorectal cancer. The signatures that associate with disease have
-shown potential to aid in diagnosis of disease throught gut microbiome
-profiling and machine learning. Taxonomic composition of microbial
-communities can be assessed using amplicon sequencing of the 16S rRNA
-gene, which is the input to classification models. Analysis of 16S rRNA
-gene sequence data generally relies on assigning sequences into
-operational taxonomic units (OTUs). The process of OTU clustering can
-either be reference-based or \emph{de novo}. The quality of OTUs
-generated with reference-based clustering is generally poor compared to
-those generated with \emph{de novo} clustering
-(\protect\hyperlink{ref-westcott2015}{1}). While \emph{de novo}
-clustering produces high-quality OTU clusters where sequences are
+and colorectal cancer (\protect\hyperlink{ref-Morgan2012}{1},
+\protect\hyperlink{ref-Sobhani2011}{2}). There is great potential
+diagnose of disease with gut microbiome sequence data and machine
+learning. Taxonomic composition of microbial communities can be assessed
+using amplicon sequencing of the 16S rRNA gene, which is the input to
+classification models. Analysis of 16S rRNA gene sequence data generally
+relies on assigning sequences into operational taxonomic units (OTUs).
+The process of OTU clustering can either be reference-based or \emph{de
+novo}. The quality of OTUs generated with reference-based clustering is
+generally poor compared to those generated with \emph{de novo}
+clustering (\protect\hyperlink{ref-westcott2015}{3}). While \emph{de
+novo} clustering produces high-quality OTU clusters where sequences are
 accurately grouped based on similarity thresholds, the resulting OTU
@@
 introduced OptiFit, a method for fitting new sequence data into existing
-OTUs (\protect\hyperlink{ref-sovacool2022}{2}). While OptiFit can
+OTUs (\protect\hyperlink{ref-sovacool2022}{4}). While OptiFit can
 effectively fit new sequence data to existing OTU clusters, it is
@@
 We compared the ability of several approaches for assigning 16S rRNA
-gene seequences to OTUs including, \emph{de novo},
-database-reference-based, and self-reference-based with OptiFit. To test
+gene sequences to OTUs including, \emph{de novo} and reference-based
+clustering. For reference-based clustering, we used closed-reference
+clustering to a public database (database-reference-based) and to OTUs
+generated froma subset of the samples (self-reference-based). To test
 how the model performance compared between these approaches, we used a
@@
 neoplasia (SRN) consisting of advanced adenoma and carcinoma (n = 229)
-(\protect\hyperlink{ref-baxter2016}{3}). For the \emph{de novo}
+(\protect\hyperlink{ref-baxter2016}{5}). For the \emph{de novo}
 workflows, all the 16S rRNA sequence data was clustered into OTUs. The
 OTU clustering was conducted using two common algorithms: 1) the
-OptiClust algorithm in mothur 1 and 2) the VSEARCH algorithm used in
-QIIME2 (\protect\hyperlink{ref-rognes2016}{4},
-\protect\hyperlink{ref-bolyen2019}{5}). For both algorithms, the
+OptiClust algorithm in mothur (\protect\hyperlink{ref-westcott2017}{6})
+and 2) the VSEARCH algorithm used in QIIME2
+(\protect\hyperlink{ref-rognes2016}{7},
+\protect\hyperlink{ref-bolyen2019}{8}). For both algorithms, the
 resulting abundance data was then split into training and testing sets,
@@
 of sequences and whether they are appropriately clustered or not
-(\protect\hyperlink{ref-westcott2015}{1}). MCC scores range between
+(\protect\hyperlink{ref-westcott2015}{3}). MCC scores range between
 negative one and one, and measure how well clustering assignment
@@
 OptiFit self-reference workflows (average MCC = 0.879, standard
-deviation = 0.002). Previously, we observed that \emph{de novo}
-clustering tends to produce higher MCC scores than reference-based
-clustering (\protect\hyperlink{ref-sovacool2022}{2}). Consistent with
-prior findings the reference-based methods produced lower MCC scores
-(OptiFit Greengenes MCC = 0.786; VSEARCH Greengenes MCC = 0.531) than
-the \emph{de novo} methods with the same algorithm (OptiClust \emph{de
-novo} MCC = 0.884; VSEARCH \emph{de novo} MCC = 0.641). Another metric
-we examined for the OptiFit workflow was the fraction of sequences from
-the test set that mapped to the reference OTUs. Since sequences that did
-not map to reference OTUs were eliminated, if a high percentage of reads
-did not map to an OTU we expected this loss of data to negatively impact
+deviation = 0.002). Consistent with prior findings, the reference-based
+methods produced lower MCC scores (OptiFit Greengenes MCC = 0.786;
+VSEARCH Greengenes MCC = 0.531) than the \emph{de novo} methods
+(OptiClust \emph{de novo} MCC = 0.884; VSEARCH \emph{de novo} MCC =
+0.641) (\protect\hyperlink{ref-sovacool2022}{4}). Another metric we
+examined for the OptiFit workflow was the fraction of sequences from the
+test set that mapped to the reference OTUs. Since sequences that did not
+map to reference OTUs were eliminated, if a high percentage of reads did
+not map to an OTU we expected this loss of data to negatively impact
 classification performance. We found that loss of data was not an issue
@@
 operating characteristic curve (AUROC) for each data split. During
-cross-validation (CV) training, the model performance was equivalent
-between OptFit self-reference and OptiClust \emph{de novo} (p-value =
-0.071; Figure 2A) while while performance for both VSEARCH methods was
-lower than the OptiClust \emph{de novo}, OptiFit self, and OptiFit
-Greengenes methods (p-values \textless{} 0.05). The trained model was
-then applied to the test data classifying samples as either control or
-SRN. Both VSEARCH methods perform slightly worse than the OptiClust *de
-novo* method (both p-values \textless{} 0.05). However the performance
-on the test data was equivalent between the OptiClust \emph{de novo},
-OptiFit Greengenes, and OptiFit self-reference approaches (p-value
+cross-validation (CV) training, the performance of the OptFit
+self-reference and OptiClust \emph{de novo} models were not
+significantly different (p-value = 0.071; Figure 2A), while performance
+for both VSEARCH methods was significantly lower than the OptiClust
+\emph{de novo}, OptiFit self, and OptiFit Greengenes methods (p-values
+\textless{} 0.05). The trained model was then applied to the test data
+classifying samples as either control or SRN. Both VSEARCH methods
+perform slightly worse than the OptiClust \emph{de novo} method (both
+p-values \textless{} 0.05). However the performance on the test data for
+the OptiClust \emph{de novo}, OptiFit Greengenes, and OptiFit
+self-reference approaches were not significantly different (p-value
 \textgreater{} 0.05; Figures 2B and 2C). These results indicate that new
@@
 downloaded from the NCBI Sequence Read Archive (accession no. SRP062005)
-(\protect\hyperlink{ref-baxter2016}{3},
-\protect\hyperlink{ref-edgar2011}{6}). This dataset contains stool
+(\protect\hyperlink{ref-baxter2016}{5},
+\protect\hyperlink{ref-edgar2011}{9}). This dataset contains stool
 samples from 490 subjects. For this analysis, samples from subjects
@@
 \textbf{Data processing.} The full dataset was preprocessed with mothur
-(v1.47) (\protect\hyperlink{ref-schloss2009}{7}) to join forward and
+(v1.47) (\protect\hyperlink{ref-schloss2009}{10}) to join forward and
 reverse reads, merge duplicate reads, align to the SILVA reference
-database (v132) (\protect\hyperlink{ref-quast2013}{8}), precluster,
-remove chimeras with UCHIME (\protect\hyperlink{ref-edgar2011}{6}),
+database (v132) (\protect\hyperlink{ref-quast2013}{11}), precluster,
+remove chimeras with UCHIME (\protect\hyperlink{ref-edgar2011}{9}),
 assign taxonomy, and remove non-bacterial reads following the Schloss
@@
 set an average of 80 times (standard deviation = 4.1) and the test set
-an average of 20 times (standard deviation = 4.1`).
+an average of 20 times (standard deviation = 4.1).
 
@@
   OptiClust, then the test set was fit to the OTUs of the training set
-  using the OptiFit algorithm (\protect\hyperlink{ref-sovacool2022}{2}).
+  using the OptiFit algorithm (\protect\hyperlink{ref-sovacool2022}{4}).
   The OptiFit algorithm was run with method open so that any sequences
@@
   OptiFit Greengenes: Reference sequences from the Greengenes database
-  v13\_8\_99 (\protect\hyperlink{ref-desantis2006}{9}) were downloaded
+  v13\_8\_99 (\protect\hyperlink{ref-desantis2006}{12}) were downloaded
   and processed with mothur by trimming to the V4 region and clustered
   \emph{de novo} with OptiClust
-
-  \begin{enumerate}
-  \def\labelenumii{\arabic{enumii}.}
-  \tightlist
-  \item
-    The preprocessed data was fit to the clustered reference data using
-    OptiFit with the method open to allow any sequences that did not map
-    to the existing reference clusters would form new OTUs. The data was
-    then subsampled to 10,000 reads and any novel OTUs from the test set
-    were removed. The dataset was then split into two sets where 80\% of
-    the samples were assigned to the training set and 20\% to the
-    testing set. This process was repeated for each of the 100 splits
-    resulting in 100 training and testing datasets.
-  \end{enumerate}
+  (\protect\hyperlink{ref-westcott2017}{6}). The preprocessed data was
+  fit to the clustered reference data using OptiFit with the method open
+  to allow any sequences that did not map to the existing reference
+  clusters would form new OTUs. The data was then subsampled to 10,000
+  reads and any novel OTUs from the test set were removed. The dataset
+  was then split into two sets where 80\% of the samples were assigned
+  to the training set and 20\% to the testing set. This process was
+  repeated for each of the 100 splits resulting in 100 training and
+  testing datasets.
 \item
   VSEARCH Greengenes: Preprocessed data was clustered using VSEARCH
-  v2.15.2 (\protect\hyperlink{ref-rognes2016}{4}) directly to
+  v2.15.2 (\protect\hyperlink{ref-rognes2016}{7}) directly to
   unprocessed Greengenes 97\% OTU reference alignment consistent with
   how VSEARCH is typically used by the QIIME2 software for
-  reference-based clustering (\protect\hyperlink{ref-bolyen2019}{5}).
+  reference-based clustering (\protect\hyperlink{ref-bolyen2019}{8}).
   The data was then subsampled to 10,000 reads and any novel OTUs from
@@
   OptiClust \emph{de novo}: All the preprocessed data was clustered
-  together with OptiClust 1 to generate OTUs. The data was subsampled to
-  10,000 reads per sample and the resulting abundance tables were split
-  into the training and testing sets. The process was repeated for each
-  of the 100 splits resulting in 100 training and testing datasets.
+  together with OptiClust (\protect\hyperlink{ref-westcott2017}{6}) to
+  generate OTUs. The data was subsampled to 10,000 reads per sample and
+  the resulting abundance tables were split into the training and
+  testing sets. The process was repeated for each of the 100 splits
+  resulting in 100 training and testing datasets.
 \item
   VSEARCH \emph{de novo}: All the preprocessed data was clustered using
-  VSEARCH v2.15.2 (\protect\hyperlink{ref-rognes2016}{4}) with 97\%
+  VSEARCH v2.15.2 (\protect\hyperlink{ref-rognes2016}{7}) with 97\%
   identity and then subsampled to 10,000 reads per sample. The process
@@
 the R package mikrompl (v 1.2.0)
-(\protect\hyperlink{ref-topuxe7uoglu2021}{10}) to predict the diagnosis
+(\protect\hyperlink{ref-topuxe7uoglu2021}{13}) to predict the diagnosis
 (SRN or normal) for the samples in the test set for each data split. The
@@
 set. Any OTUs in the test set that were not in the training set were
-removed. P values comparing model performance were calculated as
-previously described (\protect\hyperlink{ref-topuxe7uoglu2020}{11}). The
+removed. P-values comparing model performance were calculated as
+previously described (\protect\hyperlink{ref-topuxe7uoglu2020}{14}). The
 averaged ROC curves were plotted by taking the average and standard
@@
 The analysis workflow was implemented in Snakemake
-(\protect\hyperlink{ref-koster2012}{12}). Scripts for analysis were
-written in R (\protect\hyperlink{ref-R2020}{13}) and GNU bash
-(\protect\hyperlink{ref-GNUbash}{14}). The software used includes mothur
-v1.47.0 (\protect\hyperlink{ref-schloss2009}{7}), VSEARCH v2.15.2
-(\protect\hyperlink{ref-rognes2016}{4}), RStudio
-(\protect\hyperlink{ref-RStudio2019}{15}), the Tidyverse metapackage
-(\protect\hyperlink{ref-wickham2019}{16}), R Markdown
-(\protect\hyperlink{ref-xie_r_2018}{17}), the SRA toolkit
-(\protect\hyperlink{ref-noauthor_sra-tools_nodate}{18}), and conda
-(\protect\hyperlink{ref-noauthor_anaconda_2016}{19}). The complete
+(\protect\hyperlink{ref-koster2012}{15}). Scripts for analysis were
+written in R (\protect\hyperlink{ref-R2020}{16}) and GNU bash
+(\protect\hyperlink{ref-GNUbash}{17}). The software used includes mothur
+v1.47.0 (\protect\hyperlink{ref-schloss2009}{10}), VSEARCH v2.15.2
+(\protect\hyperlink{ref-rognes2016}{7}), RStudio
+(\protect\hyperlink{ref-RStudio2019}{18}), the Tidyverse metapackage
+(\protect\hyperlink{ref-wickham2019}{19}), R Markdown
+(\protect\hyperlink{ref-xie_r_2018}{20}), the SRA toolkit
+(\protect\hyperlink{ref-noauthor_sra-tools_nodate}{21}), and conda
+(\protect\hyperlink{ref-noauthor_anaconda_2016}{22}). The complete
 workflow and supporting files required to reproduce this study are
@@
 \begin{CSLReferences}{0}{1}
+\leavevmode\vadjust pre{\hypertarget{ref-Morgan2012}{}}%
+\CSLLeftMargin{1. }%
+\CSLRightInline{\textbf{Morgan XC}, \textbf{Tickle TL}, \textbf{Sokol
+H}, \textbf{Gevers D}, \textbf{Devaney KL}, \textbf{Ward DV},
+\textbf{Reyes JA}, \textbf{Shah SA}, \textbf{LeLeiko N}, \textbf{Snapper
+SB}, \textbf{Bousvaros A}, \textbf{Korzenik J}, \textbf{Sands BE},
+\textbf{Xavier RJ}, \textbf{Huttenhower C}. 2012. Dysfunction of the
+intestinal microbiome in inflammatory bowel disease and treatment.
+Genome Biol \textbf{13}:R79.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Sobhani2011}{}}%
+\CSLLeftMargin{2. }%
+\CSLRightInline{\textbf{Sobhani I}, \textbf{Tap J},
+\textbf{Roudot-Thoraval F}, \textbf{Roperch JP}, \textbf{Letulle S},
+\textbf{Langella P}, \textbf{Corthier G}, \textbf{Tran Van Nhieu J},
+\textbf{Furet JP}. 2011. Microbial dysbiosis in colorectal cancer
+({CRC}) patients. PLoS One \textbf{6}:e16393.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-westcott2015}{}}%
-\CSLLeftMargin{1. }%
+\CSLLeftMargin{3. }%
 \CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo
@@
 \leavevmode\vadjust pre{\hypertarget{ref-sovacool2022}{}}%
-\CSLLeftMargin{2. }%
+\CSLLeftMargin{4. }%
 \CSLRightInline{\textbf{Sovacool KL}, \textbf{Westcott SL},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-baxter2016}{}}%
-\CSLLeftMargin{3. }%
+\CSLLeftMargin{5. }%
 \CSLRightInline{\textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-westcott2017}{}}%
+\CSLLeftMargin{6. }%
+\CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2017.
+OptiClust, an improved method for assigning amplicon-based sequence data
+to operational taxonomic units. mSphere \textbf{2}:e00073--17.
+doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-rognes2016}{}}%
-\CSLLeftMargin{4. }%
+\CSLLeftMargin{7. }%
 \CSLRightInline{\textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bolyen2019}{}}%
-\CSLLeftMargin{5. }%
+\CSLLeftMargin{8. }%
 \CSLRightInline{\textbf{Bolyen E}, \textbf{Rideout JR}, \textbf{Dillon
@@
 \leavevmode\vadjust pre{\hypertarget{ref-edgar2011}{}}%
-\CSLLeftMargin{6. }%
+\CSLLeftMargin{9. }%
 \CSLRightInline{\textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente
@@
 \leavevmode\vadjust pre{\hypertarget{ref-schloss2009}{}}%
-\CSLLeftMargin{7. }%
+\CSLLeftMargin{10. }%
 \CSLRightInline{\textbf{Schloss PD}, \textbf{Westcott SL},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-quast2013}{}}%
-\CSLLeftMargin{8. }%
+\CSLLeftMargin{11. }%
 \CSLRightInline{\textbf{Quast C}, \textbf{Pruesse E}, \textbf{Yilmaz P},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-desantis2006}{}}%
-\CSLLeftMargin{9. }%
+\CSLLeftMargin{12. }%
 \CSLRightInline{\textbf{DeSantis TZ}, \textbf{Hugenholtz P},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-topuxe7uoglu2021}{}}%
-\CSLLeftMargin{10. }%
+\CSLLeftMargin{13. }%
 \CSLRightInline{\textbf{Topçuoğlu BD}, \textbf{Lapp Z}, \textbf{Sovacool
@@
 \leavevmode\vadjust pre{\hypertarget{ref-topuxe7uoglu2020}{}}%
-\CSLLeftMargin{11. }%
+\CSLLeftMargin{14. }%
 \CSLRightInline{\textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-koster2012}{}}%
-\CSLLeftMargin{12. }%
+\CSLLeftMargin{15. }%
 \CSLRightInline{\textbf{Koster J}, \textbf{Rahmann S}. 2012.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-R2020}{}}%
-\CSLLeftMargin{13. }%
+\CSLLeftMargin{16. }%
 \CSLRightInline{\textbf{R Core Team}. 2020.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-GNUbash}{}}%
-\CSLLeftMargin{14. }%
+\CSLLeftMargin{17. }%
 \CSLRightInline{\textbf{GNU Project}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-RStudio2019}{}}%
-\CSLLeftMargin{15. }%
+\CSLLeftMargin{18. }%
 \CSLRightInline{\textbf{RStudio Team}. 2019.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-wickham2019}{}}%
-\CSLLeftMargin{16. }%
+\CSLLeftMargin{19. }%
 \CSLRightInline{\textbf{Wickham H}, \textbf{Averick M}, \textbf{Bryan
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xie_r_2018}{}}%
-\CSLLeftMargin{17. }%
+\CSLLeftMargin{20. }%
 \CSLRightInline{\textbf{Xie Y}, \textbf{Allaire JJ}, \textbf{Grolemund
@@
 \leavevmode\vadjust pre{\hypertarget{ref-noauthor_sra-tools_nodate}{}}%
-\CSLLeftMargin{18. }%
+\CSLLeftMargin{21. }%
 \CSLRightInline{{SRA}-{Tools} - {NCBI}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-noauthor_anaconda_2016}{}}%
-\CSLLeftMargin{19. }%
+\CSLLeftMargin{22. }%
 \CSLRightInline{2016. Anaconda {Software Distribution}. Anaconda

### 2023-04-24 code/slurm/snakemake.sh
rerun vsearch to correct subsample issue

@@ -9,7 +9,7 @@
 #SBATCH --account=pschloss1
 #SBATCH --partition=standard
 #SBATCH --mail-user=armourc@umich.edu
-#SBATCH --mail-type=FAIL
+#SBATCH --mail-type=FAIL,END
 #SBATCH --output=%x-%j.out
 #SBATCH --export=ALL
 
@@ -31,4 +31,4 @@ conda activate glne
 mkdir -p logs/slurm/
 
 # Initiating snakemake and running workflow in cluster mode
-time snakemake --profile config/slurm/ --latency-wait 90 --rerun-incomplete
+time snakemake --profile config/slurm/ --latency-wait 90 --rerun-incomplete 

### 2023-04-27 submission/manuscript.Rmd
final revisions

@@
 
-\vspace{10mm}
-
-**observation format** (max 1200 words, 2 figures, 25 ref)
-
 \newpage
@@
 There is increasing evidence for an association between the composition of the gut microbiome and a variety of diseases, such as crohn's disease and colorectal cancer [@Morgan2012;@Sobhani2011].
-There is great potential diagnose of disease with gut microbiome sequence data and machine learning.
+There is great potential to diagnose disease with gut microbiome sequence data and machine learning.
 Taxonomic composition of microbial
@@
 gene sequences to OTUs including, *de novo* and reference-based clustering.
-For reference-based clustering, we used closed-reference clustering to a public database (database-reference-based) and to OTUs generated froma subset of the samples (self-reference-based).
+For reference-based clustering, we used closed-reference clustering to a public database (database-reference-based) and to OTUs generated from a subset of the samples (self-reference-based).
 To test how the model performance
@@
 `r avg_frac_mapped %>% filter(methods == "optifit_self", set == "test") %>% pull(sd_frac) %>% round(digits=2)`%)
-of sequences in the test set mapped to the reference OTUs. This number
+of sequences in the subsampled test set mapped to the reference OTUs. This number
 is higher than the average fraction of reads mapped in the OptiFit
@@
 get_pval <- function(pvals,met,g1,g2){
-  pvals %>% 
+  val <- pvals %>% 
     filter(metric == met) %>% 
     filter(group1 == g1 & group2 == g2 | group1 == g2 & group2 == g1) %>% 
-    pull(p_value) %>% round(digits = 3)
+    pull(p_value)
+
+  format(round(val,digits = 3),nsmall=3)
 }
+#get_pval(pvals,"AUC","opticlust_denovo","vsearch_gg")
 ```
@@
 operating characteristic curve (AUROC) for each data split. During
-cross-validation (CV) training, the performance of the OptFit self-reference and OptiClust *de novo* models were not significantly different (p-value =
+cross-validation (CV) training, the performance of the OptiFit self-reference and OptiClust *de novo* models were not significantly different (p-value =
 `r get_pval(pvals, "cv_metric_AUC", "opticlust_denovo", "optifit_self")`;
@@
 methods (p-values \< 0.05). The trained model was then applied to the
-test data classifying samples as either control or SRN. Both VSEARCH
-methods perform slightly worse than the OptiClust *de novo* method
-(both p-values \< 0.05). However the performance on the test data for the OptiClust *de novo*, OptiFit Greengenes, and
-OptiFit self-reference approaches were not significantly different (p-value \> 0.05; Figures 2B and 2C).
+test data classifying samples as either control or SRN.
+The VSEARCH Greengenes method performed slightly worse than the OptiClust *de novo* method (p-value = `r get_pval(pvals,"AUC","opticlust_denovo","vsearch_gg")`).
+However the performance on the test data for the OptiClust *de novo*, OptiFit Greengenes, OptiFit self-reference, and VSEARCH *de novo* approaches were not significantly different (p-values \> 0.05; Figures 2B and 2C).
 These results indicate that new data could be fit to existing OTU

### 2023-04-27 submission/manuscript.tex
final revisions

@@
 
-\vspace{10mm}
-
-\textbf{observation format} (max 1200 words, 2 figures, 25 ref)
-
 \newpage
@@
 and colorectal cancer (\protect\hyperlink{ref-Morgan2012}{1},
-\protect\hyperlink{ref-Sobhani2011}{2}). There is great potential
-diagnose of disease with gut microbiome sequence data and machine
-learning. Taxonomic composition of microbial communities can be assessed
-using amplicon sequencing of the 16S rRNA gene, which is the input to
+\protect\hyperlink{ref-Sobhani2011}{2}). There is great potential to
+diagnose disease with gut microbiome sequence data and machine learning.
+Taxonomic composition of microbial communities can be assessed using
+amplicon sequencing of the 16S rRNA gene, which is the input to
 classification models. Analysis of 16S rRNA gene sequence data generally
@@
 clustering to a public database (database-reference-based) and to OTUs
-generated froma subset of the samples (self-reference-based). To test
+generated from a subset of the samples (self-reference-based). To test
 how the model performance compared between these approaches, we used a
@@
 since on average 99.8\% (standard deviation = 0.68\%) of sequences in
-the test set mapped to the reference OTUs. This number is higher than
-the average fraction of reads mapped in the OptiFit Greengenes workflow
-( 96.8\% +/- 3.5). These results indicate that the OptiFit
-self-reference method performed as well as the OptiClust \emph{de novo}
-method and better than using an external database.
+the subsampled test set mapped to the reference OTUs. This number is
+higher than the average fraction of reads mapped in the OptiFit
+Greengenes workflow ( 96.8\% +/- 3.5). These results indicate that the
+OptiFit self-reference method performed as well as the OptiClust
+\emph{de novo} method and better than using an external database.
 
@@
 operating characteristic curve (AUROC) for each data split. During
-cross-validation (CV) training, the performance of the OptFit
+cross-validation (CV) training, the performance of the OptiFit
 self-reference and OptiClust \emph{de novo} models were not
-significantly different (p-value = 0.071; Figure 2A), while performance
+significantly different (p-value = 0.066; Figure 2A), while performance
 for both VSEARCH methods was significantly lower than the OptiClust
@@
 \textless{} 0.05). The trained model was then applied to the test data
-classifying samples as either control or SRN. Both VSEARCH methods
-perform slightly worse than the OptiClust \emph{de novo} method (both
-p-values \textless{} 0.05). However the performance on the test data for
-the OptiClust \emph{de novo}, OptiFit Greengenes, and OptiFit
-self-reference approaches were not significantly different (p-value
-\textgreater{} 0.05; Figures 2B and 2C). These results indicate that new
-data could be fit to existing OTU clusters using OptiFit without
-impacting model performance.
+classifying samples as either control or SRN. The VSEARCH Greengenes
+method performed slightly worse than the OptiClust \emph{de novo} method
+(p-value = 0.030). However the performance on the test data for the
+OptiClust \emph{de novo}, OptiFit Greengenes, OptiFit self-reference,
+and VSEARCH \emph{de novo} approaches were not significantly different
+(p-values \textgreater{} 0.05; Figures 2B and 2C). These results
+indicate that new data could be fit to existing OTU clusters using
+OptiFit without impacting model performance.
 

### 2023-04-27 submission/response_to_reviewer.Rmd
second round revision with pat
update line numbers

@@
 
-Machine learning classification using self-reference-based OTU clustering
+```{r,include=F}
+# Note: These line numbers correspond to the word document in the resubmission folder
+```
+
+Machine learning classification by fitting amplicon sequences to existing OTUs
 
@@
 We added the following methods: reference based clustring using OptiFit and the GreenGenes reference, \*de novo\* clustering using VSEARCH, and reference-based clustering using VSEARCH and the GreenGenes reference.
-See lines 47 - 64 and figure 1 for workflow descriptions and lines 87 - 98 and Figure 2 for the results.}
+See lines 53 - 74 and Figure 1 for workflow descriptions and lines 99 - 112 and Figure 2 for the results.}
 
@@
 This likely indicates that the model depends on well clustered OTUs.
-We've added some discussion on this to the paper (lines: ).}
+We've added some discussion on this to the paper (lines: 118 - 122).}

### 2023-06-15 submission/second_revision/manuscript.Rmd
second revision documents

--- 2023-04-27 submission/manuscript.Rmd
@@
 opts_chunk$set("cache" = FALSE)
+opts_chunk$set("include" = FALSE)
 
@@
 
-Machine learning classification using the gut microbiome relies on
-assigning 16S rRNA gene sequences into operational taxonomic units
-(OTUs) to quantify microbial composition. OTU abundances are then used
-to train a classification model that can be applied to classify new
-samples. The standard approaches to clustering sequences include
-reference-based and *de novo* clustering. Reference-based clustering
-requires a well-curated reference database that may not exist for all
-systems. *De novo* clustering tends to produce higher quality OTU
-assignments than reference-based, but clusters depend on the sequences
-in the dataset and therefore OTU assignments will change when new
-samples are sequenced. This lack of stability complicates machine
-learning classification since new sequences must be reclustered with the
-old data and the model retrained with the new OTU assignments. The
-OptiFit algorithm addresses these issues by fitting new sequences into
-existing OTUs. While OptiFit produces high quality OTU clusters, it is
-unclear whether this method for fitting new sequence data into existing
-OTUs will impact the performance of classification models trained with
-the older data. We used OptiFit to cluster sequences into existing OTUs
-and evaluated model performance in classifying a dataset containing
+The ability to use 16S rRNA gene sequence data to train machine learning
+classification models offers the opportunity diagnose patients based on
+the composition of their microbiome. In some applications the taxonomic
+resolution that provides the best models may require the use of *de
+novo* OTUs whose composition changes when new data are added. We
+previously developed a new reference-based approach, OptiFit, that fits
+new sequence data to existing *de novo* OTUs without changing the
+composition of the original OTUs. While OptiFit produces OTUs that are
+as high quality as *de novo* OTUs, it is unclear whether this method for
+fitting new sequence data into existing OTUs will impact the performance
+of classification models relative to models trained and tested only
+using *de novo* OTUs. We used OptiFit to cluster sequences into existing
+OTUs and evaluated model performance in classifying a dataset containing
 samples from patients with and without colonic screen relevant neoplasia
@@
 
+\newpage
+
 ## Importance
@@
 There is great potential for using microbiome data to aid in diagnosis.
-A challenge with OTU-based classification models is that 16S rRNA gene
-sequences are often assigned to OTUs based on similarity to other
-sequences in the dataset. If data are generated from new patients, the
-old and new sequences must all be reassigned to OTUs and the
+A challenge with *de novo* OTU-based classification models is that 16S
+rRNA gene sequences are often assigned to OTUs based on similarity to
+other sequences in the dataset. If data are generated from new patients,
+the old and new sequences must be reclustered to OTUs and the
 classification model retrained. Yet there is a desire to have a single,
@@
 
-There is increasing evidence for an association between the composition of the gut microbiome and a variety of diseases, such as crohn's disease and colorectal cancer [@Morgan2012;@Sobhani2011].
-There is great potential to diagnose disease with gut microbiome sequence data and machine learning.
-Taxonomic composition of microbial
-communities can be assessed using amplicon sequencing of the 16S rRNA
-gene, which is the input to classification models. Analysis of 16S rRNA
-gene sequence data generally relies on assigning sequences into
-operational taxonomic units (OTUs). The process of OTU clustering can
-either be reference-based or *de novo*. The quality of OTUs generated
-with reference-based clustering is generally poor compared to those
-generated with *de novo* clustering @westcott2015.
-While *de novo* clustering produces high-quality OTU clusters where sequences are accurately grouped based on similarity thresholds, the resulting OTU clusters depend on the sequences within the dataset and the addition of new data has the potential to redefine OTU cluster composition.
-The unstable nature of *de novo* OTU clustering complicates deployment of
-machine learning models since integration of additional data requires
-reclustering all the data and retraining the model. The ability to
-integrate new data into a validated model without reclustering and
-retraining could allow for the application of a single model that can
-continually classify new data. Recently, Sovacool *et al.* introduced
-OptiFit, a method for fitting new sequence data into existing OTUs
-@sovacool2022. While OptiFit can effectively fit new sequence data to
-existing OTU clusters, it is unknown if the use of OptiFit will have an
-impact on classification performance. Here, we tested the ability of
-OptiFit to cluster new sequence data into existing OTU clusters for the
-purpose of classifying disease based on gut microbiome composition.
+There is increasing interest in training machine learning models to
+diagnose diseases such as Crohn's disease and colorectal cancer using
+the relative abundance of clusters of similar 16S rRNA gene sequences
+[@baxter2016; @gevers2014]. These models have been used to identify
+sequence clusters that are important for distinguishing between
+individuals from different disease categories @duvallet2017. There is
+also an opportunity to train models and apply them to classify samples
+from new individuals. For example, a model for colorectal cancer could
+be trained, "locked down", and applied to samples from new patients.
 
-We compared the ability of several approaches for assigning 16S rRNA
-gene sequences to OTUs including, *de novo* and reference-based clustering.
-For reference-based clustering, we used closed-reference clustering to a public database (database-reference-based) and to OTUs generated from a subset of the samples (self-reference-based).
-To test how the model performance
-compared between these approaches, we used a publicly available dataset
-of 16S rRNA gene sequences from stool samples of healthy subjects (n =
-226) as well as subjects with screen-relevant neoplasia (SRN) consisting
-of advanced adenoma and carcinoma (n = 229) @baxter2016. For the *de
-novo* workflows, all the 16S rRNA sequence data was clustered into OTUs.
-The OTU clustering was conducted using two common algorithms: 1) the
-OptiClust algorithm in mothur [@westcott2017] and 2) the VSEARCH algorithm
-used in QIIME2 [@rognes2016; @bolyen2019]. For both algorithms, the
-resulting abundance data was then split into training and testing sets,
-where the training set was used to tune hyperparameters and ultimately
-train and select the model. The model was applied to the testing set and
-performance evaluated (Figure 1A). We also conducted reference-based OTU
-clustering using OptiFit to fit the sequence data into OTUs based on the
-greengenes reference database. To compare with another commonly used
-method, we also used the VSEARCH algorithm to fit the sequence data to
-the greengenes reference (Figure 1B). In the OptiFit self-reference
-workflow, the data was split into a training and a testing set. The
-training set was clustered into OTUs and used to train a classification
-model. The OptiFit algorithm was used to fit sequence data of samples
-not part of the original dataset into the existing OTUs, and used the
-same model to classify the samples (Figure 1C). For each of the
-workflows the process was repeated for 100 random splits of the data to
-account for variation caused by the choice of the random number
+To apply these models to new samples the composition of the clusters
+would need to be independent of the new data. For example, amplicon
+sequence variants (ASVs) are defined without consideration of sequences
+in other samples, phylotypes are defined by clustering sequences that
+have the same taxonomy (e.g., to the same family) when classified using
+a taxonomy database, and closed reference operational taxonomic units
+(OTUs) are defined by mapping sequences to a collection of reference
+OTUs. In contrast, *de novo* approaches cluster sequences based on their
+similarity to other sequences in the dataset and can change when new
+data are added. Although it would be preferrable to select an approach
+that generates stable clusters, there may be cases where OTUs generated
+by a *de novo* approach outperform those of the other taxonomic levels.
+In fact, we recently trained machine learning models for classifying
+patients with and without screen relevant neoplasias (SRNs) in their
+colons and found that OTUs generated *de novo* using the OptiClust
+algorithm performed better than those generated using ASVs or at higher
+taxonomic levels @Armour2022.
+
+It could be possible to construct reference OTUs and map new sequences
+to those OTUs to attain similar performance as was seen with the
+OptiClust-generated OTUs. The traditional approach to reference-based
+clustering of sequences to OTUs has multiple drawbacks and
+does not produce clusters as good as those generated using OptiClust
+@sovacool2022. Sovacool *et al* recently described OptiFit, a method for
+fitting new sequence data into existing OTUs that overcomes the
+limitations of traditional reference-based clustering @sovacool2022.
+OptiFit allows researchers to fit new data into existing OTUs defined
+from the same dataset resulting in clusters that are as good as if they
+had all been clustered with OptiClust. We tested whether
+OptiClust-generated OTUs could be used to train models that were then
+used to classify held out samples after clustering their sequences to
+the model's OTUs using OptiFit.
+
+To test how the model performance compared between using *de novo* and
+reference-based clustering approaches, we used a publicly available
+dataset of 16S rRNA gene sequences from stool samples of healthy
+subjects (n = 226) as well as subjects with screen-relevant neoplasia
+(SRN) consisting of advanced adenoma and carcinoma (n = 229)
+@baxter2016. For the *de novo* workflows, the 16S rRNA sequence data
+from all samples were clustered into OTUs using the OptiClust algorithm
+in mothur @westcott2017 and the VSEARCH algorithm used in QIIME2
+[@rognes2016; @bolyen2019]. For both algorithms, the resulting
+abundance data was then split into training and testing sets, where the
+training set was used to tune hyperparameters and ultimately train and
+select the model. The model was applied to the testing set and
+performance evaluated (Figure 1A). For traditional reference-based
+clustering (database-reference-based), we used OptiFit to fit the
+sequence data into OTUs based on the commonly used greengenes reference
+database. To compare with another commonly used method, we also used
+VSEARCH to map sequences to reference OTUs from the greengenes database
+with the parameters used by QIIME2. Again, the data was then split into
+training and testing sets, hyperparameters tuned, and performance
+evaluated on the testing set (Figure 1B). In the OptiFit self-reference
+workflow (self-reference-based), the data was split into a training and
+a testing set. The training set was clustered into OTUs and used to
+train a classification model. The OptiFit algorithm was used to fit
+sequence data of samples not part of the training data into the training
+OTUs and classified using the best hyperparameters (Figure 1C). For each
+of the workflows the process was repeated for 100 random splits of the
+data to account for variation caused by the choice of the random number
 generator seed.
 
-```{r mcc, include=F}
-merged_mcc <- read_csv("../results/ml/summary/merged_mcc.csv")
+```{r mcc}
+merged_mcc <- read_csv("../../results/ml/summary/merged_mcc.csv")
 
@@
 
-```{r frac_mapped, include=F}
-counts <- read_csv("../results/tables/counts.csv")
+```{r frac_mapped}
+counts <- read_csv("../../results/tables/counts.csv")
 
@@
 We first examined the quality of the resulting OTU clusters from each
-method using the Matthews correlation coefficient (MCC). MCC is a metric
-used to measure OTU cluster quality based on the similarity of all pairs
-of sequences and whether they are appropriately clustered or not
-@westcott2015. MCC scores range between negative one and one, and
-measure how well clustering assignment correlates with the distance
-between sequences. To ensure that OptiFit appropriately integrated new
-sequence data into the existing OTUs, we expected the MCC scores
-produced by the OptiFit workflow to be similar to that of *de novo*
-clustering using the OptiClust algorithm. In the OptiFit workflow the
-test data was fit to the clustered training data for each of the 100
-data splits resulting in an MCC score for each split of the data. In the
-remaining workflows, the data was only clustered once and then split
-into the training and testing sets resulting in a single MCC score for
-each method. Indeed, the MCC scores were similar between the OptiClust
-*de novo* (MCC = `r opticlust_denovo_mcc`) and OptiFit self-reference
-workflows (average MCC = `r optifit_self_avg_mcc %>% pull(mean_mcc)`,
-standard deviation = `r optifit_self_avg_mcc %>% pull(sd_mcc)`).
-Consistent with prior findings, the reference-based methods produced
-lower MCC scores (OptiFit Greengenes MCC = `r optifit_gg_mcc`; VSEARCH
-Greengenes MCC = `r vsearch_gg_mcc`) than the *de novo* methods (OptiClust *de novo* MCC = `r opticlust_denovo_mcc`;
-VSEARCH *de novo* MCC = `r vsearch_denovo_mcc`) @sovacool2022. Another metric we
-examined for the OptiFit workflow was the fraction of sequences from the
-test set that mapped to the reference OTUs. Since sequences that did not
-map to reference OTUs were eliminated, if a high percentage of reads did
-not map to an OTU we expected this loss of data to negatively impact
+method using the Matthews correlation coefficient (MCC). MCC is an
+objective metric used to measure OTU cluster quality based on the
+similarity of all pairs of sequences and whether they are appropriately
+clustered or not @westcott2015. We expected the MCC scores produced by
+the OptiFit workflow to be similar to that of *de novo* clustering using
+the OptiClust algorithm. In the OptiFit workflow the test data was fit
+to the clustered training data for each of the 100 data splits resulting
+in an MCC score for each split of the data. In the remaining workflows,
+the data was only clustered once and then split into the training and
+testing sets resulting in a single MCC score for each method. Indeed,
+the MCC scores were similar between the OptiClust *de novo* (MCC =
+`r opticlust_denovo_mcc`) and OptiFit self-reference workflows (average
+MCC = `r optifit_self_avg_mcc %>% pull(mean_mcc)`, standard deviation =
+`r optifit_self_avg_mcc %>% pull(sd_mcc)`). Consistent with prior
+findings, the reference-based methods produced lower MCC scores (OptiFit
+greengenes MCC = `r optifit_gg_mcc`; VSEARCH greengenes MCC =
+`r vsearch_gg_mcc`) than the *de novo* methods (OptiClust *de novo* MCC
+= `r opticlust_denovo_mcc`; VSEARCH *de novo* MCC =
+`r vsearch_denovo_mcc`) @sovacool2022. Another metric we examined for
+the OptiFit workflow was the fraction of sequences from the test set
+that mapped to the reference OTUs. Since sequences that did not map to
+reference OTUs were eliminated, if a high percentage of reads did not
+map to an OTU we expected this loss of data to negatively impact
 classification performance. We found that loss of data was not an issue
@@
 (standard deviation =
-`r avg_frac_mapped %>% filter(methods == "optifit_self", set == "test") %>% pull(sd_frac) %>% round(digits=2)`%)
-of sequences in the subsampled test set mapped to the reference OTUs. This number
-is higher than the average fraction of reads mapped in the OptiFit
-Greengenes workflow (
-`r avg_frac_mapped %>% filter(methods == "optifit_gg",set == "test") %>% pull(mean_frac) %>% round(digits=1)`%
-+/-
-`r avg_frac_mapped %>% filter(methods == "optifit_gg", set == "test") %>% pull(sd_frac) %>% round(digits=1)`).
+`r avg_frac_mapped %>% filter(methods == "optifit_self", set == "test") %>% pull(sd_frac) %>% round(digits=1)`%)
+of sequences in the subsampled test set mapped to the reference OTUs.
+This number is higher than the average fraction of reads mapped in the
+OptiFit greengenes workflow
+(mean = `r avg_frac_mapped %>% filter(methods == "optifit_gg",set == "test") %>% pull(mean_frac) %>% round(digits=1)`%, standard deviation = 
+`r avg_frac_mapped %>% filter(methods == "optifit_gg",set == "test") %>% pull(sd_frac) %>% round(digits=1)`).
 These results indicate that the OptiFit self-reference method performed
@@
 
-```{r}
-performance <- read_csv("../results/ml/summary/merged_performance.csv",
+```{r performance}
+performance <- read_csv("../../results/ml/summary/merged_performance.csv",
                     col_types = cols(algorithm=col_character(),
@@
 
-pvals <- read_csv("../results/tables/pvalues.csv",col_types = cols(p_value = col_double(),
+pvals <- read_csv("../../results/tables/pvalues.csv",col_types = cols(p_value = col_double(),
                                                                    .default = col_character()))
@@
 }
-#get_pval(pvals,"AUC","opticlust_denovo","vsearch_gg")
 ```
@@
 the training data from the workflows to train a model to predict SRNs
-and used the model on the held out data. Using the predicted and actual
+and used the model on the held-out data. Using the predicted and actual
 diagnosis classification, we calculated the area under the receiver
 operating characteristic curve (AUROC) for each data split. During
-cross-validation (CV) training, the performance of the OptiFit self-reference and OptiClust *de novo* models were not significantly different (p-value =
+cross-validation (CV) training, the performance of the OptiFit
+self-reference and OptiClust *de novo* models were not significantly
+different (p-value =
 `r get_pval(pvals, "cv_metric_AUC", "opticlust_denovo", "optifit_self")`;
-Figure 2A), while performance for both VSEARCH methods was significantly lower
-than the OptiClust *de novo*, OptiFit self, and OptiFit Greengenes
+Figure 2A), while performance for both VSEARCH methods was significantly
+lower than the OptiClust *de novo*, OptiFit self, and OptiFit greengenes
 methods (p-values \< 0.05). The trained model was then applied to the
-test data classifying samples as either control or SRN.
-The VSEARCH Greengenes method performed slightly worse than the OptiClust *de novo* method (p-value = `r get_pval(pvals,"AUC","opticlust_denovo","vsearch_gg")`).
-However the performance on the test data for the OptiClust *de novo*, OptiFit Greengenes, OptiFit self-reference, and VSEARCH *de novo* approaches were not significantly different (p-values \> 0.05; Figures 2B and 2C).
+test data classifying samples as either control or SRN. The VSEARCH
+greengenes method performed slightly worse than the OptiClust *de novo*
+method (p-value =
+`r get_pval(pvals,"AUC","opticlust_denovo","vsearch_gg")`). However, the
+performance on the test data for the OptiClust *de novo*, OptiFit
+greengenes, OptiFit self-reference, and VSEARCH *de novo* approaches
+were not significantly different (p-values \> 0.05; Figures 2B and 2C).
 These results indicate that new data could be fit to existing OTU
@@
 
-We tested the ability of OptiFit to integrate new data into existing
-OTUs for the purpose of machine learning classification using OTU
-relative abundance. A potential problem with using OptiFit is that any
-sequences from the new samples that do not map to the existing OTU
-clusters will be discarded, resulting in a possible loss of information.
-However, we demonstrated that OptiFit can be used to fit new sequence
-data into existing OTU clusters and it could perform as well in
-predicting SRN compared to *de novo* clustering all the sequence data
-together. In this instance, the performance of OptiFit was equivalent to
-using a database-reference-based method despite the lower quality of the
-OTU clusters in the database-reference-based approach. This likely
-indicates that the sequences that are important to the model are well
-characterized by the reference database. However, a less well studied
-system may not be as well characterized by a reference-database which
-would make the ability to utilize one's own data a reference an exciting
-possiblility. The ability to integrate data from new samples into
-existing OTUs enables the implementation of a single machine learning
-model. This is important for model implementation because not all of the
-data needs to be available or known at the time of model generation. A
-robust machine learning model can be implemented as part of a
-non-invasive and low-cost diagnostic for SRN and other diseases.
+Random forest machine learning models trained using OptiClust-generated
+OTUs and tested using OptiFit-generated OTUs performed as well as a
+model trained using entirely *de novo* OTU assignments. A potential
+problem with reference-based clustering methods is that sequences that
+do not map to the reference OTUs are discarded, resulting in a possible
+loss of information. However, we demonstrated that the training samples
+represented the most important OTUs for classifying samples. Missing
+important OTUs is more of a risk when using a database-reference-based
+method since not all environments are well represented in public
+databases. Despite this and the lower quality OTUs, the
+database-reference-based approach performed as well as the models
+generated using OptiFit. This likely indicates that the sequences that
+were important to the model were well characterized by the greengenes
+reference OTUs. However, a less well studied system may not be as well
+characterized by a reference-database which would make the ability to
+utilize one's own data a reference an exciting possibility. Our results
+highlight that OptiFit overcomes a significant limitation with machine
+learning models trained using *de novo* OTUs. This is an important
+result for those applications where models trained using *de novo* OTUs
+outperform models generated using methods that produce clusters that do
+not depend on which sequences are included in the dataset.
 
@@
 from the NCBI Sequence Read Archive (accession no. SRP062005)
-[@edgar2011; @baxter2016]. This dataset contains stool samples from 490
-subjects. For this analysis, samples from subjects identified in the
-metadata as normal, high risk normal, or adenoma were categorized as
-"normal", while samples from subjects identified as advanced adenoma or
-carcinoma were categorized as "screen relevant neoplasia" (SRN). The
-resulting dataset consisted of 261 normal samples and 229 SRN samples.
-
-```{r n_test_train,include=F}
-n_test_train <- read_csv("../results/tables/train_test_frequency.csv") %>% 
-  pivot_wider(names_from="train_test",values_from = "n") %>% 
-  summarize(avg_test = mean(test),
-            sd_test = sd(test),
-            avg_train = mean(train),
-            sd_train = sd(train))
-```
+@baxter2016. This dataset contains stool samples from 490 subjects. For
+this analysis, samples from subjects identified in the metadata as
+normal, high risk normal, or adenoma were categorized as "normal", while
+samples from subjects identified as advanced adenoma or carcinoma were
+categorized as "screen relevant neoplasia" (SRN). The resulting dataset
+consisted of 261 normal samples and 229 SRN samples.
 
@@
 training or test sets. Each sample was in the training set an average of
-`r n_test_train %>% pull(avg_train)` times (standard deviation =
-`r format_decimal(n_test_train %>% pull(sd_train),digits = 1)`) and the
-test set an average of `r n_test_train %>% pull(avg_test)` times
-(standard deviation =
-`r format_decimal(n_test_train %>% pull(sd_test),digits = 1)`).
+80 times (standard deviation = 4.1) and the test set an average of 20
+times (standard deviation = 4.1).
 
@@
     OptiClust, then the test set was fit to the OTUs of the training set
-    using the OptiFit algorithm @sovacool2022. The OptiFit algorithm
-    was run with method open so that any sequences that did not map to
-    the existing OTU clusters would form new OTUs. The data was then
+    using the OptiFit algorithm @sovacool2022. The OptiFit algorithm was
+    run with method open so that any sequences that did not map to the
+    existing OTU clusters would form new OTUs. The data was then
     subsampled to 10,000 reads and any novel OTUs from the test set were
@@
 2.  OptiFit Greengenes: Reference sequences from the Greengenes database
-    v13_8_99 @desantis2006 were downloaded and processed with mothur by
-    trimming to the V4 region and clustered *de novo* with OptiClust [@westcott2017]. The preprocessed data was fit to the clustered
+    v13_8\_99 @desantis2006 were downloaded and processed with mothur by
+    trimming to the V4 region and clustered *de novo* with OptiClust
+    @westcott2017. The preprocessed data was fit to the clustered
     reference data using OptiFit with the method open to allow any
@@
     and testing datasets.
-
 5.  VSEARCH *de novo*: All the preprocessed data was clustered using
@@
 
-***Machine Learning.*** A random forest model was trained with the R
+**Machine Learning.** A random forest model was trained with the R
 package mikrompl (v 1.2.0) @topçuoglu2021 to predict the diagnosis (SRN
@@
 
-***Code Availability.***
-
-The analysis workflow was implemented in Snakemake [@koster2012]. Scripts
-for analysis were written in R @R2020 and GNU bash @GNUbash. The
-software used includes mothur v1.47.0 @schloss2009, VSEARCH v2.15.2
-@rognes2016, RStudio @RStudio2019, the Tidyverse metapackage
-@wickham2019, R Markdown @xie_r_2018, the SRA toolkit
-@noauthor_sra-tools_nodate, and conda @noauthor_anaconda_2016. The
-complete workflow and supporting files required to reproduce this study
-are available at:
-<https://github.com/SchlossLab/Armour_OptiFitGLNE_mBio_2023>
+**Code Availability.** The analysis workflow was implemented in
+Snakemake @koster2012. Scripts for analysis were written in R @R2020 and
+GNU bash @GNUbash. The software used includes mothur v1.47.0
+@schloss2009, VSEARCH v2.15.2 @rognes2016, RStudio @RStudio2019, the
+Tidyverse metapackage @wickham2019, R Markdown @xie_r_2018, the SRA
+toolkit @noauthor_sra-tools_nodate, and conda @noauthor_anaconda_2016.
+The complete workflow and supporting files required to reproduce this
+study are available at:
+<https://github.com/SchlossLab/Armour_OptiFitGLNE_XXXX_2023>
 
@@
 
-## Figures
-
-![](./figures/Figure1.png)
+## Figure Legends
 
@@
 OptiClust with mothur and VSEARCH as is used in the QIIME pipeline.
-
-\newpage
-
-![](figures/fig2.png)
 

### 2023-06-15 submission/second_revision/manuscript.tex
second revision documents

--- 2023-04-27 submission/manuscript.tex
@@
 
-Machine learning classification using the gut microbiome relies on
-assigning 16S rRNA gene sequences into operational taxonomic units
-(OTUs) to quantify microbial composition. OTU abundances are then used
-to train a classification model that can be applied to classify new
-samples. The standard approaches to clustering sequences include
-reference-based and \emph{de novo} clustering. Reference-based
-clustering requires a well-curated reference database that may not exist
-for all systems. \emph{De novo} clustering tends to produce higher
-quality OTU assignments than reference-based, but clusters depend on the
-sequences in the dataset and therefore OTU assignments will change when
-new samples are sequenced. This lack of stability complicates machine
-learning classification since new sequences must be reclustered with the
-old data and the model retrained with the new OTU assignments. The
-OptiFit algorithm addresses these issues by fitting new sequences into
-existing OTUs. While OptiFit produces high quality OTU clusters, it is
-unclear whether this method for fitting new sequence data into existing
-OTUs will impact the performance of classification models trained with
-the older data. We used OptiFit to cluster sequences into existing OTUs
-and evaluated model performance in classifying a dataset containing
-samples from patients with and without colonic screen relevant neoplasia
-(SRN). We compared the performance of this model to standard methods
-including \emph{de novo} and database-reference-based clustering. We
-found that using OptiFit performed as well or better in classifying
-SRNs. OptiFit can streamline the process of classifying new samples by
-avoiding the need to retrain models using reclustered sequences.
+The ability to use 16S rRNA gene sequence data to train machine learning
+classification models offers the opportunity diagnose patients based on
+the composition of their microbiome. In some applications the taxonomic
+resolution that provides the best models may require the use of \emph{de
+novo} OTUs whose composition changes when new data are added. We
+previously developed a new reference-based approach, OptiFit, that fits
+new sequence data to existing \emph{de novo} OTUs without changing the
+composition of the original OTUs. While OptiFit produces OTUs that are
+as high quality as \emph{de novo} OTUs, it is unclear whether this
+method for fitting new sequence data into existing OTUs will impact the
+performance of classification models relative to models trained and
+tested only using \emph{de novo} OTUs. We used OptiFit to cluster
+sequences into existing OTUs and evaluated model performance in
+classifying a dataset containing samples from patients with and without
+colonic screen relevant neoplasia (SRN). We compared the performance of
+this model to standard methods including \emph{de novo} and
+database-reference-based clustering. We found that using OptiFit
+performed as well or better in classifying SRNs. OptiFit can streamline
+the process of classifying new samples by avoiding the need to retrain
+models using reclustered sequences.
+
+\newpage
 
@@
 There is great potential for using microbiome data to aid in diagnosis.
-A challenge with OTU-based classification models is that 16S rRNA gene
-sequences are often assigned to OTUs based on similarity to other
-sequences in the dataset. If data are generated from new patients, the
-old and new sequences must all be reassigned to OTUs and the
+A challenge with \emph{de novo} OTU-based classification models is that
+16S rRNA gene sequences are often assigned to OTUs based on similarity
+to other sequences in the dataset. If data are generated from new
+patients, the old and new sequences must be reclustered to OTUs and the
 classification model retrained. Yet there is a desire to have a single,
@@
 
-There is increasing evidence for an association between the composition
-of the gut microbiome and a variety of diseases, such as crohn's disease
-and colorectal cancer (\protect\hyperlink{ref-Morgan2012}{1},
-\protect\hyperlink{ref-Sobhani2011}{2}). There is great potential to
-diagnose disease with gut microbiome sequence data and machine learning.
-Taxonomic composition of microbial communities can be assessed using
-amplicon sequencing of the 16S rRNA gene, which is the input to
-classification models. Analysis of 16S rRNA gene sequence data generally
-relies on assigning sequences into operational taxonomic units (OTUs).
-The process of OTU clustering can either be reference-based or \emph{de
-novo}. The quality of OTUs generated with reference-based clustering is
-generally poor compared to those generated with \emph{de novo}
-clustering (\protect\hyperlink{ref-westcott2015}{3}). While \emph{de
-novo} clustering produces high-quality OTU clusters where sequences are
-accurately grouped based on similarity thresholds, the resulting OTU
-clusters depend on the sequences within the dataset and the addition of
-new data has the potential to redefine OTU cluster composition. The
-unstable nature of \emph{de novo} OTU clustering complicates deployment
-of machine learning models since integration of additional data requires
-reclustering all the data and retraining the model. The ability to
-integrate new data into a validated model without reclustering and
-retraining could allow for the application of a single model that can
-continually classify new data. Recently, Sovacool \emph{et al.}
-introduced OptiFit, a method for fitting new sequence data into existing
-OTUs (\protect\hyperlink{ref-sovacool2022}{4}). While OptiFit can
-effectively fit new sequence data to existing OTU clusters, it is
-unknown if the use of OptiFit will have an impact on classification
-performance. Here, we tested the ability of OptiFit to cluster new
-sequence data into existing OTU clusters for the purpose of classifying
-disease based on gut microbiome composition.
+There is increasing interest in training machine learning models to
+diagnose diseases such as Crohn's disease and colorectal cancer using
+the relative abundance of clusters of similar 16S rRNA gene sequences
+(\protect\hyperlink{ref-baxter2016}{1},
+\protect\hyperlink{ref-gevers2014}{2}). These models have been used to
+identify sequence clusters that are important for distinguishing between
+individuals from different disease categories
+(\protect\hyperlink{ref-duvallet2017}{3}). There is also an opportunity
+to train models and apply them to classify samples from new individuals.
+For example, a model for colorectal cancer could be trained, ``locked
+down'', and applied to samples from new patients.
 
-We compared the ability of several approaches for assigning 16S rRNA
-gene sequences to OTUs including, \emph{de novo} and reference-based
-clustering. For reference-based clustering, we used closed-reference
-clustering to a public database (database-reference-based) and to OTUs
-generated from a subset of the samples (self-reference-based). To test
-how the model performance compared between these approaches, we used a
-publicly available dataset of 16S rRNA gene sequences from stool samples
-of healthy subjects (n = 226) as well as subjects with screen-relevant
-neoplasia (SRN) consisting of advanced adenoma and carcinoma (n = 229)
-(\protect\hyperlink{ref-baxter2016}{5}). For the \emph{de novo}
-workflows, all the 16S rRNA sequence data was clustered into OTUs. The
-OTU clustering was conducted using two common algorithms: 1) the
-OptiClust algorithm in mothur (\protect\hyperlink{ref-westcott2017}{6})
-and 2) the VSEARCH algorithm used in QIIME2
-(\protect\hyperlink{ref-rognes2016}{7},
-\protect\hyperlink{ref-bolyen2019}{8}). For both algorithms, the
+To apply these models to new samples the composition of the clusters
+would need to be independent of the new data. For example, amplicon
+sequence variants (ASVs) are defined without consideration of sequences
+in other samples, phylotypes are defined by clustering sequences that
+have the same taxonomy (e.g., to the same family) when classified using
+a taxonomy database, and closed reference operational taxonomic units
+(OTUs) are defined by mapping sequences to a collection of reference
+OTUs. In contrast, \emph{de novo} approaches cluster sequences based on
+their similarity to other sequences in the dataset and can change when
+new data are added. Although it would be preferrable to select an
+approach that generates stable clusters, there may be cases where OTUs
+generated by a \emph{de novo} approach outperform those of the other
+taxonomic levels. In fact, we recently trained machine learning models
+for classifying patients with and without screen relevant neoplasias
+(SRNs) in their colons and found that OTUs generated \emph{de novo}
+using the OptiClust algorithm performed better than those generated
+using ASVs or at higher taxonomic levels
+(\protect\hyperlink{ref-Armour2022}{4}).
+
+It could be possible to construct reference OTUs and map new sequences
+to those OTUs to attain similar performance as was seen with the
+OptiClust-generated OTUs. The traditional approach to reference-based
+clustering of sequences to OTUs has multiple drawbacks and does not
+produce clusters as good as those generated using OptiClust
+(\protect\hyperlink{ref-sovacool2022}{5}). Sovacool \emph{et al}
+recently described OptiFit, a method for fitting new sequence data into
+existing OTUs that overcomes the limitations of traditional
+reference-based clustering (\protect\hyperlink{ref-sovacool2022}{5}).
+OptiFit allows researchers to fit new data into existing OTUs defined
+from the same dataset resulting in clusters that are as good as if they
+had all been clustered with OptiClust. We tested whether
+OptiClust-generated OTUs could be used to train models that were then
+used to classify held out samples after clustering their sequences to
+the model's OTUs using OptiFit.
+
+To test how the model performance compared between using \emph{de novo}
+and reference-based clustering approaches, we used a publicly available
+dataset of 16S rRNA gene sequences from stool samples of healthy
+subjects (n = 226) as well as subjects with screen-relevant neoplasia
+(SRN) consisting of advanced adenoma and carcinoma (n = 229)
+(\protect\hyperlink{ref-baxter2016}{1}). For the \emph{de novo}
+workflows, the 16S rRNA sequence data from all samples were clustered
+into OTUs using the OptiClust algorithm in mothur 1 and the VSEARCH
+algorithm used in QIIME2 (\protect\hyperlink{ref-rognes2016}{6},
+\protect\hyperlink{ref-bolyen2019}{7}). For both algorithms, the
 resulting abundance data was then split into training and testing sets,
@@
 train and select the model. The model was applied to the testing set and
-performance evaluated (Figure 1A). We also conducted reference-based OTU
-clustering using OptiFit to fit the sequence data into OTUs based on the
-greengenes reference database. To compare with another commonly used
-method, we also used the VSEARCH algorithm to fit the sequence data to
-the greengenes reference (Figure 1B). In the OptiFit self-reference
-workflow, the data was split into a training and a testing set. The
-training set was clustered into OTUs and used to train a classification
-model. The OptiFit algorithm was used to fit sequence data of samples
-not part of the original dataset into the existing OTUs, and used the
-same model to classify the samples (Figure 1C). For each of the
-workflows the process was repeated for 100 random splits of the data to
-account for variation caused by the choice of the random number
+performance evaluated (Figure 1A). For traditional reference-based
+clustering (database-reference-based), we used OptiFit to fit the
+sequence data into OTUs based on the commonly used greengenes reference
+database. To compare with another commonly used method, we also used
+VSEARCH to map sequences to reference OTUs from the greengenes database
+with the parameters used by QIIME2. Again, the data was then split into
+training and testing sets, hyperparameters tuned, and performance
+evaluated on the testing set (Figure 1B). In the OptiFit self-reference
+workflow (self-reference-based), the data was split into a training and
+a testing set. The training set was clustered into OTUs and used to
+train a classification model. The OptiFit algorithm was used to fit
+sequence data of samples not part of the training data into the training
+OTUs and classified using the best hyperparameters (Figure 1C). For each
+of the workflows the process was repeated for 100 random splits of the
+data to account for variation caused by the choice of the random number
 generator seed.
@@
 We first examined the quality of the resulting OTU clusters from each
-method using the Matthews correlation coefficient (MCC). MCC is a metric
-used to measure OTU cluster quality based on the similarity of all pairs
-of sequences and whether they are appropriately clustered or not
-(\protect\hyperlink{ref-westcott2015}{3}). MCC scores range between
-negative one and one, and measure how well clustering assignment
-correlates with the distance between sequences. To ensure that OptiFit
-appropriately integrated new sequence data into the existing OTUs, we
-expected the MCC scores produced by the OptiFit workflow to be similar
-to that of \emph{de novo} clustering using the OptiClust algorithm. In
-the OptiFit workflow the test data was fit to the clustered training
-data for each of the 100 data splits resulting in an MCC score for each
-split of the data. In the remaining workflows, the data was only
-clustered once and then split into the training and testing sets
-resulting in a single MCC score for each method. Indeed, the MCC scores
-were similar between the OptiClust \emph{de novo} (MCC = 0.884) and
-OptiFit self-reference workflows (average MCC = 0.879, standard
-deviation = 0.002). Consistent with prior findings, the reference-based
-methods produced lower MCC scores (OptiFit Greengenes MCC = 0.786;
-VSEARCH Greengenes MCC = 0.531) than the \emph{de novo} methods
-(OptiClust \emph{de novo} MCC = 0.884; VSEARCH \emph{de novo} MCC =
-0.641) (\protect\hyperlink{ref-sovacool2022}{4}). Another metric we
-examined for the OptiFit workflow was the fraction of sequences from the
-test set that mapped to the reference OTUs. Since sequences that did not
-map to reference OTUs were eliminated, if a high percentage of reads did
-not map to an OTU we expected this loss of data to negatively impact
+method using the Matthews correlation coefficient (MCC). MCC is an
+objective metric used to measure OTU cluster quality based on the
+similarity of all pairs of sequences and whether they are appropriately
+clustered or not (\protect\hyperlink{ref-westcott2015}{8}). We expected
+the MCC scores produced by the OptiFit workflow to be similar to that of
+\emph{de novo} clustering using the OptiClust algorithm. In the OptiFit
+workflow the test data was fit to the clustered training data for each
+of the 100 data splits resulting in an MCC score for each split of the
+data. In the remaining workflows, the data was only clustered once and
+then split into the training and testing sets resulting in a single MCC
+score for each method. Indeed, the MCC scores were similar between the
+OptiClust \emph{de novo} (MCC = 0.884) and OptiFit self-reference
+workflows (average MCC = 0.879, standard deviation = 0.002). Consistent
+with prior findings, the reference-based methods produced lower MCC
+scores (OptiFit greengenes MCC = 0.786; VSEARCH greengenes MCC = 0.531)
+than the \emph{de novo} methods (OptiClust \emph{de novo} MCC = 0.884;
+VSEARCH \emph{de novo} MCC = 0.641)
+(\protect\hyperlink{ref-sovacool2022}{5}). Another metric we examined
+for the OptiFit workflow was the fraction of sequences from the test set
+that mapped to the reference OTUs. Since sequences that did not map to
+reference OTUs were eliminated, if a high percentage of reads did not
+map to an OTU we expected this loss of data to negatively impact
 classification performance. We found that loss of data was not an issue
-since on average 99.8\% (standard deviation = 0.68\%) of sequences in
-the subsampled test set mapped to the reference OTUs. This number is
-higher than the average fraction of reads mapped in the OptiFit
-Greengenes workflow ( 96.8\% +/- 3.5). These results indicate that the
-OptiFit self-reference method performed as well as the OptiClust
-\emph{de novo} method and better than using an external database.
+since on average 99.8\% (standard deviation = 0.7\%) of sequences in the
+subsampled test set mapped to the reference OTUs. This number is higher
+than the average fraction of reads mapped in the OptiFit greengenes
+workflow (mean = 96.8\%, standard deviation = 3.5). These results
+indicate that the OptiFit self-reference method performed as well as the
+OptiClust \emph{de novo} method and better than using an external
+database.
 
@@
 the training data from the workflows to train a model to predict SRNs
-and used the model on the held out data. Using the predicted and actual
+and used the model on the held-out data. Using the predicted and actual
 diagnosis classification, we calculated the area under the receiver
@@
 for both VSEARCH methods was significantly lower than the OptiClust
-\emph{de novo}, OptiFit self, and OptiFit Greengenes methods (p-values
+\emph{de novo}, OptiFit self, and OptiFit greengenes methods (p-values
 \textless{} 0.05). The trained model was then applied to the test data
-classifying samples as either control or SRN. The VSEARCH Greengenes
+classifying samples as either control or SRN. The VSEARCH greengenes
 method performed slightly worse than the OptiClust \emph{de novo} method
-(p-value = 0.030). However the performance on the test data for the
-OptiClust \emph{de novo}, OptiFit Greengenes, OptiFit self-reference,
+(p-value = 0.030). However, the performance on the test data for the
+OptiClust \emph{de novo}, OptiFit greengenes, OptiFit self-reference,
 and VSEARCH \emph{de novo} approaches were not significantly different
@@
 
-We tested the ability of OptiFit to integrate new data into existing
-OTUs for the purpose of machine learning classification using OTU
-relative abundance. A potential problem with using OptiFit is that any
-sequences from the new samples that do not map to the existing OTU
-clusters will be discarded, resulting in a possible loss of information.
-However, we demonstrated that OptiFit can be used to fit new sequence
-data into existing OTU clusters and it could perform as well in
-predicting SRN compared to \emph{de novo} clustering all the sequence
-data together. In this instance, the performance of OptiFit was
-equivalent to using a database-reference-based method despite the lower
-quality of the OTU clusters in the database-reference-based approach.
-This likely indicates that the sequences that are important to the model
-are well characterized by the reference database. However, a less well
-studied system may not be as well characterized by a reference-database
-which would make the ability to utilize one's own data a reference an
-exciting possiblility. The ability to integrate data from new samples
-into existing OTUs enables the implementation of a single machine
-learning model. This is important for model implementation because not
-all of the data needs to be available or known at the time of model
-generation. A robust machine learning model can be implemented as part
-of a non-invasive and low-cost diagnostic for SRN and other diseases.
+Random forest machine learning models trained using OptiClust-generated
+OTUs and tested using OptiFit-generated OTUs performed as well as a
+model trained using entirely \emph{de novo} OTU assignments. A potential
+problem with reference-based clustering methods is that sequences that
+do not map to the reference OTUs are discarded, resulting in a possible
+loss of information. However, we demonstrated that the training samples
+represented the most important OTUs for classifying samples. Missing
+important OTUs is more of a risk when using a database-reference-based
+method since not all environments are well represented in public
+databases. Despite this and the lower quality OTUs, the
+database-reference-based approach performed as well as the models
+generated using OptiFit. This likely indicates that the sequences that
+were important to the model were well characterized by the greengenes
+reference OTUs. However, a less well studied system may not be as well
+characterized by a reference-database which would make the ability to
+utilize one's own data a reference an exciting possibility. Our results
+highlight that OptiFit overcomes a significant limitation with machine
+learning models trained using \emph{de novo} OTUs. This is an important
+result for those applications where models trained using \emph{de novo}
+OTUs outperform models generated using methods that produce clusters
+that do not depend on which sequences are included in the dataset.
 
@@
 downloaded from the NCBI Sequence Read Archive (accession no. SRP062005)
-(\protect\hyperlink{ref-baxter2016}{5},
-\protect\hyperlink{ref-edgar2011}{9}). This dataset contains stool
+(\protect\hyperlink{ref-baxter2016}{1}). This dataset contains stool
 samples from 490 subjects. For this analysis, samples from subjects
@@
 \textbf{Data processing.} The full dataset was preprocessed with mothur
-(v1.47) (\protect\hyperlink{ref-schloss2009}{10}) to join forward and
+(v1.47) (\protect\hyperlink{ref-schloss2009}{9}) to join forward and
 reverse reads, merge duplicate reads, align to the SILVA reference
-database (v132) (\protect\hyperlink{ref-quast2013}{11}), precluster,
-remove chimeras with UCHIME (\protect\hyperlink{ref-edgar2011}{9}),
+database (v132) (\protect\hyperlink{ref-quast2013}{10}), precluster,
+remove chimeras with UCHIME (\protect\hyperlink{ref-edgar2011}{11}),
 assign taxonomy, and remove non-bacterial reads following the Schloss
@@
   OptiClust, then the test set was fit to the OTUs of the training set
-  using the OptiFit algorithm (\protect\hyperlink{ref-sovacool2022}{4}).
+  using the OptiFit algorithm (\protect\hyperlink{ref-sovacool2022}{5}).
   The OptiFit algorithm was run with method open so that any sequences
@@
   \emph{de novo} with OptiClust
-  (\protect\hyperlink{ref-westcott2017}{6}). The preprocessed data was
-  fit to the clustered reference data using OptiFit with the method open
-  to allow any sequences that did not map to the existing reference
-  clusters would form new OTUs. The data was then subsampled to 10,000
-  reads and any novel OTUs from the test set were removed. The dataset
-  was then split into two sets where 80\% of the samples were assigned
-  to the training set and 20\% to the testing set. This process was
-  repeated for each of the 100 splits resulting in 100 training and
-  testing datasets.
+
+  \begin{enumerate}
+  \def\labelenumii{\arabic{enumii}.}
+  \tightlist
+  \item
+    The preprocessed data was fit to the clustered reference data using
+    OptiFit with the method open to allow any sequences that did not map
+    to the existing reference clusters would form new OTUs. The data was
+    then subsampled to 10,000 reads and any novel OTUs from the test set
+    were removed. The dataset was then split into two sets where 80\% of
+    the samples were assigned to the training set and 20\% to the
+    testing set. This process was repeated for each of the 100 splits
+    resulting in 100 training and testing datasets.
+  \end{enumerate}
 \item
   VSEARCH Greengenes: Preprocessed data was clustered using VSEARCH
-  v2.15.2 (\protect\hyperlink{ref-rognes2016}{7}) directly to
+  v2.15.2 (\protect\hyperlink{ref-rognes2016}{6}) directly to
   unprocessed Greengenes 97\% OTU reference alignment consistent with
   how VSEARCH is typically used by the QIIME2 software for
-  reference-based clustering (\protect\hyperlink{ref-bolyen2019}{8}).
+  reference-based clustering (\protect\hyperlink{ref-bolyen2019}{7}).
   The data was then subsampled to 10,000 reads and any novel OTUs from
@@
 \setcounter{enumi}{3}
+\tightlist
 \item
   OptiClust \emph{de novo}: All the preprocessed data was clustered
-  together with OptiClust (\protect\hyperlink{ref-westcott2017}{6}) to
-  generate OTUs. The data was subsampled to 10,000 reads per sample and
-  the resulting abundance tables were split into the training and
-  testing sets. The process was repeated for each of the 100 splits
-  resulting in 100 training and testing datasets.
+  together with OptiClust 1 to generate OTUs. The data was subsampled to
+  10,000 reads per sample and the resulting abundance tables were split
+  into the training and testing sets. The process was repeated for each
+  of the 100 splits resulting in 100 training and testing datasets.
 \item
   VSEARCH \emph{de novo}: All the preprocessed data was clustered using
-  VSEARCH v2.15.2 (\protect\hyperlink{ref-rognes2016}{7}) with 97\%
+  VSEARCH v2.15.2 (\protect\hyperlink{ref-rognes2016}{6}) with 97\%
   identity and then subsampled to 10,000 reads per sample. The process
@@
 
-\textbf{\emph{Machine Learning.}} A random forest model was trained with
-the R package mikrompl (v 1.2.0)
+\textbf{Machine Learning.} A random forest model was trained with the R
+package mikrompl (v 1.2.0)
 (\protect\hyperlink{ref-topuxe7uoglu2021}{13}) to predict the diagnosis
@@
 
-\textbf{\emph{Code Availability.}}
-
-The analysis workflow was implemented in Snakemake
-(\protect\hyperlink{ref-koster2012}{15}). Scripts for analysis were
-written in R (\protect\hyperlink{ref-R2020}{16}) and GNU bash
+\textbf{Code Availability.} The analysis workflow was implemented in
+Snakemake (\protect\hyperlink{ref-koster2012}{15}). Scripts for analysis
+were written in R (\protect\hyperlink{ref-R2020}{16}) and GNU bash
 (\protect\hyperlink{ref-GNUbash}{17}). The software used includes mothur
-v1.47.0 (\protect\hyperlink{ref-schloss2009}{10}), VSEARCH v2.15.2
-(\protect\hyperlink{ref-rognes2016}{7}), RStudio
+v1.47.0 (\protect\hyperlink{ref-schloss2009}{9}), VSEARCH v2.15.2
+(\protect\hyperlink{ref-rognes2016}{6}), RStudio
 (\protect\hyperlink{ref-RStudio2019}{18}), the Tidyverse metapackage
@@
 available at:
-\url{https://github.com/SchlossLab/Armour_OptiFitGLNE_mBio_2023}
+\url{https://github.com/SchlossLab/Armour_OptiFitGLNE_XXXX_2023}
 
@@
 \begin{CSLReferences}{0}{1}
-\leavevmode\vadjust pre{\hypertarget{ref-Morgan2012}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-baxter2016}{}}%
 \CSLLeftMargin{1. }%
-\CSLRightInline{\textbf{Morgan XC}, \textbf{Tickle TL}, \textbf{Sokol
-H}, \textbf{Gevers D}, \textbf{Devaney KL}, \textbf{Ward DV},
-\textbf{Reyes JA}, \textbf{Shah SA}, \textbf{LeLeiko N}, \textbf{Snapper
-SB}, \textbf{Bousvaros A}, \textbf{Korzenik J}, \textbf{Sands BE},
-\textbf{Xavier RJ}, \textbf{Huttenhower C}. 2012. Dysfunction of the
-intestinal microbiome in inflammatory bowel disease and treatment.
-Genome Biol \textbf{13}:R79.}
+\CSLRightInline{\textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers
+MAM}, \textbf{Schloss PD}. 2016. Microbiota-based model improves the
+sensitivity of fecal immunochemical test for detecting colonic lesions.
+Genome Medicine \textbf{8}:37.
+doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.}
 
-\leavevmode\vadjust pre{\hypertarget{ref-Sobhani2011}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-gevers2014}{}}%
 \CSLLeftMargin{2. }%
-\CSLRightInline{\textbf{Sobhani I}, \textbf{Tap J},
-\textbf{Roudot-Thoraval F}, \textbf{Roperch JP}, \textbf{Letulle S},
-\textbf{Langella P}, \textbf{Corthier G}, \textbf{Tran Van Nhieu J},
-\textbf{Furet JP}. 2011. Microbial dysbiosis in colorectal cancer
-({CRC}) patients. PLoS One \textbf{6}:e16393.}
+\CSLRightInline{\textbf{Gevers D}, \textbf{Kugathasan S}, \textbf{Denson
+LA}, \textbf{Vázquez-Baeza Y}, \textbf{Van~Treuren W}, \textbf{Ren B},
+\textbf{Schwager E}, \textbf{Knights D}, \textbf{Song SJ},
+\textbf{Yassour M}, \textbf{Morgan XC}, \textbf{Kostic AD}, \textbf{Luo
+C}, \textbf{González A}, \textbf{McDonald D}, \textbf{Haberman Y},
+\textbf{Walters T}, \textbf{Baker S}, \textbf{Rosh J}, \textbf{Stephens
+M}, \textbf{Heyman M}, \textbf{Markowitz J}, \textbf{Baldassano R},
+\textbf{Griffiths A}, \textbf{Sylvester F}, \textbf{Mack D}, \textbf{Kim
+S}, \textbf{Crandall W}, \textbf{Hyams J}, \textbf{Huttenhower C},
+\textbf{Knight R}, \textbf{Xavier RJ}. 2014. The treatment-naive
+microbiome in new-onset crohn{'}s disease. Cell Host \& Microbe
+\textbf{15}:382--392.
+doi:\href{https://doi.org/10.1016/j.chom.2014.02.005}{10.1016/j.chom.2014.02.005}.}
 
-\leavevmode\vadjust pre{\hypertarget{ref-westcott2015}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-duvallet2017}{}}%
 \CSLLeftMargin{3. }%
-\CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo
-clustering methods outperform reference-based methods for assigning 16S
-rRNA gene sequences to operational taxonomic units. PeerJ
-\textbf{3}:e1487.
-doi:\href{https://doi.org/10.7717/peerj.1487}{10.7717/peerj.1487}.}
+\CSLRightInline{\textbf{Duvallet C}, \textbf{Gibbons SM}, \textbf{Gurry
+T}, \textbf{Irizarry RA}, \textbf{Alm EJ}. 2017. Meta-analysis of gut
+microbiome studies identifies disease-specific and shared responses.
+Nature Communications \textbf{8}:1784.
+doi:\href{https://doi.org/10.1038/s41467-017-01973-8}{10.1038/s41467-017-01973-8}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Armour2022}{}}%
+\CSLLeftMargin{4. }%
+\CSLRightInline{\textbf{Armour CR}, \textbf{Topçuoğlu BD},
+\textbf{Garretto A}, \textbf{Schloss PD}. 2022. A goldilocks principle
+for the gut microbiome: Taxonomic resolution matters for
+microbiome-based classification of colorectal cancer. MBio
+\textbf{13}:e0316121.}
 
 \leavevmode\vadjust pre{\hypertarget{ref-sovacool2022}{}}%
-\CSLLeftMargin{4. }%
+\CSLLeftMargin{5. }%
 \CSLRightInline{\textbf{Sovacool KL}, \textbf{Westcott SL},
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-baxter2016}{}}%
-\CSLLeftMargin{5. }%
-\CSLRightInline{\textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers
-MAM}, \textbf{Schloss PD}. 2016. Microbiota-based model improves the
-sensitivity of fecal immunochemical test for detecting colonic lesions.
-Genome Medicine \textbf{8}:37.
-doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-westcott2017}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-rognes2016}{}}%
 \CSLLeftMargin{6. }%
-\CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2017.
-OptiClust, an improved method for assigning amplicon-based sequence data
-to operational taxonomic units. mSphere \textbf{2}:e00073--17.
-doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-rognes2016}{}}%
-\CSLLeftMargin{7. }%
 \CSLRightInline{\textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bolyen2019}{}}%
-\CSLLeftMargin{8. }%
+\CSLLeftMargin{7. }%
 \CSLRightInline{\textbf{Bolyen E}, \textbf{Rideout JR}, \textbf{Dillon
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-edgar2011}{}}%
-\CSLLeftMargin{9. }%
-\CSLRightInline{\textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente
-JC}, \textbf{Quince C}, \textbf{Knight R}. 2011. UCHIME improves
-sensitivity and speed of chimera detection. Bioinformatics
-\textbf{27}:2194--2200.
-doi:\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.}
+\leavevmode\vadjust pre{\hypertarget{ref-westcott2015}{}}%
+\CSLLeftMargin{8. }%
+\CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo
+clustering methods outperform reference-based methods for assigning 16S
+rRNA gene sequences to operational taxonomic units. PeerJ
+\textbf{3}:e1487.
+doi:\href{https://doi.org/10.7717/peerj.1487}{10.7717/peerj.1487}.}
 
 \leavevmode\vadjust pre{\hypertarget{ref-schloss2009}{}}%
-\CSLLeftMargin{10. }%
+\CSLLeftMargin{9. }%
 \CSLRightInline{\textbf{Schloss PD}, \textbf{Westcott SL},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-quast2013}{}}%
-\CSLLeftMargin{11. }%
+\CSLLeftMargin{10. }%
 \CSLRightInline{\textbf{Quast C}, \textbf{Pruesse E}, \textbf{Yilmaz P},
@@
 doi:\href{https://doi.org/10.1093/nar/gks1219}{10.1093/nar/gks1219}.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-edgar2011}{}}%
+\CSLLeftMargin{11. }%
+\CSLRightInline{\textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente
+JC}, \textbf{Quince C}, \textbf{Knight R}. 2011. UCHIME improves
+sensitivity and speed of chimera detection. Bioinformatics
+\textbf{27}:2194--2200.
+doi:\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.}
 
@@
 
-\hypertarget{figures}{%
-\subsection{Figures}\label{figures}}
-
-\includegraphics{./figures/Figure1.png}
+\hypertarget{figure-legends}{%
+\subsection{Figure Legends}\label{figure-legends}}
 
@@
 pipeline.
-
-\newpage
-
-\includegraphics{figures/fig2.png}
 

### 2023-07-05 submission/second_revision/manuscript.tex
fix citation error

@@
 \usepackage{amsmath,amssymb}
-\usepackage{lmodern}
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math}
+  \usepackage{unicode-math} % this also loads fontspec
   \defaultfontfeatures{Scale=MatchLowercase}
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else
+  % xetex/luatex font selection
 \fi
@@
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
+\urlstyle{same}
 \hypersetup{
@@
 workflows, the 16S rRNA sequence data from all samples were clustered
-into OTUs using the OptiClust algorithm in mothur 1 and the VSEARCH
-algorithm used in QIIME2 (\protect\hyperlink{ref-rognes2016}{6},
-\protect\hyperlink{ref-bolyen2019}{7}). For both algorithms, the
+into OTUs using the OptiClust algorithm in mothur
+(\protect\hyperlink{ref-westcott2017}{6}) and the VSEARCH algorithm used
+in QIIME2 (\protect\hyperlink{ref-rognes2016}{7},
+\protect\hyperlink{ref-bolyen2019}{8}). For both algorithms, the
 resulting abundance data was then split into training and testing sets,
@@
 similarity of all pairs of sequences and whether they are appropriately
-clustered or not (\protect\hyperlink{ref-westcott2015}{8}). We expected
+clustered or not (\protect\hyperlink{ref-westcott2015}{9}). We expected
 the MCC scores produced by the OptiFit workflow to be similar to that of
@@
 \textbf{Data processing.} The full dataset was preprocessed with mothur
-(v1.47) (\protect\hyperlink{ref-schloss2009}{9}) to join forward and
+(v1.47) (\protect\hyperlink{ref-schloss2009}{10}) to join forward and
 reverse reads, merge duplicate reads, align to the SILVA reference
-database (v132) (\protect\hyperlink{ref-quast2013}{10}), precluster,
-remove chimeras with UCHIME (\protect\hyperlink{ref-edgar2011}{11}),
+database (v132) (\protect\hyperlink{ref-quast2013}{11}), precluster,
+remove chimeras with UCHIME (\protect\hyperlink{ref-edgar2011}{12}),
 assign taxonomy, and remove non-bacterial reads following the Schloss
@@
   OptiFit Greengenes: Reference sequences from the Greengenes database
-  v13\_8\_99 (\protect\hyperlink{ref-desantis2006}{12}) were downloaded
+  v13\_8\_99 (\protect\hyperlink{ref-desantis2006}{13}) were downloaded
   and processed with mothur by trimming to the V4 region and clustered
   \emph{de novo} with OptiClust
-
-  \begin{enumerate}
-  \def\labelenumii{\arabic{enumii}.}
-  \tightlist
-  \item
-    The preprocessed data was fit to the clustered reference data using
-    OptiFit with the method open to allow any sequences that did not map
-    to the existing reference clusters would form new OTUs. The data was
-    then subsampled to 10,000 reads and any novel OTUs from the test set
-    were removed. The dataset was then split into two sets where 80\% of
-    the samples were assigned to the training set and 20\% to the
-    testing set. This process was repeated for each of the 100 splits
-    resulting in 100 training and testing datasets.
-  \end{enumerate}
+  (\protect\hyperlink{ref-westcott2017}{6}). The preprocessed data was
+  fit to the clustered reference data using OptiFit with the method open
+  to allow any sequences that did not map to the existing reference
+  clusters would form new OTUs. The data was then subsampled to 10,000
+  reads and any novel OTUs from the test set were removed. The dataset
+  was then split into two sets where 80\% of the samples were assigned
+  to the training set and 20\% to the testing set. This process was
+  repeated for each of the 100 splits resulting in 100 training and
+  testing datasets.
 \item
   VSEARCH Greengenes: Preprocessed data was clustered using VSEARCH
-  v2.15.2 (\protect\hyperlink{ref-rognes2016}{6}) directly to
+  v2.15.2 (\protect\hyperlink{ref-rognes2016}{7}) directly to
   unprocessed Greengenes 97\% OTU reference alignment consistent with
   how VSEARCH is typically used by the QIIME2 software for
-  reference-based clustering (\protect\hyperlink{ref-bolyen2019}{7}).
+  reference-based clustering (\protect\hyperlink{ref-bolyen2019}{8}).
   The data was then subsampled to 10,000 reads and any novel OTUs from
@@
   OptiClust \emph{de novo}: All the preprocessed data was clustered
-  together with OptiClust 1 to generate OTUs. The data was subsampled to
-  10,000 reads per sample and the resulting abundance tables were split
-  into the training and testing sets. The process was repeated for each
-  of the 100 splits resulting in 100 training and testing datasets.
+  together with OptiClust (\protect\hyperlink{ref-westcott2017}{6}) to
+  generate OTUs. The data was subsampled to 10,000 reads per sample and
+  the resulting abundance tables were split into the training and
+  testing sets. The process was repeated for each of the 100 splits
+  resulting in 100 training and testing datasets.
 \item
   VSEARCH \emph{de novo}: All the preprocessed data was clustered using
-  VSEARCH v2.15.2 (\protect\hyperlink{ref-rognes2016}{6}) with 97\%
+  VSEARCH v2.15.2 (\protect\hyperlink{ref-rognes2016}{7}) with 97\%
   identity and then subsampled to 10,000 reads per sample. The process
@@
 package mikrompl (v 1.2.0)
-(\protect\hyperlink{ref-topuxe7uoglu2021}{13}) to predict the diagnosis
+(\protect\hyperlink{ref-topuxe7uoglu2021}{14}) to predict the diagnosis
 (SRN or normal) for the samples in the test set for each data split. The
@@
 removed. P-values comparing model performance were calculated as
-previously described (\protect\hyperlink{ref-topuxe7uoglu2020}{14}). The
+previously described (\protect\hyperlink{ref-topuxe7uoglu2020}{15}). The
 averaged ROC curves were plotted by taking the average and standard
@@
 \textbf{Code Availability.} The analysis workflow was implemented in
-Snakemake (\protect\hyperlink{ref-koster2012}{15}). Scripts for analysis
-were written in R (\protect\hyperlink{ref-R2020}{16}) and GNU bash
-(\protect\hyperlink{ref-GNUbash}{17}). The software used includes mothur
-v1.47.0 (\protect\hyperlink{ref-schloss2009}{9}), VSEARCH v2.15.2
-(\protect\hyperlink{ref-rognes2016}{6}), RStudio
-(\protect\hyperlink{ref-RStudio2019}{18}), the Tidyverse metapackage
-(\protect\hyperlink{ref-wickham2019}{19}), R Markdown
-(\protect\hyperlink{ref-xie_r_2018}{20}), the SRA toolkit
-(\protect\hyperlink{ref-noauthor_sra-tools_nodate}{21}), and conda
-(\protect\hyperlink{ref-noauthor_anaconda_2016}{22}). The complete
+Snakemake (\protect\hyperlink{ref-koster2012}{16}). Scripts for analysis
+were written in R (\protect\hyperlink{ref-R2020}{17}) and GNU bash
+(\protect\hyperlink{ref-GNUbash}{18}). The software used includes mothur
+v1.47.0 (\protect\hyperlink{ref-schloss2009}{10}), VSEARCH v2.15.2
+(\protect\hyperlink{ref-rognes2016}{7}), RStudio
+(\protect\hyperlink{ref-RStudio2019}{19}), the Tidyverse metapackage
+(\protect\hyperlink{ref-wickham2019}{20}), R Markdown
+(\protect\hyperlink{ref-xie_r_2018}{21}), the SRA toolkit
+(\protect\hyperlink{ref-noauthor_sra-tools_nodate}{22}), and conda
+(\protect\hyperlink{ref-noauthor_anaconda_2016}{23}). The complete
 workflow and supporting files required to reproduce this study are
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-westcott2017}{}}%
+\CSLLeftMargin{6. }%
+\CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2017.
+OptiClust, an improved method for assigning amplicon-based sequence data
+to operational taxonomic units. mSphere \textbf{2}:e00073--17.
+doi:\href{https://doi.org/10.1128/mSphereDirect.00073-17}{10.1128/mSphereDirect.00073-17}.}
+
 \leavevmode\vadjust pre{\hypertarget{ref-rognes2016}{}}%
-\CSLLeftMargin{6. }%
+\CSLLeftMargin{7. }%
 \CSLRightInline{\textbf{Rognes T}, \textbf{Flouri T}, \textbf{Nichols
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bolyen2019}{}}%
-\CSLLeftMargin{7. }%
+\CSLLeftMargin{8. }%
 \CSLRightInline{\textbf{Bolyen E}, \textbf{Rideout JR}, \textbf{Dillon
@@
 \leavevmode\vadjust pre{\hypertarget{ref-westcott2015}{}}%
-\CSLLeftMargin{8. }%
+\CSLLeftMargin{9. }%
 \CSLRightInline{\textbf{Westcott SL}, \textbf{Schloss PD}. 2015. De novo
@@
 \leavevmode\vadjust pre{\hypertarget{ref-schloss2009}{}}%
-\CSLLeftMargin{9. }%
+\CSLLeftMargin{10. }%
 \CSLRightInline{\textbf{Schloss PD}, \textbf{Westcott SL},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-quast2013}{}}%
-\CSLLeftMargin{10. }%
+\CSLLeftMargin{11. }%
 \CSLRightInline{\textbf{Quast C}, \textbf{Pruesse E}, \textbf{Yilmaz P},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-edgar2011}{}}%
-\CSLLeftMargin{11. }%
+\CSLLeftMargin{12. }%
 \CSLRightInline{\textbf{Edgar RC}, \textbf{Haas BJ}, \textbf{Clemente
@@
 \leavevmode\vadjust pre{\hypertarget{ref-desantis2006}{}}%
-\CSLLeftMargin{12. }%
+\CSLLeftMargin{13. }%
 \CSLRightInline{\textbf{DeSantis TZ}, \textbf{Hugenholtz P},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-topuxe7uoglu2021}{}}%
-\CSLLeftMargin{13. }%
+\CSLLeftMargin{14. }%
 \CSLRightInline{\textbf{Topçuoğlu BD}, \textbf{Lapp Z}, \textbf{Sovacool
@@
 \leavevmode\vadjust pre{\hypertarget{ref-topuxe7uoglu2020}{}}%
-\CSLLeftMargin{14. }%
+\CSLLeftMargin{15. }%
 \CSLRightInline{\textbf{Topçuoğlu BD}, \textbf{Lesniak NA},
@@
 \leavevmode\vadjust pre{\hypertarget{ref-koster2012}{}}%
-\CSLLeftMargin{15. }%
+\CSLLeftMargin{16. }%
 \CSLRightInline{\textbf{Koster J}, \textbf{Rahmann S}. 2012.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-R2020}{}}%
-\CSLLeftMargin{16. }%
+\CSLLeftMargin{17. }%
 \CSLRightInline{\textbf{R Core Team}. 2020.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-GNUbash}{}}%
-\CSLLeftMargin{17. }%
+\CSLLeftMargin{18. }%
 \CSLRightInline{\textbf{GNU Project}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-RStudio2019}{}}%
-\CSLLeftMargin{18. }%
+\CSLLeftMargin{19. }%
 \CSLRightInline{\textbf{RStudio Team}. 2019.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-wickham2019}{}}%
-\CSLLeftMargin{19. }%
+\CSLLeftMargin{20. }%
 \CSLRightInline{\textbf{Wickham H}, \textbf{Averick M}, \textbf{Bryan
@@
 \leavevmode\vadjust pre{\hypertarget{ref-xie_r_2018}{}}%
-\CSLLeftMargin{20. }%
+\CSLLeftMargin{21. }%
 \CSLRightInline{\textbf{Xie Y}, \textbf{Allaire JJ}, \textbf{Grolemund
@@
 \leavevmode\vadjust pre{\hypertarget{ref-noauthor_sra-tools_nodate}{}}%
-\CSLLeftMargin{21. }%
+\CSLLeftMargin{22. }%
 \CSLRightInline{{SRA}-{Tools} - {NCBI}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-noauthor_anaconda_2016}{}}%
-\CSLLeftMargin{22. }%
+\CSLLeftMargin{23. }%
 \CSLRightInline{2016. Anaconda {Software Distribution}. Anaconda

### 2023-07-05 submission/second_revision/manuscript.Rmd
fix citation error
update repo link

@@
     v13_8\_99 @desantis2006 were downloaded and processed with mothur by
-    trimming to the V4 region and clustered *de novo* with OptiClust
-    @westcott2017. The preprocessed data was fit to the clustered
+    trimming to the V4 region and clustered *de novo* with OptiClust @westcott2017. The preprocessed data was fit to the clustered
     reference data using OptiFit with the method open to allow any
@@
 The complete workflow and supporting files required to reproduce this
-study are available at:
-<https://github.com/SchlossLab/Armour_OptiFitGLNE_XXXX_2023>
+study are available at: [https://github.com/SchlossLab/Armour_OptiFitGLNE_mSphere_2023](https://github.com/SchlossLab/Armour_OptiFitGLNE_mSphere_2023)
 
